---
title: "07: Econometric Modeling"
format: 
  revealjs: 
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to course website with lecture slides](https://tmieno2.github.io/Applied-Econometrics-MS-Quarto/lectures/)"
execute:
  echo: true
  out-width: 55%
webr:
  packages: ['fixest', 'dplyr', 'data.table', 'ggplot2', 'wooldridge', 'car', 'broom']
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 10
    fig-height: 6
    out-width: 100%
filters:
  - webr
---

# Functional form

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

```{r packages, include = F, cache = F}
#--- load packages ---#
library(broom)
library(fixest)
library(ggplot2)
library(data.table)
library(dplyr)
library(modelsummary)
library(car)

source(here::here("lectures/_lecture-theme.R"))
```

## Functional form

::: {.panel-tabset}

### Introduction

<details class="transcript">
<summary>Transcript</summary>

This section is about functional form, meaning the mathematical shape we use to connect an outcome to its explanatory variables. The first bullet gives us the key rule: you may transform the variables without leaving the ordinary least squares framework, provided the model remains linear in the unknown parameters. Linear in parameters means that beta zero, beta one, and the other unknown coefficients enter as an additive weighted sum. The variables themselves do not have to enter linearly.

Look at the four examples in the callout. In the log-linear model, we take the log of the outcome but leave x in levels. In the linear-log model, the outcome stays in levels and we take the log of x. In the log-log model, both variables are logged. In the quadratic model, x and x squared both appear. Each expression can still be written as beta zero times a column of ones, plus beta one times one observed column, and, where needed, beta two times another observed column. OLS can therefore estimate all four.

The second bullet warns that transformation changes interpretation. Beta one is no longer automatically the unit change in y caused by a one-unit increase in x. Depending on the model, it can describe a percentage change, an elasticity, or a marginal effect that varies with x.

</details>

+ Transformation of variables is allowed without disturbing our analytical framework as long as the model is linear in <span style = "color: blue;"> parameter </span>.

+ Transformation of variables change the interpretation of the coefficients estimates

:::{.callout-note title="Example models"}

::: {.columns}

::: {.column width="50%"}
**log-linear**

$log(y_i)= \beta_0+\beta_1 x_i + u_i$ 

<br>

**linear-log**

$y_i= \beta_0+\beta_1 log(x_i) + u_i$
:::
<!--end of the 1st column-->
::: {.column width="50%"}
**log-log**

$log(y_i)= \beta_0+\beta_1 log(x_i) + u_i$ 

<br>

**quadratic**

$y_i= \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + u_i$ 
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

:::

### Linear or Non-linear?

<details class="transcript">
<summary>Transcript</summary>

The question at the top is reasonable: if a transformed model produces a curved relationship between x and y, why do we still call it linear? The blue words identify the answer. In a linear regression model, linear describes the way the unknown parameters enter, not necessarily the way the observed variables enter.

For example, in a quadratic model, x squared may trace a curve, but once you observe x you can calculate x squared for every observation. It is simply another regressor column. The conditional mean is still beta zero plus beta one times x plus beta two times x squared, an additive linear combination of the unknown betas. The same logic applies to log of x, log of y, or a product such as education times experience. As the callout says, transforming dependent or independent variables does not by itself disturb the OLS properties we have developed, as long as the resulting model is linear in its parameters and the other OLS assumptions are satisfied.

Now contrast that with the two displayed nonlinear models. In the first, beta one is an exponent on x. In the second, beta zero and beta one appear inside a denominator. Those parameters do not act as simple additive weights on known regressor columns. Ordinary least squares cannot estimate them by the usual linear projection calculation. So do not diagnose linearity by looking for a straight line in a graph. Inspect the parameters. If the betas enter additively and to the first power, the model is linear in parameters even when the fitted relationship in the original variables is curved.

</details>

+ In the models we just saw, the dependent variable and independent variable are non-linearly related, how come are these models called simple <span style = "color: blue;"> linear </span> model?

+ <span style = "color: blue;"> "linear" </span> in simple <span style = "color: blue;"> linear </span> model means that the model is linear in <span style = "color: blue;"> parameter </span>, but not in <span style = "color: blue;"> variable </span>


**Examples: Non-linear models**

```{=tex}
\begin{align*}
  y_i=\beta_0+x_i^{\beta_1}+u_i \\
  y_i=\frac{x_i}{\beta_0+\beta_1 x_i}+u_i
\end{align*}
```

<br>

:::{.callout-note}
Transformation of the dependent and independent variables would not affect the properties of the OLS estimator as long as the model is linear in parameter.
:::

### Why bother?

<details class="transcript">
<summary>Transcript</summary>

Here is why functional form is an economic issue rather than a cosmetic one. The displayed equation makes corn yield equal to an intercept, plus beta one times fertilizer, plus an unobserved component mu. Because fertilizer enters only as a single level term, the marginal effect is beta one everywhere. The model therefore says that one additional unit of fertilizer changes expected yield by exactly the same amount whether a farmer is moving from zero to one unit, from fifty to fifty-one, or from one hundred to one hundred and one.

That constant response is the problem. Agronomic reasoning suggests diminishing marginal productivity: early fertilizer applications may raise yield substantially, later applications add less, and excessive fertilizer can eventually reduce yield or create other damage. A straight line cannot flatten and cannot turn downward.

Open the answer to see the same point stated compactly. The phrase "the impact becomes almost zero at some point" tells us what a better model must permit: a marginal effect that varies with fertilizer. A quadratic term is one possible choice because it allows the slope to decline and even change sign.

</details>

Consider a following model:

```{=tex}
\begin{align*}
\text{corn yield} = \beta_0 + \beta_1 \cdot \text{fertilizer} + \mu
\end{align*}
```

**Question**

What is wrong with this model?

<br>
<details>
  <summary>Answer</summary>
It assumes that fertilizer affects corn yield the same way no matter how much fertilizer you apply. It does not reflect the reality that the impact of fertilizer becomes almost zero at some point. A better model?
</details>

:::
<!--end of panel-->

## Various functional forms

::: {.panel-tabset}

### Log-linear

::: {.panel-tabset}

#### Basics

<details class="transcript">
<summary>Transcript</summary>

In a log-linear model, only the dependent variable is transformed. Read the first equation as log y sub i equals beta zero plus beta one times x sub i plus u sub i. Here y must be positive because its logarithm must exist, while x remains in its original units.

Look at the calculus line. Differentiate both sides with respect to x sub i. By the chain rule, the derivative of log y with respect to x is one over y times the derivative of y with respect to x. The derivative of beta zero is zero, the derivative of beta one x is beta one, and the usual marginal-effect calculation holds u fixed. That gives one over y times partial y over partial x equals beta one. For a small finite change, we write the proportional change delta y over y as approximately beta one times delta x.

Now translate that result into speech. If x rises by one unit, so delta x equals one, y changes by approximately one hundred times beta one percent. Also notice that the slide's delta expression is a local approximation. For a one-unit increase in x, the fitted conditional median of y is multiplied exactly by exponential of beta one, so the exact percentage change is one hundred times exponential beta one minus one. The approximation one hundred beta one is very close when beta one is small.

</details>

**Model**
```{=tex}
\begin{align*}
  log(y_i)= \beta_0+\beta_1 x_i + u_i \notag
\end{align*}
```

**Calculus**

Differentiating the both sides wrt $x_i$,

```{=tex}
\begin{align*}
  \frac{1}{y_i}\cdot\frac{\partial y_i}{\partial x_i} = \beta_1 \Rightarrow \frac{\Delta y_i}{y_i} = \beta_1 \Delta x_i \notag
\end{align*}
```

**Interpretation**

$\beta_1$ measures a percentage change in $y_i$ (once multiplied by 100) when $x_i$ is increased by one unit

#### Visualization

<details class="transcript inline">
<summary>Transcript</summary>

This figure translates the log-linear equation back into the original units of y. The horizontal axis is x, running from one to two, and the vertical axis is y. The legend labels each curve by the pair beta zero, beta one. The curve associated with two comma one uses y equals exponential of two plus x. The curve associated with four comma negative one uses y equals exponential of four minus x.

Why exponential? Starting from log y equals beta zero plus beta one x, exponentiate both sides. Ignoring the disturbance for the moment, y equals exponential of beta zero plus beta one x. That is why a model that is linear in log y becomes curved when we plot y itself. For the positive beta one curve, y rises as x rises, and the increase becomes steeper because the marginal effect in levels is beta one times y. A larger current y produces a larger absolute change. For the negative beta one curve, y falls as x rises and approaches zero from above.

The graph illustrates two restrictions. First, the sign of beta one determines whether the curve always rises or always falls; this simple form cannot turn around. Second, the fitted level is always positive because an exponential is positive.

</details>

```{r }
#| echo: false
N <- 1000
x <- seq(1, 2, length = N)
y_p <- exp(2 + x)
y_n <- exp(4 - x)

y_p_data <-
  data.table(
    y = y_p,
    x = x,
    type = "(2,1)"
  )

y_n_data <-
  data.table(
    y = y_n,
    x = x,
    type = "(4,-1)"
  )

plot_data <- rbind(y_p_data, y_n_data)

(
  g_log_linear <-
    ggplot(data = plot_data) +
    geom_line(aes(y = y, x = x, color = type)) +
    scale_color_discrete(name = expression(list(beta[0], beta[1]))) +
    theme(
      legend.position = "bottom"
    ) +
    theme_lecture
)
```

#### Example

<details class="transcript">
<summary>Transcript</summary>

Now apply the general log-linear derivation to a familiar labor-economics model. The first line says log wage equals beta zero plus beta one times education plus u. Wage is the hourly wage in positive level units, and education is measured in years. Beta zero is the fitted log wage at zero years of education, which may not be the quantity we care most about. Beta one is the slope of log wage with respect to education.

The calculus line makes that interpretation explicit. Differentiating log wage with respect to education gives one over wage times partial wage over partial education on the left and beta one on the right. For a small change, delta wage over wage is approximately beta one times delta education. If education rises by one year, delta education equals one, so the proportional wage change is approximately beta one. Multiplying by one hundred converts that proportion to a percentage. Thus we say that one additional year of education is associated with an approximately one hundred beta one percent change in hourly wage, holding the other included regressors fixed.

</details>

**Model**

```{=tex}
\begin{align*}
    log(wage)=\beta_0 + \beta_1 educ + u \notag
\end{align*}
```

**Calculus**

Differentiating both sides with respect to $educ$,

```{=tex}
\begin{align*}
    \frac{1}{wage} \frac{\partial wage}{\partial educ} = \beta_1 \Rightarrow \frac{\Delta wage}{wage} = \beta_1\Delta educ\notag
\end{align*}
```

**Interpretation**

If education increases by 1 year $(\Delta educ=1)$, then wage increases by $\beta_1*100\%$ $(\frac{\Delta wage}{wage}=\beta_1)$

#### Estimated model

<details class="transcript inline">
<summary>Transcript</summary>

The left side gives the estimated log-linear model. The first fitted equation says predicted log wage equals zero point five eight four plus zero point zero eight three times education. The slope means that one additional year of education is associated with approximately an eight point three percent higher wage. Using the exact calculation gives one hundred times exponential of zero point zero eight three minus one, which is about eight point seven percent. The second equation exponentiates the fitted log value, producing exponential of zero point five eight four plus zero point zero eight three education in wage units.

Now look at the figure. The horizontal axis is years of education and the vertical axis is hourly wage, limited on screen to zero through fifteen dollars. Each dot is an observation from the wage data. The red "OLS Regression Line" comes from regressing wage in levels directly on education, so it is straight. The blue "Log Fit" is the exponentiated log-linear prediction, so it curves upward. The two fits can be similar near the center of the education distribution but diverge near the ends because one imposes a constant dollar slope and the other imposes a constant proportional slope.

The callout is essential. If log wage equals its fitted value plus u, exponentiating gives wage equals exponential of the fitted value times exponential u. Simply exponentiating the fitted line omits the conditional mean of exponential u. Under a symmetric log error, the fitted log line corresponds to the conditional median of wage, not generally its conditional mean. Jensen's inequality gives expected exponential u greater than exponential of expected u, so recovering the mean requires a retransformation factor. The slide deliberately labels the plotted blue curve as the exponentiated fit, not an automatically unbiased mean prediction.

</details>

::: {.columns}

::: {.column width="40%"}
When you estimate the following model using the wage dataset:

$$log(wage)=\beta_0 + \beta_1 educ + u \notag$$

<br>

Then, the estimated equation is the following:

```{=tex}
\begin{align*}
  \widehat{log(wage)}=0.584+0.083 educ \notag
\end{align*}
```

```{=tex}
\begin{align*}
  \widehat{wage}=e^{0.584+0.083 educ}
\end{align*}
```

:::{.callout-note title="Careful"}
Exponentiating the fitted line gives the predicted <span style = "color: blue;">median</span> of $wage$, not its mean. Because $E[e^u]>e^{E[u]}$, recovering $E[wage\mid educ]$ requires an extra retransformation factor.
:::
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r g_wage_log}
#| echo: false
data(wage1, package = "wooldridge")
wage <- data.table(wage1)

min_educ <- wage[, educ] %>% min()
max_educ <- wage[, educ] %>% max()
x <- seq(min_educ, max_educ, length = 1000)
y <- exp(0.584 + 0.083 * x)
log_fit <- data.table(y = y, x = x)
col <- c("OLS Regression Line" = "red", "Log Fit" = "blue")
ggplot(data = wage, aes(y = wage, x = educ)) +
  geom_point(size = 0.7) +
  geom_smooth(method = "lm", se = F, aes(color = "OLS Regression Line")) +
  geom_line(data = log_fit, aes(y = y, x = x, color = "Log Fit"), linewidth = 1) +
  scale_color_manual(values = col, name = "") +
  ylim(0, 15) +
  theme_lecture +
  theme(
    legend.position = "bottom"
  )
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


:::
<!--end of panel-->

### Linear-log

::: {.panel-tabset}

#### Basics

<details class="transcript">
<summary>Transcript</summary>

In the linear-log model, y remains in its original units and only x is logged. The model line reads y sub i equals beta zero plus beta one times log x sub i plus u sub i. Because log x appears, x must be positive. Beta zero is the fitted value when log x is zero, which means x equals one, not when x equals zero.

Follow the calculus line carefully. The derivative of y with respect to x is the quantity we want. The derivative of beta one log x is beta one divided by x, so partial y over partial x equals beta one over x. For a small finite change, multiply by delta x: delta y is approximately beta one times delta x over x. The fraction delta x over x is a proportional change in x.

This produces a level response to a percentage change. If x rises by one percent, delta x over x is zero point zero one, and y changes by approximately zero point zero one times beta one units of y. Do not say y changes by beta one percent. Y is not logged, so its change is measured in y's original units. Do not say a one-unit increase in x changes y by beta one either. The one-unit marginal effect is beta one over the current x and therefore varies with x.

That dependence creates diminishing absolute effects when beta one is positive: the same one-unit increase in x matters more when x is small than when x is large. Yet the effect of a fixed percentage increase is constant.

</details>

**Model**

```{=tex}
\begin{align*}
  y_i= \beta_0+\beta_1 log(x_i) +u_i \notag
\end{align*}
```

**Calculus**

Differentiating the both sides wrt $x_i$,

```{=tex}
\begin{align*}
  \frac{\partial y_i}{\partial x_i} = \frac{\beta_1}{x_i} \Rightarrow \Delta y_i = \beta_1\frac{\Delta x_i}{x_i} \notag
\end{align*}
```

**Interpretation**

When $x$ increases by $1\%$ (that is, $\frac{\Delta x_i}{x_i} = 0.01$), $y$ increases by $\beta_1 \times 0.01$ units.

#### Visualization

<details class="transcript inline">
<summary>Transcript</summary>

The equation above the plot fixes beta zero at one and beta one at two, so the curve is y equals one plus two times log x. The horizontal axis is x, shown from one to twenty, and the vertical axis is y in level units. At x equal to one, log one is zero, so the curve begins at y equal to one. As x rises, y rises, but the line becomes progressively flatter.

The derivative explains that shape. Partial y over partial x equals two over x. At x equal to one, the slope is two. At x equal to ten, it is zero point two. At x equal to twenty, it is zero point one. Thus equal one-unit increases in x have smaller effects at larger starting values. By contrast, equal percentage increases have the same effect on y because beta one multiplies the proportional change in x.

This graph also shows what the linear-log form cannot do. With positive beta one it is always increasing over positive x, although at a diminishing rate. It never reaches a peak and turns downward.

</details>

$$y = \beta_0 + \beta_1 log(x) = 1 + 2 \times log(x)$$

```{r linear_log_vis}
#| echo: false

N <- 1000
x <- seq(1, 20, length = N)
y <- 1 + 2 * log(x)
plot_data <- data.table(y = y, x = x, type = "(0,1)")

ggplot(data = plot_data) +
  geom_line(aes(y = y, x = x)) +
  scale_color_discrete(name = expression(list(beta[0], beta[1]))) +
  theme(
    legend.position = "bottom"
  ) +
  theme_lecture
```

:::
<!--end of panel-->

### Log-log

<details class="transcript">
<summary>Transcript</summary>

In the log-log model, both the dependent variable and the regressor are transformed. The first equation reads log y sub i equals beta zero plus beta one times log x sub i plus u sub i. Both x and y must therefore be positive.

The calculus line is written as a ratio of proportional derivatives. Differentiating gives one over y times partial y over partial x equals beta one times one over x. Rearranging, the percentage change in y divided by the percentage change in x equals beta one. In finite-change language, delta y over y is approximately beta one times delta x over x. If x rises by one percent, y changes by approximately beta one percent. A negative beta one means the variables move in opposite directions, as in a downward-sloping demand curve; a positive beta one means they move together.

Beta one is therefore an elasticity. It is unit-free, so measuring x in dollars or cents and y in pounds or kilograms does not change it. The phrase "constant elasticity" on the slide is a restriction, not merely a convenient interpretation. The model requires the same percentage response at every positive value of x. In level form, ignoring u, y equals exponential beta zero times x raised to beta one, so the absolute slope can change even though the elasticity remains fixed.

Be precise about the approximation. For a small percentage change, beta one times that percentage is accurate.

</details>

**Model**

```{=tex}
\begin{align*}
  log(y_i)= \beta_0+\beta_1 log(x_i) +u_i \notag
\end{align*}
```

**Calculus**

Differentiating the both sides wrt $x_i$,

```{=tex}
\begin{align*}
  \frac{\partial y_i}{y_i}/\frac{\partial x_i}{x_i} = \beta_1 \Rightarrow \frac{\Delta y_i}{y_i} = \beta_1 \frac{\Delta x_i}{x_i}\notag
\end{align*}
```

**Interpretation**

A <span style = "color: blue;"> percentage </span> change in $x$ would result in a $\beta_1$ <span style = "color: blue;"> percentage </span> change in $y_i$ (constant elasticity)

### Quadratic

```{r}
#| echo: false
N <- 1000
x <- seq(0, 2, length = N)
y_1 <- x + x^2
y_2 <- 3 * x - 2 * x^2
y_1_data <- data.table(
  y = y_1,
  x = x
)
y_2_data <- data.table(
  y = y_2,
  x = x
)

g_quad_1 <- ggplot(data = y_1_data) +
  geom_line(aes(y = y, x = x), color = "red") +
  theme_lecture

g_quad_2 <- ggplot(data = y_2_data) +
  geom_line(aes(y = y, x = x), color = "red") +
  theme_lecture
```

::: {.panel-tabset}

#### Basics

<details class="transcript">
<summary>Transcript</summary>

The quadratic model adds x squared as a second regressor: y sub i equals beta zero plus beta one x sub i plus beta two x sub i squared plus u sub i. It is curved in x but still linear in the parameters because beta one and beta two simply multiply two known columns, x and x squared.

Look at the calculus line. Differentiate term by term with respect to x. Beta zero disappears, beta one x contributes beta one, and beta two x squared contributes two beta two x. The marginal effect is therefore beta one plus two times beta two times x. For a one-unit increase in x, the small-change approximation says y changes by that expression, evaluated at the starting value of x.

This dependence on x is the central feature. There is no single constant effect of x to report. Beta one alone is the marginal effect only at x equal to zero. If zero is outside the observed data or has no meaningful interpretation, treating beta one as "the effect of x" is misleading. Beta two controls how the slope changes: each one-unit increase in x changes the marginal effect by two beta two. A positive beta two makes the slope increase with x; a negative beta two makes it decrease.

</details>

**Model**

$y_i= \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + u_i$ 

<br>

**Calculus**

Differentiating the both sides wrt $x_i$,

$\frac{\partial y_i}{\partial x_i} = \beta_1 + 2*\beta_2 x_i\Rightarrow  \Delta y_i = (\beta_1 + 2*\beta_2 x_i)\Delta x_i$

<br>

**Interpretation**

When $x$ increases by 1 unit $(\Delta x_i=1)$, $y$ increases by $\beta_1 + 2*\beta_2 x_i$

#### Visualization

<details class="transcript inline">
<summary>Transcript</summary>

These two panels plot quadratic functions over x from zero to two. In both panels, the horizontal axis is x and the vertical axis is y. The left equation is y equals x plus x squared, so beta one and beta two both equal one. Its marginal effect is one plus two x, which is positive and increasing throughout the plotted range. The curve therefore rises and becomes steeper as x increases.

The right equation is y equals three x minus two x squared. Here beta one is three and beta two is negative two. Its marginal effect is three minus four x. At x equal to zero the slope is three; it falls as x increases, reaches zero at x equal to three quarters, and becomes negative after that. The red curve consequently rises, peaks, and then falls. In general, a quadratic turning point occurs where beta one plus two beta two x equals zero, so x star is minus beta one divided by two beta two. When beta two is negative, that point is a maximum; when beta two is positive, it is a minimum.

This is why the slide calls the form flexible. It can represent increasing returns, diminishing returns, a peak, or a trough.

</details>

Quadratic functional form is quite flexible.

::: {.columns}

::: {.column width="50%"}
$y = x + x^2$ $(\beta_1 = 1, \beta_2 = 1)$

```{r}
#| echo: false
g_quad_1
```
:::
<!--end of the 1st column-->
::: {.column width="50%"}
$y = 3x-2x^2$ $(\beta_1 = 3, \beta_2 = -2)$

```{r}
#| echo: false
g_quad_2
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

#### Example

<details class="transcript">
<summary>Transcript</summary>

Use the three questions to see why a constant return to education may be too restrictive. The outcome of interest is income or wage, and the marginal impact of education means the change in expected income associated with a small increase in years of schooling, holding the other modeled factors fixed.

The first bullet asks about adding two years after elementary school. Those years might move a person toward completing a credential or acquiring foundational skills. The second asks about two years after college, perhaps a master's degree or professional training. The third asks about adding two years after already spending six years in a doctoral program. Those increments all have the same numerical size, two years, but there is no strong reason to assume that the labor market values them identically. Credentials, field, opportunity cost, and the changing scarcity of skills can all make the return depend on the starting education level.

A model with only one education term forces all three cases to share the same slope. That equality comes from the specification, not from evidence. The observation at the bottom says the marginal impact does not appear to be linear, meaning more precisely that the level relationship may not have a constant marginal effect. Adding education squared is the simplest way to let the slope change smoothly with education. The next tab shows how to estimate it in R.

</details>

**Education impacts on income**

The marginal impact of education (the impact of a small change in education on income) may differ depending on the level of education you have had:

+ How much does it help to have two more years of education when you have had education until elementary school?

+ How much does it help to have two more years of education when you have graduated a college?

+ How much does it help to spend two more years as a Ph.D student if you have already spent six years in a Ph.D program

<br>

**Observation**

The marginal impact of education does not seem to be linear.

#### Implementation in R

<details class="transcript">
<summary>Transcript</summary>

The sentence above the code gives the formula rule to remember. Inside an R model formula, operators can have special formula meanings, so wrap literal arithmetic in capital I. Here I of education caret two tells R to calculate education squared observation by observation and include that numeric column. Without I, the caret is interpreted by the formula language rather than simply as the arithmetic square you intend.

The runnable regression call is fixest colon colon feols. Its first argument specifies wage as the dependent variable and includes female, education, and I of education squared on the right. Female controls for the average level difference associated with the binary indicator. The two education terms together create the quadratic. The vcov argument is set to "hetero," asking fixest for heteroskedasticity-robust standard errors. The data argument points to wage1, which the preceding data call loads from the wooldridge package. Assigning the result to quad_reg lets later tabs reuse the same fitted model. Finally, broom colon colon tidy converts the model output into one row per coefficient with the estimate, standard error, test statistic, and p-value.

When you run the cell, the fitted education coefficient is about negative zero point four one six and the coefficient on education squared is about positive zero point zero three nine five. Those signs imply a U-shaped fitted parabola, and the marginal effect is negative zero point four one six plus two times zero point zero three nine five times education. Do not interpret the negative coefficient on education alone as education generally reducing wage. It is the slope at education equal to zero. The combination of both coefficients determines the slope at every education level. The setup code above prepares the WebR plotting theme and packages for the browser; it does not change the econometric specification.

</details>

When you want to include a variable that is a transformation of an existing variable, you can use `I()` function in which you write the mathematical expression of the desired transformation.

```{webr-r}
#| context: setup
# NOTE: a webr setup cell renders a loading widget wherever it sits, so keep it
# inside a content slide rather than on a section divider (or, worse, above the
# first heading, where it becomes an untitled slide of its own).
# Mirror of lecture_theme() in lectures/_lecture-theme.R. WebR runs in the
# browser, against a virtual filesystem that does not hold the repo, so it
# cannot source that file. Edit the file first, then run
#     Rscript lectures/check-lecture-theme.R
# which fails if this copy has drifted from it.
lecture_theme <- function(base_size = 16) {
  ggplot2::theme_bw(base_size = base_size) +
    ggplot2::theme(
      axis.text = ggplot2::element_text(size = base_size * 0.9),
      axis.title = ggplot2::element_text(size = base_size),
      legend.text = ggplot2::element_text(size = base_size * 0.9),
      legend.title = ggplot2::element_text(size = base_size * 0.9),
      plot.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.key = ggplot2::element_rect(fill = "transparent", colour = NA),
      # theme_bw's grey92 grid was chosen against a white panel and all but
      # disappears on the warm paper once the panel stops being opaque.
      panel.grid = ggplot2::element_line(colour = "grey87")
    )
}

# The WebR canvas is 10x6in at out-width 100% of its output track
# (lectures/_metadata.yml), the same canvas the knitr figures use. It asks for
# the whole track because the 420px height cap in notebook.scss is what really
# sizes it. 18pt rather than the 16pt knitr default: on a full slide both land
# at 700x420, where 18pt reads at 17.5px and 16pt at 15.5px, and these decks
# put figures in half-width columns where the smaller size gets hard to read.
# webR draws its canvas at 72 dpi no matter what `dpi` says, so on a 216-dpi
# canvas everything ggplot measures in points or mm comes out a third of its
# size. Scale ggplot2's mm-to-point constants and the theme by the same
# factor and the figure matches the knitr ones. Keep in step with `dpi` above.
webr_scale <- 216 / 72
assignInNamespace(".pt", ggplot2::.pt * webr_scale, "ggplot2")
assignInNamespace(".stroke", ggplot2::.stroke * webr_scale, "ggplot2")
theme_lecture <- lecture_theme(18 * webr_scale) +
  # base_size also sets the theme line widths, which are in mm and already
  # pass through the scaled .pt; pin them to what 18pt gives or they come
  # out 9x thick.
  ggplot2::theme(
    line = ggplot2::element_line(linewidth = 18 / 22),
    rect = ggplot2::element_rect(linewidth = 18 / 22)
  )
ggplot2::theme_set(theme_lecture)
```

```{webr-r}
#| autorun: true
#--- prepare a dataset ---#
data("wage1", package = "wooldridge")

#--- run a regression ---#
quad_reg <- fixest::feols(wage ~ female + educ + I(educ^2), vcov = "hetero", data = wage1)

#--- look at the results ---#
broom::tidy(quad_reg)
```

#### Estimated Model

<details class="transcript inline">
<summary>Transcript</summary>

The displayed fitted equation summarizes the quadratic regression: wage equals about five point six zero, minus two point one two times female, minus zero point four one six times education, plus zero point zero three nine times education squared. The intercept is the fitted wage for a man, because female equals zero, at zero years of education. The female coefficient shifts the fitted curve downward when female equals one. The final two terms determine how the curve changes with schooling.

Now read the plotting code and figure. The sequence x runs from the minimum to the maximum education observed in wage1. The new data table sets education equal to each value in that sequence and female equal to one, so the plotted line is specifically the fitted relationship for women. Predict with newdata calculates the model's fitted wage at each of those combinations. In the graph, the horizontal axis is education in years and the vertical axis is hourly wage. The points are the observed wage and education pairs for everyone in the sample, while the line is the female-equals-one prediction. Because the points include both women and men but the line is for women, use the scatter mainly to see support and overall fit, not as a within-group residual display.

The line falls slightly at very low education and then rises increasingly steeply. Its turning point solves negative zero point four one six plus two times zero point zero three nine five times education equals zero, which is about five point three years. Very few observations lie at the lowest schooling levels, so the apparent decline rests on sparse support and the imposed parabolic shape. The safe conclusion is that the fitted marginal effect becomes more positive as education increases over the well-populated part of the sample.

</details>

**Estimated Model**

$wage = 5.60 - 2.12\times female -0.416\times educ + 0.039\times educ^2$

```{webr-r}
#| context: output
N <- 1000
x <- seq(min(wage1$educ), max(wage1$educ), length = N)
data <- data.table(educ = x, female = 1) %>%
  mutate(y = predict(quad_reg, newdata = .))

g_quad_ex <-
  ggplot() +
  geom_point(data = wage1, aes(y = wage, x = educ), size = 0.4) +
  geom_line(data = data, aes(y = y, x = x)) +
  theme_lecture

g_quad_ex
```

#### Marginal impact of $educ$

<details class="transcript">
<summary>Transcript</summary>

Start with the fitted equation from the previous tab and differentiate it with respect to education. The constant and female terms disappear. The derivative of negative zero point four one six times education is negative zero point four one six, and the derivative of zero point zero three nine times education squared is two times zero point zero three nine times education. That gives the displayed marginal effect: negative zero point four one six plus zero point zero seven eight times education, measured in dollars of hourly wage per additional year of schooling.

Evaluate the expression at the two values shown. At education equal to four, the fitted marginal effect is negative zero point four one six plus zero point zero seven eight times four, about negative zero point one zero four dollars per hour for an additional year. At education equal to ten, it is negative zero point four one six plus zero point zero seven eight times ten, about positive zero point three six four dollars per hour. The sign changes because the quadratic slope rises with education. These are local marginal effects, not the exact fitted change from four to five or ten to eleven; for a discrete one-year change, the squared term also contributes beta education squared times one squared. The derivative is the standard marginal-effect summary.

</details>

According to the estimated model, the marginal impact of $educ$ is:

$\frac{\partial wage}{\partial educ} = -0.416+0.039\times 2\times educ$

+ When $educ = 4$, additional year of education is going to increase hourly wage by `r -0.416+0.078*4` on average 

+ When $educ = 10$, additional year of education is going to increase hourly wage by `r -0.416+0.078*10` on average

:::
<!--end of panel-->

:::
<!--end of panel-->

# Statistical significance of the marginal impact

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Statistical significance of the marginal impact

::: {.panel-tabset}

### Marginal impact

<details class="transcript">
<summary>Transcript</summary>

This slide turns the quadratic marginal effect into an inference question. The expression at the top is negative zero point four one six plus two times zero point zero three nine times education. The first number estimates the coefficient on education, and the second estimates the coefficient on education squared. Because education appears in the weight on the squared-term coefficient, the estimated marginal effect changes with the education level at which you evaluate it.

The two bullets report the individual coefficients and their t-statistics. Be clear about what those separate tests ask. The t-test on beta education asks whether the coefficient on the linear education term is zero. In a quadratic model, that coefficient is the marginal effect only at education equal to zero. The t-test on beta education squared asks whether there is curvature, meaning whether the marginal effect changes with education. Neither test directly asks whether the marginal effect is zero at four, ten, or any other nonzero level.

So the question at the bottom, "is the marginal impact statistically significantly different from zero," is incomplete until we specify education. A quadratic can yield an effect that is indistinguishable from zero at one schooling level and clearly positive at another.

</details>

Let's work with the income model, in which the marginal impact of $educ$ is:

```{=tex}
\begin{align*}
\frac{\partial wage}{\partial educ} = -0.416+0.039\times 2\times educ
\end{align*}
```

+ $\beta_{educ}$: $-0.416$ $(t$-stat $= -1.80)$
+ $\beta_{educ^2}$: $0.039$ $(t$-stat $= 4.10)$

<br>

**Question**

So, is the marginal impact of $educ$ statistically significantly different from $0$?

### In the linear case

::: {.panel-tabset}

#### Model and marginal impact

<details class="transcript">
<summary>Transcript</summary>

Start with the linear specification as a benchmark. The code calls fixest colon colon feols with wage on the left and female plus education on the right. The data argument uses wage1. The vcov argument requests heteroskedasticity-robust standard errors, so the uncertainty calculation does not assume a constant error variance. Broom tidy prints the intercept, female coefficient, and education coefficient with their robust standard errors, test statistics, and p-values. Autorun means the browser evaluates the cell automatically when it is ready.

The estimated equation below the output is wage equals zero point six two, minus two point two seven times female, plus zero point five one times education. Holding education fixed, the female coefficient is the fitted wage difference between observations coded one and the male baseline coded zero. Holding female fixed, one additional year of education is associated with about zero point five one dollars more in hourly wage.

Differentiate this linear equation with respect to education. The derivative is simply beta education, estimated at about zero point five zero six. It contains no education value, so the model imposes the same marginal effect at four, ten, and eighteen years.

</details>

**Regression**

```{webr-r}
#| autorun: true
linear_reg <- fixest::feols(wage ~ female + educ, vcov = "hetero", data = wage1)
broom::tidy(linear_reg)
```

<br>

**Estimated model**

$wage = 0.62 - 2.27 \times female + 0.51 \times educ$

<br>

::: {.panel-tabset}

##### Question 1

<details class="transcript">
<summary>Transcript</summary>

Question one asks for the marginal impact of education in the fitted linear model. Look at the equation immediately above this tab: wage equals zero point six two, minus two point two seven times female, plus zero point five one times education. Differentiate with respect to education. The intercept and female term do not change, and the derivative of zero point five one times education is zero point five one.

So the answer is about zero point five one dollars of hourly wage per additional year of education, holding female fixed. The units are important. Wage is in dollars per hour and education is in years, so this is not a percentage and not five point one percent.

The answer is simple only because education enters once and linearly. There is no squared term, log, or interaction that would add another coefficient or make the derivative depend on another variable.

</details>

What is the marginal impact of $educ$?

<br>
<details>
  <summary>Answer</summary>
0.51
</details>

##### Question 2

<details class="transcript">
<summary>Transcript</summary>

Question two asks whether the marginal impact varies with the level of education. In this fitted equation the answer is no. The derivative with respect to education is the constant coefficient zero point five one, and education itself does not appear in that derivative. The model therefore assigns the same fitted increase in hourly wage to an extra year after eight years of schooling as to an extra year after eighteen.

Notice what kind of statement this is. We did not use the data to test whether those two marginal effects are equal. We made them equal by writing a model with only one linear education term. A straight-line specification has a constant slope by construction. The answer panel correctly says the model assumed the marginal impact is constant.

That distinction matters whenever you interpret a regression. A restriction implied by functional form is not an empirical finding.

</details>

Does the marginal impact of education vary depending on the level of education? 

<br>
<details>
  <summary>Answer</summary>
No, the model we estimated assumed that the marginal impact of education is constant.
</details>

:::
<!--end of panel-->

#### Testing?

<details class="transcript">
<summary>Transcript</summary>

In the linear model, the marginal effect of education is beta education itself. To ask whether education has a nonzero marginal impact, write the null as beta education equals zero and the two-sided alternative as beta education does not equal zero. That is a restriction on one population coefficient, so it is the ordinary t-test printed on the education row of the regression table.

The t-statistic is beta hat education minus the null value zero, divided by the standard error of beta hat education. Its sign tells you the estimated direction, and its magnitude tells you how many standard errors the estimate lies from zero. The p-value measures how surprising a statistic at least that extreme would be under the null. In the output from the model tab, the robust t-statistic is about eight point four six and the p-value is essentially zero at conventional display precision, so we reject the null.

Nothing extra is required because the parameter of interest and one reported coefficient are the same object.

</details>

You can just test if $\hat{\beta}_{educ}$ (the marginal impact of education) is statistically significantly different from $0$, which is just a t-test.

:::
<!--end of panel-->

### The quadratic case

::: {.panel-tabset}

#### Issue

<details class="transcript">
<summary>Transcript</summary>

The three bullets summarize what changes in the quadratic specification. First, the marginal impact of education depends on the current education level because the derivative is beta education plus two times beta education squared times education. Two people at different schooling levels generally have different fitted slopes even though they are evaluated using the same estimated model.

Second, there is no single test of "the marginal effect" that applies universally. A test at education equal to four concerns beta education plus eight beta education squared. A test at education equal to ten concerns beta education plus twenty beta education squared. Those are different population quantities and have different estimates and standard errors. Rejecting one null does not logically imply rejecting the other.

Third, you need to choose evaluation levels and test the relevant combination at each one.

</details>

With the quadratic specification

+ The marginal impact of education varies depending on your education level

+ There is no single test that tells you whether the marginal impact of education is statistically significant universally

+ Indeed, you need different tests for different education levels


#### Example 1

<details class="transcript">
<summary>Transcript</summary>

The research question here is local and two-sided: for a person currently at four years of education, does one additional year have a positive or negative marginal impact that differs from zero? Begin with the derivative beta education plus two times beta education squared times education. Substituting four gives beta education plus eight beta education squared.

The null is therefore beta education plus eight beta education squared equals zero, and the alternative says that same combination does not equal zero. The callout emphasizes that hypotheses concern the population parameters beta, without hats. We do not hypothesize that the random estimates equal zero. We use the estimates to form evidence about the unknown population combination.

Now answer the multiple-choice prompt. This is not a test of one coefficient because two coefficients appear. It is also not a test of multiple equations. It is one equation, or one linear restriction, involving multiple coefficients. A single restriction can be tested with a t-statistic. An equivalent one-restriction F-statistic is the square of that t-statistic when both use the same variance estimate and denominator degrees of freedom.

Look at the t formula. The numerator is beta hat education plus eight beta hat education squared, the estimated marginal effect at four under a zero null. The denominator is the standard error of that whole combination. If b one and b two denote the two estimates, its variance is variance b one plus sixty-four times variance b two plus sixteen times covariance of b one and b two. That covariance is why dividing the numerator by either individual standard error would be wrong. The next tab uses software to perform this linear-restriction calculation.

</details>

**Marginal impact of education**

$\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 2 \times educ$

<br>

**Hypothesis testing**

Does an additional year of education have a statistically significant impact (positive or negative) if your current education level is 4?

+ $H_0$: $\beta_{educ} + \beta_{educ^2} \times 2 \times 4 =0$

+ $H_1$: $\beta_{educ} + \beta_{educ^2} \times 2 \times 4 \ne 0$

:::{.callout-note}
Hypotheses are statements about the <span style = "color: blue;">population</span> parameters $\beta$, never about the estimates $\hat{\beta}$. We use the estimates to build the test statistic below.
:::

<br>

**Question**

Is this

+ test of a single coefficient? (t-test)
+ test of a single equation with multiple coefficients? (t-test)
+ test of multiple equations with multiple coefficients? (F-test)

**t-statistic**

$t = \frac{\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 2 \times 4}{se(\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 2 \times 4)} = \frac{\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 8}{se(\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 8)}$

#### R implementation

<details class="transcript">
<summary>Transcript</summary>

The car colon colon linearHypothesis function tests restrictions on coefficients from an estimated model. Its first argument is quad_reg, the quadratic wage model. Its second argument is a character string representing the null restriction. The intended string says the education coefficient plus eight times the squared-education coefficient equals zero. The coefficient names in such a string must match the names stored in the fitted model.

The note above the cell explains the one-restriction equivalence. A t-test directly standardizes one estimated linear combination. An F-test for exactly one restriction squares that t-statistic, so the two-sided t-test and the one-restriction F-test give the same p-value when constructed from the same covariance estimate. With heteroskedasticity-robust covariance estimation, software may label the large-sample squared statistic as chi-squared rather than finite-sample F, but the restriction being tested is still the one displayed on the previous tab.

The slide reports a p-value of zero point three nine four. That is much larger than common significance levels such as zero point zero five, so we do not reject the null. This does not prove that the marginal effect at four is exactly zero. It says the estimate is not far enough from zero, relative to its uncertainty, to distinguish it statistically in this sample. The conclusion is specific to education equal to four. It does not tell us whether the effect is significant at ten, which is the next example.

</details>

Remember, a trick to do this test using R is take advantage of the fact that $F_{1, n-k-1} \sim t_{n-k-1}^2$.

```{webr-r}
car::linearHypothesis(quad_reg, "educ + 8*I(educ^2)=0")
```

The p-value is $0.394$, so we do not reject the null: at an education level of 4, the marginal impact of education is not statistically distinguishable from zero.

#### Example 2

<details class="transcript">
<summary>Transcript</summary>

Now evaluate the same quadratic derivative at education equal to ten. Start with beta education plus two times beta education squared times education. Substituting ten changes the multiplier on the squared-term coefficient to twenty, so the marginal effect parameter is beta education plus twenty beta education squared.

The null says that one linear combination equals zero, and the two-sided alternative says it does not. As in example one, this is one equation involving multiple coefficients. It is therefore a single-restriction t-test, or equivalently a one-restriction F-type test. The fact that two coefficients appear does not by itself make the null a multiple-restriction hypothesis.

The t-statistic shown at the bottom divides beta hat education plus twenty beta hat education squared by the standard error of that same combination. Its variance is variance of beta hat education, plus four hundred times the variance of beta hat education squared, plus forty times their covariance. Notice how changing the evaluation point changes both the estimated marginal effect and its standard error. We are not reinterpreting the test from education equal to four. We are testing a different population quantity using the same underlying regression estimates and covariance matrix.

</details>

**Marginal impact of education**

$\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 2 \times educ$

<br>

**Hypothesis testing**

Does an additional year of education have a statistically significant impact (positive or negative) if your current education level is 10?    

+ $H_0$: $\beta_{educ} + \beta_{educ^2} \times 2 \times 10 =0$

+ $H_1$: $\beta_{educ} + \beta_{educ^2} \times 2 \times 10 \ne 0$

**Question**

Is this

+ test of a single coefficient? (t-test)
+ test of a single equation with multiple coefficients? (t-test)
+ test of multiple equations with multiple coefficients? (F-test)

**t-statistic**

$t = \frac{\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 2 \times 10}{se(\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 2 \times 10)} = \frac{\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 20}{se(\hat{\beta}_{educ} + \hat{\beta}_{educ^2} \times 20)}$

#### R implementation

<details class="transcript">
<summary>Transcript</summary>

This cell uses the same linearHypothesis approach as example one, replacing the weight eight with twenty because the evaluation point is now ten years of education. The first argument remains quad_reg, so we are not re-estimating the wage equation. We are asking a new question of the same coefficient vector and covariance matrix. The character string represents the null that the education coefficient plus twenty times the squared-education coefficient equals zero.

The reported p-value is far below zero point zero one. We therefore reject the null at the one percent level and conclude that the marginal effect at ten years is statistically distinguishable from zero. The fitted combination is positive, so the estimated direction is a positive association between an additional year of education and hourly wage at that education level.

</details>

```{webr-r}
car::linearHypothesis(quad_reg, "educ + 20*I(educ^2)=0")
```

The p-value is far below $0.01$, so we reject the null at the 1% level: at an education level of 10, the marginal impact of education is statistically significant.

:::
<!--end of panel-->

:::
<!--end of panel-->

# Interaction terms

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Interaction terms

::: {.panel-tabset}

### What is it?

<details class="transcript">
<summary>Transcript</summary>

An interaction term is a new regressor formed by multiplying two variables. The example is education times experience. Including this product allows the marginal effect of one variable to depend on the level of the other, which a purely additive model cannot do.

Read the displayed wage model carefully. It contains an intercept, experience by itself, and education times experience. Differentiate with respect to experience and you will obtain beta one plus beta two times education, so the return to another year of experience changes with schooling. Equivalently, differentiating with respect to education gives beta two times experience, so the association with schooling changes with experience. Interaction effects are symmetric in that algebraic sense even if the research question emphasizes one direction.

The important callout warns that this is a deliberately stripped-down example. Education does not enter on its own. That omission forces education to affect wage only through experience. When experience equals zero, the fitted marginal effect of education must be exactly zero. It also makes beta one the experience slope at education equal to zero. Those are strong restrictions chosen for a short derivation, not general modeling advice.

The rule of thumb in blue is the hierarchy principle: when you include x one times x two, ordinarily include x one and x two as separate main effects too. Then each main-effect coefficient describes a slope when the other variable is zero, and the interaction coefficient describes how that slope changes.

</details>

A variable that is a multiplication of two variables 

<br>

**Example**

$educ\times exper$

<br>

**A model with an interaction term**

$wage = \beta_0 + \beta_1 exper + \beta_2 educ \times exper + u$

<br>

:::{.callout-important title="A deliberately stripped-down model"}
Notice that $educ$ does <span style = "color: red;">not</span> enter on its own here. This keeps the algebra on the next slides as simple as possible, but it is <span style = "color: blue;">not</span> what you should do in your own work.

**Rule of thumb:** whenever you include an interaction $x_1 \times x_2$, include $x_1$ and $x_2$ as separate terms too. Omitting a main effect forces the marginal impact of $exper$ to be exactly $\beta_1$ when $educ = 0$ and constrains $educ$ to affect $wage$ only through $exper$ — a restriction you almost never want to impose.
:::

### Marginal impact

<details class="transcript">
<summary>Transcript</summary>

Start from the stripped-down model on the previous tab: wage equals beta zero plus beta one times experience plus beta two times education times experience plus u. Differentiate with respect to experience. Beta zero drops out, beta one experience contributes beta one, and beta two education experience contributes beta two times education. The marginal impact is therefore beta one plus beta two education, exactly as displayed.

The first implication follows immediately: the experience slope is not constant. Evaluate it at a particular schooling level. Beta one is the marginal effect of experience when education equals zero. That is the reference value built into the parameterization, not an overall experience effect. If nobody in the relevant population has zero education, beta one alone is an extrapolated slope and may have little substantive meaning.

Beta two tells you how much the experience slope changes when education increases by one year. Its units are dollars of hourly wage per year of experience per year of education. If beta two is positive, another year of experience is associated with a larger wage increment for people with more schooling, so the two inputs reinforce each other in the fitted model. If beta two is negative, the experience return declines with schooling.

</details>

**Marginal impact of experience**:

$\frac{\partial wage}{\partial exper} = \beta_1+\beta_2\times educ$

<br>

**Implications**
  
The marginal impact of experience depends on education

+ $\beta_1$: the marginal impact of experience when $educ=0$

+ if $\beta_2>0$: additional year of experience is worth more when you have more years of education

### R implementation

<details class="transcript">
<summary>Transcript</summary>

The code implements the stripped-down interaction model exactly. Fixest colon colon feols estimates wage on female, experience, and the literal product of experience and education. Female is included as a level control. Experience is the main effect shown in the algebra. Capital I around experience times education tells the formula parser to perform arithmetic multiplication and create only that product column. Without I, an asterisk in an R formula expands to both main effects plus their interaction.

The assignment stores the fitted model as reg_int so later tabs can extract coefficients, make predictions, and test restrictions. There is no vcov argument in this call, so fixest reports its default IID standard errors rather than the heteroskedasticity-robust standard errors used in the quadratic example. The surrounding parentheses print the model immediately after assigning it. Autorun evaluates the cell automatically.

When run on wage1, the estimates are approximately six point one two for the intercept, negative two point four one eight for female, negative zero point one eight eight for experience, and positive zero point zero two zero four for experience times education. The interaction estimate means each additional year of education raises the fitted marginal effect of a year of experience by about two cents per hour. The negative experience coefficient is the slope at education equal to zero, so it should not be described as the experience effect for a typical worker.

The hidden knitr chunk fits the identical model for the static figures on the next tab; it does not add another specification.

</details>

Just like the quadratic case with $educ^2$, you can use `I()`.

```{webr-r}
#| autorun: true
(
  reg_int <- fixest::feols(wage ~ female + exper + I(exper * educ), data = wage1)
)
```

```{r}
#| include: false
reg_int <- fixest::feols(wage ~ female + exper + I(exper * educ), data = wage1)
```

### Estimated model

<details class="transcript inline">
<summary>Transcript</summary>

The equation at the top reports the fitted model from the previous tab. Wage equals six point one two one, minus two point four one eight times female, minus zero point one eight eight times experience, plus zero point zero two zero times education times experience. Differentiating with respect to experience gives the displayed marginal effect, negative zero point one eight eight plus zero point zero two zero times education.

The code creates one thousand education values from the sample minimum to maximum and evaluates that expression at each value. In the left panel, the horizontal axis is education in years and the vertical axis is the marginal impact of experience, measured as dollars of hourly wage per additional year of experience. The horizontal reference line marks zero. The fitted effect begins negative at low education, increases linearly because the interaction coefficient is positive, and crosses zero at roughly zero point one eight eight divided by zero point zero two zero, about nine point two years of education. Above that value, the fitted experience slope is positive.

The right panel is a histogram of education. Its horizontal axis is again years of education and its vertical axis is the count of sample observations in each bin. Read the two panels together. The effect curve tells you what the model implies at each education value; the histogram tells you how much data support those evaluations. An effect near an education level with very few observations depends heavily on functional-form extrapolation.

</details>

**Estimated Model**

$wage = 6.121 - 2.418 \times female - 0.188 \times exper + 0.020 \times educ \times exper$ 

<br>

**Marginal impact of experience**

$\frac{\partial wage}{\partial exper} = - 0.188 + 0.020 \times educ$

```{r echo = F}
N <- 1000
educ_temp <- seq(min(wage1$educ), max(wage1$educ), length = N)
min_exper <- reg_int$coefficients["exper"] + reg_int$coefficients["I(exper * educ)"] * educ_temp
data_exper <- data.table(
  x = educ_temp,
  y = min_exper
)

g_marginal <-
  ggplot(data = data_exper) +
  geom_line(aes(x = x, y = y)) +
  ylab("Marginal impact\nof experience") +
  xlab("Education (years)") +
  geom_hline(yintercept = 0) +
  theme_lecture

g_hist <-
  ggplot(data = wage1) +
  geom_histogram(aes(educ)) +
  ylab("Count") +
  xlab("Education (years)") +
  theme_lecture
```

::: {.columns}

::: {.column width="50%"}
Marginal impact of $exper$:

```{r}
#| echo: false
#| out-width: "90%"
#| fig-width: 7.5
#| fig-height: 5.5
g_marginal
```
:::
<!--end of the 1st column-->
::: {.column width="50%"}
Histogram of education:

```{r}
#| echo: false
#| out-width: "90%"
#| fig-width: 7.5
#| fig-height: 5.5
g_hist
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

### Hypothesis testing

::: {.panel-tabset}

#### Hypothesis

<details class="transcript">
<summary>Transcript</summary>

This slide applies the quadratic-section logic to an interaction. Because the marginal impact of experience is negative zero point one eight eight plus zero point zero two zero times education, it varies with schooling. Substituting ten years gives about zero point zero one two dollars per hour for an additional year of experience using the rounded coefficients on screen. Substituting fifteen gives about zero point one one two dollars. These calculations are fitted marginal effects, not significance tests.

The research question below selects education equal to ten and asks whether the experience effect is positive or negative and statistically different from zero. In population notation, the null is beta experience plus ten times beta experience-education equals zero. The alternative says that combination is not zero. The interaction coefficient's label stands for the coefficient on experience times education.

This is one linear restriction involving two population coefficients. A t-test of the combination is appropriate, or an equivalent one-restriction F-type test. Testing beta experience alone would ask about the experience slope at zero education. Testing the interaction alone would ask whether the experience slope changes with education. Neither is the question posed here.

The same framework can test fifteen years by replacing ten with fifteen. The next tab asks R for the ten-year restriction.

</details>

Just like the case of the quadratic specification of education, marginal impact of experience is not constant

We can test if the marginal impact of experience is statistically significant for a given level of education

  * When $educ=10$, $\frac{\partial wage}{\partial exper} = - 0.188 + 0.020 \times 10=0.012$
  * When $educ=15$, $\frac{\partial wage}{\partial exper} = - 0.188 + 0.020 \times 15=0.112$

**Question**

Does an additional year of experience have a statistically significant impact (positive or negative) if your current education level is 10

<br>

**Hypothesis**

+ $H_0$: $\beta_{exper} + \beta_{exper\_educ} \times 10=0$

+ $H_1$: $\beta_{exper} + \beta_{exper\_educ} \times 10\ne 0$

#### R implementation

<details class="transcript">
<summary>Transcript</summary>

The code passes reg_int and one restriction string to car colon colon linearHypothesis. The string says the experience coefficient plus ten times the coefficient named I of experience times education equals zero. That is exactly the null from the previous tab at education equal to ten. Coefficient names inside the string must match the fitted model's labels, including the capital I and the product term.

Using the unrounded estimates, the marginal effect at ten is about negative zero point one eight seven six plus ten times zero point zero two zero three seven, or positive zero point zero one six one dollars per hour. The test output gives a p-value of about zero point one one seven. At the five percent level, we therefore fail to reject the null that the experience effect is zero for a worker with ten years of education. The point estimate is positive, but it is not estimated precisely enough to distinguish from zero at that conventional level.

Do not turn "fail to reject" into proof of no effect. It means the sample and model do not provide sufficiently strong evidence against zero for this particular linear combination.

</details>

```{webr-r}
#| autorun: true
car::linearHypothesis(reg_int, "exper+10*I(exper * educ)=0")
```

:::
<!--end of panel-->


:::
<!--end of panel-->

# Including qualitative information

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Including qualitative information

::: {.panel-tabset}

### Issue

<details class="transcript">
<summary>Transcript</summary>

So far our explanatory variables have naturally been numerical, such as education and experience. Many economically relevant characteristics are qualitative categories rather than measured amounts. The examples on screen include sex, marital status, and highest degree. Other examples would be region, crop type, treatment status, or occupation.

A regression still needs numerical columns, so we represent categories with indicator variables.

</details>

**Issue**

How do we include qualitative information as an independent variable?

<br>

**Examples**

+ male or female (binary)

+ married or single (binary)

+ high-school, college, masters, or Ph.D (more than two states)

### Binary variables


::: {.panel-tabset}

#### Dummy variable

<details class="transcript">
<summary>Transcript</summary>

The callout defines a dummy variable, also called an indicator or binary variable. It takes exactly two numerical values. One marks the category named by the variable and zero marks the other state. For example, in this dataset female equals one for women and zero for men; married equals one for married respondents and zero otherwise. The choice of which state receives one determines the sign and wording of the comparison, not the model's fitted values.

The code beneath the definition uses dplyr colon colon select to keep six columns from wage1: wage, education, experience, female, and married, along with the listed variables, then pipes the result to head. Head displays the first six rows by default. Look across a row to see that continuous variables retain their measured values while female and married appear as zeros and ones. The pipe passes the selected table into head, making this a quick inspection rather than a transformation saved back to the data.

</details>

:::{.callout-note title="Dummy variable"}
+ Relevant information in binary variables can be captured by a <span style = "color: red;">zero-one</span> variable that takes the value of $1$ for one state and $0$ for the other state

+ We use "dummy variable" to refer to a binary (zero-one) variable
:::

<br>

**Example**

```{webr-r}
dplyr::select(wage1, wage, educ, exper, female, married) %>%
  head()
```

#### Model with a dummy variable

<details class="transcript">
<summary>Transcript</summary>

The model is wage equals beta zero plus sigma f times female plus beta two times education plus u. Under the zero conditional mean assumption, the conditional expectation drops u. To interpret sigma f, evaluate the equation for each value of the dummy while keeping education at the same value.

For female equal to one, the conditional expected wage is beta zero plus sigma f plus beta two education. For female equal to zero, the sigma term disappears and expected wage is beta zero plus beta two education. Because zero identifies men in this coding, men are the baseline group. Subtract the second expression from the first. Beta zero and beta two education cancel, leaving sigma f. That is the equation displayed in the middle of the slide.

Verbally, sigma f is the expected hourly-wage difference between women and men who have the same education in this model. If sigma f is negative, women have the lower fitted conditional mean by its absolute value; if positive, they have the higher one. The comparison is conditional because education is held fixed. It is not the raw difference between average female and male wages, which could also reflect different education distributions.

Notice another restriction. The education coefficient beta two is identical for both groups, so this model permits different intercepts but forces parallel wage-education lines.

</details>

**Model**

$wage = \beta_0 +\sigma_f female +\beta_2 educ + u$

<br>

**Interpretation**

+ `female`: $E[wage|female=1,educ] = \beta_0 + \sigma_f +\beta_2 educ$

+ `male`: $E[wage|female=0,educ] = \beta_0 + \beta_2 educ$

<br>

This means that 

$\sigma_f = E[wage|female=1,educ]-E[wage|female=0,educ]$

Verbally,

+ $\sigma_f$ is the difference in the expected wage conditional on education between female and male

+ $\sigma_f$ measures how much more (less) female workers make compared to male workers (<span style = "color: blue;"> baseline </span>) if they were to have the same education level 

#### R implementation

<details class="transcript">
<summary>Transcript</summary>

The code estimates the dummy-variable model with fixest colon colon feols. Wage is the dependent variable, and female plus education are the regressors. The data argument uses wage1. The result is stored as reg_df and then printed. Because no vcov option appears, the output uses fixest's default IID standard errors.

The fitted female coefficient is about negative two point two seven three. Since female equals one for women and zero for men, read it as follows: among observations with the same education, the fitted hourly wage for women is about two dollars and twenty-seven cents lower than for men, on average. The inline R expression below the output pulls that exact coefficient from reg_df by its name, so the prose stays synchronized with the estimate rather than hardcoding a number.

The education coefficient is about zero point five zero six dollars per hour per additional year, and the intercept is about zero point six two three. Those terms define the baseline male line. Adding negative two point two seven three shifts that entire line downward for women. Because there is no female-times-education interaction, both groups share the same education slope.

</details>

**R implementation**

```{webr-r}
reg_df <- fixest::feols(wage ~ female + educ, data = wage1)

reg_df
```

```{r}
#| include: false
reg_df <- fixest::feols(wage ~ female + educ, data = wage1)

reg_df
```

<br>

**Interpretation**

Female workers make `r reg_df$coefficients["female"]` ($/hour) less than male workers on average even though they have the same education level. 

#### Visualization

<details class="transcript inline">
<summary>Transcript</summary>

This figure draws the two fitted regression lines implied by the dummy-variable model. The horizontal axis is education in years and the vertical axis is hourly wage in dollars. The legend distinguishes female and male. The male line is beta zero plus beta two times education because female equals zero. The female line is beta zero plus sigma f plus beta two times education because female equals one. The text labels on the plot write those two equations next to their lines.

Both lines have slope beta two, so they are parallel. Their vertical separation at every education value is sigma f. In this fitted model sigma f is negative, so the female line lies below the male line by about two point two seven dollars per hour everywhere. The code constructs education values from zero to twenty, evaluates each equation, combines the two data tables, and maps line color to group. The vertical and horizontal reference lines mark zero on the axes; they are coordinate guides, not estimated effects.

The picture reveals the restriction more clearly than the table. Adding a dummy by itself allows the intercept to differ across groups but does not allow the education slope to differ.

</details>

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6.5
beta_0 <- reg_df$coefficients[1] # intercept
sigma_0 <- reg_df$coefficients["female"] # coef on female
beta_1 <- reg_df$coefficients["educ"] # coef on educ
x <- seq(0, 20, length = 1000)
fe_data <- data.table(x = x, y = beta_1 * x + beta_0 + sigma_0, type = "female")
ma_data <- data.table(x = x, y = beta_1 * x + beta_0, type = "male")
plot_data <- rbind(fe_data, ma_data)

text_data <- data.table(
  x = c(5, 12), y = c(6, 1),
  label = c(
    "wage == beta[0]+beta[2]*educ", "wage == beta[0]+sigma[f]+beta[2]*educ"
  )
)

g_comp <- ggplot(data = plot_data) +
  geom_line(aes(x = x, y = y, color = type)) +
  geom_text(
    data = text_data, aes(x, y, label = label),
    size = 4.3, parse = TRUE, family = "Times"
  ) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  ylab("Hourly Wage ($)") +
  xlab("Education (years)") +
  scale_color_discrete(name = "") +
  theme_lecture +
  theme(
    legend.position = "bottom"
  )
g_comp
```

:::
<!--end of panel-->

### Use `male` instead?

::: {.panel-tabset}

#### Model and theory

<details class="transcript">
<summary>Transcript</summary>

Now recode the same two categories with male equal to one and female equal to zero. The model is wage equals beta zero plus sigma m times male plus beta two education plus u. The algebra is identical, but the baseline changes.

For male equal to one, expected wage is beta zero plus sigma m plus beta two education. For male equal to zero, which now means female, expected wage is beta zero plus beta two education. Subtracting the female expression from the male expression leaves sigma m. Thus sigma m is the conditional expected wage difference, male minus female, at the same education level.

Compare this with the earlier female coding. Sigma f represented female minus male, so sigma m must be the negative of sigma f. The new intercept is the old intercept plus sigma f because the baseline line has switched from men to women. The education slope and every fitted wage remain unchanged. Coding changes the parameter labels and signs, not the substantive model.

The callout gives the general rule: whichever category receives zero is the baseline.

</details>

**Model**

$wage = \beta_0 +\sigma_m male +\beta_2 educ + u$

<br>

**Interpretation**

+ `male`: $E[wage|male = 1,educ] = \beta_0 + \sigma_m +\beta_2 educ$

+ `female`: $E[wage|male = 0,educ] = \beta_0 + \beta_2 educ$

<br>

This means that 

$\sigma_m = E[wage|male=1,educ]-E[wage|male=0,educ]$

Verbally,

+ $\sigma_m$ is the difference in the expected wage conditional on education between male and female

+ $\sigma_m$ measures how much more (less) male workers make compared to female workers (<span style = "color: blue;">baseline</span>) if they were to have the same education level 

:::{.callout-important}
Whichever status that is given the value of $0$ becomes the baseline
:::

#### Regression results

<details class="transcript">
<summary>Transcript</summary>

The first code line creates male as one minus female. Because female is zero or one, this reverses the coding exactly: a female value of zero becomes male equal to one, and a female value of one becomes male equal to zero. Dplyr mutate returns the updated wage1 data with the new column.

The feols call then regresses wage on male and education. Wage1 is supplied through the data argument, and the fitted object is again named reg_df. The output shows a male coefficient of about positive two point two seven three, exactly the opposite sign of the earlier female coefficient. The intercept is about negative one point six five one rather than positive zero point six two three because it now describes the fitted female wage at zero education. The education coefficient remains about zero point five zero six.

The inline interpretation extracts the male coefficient by name. With women as the zero-coded baseline, men have a fitted hourly wage about two dollars and twenty-seven cents higher at the same education level. This says the same thing as women having a fitted wage two dollars and twenty-seven cents lower than men.

</details>

**Regression results**

```{webr-r}
#| autorun: true
wage1 <- dplyr::mutate(wage1, male = 1 - female)

reg_df <- fixest::feols(wage ~ male + educ, data = wage1)

reg_df
```

<br>

**Interpretation**

Male workers make `r reg_df$coefficients["male"]` ($/hour) more than female workers on average even though they have the same education level. 

#### Question

<details class="transcript">
<summary>Transcript</summary>

Before opening the answer, write the identity connecting these columns. For every observation, male plus female equals one. A regression intercept is itself a column of ones. Therefore the intercept column equals male plus female exactly.

If the model includes an intercept, male, and female, one regressor column is a perfect linear combination of the others. This is perfect multicollinearity. The design matrix does not have full column rank, so the normal equations cannot identify a unique coefficient on all three columns. Many coefficient triples produce the same fitted values because you can shift the intercept and offset both dummy coefficients without changing a prediction.

The answer panel lists the consequences. Male and female contain redundant information once the intercept is present. To estimate the model, drop either male or female and interpret the remaining indicator relative to the omitted group.

</details>

What do you think will happen if  we include both male and female dummy variables?

<br>
<details>
  <summary>Answer</summary>
+ They contain redundant information

+ Indeed, including both of them along with the intercept would cause <span style = "color: blue;">perfect collinearity problem</span>

+ So, you <span style = "color: blue;"> need to </span>drop either one of them
</details>

#### Perfect Collinearity

<details class="transcript">
<summary>Transcript</summary>

The equation above the code states the exact dependence: intercept equals male plus female. The feols formula nevertheless requests wage on male, female, and education. Because those columns are perfectly collinear, all requested coefficients cannot be estimated uniquely.

When this cell runs, fixest detects the collinearity and removes female. It estimates the remaining model with an intercept, male, and education, then prints a message that one variable was removed because of collinearity. The reported coefficients match the male-baseline-choice regression from the previous tab because that is the estimable parameterization fixest retained.

</details>

In the model, $intercept = male + female$, which causes perfect collinearity.

Here is what happens if you include both:

```{webr-r}
#| autorun: true
reg_dmf <- fixest::feols(wage ~ male + female + educ, data = wage1)

reg_dmf
```

<br>

One of the variables that cause perfect collinearity is automatically dropped.

:::
<!--end of panel-->

:::
<!--end of panel-->

# Interactions with a dummy variable

## Interactions with a dummy variable

::: {.panel-tabset}

### Issue

<details class="transcript">
<summary>Transcript</summary>

The first bullet recalls the restriction in the dummy-only model: women and men could have different fitted wage levels, but education had exactly the same slope for both. The question here is whether we can allow the marginal association between education and wage to differ by group. We can, by interacting the female indicator with education while retaining both main effects.

Read the flexible model term by term. Beta zero is the male intercept because male is the female-equals-zero baseline. Sigma f times female allows the female intercept to differ. Beta two times education is the male education slope. Gamma times female times education allows that slope to change for women.

Substitute female equal to zero. Both terms containing female disappear, leaving expected wage for men equal to beta zero plus beta two education. Substitute female equal to one. The intercept becomes beta zero plus sigma f, and the two education terms combine into beta two plus gamma times education. Thus the model gives each group its own intercept and slope.

Gamma is the female education slope minus the male education slope. If gamma is positive, the fitted return to another year of schooling is larger for women; if negative, it is smaller. Sigma f now has a narrower interpretation than in the parallel-lines model: it is the female-minus-male wage difference at education equal to zero. At education level e, the fitted group gap is sigma f plus gamma times e, so it can widen, narrow, or cross zero.

</details>

+ In the previous example, the impact of education on wage was modeled to be exactly the same

+ Can we build a more flexible model that allows us to estimate the differential impacts of education on wage between male and female?

**A more flexible model**

$wage = \beta_0 + \sigma_f female +\beta_2 educ + \gamma female\times educ + u$

+ `female`: $E[wage|female=1,educ] = \beta_0 + \sigma_f +(\beta_2+\gamma) educ$
+ `male`: $E[wage|female=0,educ] = \beta_0 + \beta_2 educ$

<br>

**Interpretation**

For female, education is more effective by $\gamma$ than it is for male.

### Example using R

<details class="transcript">
<summary>Transcript</summary>

The feols formula includes female, education, and I of female times education. Including all three respects the hierarchy rule: the two main effects remain alongside their product. Capital I tells the formula parser to multiply the two columns literally. The data argument uses wage1, and the surrounding parentheses cause the fitted model, stored as reg_di, to print immediately.

The output gives an intercept of about zero point two zero zero, a female coefficient of about negative one point one nine nine, an education coefficient of about zero point five three nine, and an interaction coefficient of about negative zero point zero eight six. No vcov argument is supplied, so these are paired with the default IID standard errors.

Interpret the slopes by substituting dummy values. For men, female equals zero, so the marginal effect of education is beta education, about fifty-four cents of hourly wage per additional year. For women, female equals one, so the marginal effect is beta education plus beta interaction, about zero point five three nine minus zero point zero eight six, or roughly forty-five cents. The interaction estimate is the difference between those slopes: the fitted education slope is about eight point six cents lower for women.

The female main-effect coefficient is not an overall group gap. In this interacted model it is the fitted female-minus-male difference at zero years of education. At education e, the gap is negative one point one nine nine minus zero point zero eight six times e. Also check uncertainty. The interaction t-statistic is about negative zero point eight three with p about zero point four zero seven, so this sample does not provide evidence that the slopes differ from zero at conventional levels.

</details>

```{webr-r}
(
  reg_di <- fixest::feols(wage ~ female + educ + I(female * educ), data = wage1)
)
```


### Interpretation

<details class="transcript inline">
<summary>Transcript</summary>

The statement above the graph interprets gamma, the coefficient on female times education. Its estimate is negative zero point zero eight six, so the fitted marginal benefit of one more year of education is about eight point six cents per hour lower for women than for men. This is a difference in slopes, not the wage gap at a particular schooling level.

The graph has education in years on the horizontal axis and hourly wage in dollars on the vertical axis. The male line uses beta zero plus beta two education. The female line uses beta zero plus sigma f plus beta two plus gamma times education. The labels printed beside the lines show those equations, and the color legend identifies the groups. The vertical and horizontal lines mark zero on the axes.

Unlike the earlier dummy-only picture, these lines are not parallel. The male slope is about zero point five three nine, while the female slope is about zero point four five three. Because gamma is negative, the female line rises more slowly. Sigma f is also negative, so the female line starts below the male line at zero education and the fitted gap becomes more negative as education rises. At any education e, that gap is sigma f plus gamma e.

</details>

The marginal benefit of education is 0.086 ($/hour) less for females workers than for male workers on average.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6.5
reg_di <- fixest::feols(wage ~ female + educ + I(female * educ), data = wage1)

beta_0 <- reg_di$coefficients[1] # intercept
sigma_0 <- reg_di$coefficients["female"] # coef on female
beta_1 <- reg_di$coefficients["educ"] # coef on educ
gamma <- reg_di$coefficients["I(female * educ)"]
x <- seq(0, 20, length = 1000)
fe_data <- data.table(x = x, y = (beta_1 + gamma) * x + beta_0 + sigma_0, type = "female")
ma_data <- data.table(x = x, y = beta_1 * x + beta_0, type = "male")
plot_data <- rbind(fe_data, ma_data)

text_data <- data.table(
  x = c(5, 12), y = c(6, 1),
  label = c(
    "wage == beta[0]+beta[2]*educ", "wage == beta[0]+sigma[f]+(beta[2]+gamma)*educ"
  )
)

g_comp_i <- ggplot(data = plot_data) +
  geom_line(aes(x = x, y = y, color = type)) +
  geom_text(
    data = text_data, aes(x, y, label = label),
    size = 4.3, parse = TRUE, family = "Times"
  ) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  ylab("Hourly Wage ($)") +
  xlab("Education (years)") +
  scale_color_discrete(name = "") +
  theme_lecture +
  theme(
    legend.position = "bottom"
  )

g_comp_i
```


:::
<!--end of panel-->


## Categorical variable: more than two states

::: {.panel-tabset}

### Issue

<details class="transcript">
<summary>Transcript</summary>

Now move from a binary variable to a categorical variable with three states. The example degree variable records whether the highest degree is college, master's, or doctoral. A single zero-one variable cannot distinguish all three states, so we define one indicator for each category: college equals one only for college graduates, master equals one only for master's graduates, and doctor equals one only for Ph.D. graduates.

For every observation exactly one of those indicators equals one, so college plus master plus doctor equals one. With an intercept, including all three would recreate perfect collinearity because their sum equals the intercept column. The rule is therefore to include the number of categories minus one. With three states, include two indicators and omit one.

</details>

+ Consider a variable called $degree$ which has three status values:  college, master, and doctor.

+ Unlike a binary variable, there are three status values. 

+ How do we include a categorical variable like this in a model?

**What do we do about this?**

You can create three dummy variables like below:

+ `college`: 1 if the highest degree is college, 0 otherwise
+ `master`: 1 if the highest degree is Master's, 0 otherwise
+ `doctor`: 1 if the highest degree is Ph.D., 0 otherwise

You then include two (the number of status values - 1) of the three dummy variables:

### Model

<details class="transcript">
<summary>Transcript</summary>

The model includes master and doctor indicators but omits college, so college is the baseline. It also includes education as a continuous regressor. Under the conditional mean assumption, evaluate the equation for each feasible indicator pattern while holding education fixed.

For a college graduate, master and doctor both equal zero. Expected wage is beta zero plus beta one education. For a master's graduate, master equals one and doctor equals zero, so expected wage is beta zero plus sigma m plus beta one education. For a Ph.D. graduate, master equals zero and doctor equals one, giving beta zero plus sigma d plus beta one education. The impossible combination master equals one and doctor equals one should not occur if highest degree is coded correctly.

Subtract the college expression from the master's expression and only sigma m remains. It is the fitted wage difference between master's and college graduates at the same education level. Sigma d similarly compares Ph.D. with college. Neither coefficient directly compares master's with Ph.D. That difference is sigma d minus sigma m.

The common beta one term means this model forces the same education slope across all degree categories. The red phrases emphasize relativity: every dummy coefficient must be interpreted against the omitted college baseline.

</details>

**Model**

$wage = \beta_0 + \sigma_m master +\sigma_d doctor + \beta_1 educ + u$

+ $college$: $E[wage|master=0, doctor = 0, educ] = \beta_0 + \beta_1 educ$
+ $master$: $E[wage|master=1, doctor = 0, educ] = \beta_0 + \sigma_m + \beta_1 educ$
+ $doctor$: $E[wage|master=0, doctor = 1, educ] = \beta_0 + \sigma_d + \beta_1 educ$

<br>

**Interpretation**

$\sigma_m$: the impact of having a MS degree <span style = "color: red;">relative to</span> having a <span style = "color: red;">college degree</span> 

$\sigma_d$: the impact of having a Ph.D. degree <span style = "color: red;">relative to</span> having a <span style = "color: red;">college degree</span>

<br>

**Important**

The omitted category (here, `college`) becomes the baseline.

:::
<!--end of panel-->

## Structural differences across groups

::: {.panel-tabset}

### Definition

<details class="transcript">
<summary>Transcript</summary>

Structural difference means that the conditional relationship itself can differ across groups, not merely that one group has a constant level shift. The example models cumulative college GPA separately for male and female athletes. Cumgpa is cumulative grade point average, sat is SAT score, hsperc is high-school rank percentile, and tothrs is total completed college credit hours.

Look at the two equations. The male model uses alpha zero through alpha three. The female model uses beta zero through beta three. Allowing every coefficient to differ means the groups may have different intercepts, different SAT slopes, different high-school-rank slopes, and different credit-hour slopes.

If the true group relationships differ this broadly, a single pooled equation with common slopes imposes false equality restrictions. Its common coefficients can become a weighted compromise between two functions and may describe neither group well.

</details>

Structural difference refers to the fundamental differences in the model of a phenomenon in the population:

<br>

**Example**

Male: $cumgpa = \alpha_0 + \alpha_1 sat + \alpha_2 hsperc + \alpha_3 tothrs + u$

Female: $cumgpa = \beta_0 + \beta_1 sat + \beta_2 hsperc + \beta_3 tothrs + u$

+ $cumgpa$: college grade points averages for male and female college athletes

+ $sat$: SAT score

+ $hsperc$: high school rank percentile

+ $tothrs$: total hours of college courses

<br>

**In this example,**

$cumgpa$ are determined in a fundamentally different manner between female and male students.

You do not want to run a single regression that fits a single model for both female and male students. 

### What to do?

<details class="transcript">
<summary>Transcript</summary>

If you suspect structural differences, do not simply assume separate models. Estimate a flexible unrestricted specification and test whether the extra group-difference terms are needed. The displayed equation does this inside one regression by using female as the group indicator and interacting it with every explanatory variable.

Read the equation line by line. Beta zero is the male intercept, and sigma zero times female permits the female intercept to differ. Beta one sat is the male SAT slope, while sigma one times sat times female changes that slope for women. The same pairing appears for hsperc and tothrs. All main effects are retained alongside their interactions, which satisfies the hierarchy rule.

Set female equal to zero and every sigma term disappears. You recover the male equation with beta zero, beta one, beta two, and beta three. Set female equal to one and collect terms. The female intercept becomes beta zero plus sigma zero; the SAT slope becomes beta one plus sigma one; the hsperc slope becomes beta two plus sigma two; and the credit-hours slope becomes beta three plus sigma three.

This parameterization is useful because each sigma is a female-minus-male difference in one structural coefficient. If all sigmas equal zero, the same regression function applies to both groups. If at least one differs, the functions are structurally different under this specification.

</details>

If you suspect that the underlying process of how the dependent variable is determined vary across groups, then you should test that hypothesis!

<br>

**To do so,**

You estimate the model that allows to estimate separate models across groups within a single regression analysis.

<br>

**A more flexible model**

$$cumgpa = \beta_0 + \sigma_0 female + \beta_1 sat + \sigma_1 (sat \times female)$$
          $$\;\; + \beta_2 hsperc + \sigma_2 (hsperc \times female)$$
          $$\qquad + \beta_3 tothrs + \sigma_3 (tothrs \times female) + u$$


### Interpretation

<details class="transcript">
<summary>Transcript</summary>

The first two lines are the group-specific conditional means obtained from the flexible model. For men, female equals zero, so expected cumulative GPA is beta zero plus beta one SAT plus beta two high-school percentile plus beta three total hours. For women, female equals one, so each corresponding coefficient is beta plus sigma. This confirms that one pooled interacted equation reproduces two fully separate linear functions.

The bullets summarize the parameter roles. The betas are the baseline male coefficients and also the common components of the female coefficients. Each sigma measures a female-minus-male difference: sigma zero for the intercept, sigma one for the SAT slope, sigma two for the hsperc slope, and sigma three for the tothrs slope.

The verbal null says the male and female GPA models are not structurally different. Mathematically, that means sigma zero, sigma one, sigma two, and sigma three all equal zero simultaneously. Count the restrictions: there are four equations. That makes this a joint test, conventionally presented as an F-test in a finite-sample linear model, rather than a t-test of one coefficient.

Four separate t-tests would not answer the same single joint question. They ignore covariance among estimates and create a multiple-testing issue. It is possible for no individual coefficient to cross the five percent threshold while the four coefficients are jointly far from zero. Open the answer and you see "F-test."

</details>

Male: $E[cumgpa] = \beta_0 + \beta_1 sat + \beta_2 hsperc + \beta_3 tothrs$
Female: $E[cumgpa] = (\beta_0 +\sigma_0) + (\beta_1+\sigma_1) sat + (\beta_2+\sigma_2) hsperc + (\beta_3+\sigma_3) tothrs$

<br>

**Interpretation**

+ $\beta$s are commonly shared by female and male students 
+ $\sigma$s capture the differences between female and male students 

**Null Hypothesis**

+  (verbally) The model of GPA for male and female students are not structurally different.
+ (mathematically) $H_0: \;\; \sigma_0=0,\;\; \sigma_1=0, \;\; \sigma_2=0, \;\; \text{and} \;\; \sigma_3=0$

**Question**

What test do we do? t-test or F-test?

<br>
<details>
  <summary>Answer</summary>
F-test.
</details>

### R code

<details class="transcript">
<summary>Transcript</summary>

This code prepares the GPA data and estimates the unrestricted group-specific model. Data with the name gpa3 is loaded from the wooldridge package. The pipeline starts from gpa3 and uses dplyr filter with not is dot n a of ctothrs, retaining only observations for which ctothrs is observed.

Dplyr mutate then creates three products explicitly. Female_sat is female times sat, female_hsperc is female times hsperc, and female_tothrs is female times tothrs. The colon-equals syntax supplies the new column names. Storing the resulting table as gpa lets the regression reuse those terms.

Fixest feols estimates cumgpa on seven regressors. Female allows an intercept difference. Sat, hsperc, and tothrs are the baseline male slopes. Each female product allows the corresponding female slope to differ. The formula is split across lines only for readability. The data argument points to gpa, and the fitted unrestricted model is stored as reg_full for later output and testing. No vcov argument is specified, so fixest uses IID standard errors.

The cell does not print reg_full, so its main purpose is preparation. Autorun ensures later cells can use the object.

</details>

Run the unrestricted model with all the interaction terms:

```{webr-r}
#| autorun: true
data("gpa3", package = "wooldridge")

gpa <-
  gpa3 %>%
  dplyr::filter(!is.na(ctothrs)) %>%
  #--- create interaction terms ---#
  dplyr::mutate(
    female_sat := female * sat,
    female_hsperc := female * hsperc,
    female_tothrs := female * tothrs
  )

#--- regression with female dummy ---#
reg_full <-
  fixest::feols(
    cumgpa ~
      female + sat + female_sat + hsperc + female_hsperc +
      tothrs + female_tothrs,
    data = gpa
  )
```


### What do you see?

<details class="transcript">
<summary>Transcript</summary>

Summary of reg_full prints the unrestricted coefficient table and fit information. There are 366 observations in the filtered sample. The baseline male coefficients are the intercept and the coefficients on sat, hsperc, and tothrs. The female coefficient changes the intercept, while female_sat, female_hsperc, and female_tothrs change the three slopes for women.

Focus on the four female-related rows. The female intercept difference has p about zero point three nine. Female_sat has p about zero point zero five two, close to but still above five percent. Female_hsperc has p about zero point eight six, and female_tothrs has p about zero point nine four. Thus none is individually significant at the five percent level.

The tempting conclusion is that men and women have the same regression function. That does not follow. Each t-test asks whether one difference coefficient is zero while leaving the others unrestricted. Structural equality asks whether all four are zero at once. The coefficient estimates are correlated because each interaction is built from the same group indicator and related covariates. A collection of moderately unusual estimates can provide strong joint evidence even when no single standardized estimate passes its individual cutoff.

This is not a request to reinterpret near-significant p-values. It is a request to test the actual multi-parameter hypothesis. We need a joint Wald or F-type test using the covariance matrix of all four estimates. The next tab performs it in one call.

</details>

**Regression results**

```{webr-r}
#| autorun: true
summary(reg_full)
```

<br>

**What do you see?**

+ None of the variables that involve $female$ are statistically significant at the 5% level individually. 

+ Does this mean that $male$ and $female$ students have the same regression function?

+ No, we are testing the joint significance of the coefficients. We need to do an $F$-test!


### Test of joint significance

<details class="transcript">
<summary>Transcript</summary>

The linearHypothesis call supplies reg_full as the fitted unrestricted model and a character vector containing four restrictions. The c function combines the strings. They set the female intercept difference, female_hsperc, female_sat, and female_tothrs equal to zero. The order of the restrictions does not matter, but each stored coefficient name must match the model output.

Together these restrictions state that the female equation has the same intercept and all three of the same slopes as the male equation. The function uses the model's coefficient covariance matrix to construct one joint statistic. With the current fixest and car methods, the output is labeled chi-squared: about thirty-two point seven on four restrictions, with p about one point four times ten to the minus six. In the classical finite-sample presentation, the corresponding joint test is discussed as an F-test. Both formulations address the same structural-equality null.

The p-value is far below one percent, so we reject the joint null. The data provide strong evidence that at least one of the four group-difference parameters is nonzero under this model. That does not tell us that all four differ, nor does it identify which difference is causal. It says the complete male and female regression functions should not be treated as identical.

</details>

```{webr-r}
#| autorun: true
car::linearHypothesis(
  reg_full,
  c(
    "female = 0",
    "female_hsperc = 0",
    "female_sat = 0",
    "female_tothrs = 0"
  )
)
```

:::
<!--end of panel--> 

## R coding tips: categorical variables and interaction terms

::: {.panel-tabset}

### Prepare a dataset

<details class="transcript">
<summary>Transcript</summary>

This practical section shows how categorical data are commonly represented in R. The original big9salary data contain separate zero-one columns for universities. Real analysis is usually easier when one column stores the category label, so the code reconstructs a university column before using fixest's factor helpers.

The data call loads big9salary from wooldridge. As_tibble converts it to a tibble for readable dplyr work. Mutate creates university with case_when. Each condition checks whether one university indicator equals one; the value to the right of the tilde is the label assigned when that condition is true. For example, osu equal to one becomes "Ohio State U," and msu equal to one becomes "Michigan State U." Case_when evaluates conditions in order, so mutually exclusive indicators are expected.

There is an important data-preparation detail visible in the code: only eight named indicators are mapped and there is no final default condition. Any row for the original omitted university, where all eight displayed indicators equal zero, receives a missing university value. Later regressions using university will omit such rows.

Finally, relocate moves id, year, salary, pubindx, and university to the front of the table without dropping the remaining columns. The transformed object is stored as big9salary_c. The goal is a readable single category column that functions can expand into indicators consistently. The next tab inspects the beginning and end of this result before modeling it.

</details>

```{webr-r}
#--- get big9salary ---#
data("big9salary", package = "wooldridge")

#--- create a variable that indicates university ---#
#--- this is how the data would look most of the time (instead of having a bunch of dummy variables) ---#
big9salary_c <-
  as_tibble(big9salary) %>%
  dplyr::mutate(
    university =
      case_when(
        osu == 1 ~ "Ohio State U",
        iowa == 1 ~ "U of Iowa",
        indiana == 1 ~ "Indiana U",
        purdue == 1 ~ "Purdue U",
        msu == 1 ~ "Michigan State U",
        mich == 1 ~ "Michigan U",
        wisc == 1 ~ "U of Wisconsin",
        illinois == 1 ~ "U of Illinois"
      )
  ) %>%
  dplyr::relocate(id, year, salary, pubindx, university)
```

### Data

<details class="transcript inline">
<summary>Transcript</summary>

This cell inspects the reshaped data from both ends. Head of big9salary_c prints the first six rows by default, and tail prints the last six. Because the two expressions are separate, the output shows two small tables one after the other.

Focus on the columns relocated to the front. Id identifies the faculty member, year gives the observation year, salary is the annual salary outcome, pubindx is the publication index, and university is the new text category. For rows matched by case_when, one university label now carries information that was previously scattered across several dummy columns. The original columns still remain later in the tibble because relocate changes order rather than selecting a subset.

</details>

Take a look at the data,

```{webr-r}
head(big9salary_c)

tail(big9salary_c)
```

### `i()`

<details class="transcript">
<summary>Transcript</summary>

The formula uses fixest's i function to handle university as a categorical regressor. Salary is the dependent variable. Pubindx and female enter as ordinary controls. I of university creates the needed category indicators, while ref equals "Indiana U" explicitly selects Indiana as the omitted baseline. The data argument uses big9salary_c. The fitted model is piped into broom tidy, which returns one row per coefficient with estimates, standard errors, test statistics, and p-values.

Each university coefficient is an intercept difference relative to Indiana, holding publication index and female fixed. A negative Michigan State coefficient means faculty at Michigan State have a lower fitted annual salary than otherwise comparable faculty at Indiana by the coefficient's absolute dollar amount. It is not a comparison with the overall average or with every other university.

The ref argument makes the baseline visible and reproducible. That is safer than manually creating dummies, accidentally including all categories with an intercept, or relying on an implicit alphabetical reference. The function automatically includes c minus one estimable indicators and adapts if verified categories change.

Read the output's sample note too. Observations with missing salary, female, pubindx, or university are removed. Because the preparation code leaves the original all-zero category unmapped, university missingness contributes to that omission. Once the category column is correct, i of category variable is the concise approach to use.

</details>

You can use the `i()` function inside `fixest::feols()` like below:

```{webr-r}
fixest::feols(salary ~ pubindx + female + i(university, ref = "Indiana U"), data = big9salary_c) %>%
  broom::tidy()
```

<br>

`ref = "Indiana U"` sets the base category to `"Indiana U"`. 

So, for example, the highlighted line means that faculty members at Michigan State U make $9,118$ USD less annually than those at Indiana U.

<br>

**Key**

You do not have to make a bunch of dummy variables like the original dataset. Just use `i(category_variable)`.

### Interaction terms 

<details class="transcript">
<summary>Transcript</summary>

The research question now asks whether the association between publication index and salary varies by university. The formula retains female and pubindx as main effects. The first i term, i of university with Indiana as reference, allows university intercepts to differ. The second i term passes pubindx as the continuous variable and again sets Indiana as the reference; it creates university-by-pubindx interaction terms that allow slopes to differ.

Interpret the main pubindx coefficient first. It is the fitted salary change for a one-unit increase in publication index at the baseline Indiana University. Each interaction coefficient is an additional slope relative to Indiana. For Michigan State, the interaction estimate is about four hundred thirty-six dollars. Its fitted publication-index slope is therefore the Indiana slope plus four hundred thirty-six, not four hundred thirty-six alone. The slide's final sentence correctly frames the interaction as a marginal-impact difference.

The ordinary university indicators are still needed. Without them, the specification would force all university lines to share an intercept at pubindx equal to zero while allowing only slopes to differ. Including both sets gives each nonbaseline university an intercept adjustment and a slope adjustment, the categorical analogue of interacting a dummy with education.

Broom tidy exposes all coefficients and their uncertainty. Be precise that the variable is named pubindx in the data and code.

</details>

You can use `i()` for creating interactions of a categorical variable and a continuous variable.

Suppose you are interested in understanding the impact of `pubindx` (continuous) by `university` (categorical), then

```{webr-r}
fixest::feols(salary ~ female + pubindx + i(university, ref = "Indiana U") + i(university, pubindx, ref = "Indiana U"), data = big9salary_c) %>%
  broom::tidy()
```

<br>

So, the marginal impact of `pubindex` is $436$ greater for those at Michigan State U than those at Indiana U.

:::
<!--end of panel-->

# Other miscellaneous topics

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Goodness of fit: $R^2$

::: {.panel-tabset}

### $R^2$ does not matter (often)

<details class="transcript">
<summary>Transcript</summary>

The callout makes a deliberately strong statement: a small R squared is often not a serious problem. To understand it, separate goodness of fit from identification. R squared is one minus the residual sum of squares divided by the total sum of squares when the usual intercept is present. It measures the fraction of sample variation in the dependent variable accounted for by the fitted values. A value near one means observations lie close to the fitted regression surface; a value near zero means much outcome variation remains in the residuals.

That statistic does not tell you whether a coefficient is unbiased, causal, or even economically sensible. A model can fit extremely well because it includes variables correlated with the outcome while still suffer severe endogeneity. Conversely, a randomized regressor can identify a causal effect cleanly even when many other determinants of the outcome remain unexplained and R squared is low.

Whether fit matters depends on the goal. For prediction, unexplained variation and out-of-sample performance are central, although in-sample R squared alone is still not enough. For estimating a particular causal or structural parameter, the crucial questions concern exogeneity, research design, correct specification of the relevant relationship, and valid uncertainty.

</details>

:::{.callout-important}
Small value of $R^2$ does not mean the end of the world (In fact, we could not care less about it.)
:::

### Why?

<details class="transcript">
<summary>Transcript</summary>

The example regression has ecolabs, described in the bullets as the hypothetical pounds of eco-labeled apples a family would demand, on the left. Regprc is the price of regular apples, and ecoprc is the experimentally assigned price of eco-labeled apples. Beta one captures how demand changes with the regular-apple price, holding the eco price fixed. Beta two captures how demand changes with the eco-apple price, holding the regular price fixed.

The key design fact is in the first bullet: the researcher randomly set ecoprc when constructing the survey scenarios. Random assignment breaks systematic association between the assigned price and unobserved demand determinants, supporting the zero conditional mean condition for causal interpretation of that price coefficient. The slide also states the study's narrow objective, understanding how eco-apple demand responds to the eco-apple price.

Family demand can still vary for many reasons not in this short equation: income, household size, preferences, environmental attitudes, and ordinary idiosyncratic choices. Those omitted determinants increase residual variation and lower R squared. They do not bias the randomized price coefficient merely because they are omitted, provided randomization makes them unrelated to assigned price in expectation.

That is the central distinction. R squared asks how much individual demand variation the model explains. Identification asks whether variation in ecoprc is exogenous enough to recover its causal effect. Randomization can answer the second question even when the first answer is "very little."

</details>

**Example**

$$ecolabs = \beta_0 + \beta_1 regprc + \beta_2 ecoprc$$

+ $ecolabs$: the (hypothetical) pounds of ecologically friendly (eco-labeled) apples a family would demand
+ $regprc$: prices of regular apples 
+ $ecoprc$: prices of the hypothetical eco-labeled apples

<br>

**Key**

+ The data was obtained via survey and $ecoprc$ was set randomly (So, we know $E[u|x] = 0$) by the researcher.
+ The (only) objective of the study is to understand the impact of the price of eco-labeled apple on the demand for eco-labeled apples.

### Regression results

<details class="transcript">
<summary>Transcript</summary>

The code loads apple from the wooldridge package, then estimates ecolbs on regprc and ecoprc with fixest feols. Notice that the dataset's actual outcome name in code is ecolbs. The surrounding parentheses print the fitted model immediately. No vcov argument is shown, so the table uses IID standard errors.

The fitted intercept is about one point nine six five. The regular-apple price coefficient is positive three point zero two nine, consistent with regular and eco-labeled apples acting as substitutes in this survey: a higher regular price is associated with greater demand for eco apples. The eco-apple price coefficient is negative two point nine two six, so raising the eco price by one price unit reduces hypothetical eco-apple demand by about two point nine three pounds, holding regular price fixed. Both price coefficients have very small p-values in the displayed regression.

The ordinary R squared is about zero point zero three six, and the adjusted R squared shown by fixest is about zero point zero three three. Thus the two price variables explain only a few percent of the sample variation in stated demand. Open the answer: that low value is not a problem for the stated objective. The study seeks the causal response to assigned price, not a high-accuracy prediction of each family's pounds demanded.

Randomizing the survey price provides credibility that a higher-demand family was not systematically assigned a different ecoprc.

</details>

```{webr-r}
data(apple, package = "wooldridge")

(
  fixest::feols(ecolbs ~ regprc + ecoprc, data = apple)
)
```

<br>

:::{.callout-note title="Question"}
Note that $R^2$ is very small. Is this a problem?
:::

<br>
<details>
  <summary>Answer</summary>
No.

+ Their goal is not predicting the demand of eco-labeled apple. Understanding the <span style = "color: blue;"> causal </span> impact of price on demand.
+ The goal is achieved via randomization of the price variables at the stage of designing the survey!

</details>

:::
<!--end of panel-->

## Scaling

::: {.panel-tabset}


### Questions

<details class="transcript">
<summary>Transcript</summary>

This section asks what changes when you rescale a variable used in a regression. Think of measuring education in months instead of years, income in thousands of dollars instead of dollars, or land in hectares instead of acres. The underlying observations and economic relationship have not changed; only the numerical unit has.

The four bullets ask about the coefficient, its standard error, the t-statistic, and R squared.

</details>

What happens if you scale up/down variables used in regression? 

+ coefficients
+ standard errors
+ t-statistics
+ $R^2$

### Let's see

<details class="transcript inline">
<summary>Transcript</summary>

Compare the two code cells side by side. The left feols call regresses wage on female and education in its original unit, years. The right uses I of education times twelve, creating the same schooling amounts measured in months. Capital I makes the multiplication literal inside the formula. Both calls use wage1 and the default IID standard errors, and autorun evaluates both.

In the left output, the education coefficient is about zero point five zero six and its standard error is about zero point zero five zero four. In the right output, the coefficient on education times twelve is about zero point zero four two two and its standard error is about zero point zero zero four two. Each is exactly one twelfth of its original-scale counterpart apart from displayed rounding.

The t-statistic stays about ten point zero five because dividing both the estimate and standard error by twelve leaves their ratio unchanged. The p-value is therefore unchanged too. Ordinary R squared is about zero point two five eight eight in both models, and adjusted R squared is about zero point two five six zero. The intercept, female coefficient, fitted wages, residuals, and residual standard error are also unchanged because the education contribution to each prediction is numerically identical.

The callout summarizes the requested four quantities: coefficient one twelfth, standard error one twelfth, t-statistic the same, and R squared the same. This is a unit conversion, not new information or a different fit.

</details>

::: {.columns}

::: {.column width="50%"}
```{webr-r}
#| autorun: true
#--- regression with original scale ---#
fixest::feols(wage ~ female + educ, data = wage1)
```
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{webr-r}
#| autorun: true
#--- regression with scaled educ ---#
fixest::feols(wage ~ female + I(educ * 12), data = wage1)
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

<br>

:::{.callout-note title="So,"}
+ coefficient: 1/12
+ standard error: 1/12
+ t-stat: the same
+ $R^2$: the same
:::

### Interpretation

<details class="transcript">
<summary>Transcript</summary>

The first bullet interprets the unscaled coefficient. With education measured in years, one additional year is associated with about zero point five zero six dollars more hourly wage, holding female fixed. The coefficient's units are dollars per hour of wage per year of education.

The second bullet interprets the scaled model. Education times twelve records months, so an increase of one in that regressor means one additional month. Its coefficient, about zero point zero four two two dollars per hour, is the fitted wage change per month of education. The parenthetical example helps read the scale: a value of forty-eight represents four years because forty-eight months divided by twelve equals four.

These statements are numerically equivalent. A one-year increase is a twelve-month increase, so the scaled model predicts zero point zero four two two times twelve, approximately zero point five zero six dollars per hour. That matches the original coefficient up to rounding. Neither model implies a different return to education, different predictions, or different significance.

</details>

**Interpretation**

+ Regression <span style = "color: blue;"> without </span> scaling 

hourly wage increases by $0.506$ if education increases by a <span style = "color: blue;"> year </span>

+ Regression <span style = "color: blue;"> with </span> scaling (e.g., 48 means 4 years)

hourly wage increases by $0.0422$ if education increases by a <span style = "color: blue;"> month </span>

<br>

**Note**

According to the scaled model, hourly wage increases by $0.0422 * 12$ if education increases by a year (12 months).

That is, the estimated marginal impact of education on wage from the scaled model is the same as that from the non-scaled model.

### Summary

<details class="transcript">
<summary>Transcript</summary>

The final slide collects the invariance result for an independent variable. If you multiply a regressor by a factor a, its coefficient estimate is divided by a. The first bullet's phrase "scaled up/back to the exact degree" means the coefficient moves inversely with the regressor's numerical scale so their product, and therefore every fitted value, stays the same. Its standard error is divided by the same factor because the estimator's sampling distribution is rescaled identically.

The t-statistic for testing a zero coefficient remains unchanged. Algebraically, beta hat divided by a over standard error divided by a equals beta hat over standard error. The p-value and significance stars therefore remain unchanged as well. R squared stays the same because fitted values and residuals do not change. Simply converting years to months cannot improve or worsen the model.

If instead you divide the regressor by a, such as expressing dollars in thousands, the coefficient and standard error are multiplied by a. The same inverse-unit logic applies.

</details>

When an independent variable is scaled, 

+ its coefficient estimate and standard error are going to be scaled up/back to the exact degree the variable is scaled up/back
+ t-statistics stays the same (as it should be)
+ $R^2$ stays the same (the model does not improve by simply scaling independent variables)

:::
<!--end of panel-->
