```{r}
#| include: false
# Links and dates come from course-dates.json -- never type a Dropbox URL or a
# date into this page. See CLAUDE.md.
source(here::here("course-dates.R"))

cal <- read_course_dates()
year <- format(cal$first_class_day, "%Y")
```

All assignment files for `r cal$label` are in one Dropbox folder:
**[download the course materials here](`r materials_url()`)**.
Open the folder for the assignment you need — each one contains the `.qmd` to
work on and every dataset it uses.

Submit through the same link every time:
**[submission folder](`r submission_url()`)**.
Submit both the `.qmd` and the rendered `.html`, and name your file
`lastname_assignment_1.qmd` (`Mieno_assignment_1.qmd`, for example) so it can be
filed automatically.

```{r}
#| echo: false
#| results: asis

ids <- c("assignment-1", "assignment-2", "assignment-3")

for (id in ids) {
  n <- sub("assignment-", "", id)
  due <- format_course_date(paste0(id, "-due"), "%A, %B %e")
  cat(sprintf("\n# Assignment %s\n\n", n))
  cat(sprintf("Due **%s**, before the lab session in which the TA reviews it.\n\n", due))
  cat(sprintf("1. Download the `%s/` folder from the [course materials](%s)\n",
              id, materials_url()))
  cat(sprintf("2. Work on **%s.qmd**, leaving all your R code and output in place\n", id))
  cat(sprintf("3. Submit the qmd and the html [here](%s)\n", submission_url()))
}
```

# Final project

Due **`r format_course_date('final-paper', '%A, %B %e')`**.

1. Download the final project template from the
   [course materials folder](`r materials_url()`)
2. Work on **final-project-template.qmd** to complete your project (present all
   the R code)
3. Submit the qmd and the html [here](`r submission_url()`)

Milestones: research topic and dataset approved by
**`r format_course_date('paper-topic-approval', '%A, %B %e')`**; paper proposal
due **`r format_course_date('paper-proposal', '%A, %B %e')`**.
