---
title: "Assignment 3 (<span style='color:blue'>Score: /15 </span>)"
author: "Your Name Here"
format:
  html:
    number-sections: true
    number-depth: 1
    theme: flatly
    toc: true
    include-in-header:
      text: |
        <style> .answer { color: blue; } </style>
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| keep: true
#| echo: false
#| include: false
library(knitr)

opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  #--- figure related ---#
  fig.align = "center",
  fig.width = 5,
  fig.height = 4
)
```

# Instruction

The following problems are from Chapters 6, 7, and 8 of the Wooldridge book (Introductory Econometrics, 5th Edition). All the problems are typed out on this qmd (and in html) file. So, it is not necessary to look at the book for the corresponding problems. Indeed, you should not look at the book because problems here are slightly modified versions of those from the book.

Due date is <span style='color:blue'>{{assignment-3-due|%B %e}}</span> <span style='color:red'>before</span> the lab session on that day, in which this assignment is reviewed by the TA. Submission after the deadline would result in a heavy discount on the score.

Please submit your assignment (both the qmd and the resulting html files) using the following [submission link]({{submission}}). Name your qmd file using the following convention: `lastname_assignment_3.qmd`. For example, it would be `Mieno_assignment_3.qmd` for me.

I do not accept any hand-written answers (you will get score of zero for that problem). Leave all the R codes and results in your answer (where appropriate) for the computer exercises. If your R codes are not provided, you will get score of 0 for that question.

You are encouraged to discuss the problems with your classmates. However, I would recommend that you do so after you try all the problems yourself. You would be simply doing disservice to yourself if you resort to your friends' help immediately, depriving you of important opportunities to think carefully about what we have learned and understand them through solving the problems.

------

::: {.answer}
Here, I load all the packages I will be using:
:::

```{r load_libraries}
#| echo: false
#| message: false
#| warning: false
library(modelsummary)
library(car)
library(fixest)
library(tidyverse)
library(broom)
```

<!-- Non-computer Exercises -->

# Non-computer Exercises

------

## Problem 3 (Chapter 7)

Using the data in GPA2.RAW, the following equation was estimated:

$$
\begin{align}
sat = & \beta_0 + \beta_1 hsize + \beta_2 hsize^2 +\beta_3 female \\ & +  \beta_4 black + \beta_5 female \times black
\end{align}
$$

The table below presents the regression results.

<br>

```{r problem_3_given}
#| keep: true
#| echo: false
# Given setup for Problem 3. This chunk is handed to students, so it loads
# everything it needs rather than relying on a chunk that gets stripped out.
library(tidyverse)
library(data.table)
library(fixest)
library(modelsummary)

gpa_data <- haven::zap_labels(haven::read_dta("gpa2.dta")) %>%
  data.table() %>%
  .[, hsize_2 := hsize^2] %>%
  .[, female_black := female * black]

reg_gpa <- feols(sat ~ hsize + hsize_2 + female + black + female_black, data = gpa_data)

msummary(
  reg_gpa,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```

<br>

Note that the standard errors of coefficients estimators are reported in parentheses.

The variable $sat$ is the combined SAT score, $hsize$ is size of the student's high school graduating class, in hundreds, $female$ is a gender dummy variable, $black$ is a race dummy
variable equal to one for blacks and zero otherwise, and $female\_black$ is the interaction term between $female$ and $black$.

------

(i) Is there strong evidence that $hsize^2$ should be included in the model? From this equation, what is the optimal high school size?

::: {.answer}
Answer:
:::

::: {.answer}
$hsize^2$ should be included because it is statistically significant even at the 1% level.
:::

::: {.answer}
$\frac{\partial sat}{\partial hsize}  = \beta_1 + 2\times \beta_2 hsize = `r round(reg_gpa$coefficient['hsize'],digits=2)` + 2\times (`r round(reg_gpa$coefficient['hsize_2'],digits=2)`) hsize$. Therefore, when $hsize=`r round(-reg_gpa$coefficient['hsize']/reg_gpa$coefficient['hsize_2']/2,digits=2)`$, $sat% is maximized $ceteris$ $paribus$.
:::

------

(ii) Holding $hsize$ fixed, what is the estimated difference in SAT score between nonblack females and nonblack males? How statistically significant is this estimated difference?

::: {.answer}
Answer:
:::

::: {.answer}
**Non-black females** ($black=0$ and $female=1$)
$$
\begin{align}
\hat{sat} = & \beta_0 + \beta_1 hsize + \beta_2 hsize^2 +\beta_3 female (=1) \\ & +  \beta_4 black (=0) + \beta_5 female \times black (=1\times 0) \\
& = \beta_0 + \beta_1 hsize + \beta_2 hsize^2 + \beta_3
\end{align}
$$
:::

::: {.answer}
**Non-black males** ($black=0$ and $female=0$)
$$
\begin{align}
\hat{sat} = & \beta_0 + \beta_1 hsize + \beta_2 hsize^2 +\beta_3 female (=0) \\ & +  \beta_4 black (=0) + \beta_5 female \times black (=0\times 0) \\
& = \beta_0 + \beta_1 hsize + \beta_2 hsize^2
\end{align}
$$
:::

::: {.answer}
Therefore the estimated difference is $\hat{\beta}_3$, which is statistically significant at the 1\% level. The estimated difference is $45$ points.
:::

------

(iii) What is the estimated difference in SAT score between nonblack males and black males? Test the null hypothesis that there is no difference between their scores, against the alternative that there is a difference.

::: {.answer}
Answer:
:::

::: {.answer}
**Non-black males** ($black=0$ and $female=0$)
:::

::: {.answer}
see above
:::

::: {.answer}
**Black males** ($black=1$ and $female=0$)
$$
\begin{align}
\hat{sat} = & \beta_0 + \beta_1 hsize + \beta_2 hsize^2 +\beta_3 female (=0) \\ & +  \beta_4 black (=1) + \beta_5 female \times black (=0\times 1) \\
& = \beta_0 + \beta_1 hsize + \beta_2 hsize^2 + \beta_4
\end{align}
$$
:::

::: {.answer}
Therefore the estimated difference is $\hat{\beta}_4$, which is statistically significant at the 1\% level. The estimated difference is about $170$.
:::

------

(iv) What is the estimated difference in SAT score between black females and nonblack females? What would you need to do to test whether the difference is statistically significant?

::: {.answer}
Answer:
:::

::: {.answer}
**Black females** ($black=1$ and $female=1$)
$$
\begin{align}
\hat{sat} = & \beta_0 + \beta_1 hsize + \beta_2 hsize^2 +\beta_3 female (=1) \\ & +  \beta_4 black (=1) + \beta_5 female \times black (=1\times 1) \\
& = \beta_0 + \beta_1 hsize + \beta_2 hsize^2 + \beta_3+ \beta_4+ \beta_5
\end{align}
$$
:::

::: {.answer}
**Non-black females** ($black=0$ and $female=1$)
:::

::: {.answer}
see above
:::

::: {.answer}
Therefore the estimated difference is $\hat{\beta}_4+\hat{\beta}_5=`r round(reg_gpa$coefficient['black'],digits=2)` + `r round(reg_gpa$coefficient['female_black'],digits=2)`=`r round(reg_gpa$coefficient['black']+reg_gpa$coefficient['female_black'],digits=2)`$. To test the statistical significance of the difference, we need to calculate $se(\hat{\beta}_4+\hat{\beta}_5)$ based on the estimated variance covariance matrix of the coefficient estimators. Then, we can do $t$-test on the $t$-statistic $(\hat{\beta}_4+\hat{\beta}_5)/se(\hat{\beta}_4+\hat{\beta}_5)$.
:::


------

## Problem 6 (Chapter 7)

To test the effectiveness of a job training program on the subsequent wages of workers, we specify the model

$$
\begin{align}
    log(wage) = \beta_0 + \beta_1 train + \beta_2 educ + \beta_3 exper + u
\end{align}
$$

where $train$ is a binary variable equal to unity if a worker participated in the program. Think of the error term $u$ as containing unobserved worker ability. If less able workers have a greater chance of being selected for the program, and you use an OLS analysis, what can you say about the likely bias in the OLS estimator of $\beta_1$?

::: {.answer}
Answer:
:::

::: {.answer}
Let $\beta_4$ denote the impact of ability on $log(wage)$ after you control for the other independent variables. For whatever unit we measure worker ability, let higher ability be associated with higher number. Then, $\beta_4$ should be positive because a unit increase in ability should result in a higher wage $ceteris$ $paribus$.
:::

::: {.answer}
When less able workers have a greater chance of being selected for the program, $train$ is negatively correlated with ability.
:::

::: {.answer}
Using the formula for the direction of omitted variable bias,
:::

::: {.answer}
$$
\begin{align}
 sign(bias) & = sign(Cor(train,ability))\times sign(\beta_4) \\
& = - \times + = -
\end{align}
$$
:::

::: {.answer}
Therefore, we would understate the impact of $train$ on $wage$.
:::


------

<!-- Computer Exercises -->

# Computer Exercises

::: {.answer}
First, I load all the packages I will be using in this assignment:
:::

```{r library, message=FALSE}
library(tidyverse)
```

## Problem C1 (Chapter 6)

Use the data in **KIELMC.dta**, only for the year 1981, to answer the following questions. The data are for houses that sold during 1981 in North Andover, Massachusetts; 1981 was the year construction began on a local garbage incinerator.

------

(i) To study the effects of the incinerator location on housing price, consider the simple regression model
$$
log(price) = \beta_0 + \beta_1 log(dist) + u
$$
where $price$ is housing price in dollars and $dist$ is distance from the house to the incinerator measured in feet. Interpreting this equation causally, what sign do you expect for $\beta_1$ if the presence of the incinerator depresses housing prices? Estimate this equation and interpret the results.

::: {.answer}
Answer:
:::

::: {.answer}
Given the people do not like having an incinerator in their neighborhood, I would expect that $\beta_1$ be positive because the further away you are from the incinerator, the higher the house price should be on average.
:::

```{r c1_6_1, cache=FALSE}
#--- load the data ---#
data <- haven::zap_labels(haven::read_dta("KIELMC.dta"))

#--- subset the data to 1981 ---#
data_81 <- filter(data, year == 1981)

#--- OLS ---#
reg_c1_6_1 <- feols(log(price) ~ log(dist), data = data_81)

#--- msummary ---#
msummary(
  reg_c1_6_1,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
According to the regression results, if the distance to the incinerator increases by $1\%$, house price increases by $`r round(reg_c1_6_1$coefficient[2],digits=3)` \%$ on average $ceteris$ $paribus$
:::

------

(ii) To the simple regression model in part (i), add the variables log($intst$), log($area$), log($land$), $rooms$, $baths$, and $age$, where $intst$ is distance from the home to the interstate, $area$ is square footage of the house, $land$ is the lot size in square feet, $rooms$ is total number of rooms, $baths$ is number of bathrooms, and $age$ is age of the house in years. Now, what do you conclude about the effects of the incinerator? Explain why (i) and (ii) give conflicting results.

::: {.answer}
Answer:
:::

```{r c1_6_2, cache=FALSE}
#--- OLS ---#
reg_c1_6_2 <- feols(log(price) ~ log(dist) + log(intst) + log(area) +
  log(land) + rooms + baths + age, data = data_81)

#--- msummary ---#
msummary(
  reg_c1_6_2,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
According to the new regression results, the incinerator does not have any statistically significant impact on house price, which is a different conclusion we drew earlier with the simple regression. This is because all the added variables that stayed in the error term in the simple regression are correlated with the incinerator placement. The table below shows the correlation coefficient between $dist$ and the other included variables.
:::

```{r cor}
round(cor(select(data_81, dist, intst, area, land, rooms, baths, age)), digits = 2)
```

::: {.answer}
You can see the $dist$ is positively correlated with favorable conditions ($area$, $land$, $rooms$, $baths$) and negatively correlated with unfavorable conditions ($age$). That is, the incinerator seemed to be placed in a poorer area. Consequently, when the new variables are not included, we overstated the negative impact of the incinerator on house price.
:::


------

(iii) Add $[log(intst)]^2$ to the model from part (ii). Now what happens? What do you conclude about the importance of functional form?

::: {.answer}
Answer:
:::

```{r c1_6_3}
#--- define a new variable ---#
data_81 <- mutate(data_81, log_intst_2 = log(intst)^2)

#--- OLS ---#
reg_3 <- feols(log(price) ~ log(dist) + log(intst) + log_intst_2 +
  log(area) + log(land) + rooms + baths + age, data = data_81)

#--- msummary ---#
msummary(
  reg_3,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
$[log(intst)]^2$ is highly statistically significant. More important, after including $[log(intst)]^2$, $log(dist)$ has become significant again. This means that the previous model was misspecified leaving $[log(intst)]^2$ left in the error term that should have been included. As you can see below, $log(dist)$ and $[log(intst)]^2$ are very highly correlated. Therefore, the bias of not including $[log(intst)]^2$ was severe. This example shows clearly that functional misspecification can lead to a significant bias on the coefficient estimate on the variable of interest ($log(dist)$).
:::

```{r cor_2}
round(cor(log(data_81$dist), log(data_81$intst)^2), digits = 3)
```

------

(iv) Is the square of log($dist$) significant when you add it to the model from part (iii)?

::: {.answer}
Answer:
:::

```{r c1_6_4}
#--- define a new variable ---#
data_81 <- mutate(data_81, log_dist_2 = log(dist)^2)

#--- OLS ---#
reg_3 <- feols(log(price) ~ log(dist) + log_dist_2 + log(intst) + log_intst_2 +
  log(area) + log(land) + rooms + baths + age, data = data_81)

#--- msummary ---#
msummary(
  reg_3,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
No, it is not.
:::

------

## Problem C3 (Chapter 6)

Consider a model where the return to education depends upon the amount of work experience (and vice versa):

$$
\begin{align}
 log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 educ\times exper + u
\end{align}
$$

------

(i) Show that the return to another year of education (in decimal form), holding $exper$ fixed, is $\beta_1 + \beta_3 exper$.

::: {.answer}
Answer:
:::

::: {.answer}
Differentiating the both sides of the equation with respect to $educ$,
:::

::: {.answer}
$$
\begin{align}
\frac{\partial log(wage)}{\partial educ} & = \beta_1+\beta_3 exper
\end{align}
$$
:::

------

(ii) State the null hypothesis that the return to education does not depend on the level of $exper$. What do you think is the appropriate alternative?

::: {.answer}
Answer:
:::

$H_0:\;\; \beta_3=0$

::: {.answer}
Since I would expect more experience would lead to a greater marginal return of education, I would set the following one-sided alternative:
:::

$H_1:\;\; \beta_3 > 0$

------

(iii) Use the data in **WAGE2.dta** to test the null hypothesis in (ii) against your stated alternative.

::: {.answer}
Answer:
:::

::: {.answer}
Here is the regression results:
:::

```{r c3_6_3, results='asis'}
#--- load the data ---#
data_c3_6 <- haven::zap_labels(haven::read_dta("WAGE2.dta"))

#--- create the interaction term ---#
data_dt_c3_6 <- mutate(data_c3_6, educ_exper = educ * exper)

#--- OLS ---#
reg_c3_6 <- feols(log(wage) ~ educ + exper + educ_exper, data = data_dt_c3_6)

#--- msummary ---#
msummary(
  reg_c3_6,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
  Note that $t$-statistic used to do one-sided and two-sided tests are exactly the same. They only differ in the critical values. Therefore, we can use the $t$-static reported in the regression results for the two-sided tests by default.
:::

```{r c3_6_3_2}
#--- get the t-stat ---#
(
  t_stat <- tidy(reg_c3_6) %>%
    filter(term == "educ_exper") %>%
    pull(statistic)
)

#--- critical value (5\% significance level) ---#
(
  c_value <- qt(0.95, degrees_freedom(reg_c3_6, type = "resid"))
)

#--- test ---#
t_stat > c_value
```

::: {.answer}
Therefore, the null hypothesis was rejected.
:::

------

(iv) Let $\theta_1$ denote the return to education (in decimal form), when $exper=10$: $\theta_1=\beta_1+10\beta_3$. Obtain $\hat{\theta}_1$. Test if $\theta_1$ statistically significantly different from 0.

::: {.answer}
Answer:
:::

::: {.answer}
The appropriate $t$-static for this test is $\hat{\theta}_1/se(\hat{\theta}_1)$. Now,
:::

::: {.answer}
$$
\begin{align}
 se(\hat{\theta}_1) & =\sqrt{Var(\hat{\theta}_1)} \\
 & = \sqrt{Var(\hat{\beta}_1)+ 10\times 2 Cov(\hat{\beta}_1,\hat{\beta}_3) +100*Var(\hat{\beta}_3)}
\end{align}
$$
:::

```{r c3_6_4}
linearHypothesis(reg_c3_6, "educ+10*educ_exper=0")
```

------

## Problem C2 (Chapter 7)

Use the data in **WAGE2.dta** for this exercise.

(i) Estimate the model
$$
\begin{align}
  log(wage) = & \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 tenure \\
  & + \beta_4 married + \beta_5 black + \beta_6 south + \beta_7 urban + u
\end{align}
$$
and report the results in the usual form. Holding other factors fixed, what is the approximate difference in monthly salary between blacks and nonblacks? Is this difference statistically significant?

::: {.answer}
Answer:
:::

```{r c2_7_1, results='asis'}
#--- load the data ---#
# you can also use the previously loaded data
data_c2_7 <- haven::zap_labels(haven::read_dta("WAGE2.dta"))

#--- OLS ---#
reg_c2_7_1 <- feols(log(wage) ~ educ + exper + tenure + married
  + black + south + urban, data = data_c2_7)

#--- msummary ---#
msummary(
  reg_c2_7_1,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
The Average salary of blacks is `r round(-reg_c2_7_1$coefficient['black']*100,digits=2)` % less than that of non-blacks. The difference is statistically significant even at the 1% level.
:::


------

(ii) Add the variables $exper^2$ and $tenure^2$ to the equation and show that they are jointly insignificant at the 20% level.

::: {.answer}
Answer:
:::

```{r c2_7_2}
#--- define quadratic variables ---#
data_c2_7 <- data_c2_7 %>%
  mutate(
    exper2 = exper^2,
    tenure2 = tenure^2
  )

#--- OLS ---#
reg_c2_7_2 <- feols(log(wage) ~ educ + exper + exper2 + tenure + tenure2 + married + black + south + urban, data = data_c2_7)

#--- msummary ---#
msummary(
  reg_c2_7_2,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
We have already estimated both the restricted and unrestricted models. The codes below conduct $F$-test.
:::

```{r c2_7_2_F}
linearHypothesis(reg_c2_7_2, c("exper2=0", "tenure2=0"))
```

::: {.answer}
Therefore, they are jointly insignificant even at the 20% level.
:::

------

(iii) Extend the original model to allow the return to education to depend on race and test whether the return to education does depend on race. Report the regression results.

::: {.answer}
Answer:
:::

```{r c2_7_3, results='asis'}
#--- generate educ-black dummy ---#
data_c2_7 <- data_c2_7 %>%
  mutate(
    educ_black = educ * black
  )

#--- OLS ---#
reg_c2_7_3 <- feols(log(wage) ~ educ + educ_black + exper +
  tenure + married + black + south + urban, data = data_c2_7)

#--- msummary ---#
msummary(
  reg_c2_7_3,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

::: {.answer}
Return to the education does not depend on race as the coefficient on $educ\_black$ is highly statistically insignificant.
:::

------

(iv) Again, start with the original model, but now allow wages to differ across four groups of people: married and black, married and nonblack, single and black, and single and nonblack. What is the estimated wage differential between married blacks and married nonblacks?

::: {.answer}
Answer:
:::

::: {.answer}
I first create three dummy variables: married black, single black, and single non-black.
:::

```{r dummy_create, results='asis'}
#--- create dummy variables ---#
data_c2_7 <- data_c2_7 %>%
  mutate(
    married_black = married * black, # married black
    single_black = (1 - married) * black, # single black
    single_nonblack = (1 - married) * (1 - black) # single non-black
  )
```

::: {.answer}
I then include the three dummy variables with $married$ and $black$ taken out from the original model:
:::

```{r c2_7_4, results='asis',cache=TRUE}
#--- ols with the dummy variables ---#
reg_c2_7_4 <- lm(log(wage) ~ educ + exper + tenure +
  married_black + single_black + single_nonblack +
  south + urban, data = data_c2_7)

#--- msummary ---#
msummary(
  reg_c2_7_4,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
<br>

------

## Problem C14 (Chapter 8)

Use **BEAUTY.dta** for this question. Here are the list of definitions of some variables that are not self-explanatory:

+ $looks$: score of how attractive observed individuals are in terms of appearance
+ $belavg$: 1 if $looks\leq 2$
+ $abvavg$: 1 if $looks\geq 4$

For question (ii), you do not have to do heteroskedasticity-robust F-test.

------

::: {.answer}
First, I load the data and define a new variable:
:::

```{r c14_8}
#--- load the data ---#
data_c14_8 <- haven::zap_labels(haven::read_dta("BEAUTY.dta"))

#--- create the square of age ---#
data_c14_8 <- data_c14_8 %>%
  mutate(exper2 = exper^2)
```
------

(i) Using the data pooled for men and women, estimate the equation
$$
\begin{align}
  lwage = & \beta_0 + \beta_1 belavg + \beta_2 abvavg + \beta_3 female \\
  & + \beta_4 educ + \beta_5 exper + \beta_6 exper^2 + u
\end{align}
$$
and report the results using heteroskedasticity-robust standard errors. Are any of the coefficients surprising in either their signs or magnitudes? Is the coefficient on female practically large and statistically significant?

::: {.answer}
Answer:
:::

```{r c14_8_1, results='asis'}
#--- ols ---#
reg_c14_8 <-
  feols(
    lwage ~ belavg + abvavg + female + educ + exper + exper2,
    vcov = "hetero",
    data = data_c14_8
  )

#--- msummary ---#
msummary(
  reg_c14_8,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```

<br>

::: {.answer}
All the signs of the coefficient estimates seem to have expected signs. Even though the coefficient on $abvavg$ is negative, it is not statistically significant. The difference in female and male workers is statistically significant and is quite large in magnitude.
:::

------

(ii) Add interactions of female with all other explanatory variables in the equation from part (i) (five interactions in all). Compute the usual F test of joint significance of the five interactions and a heteroskedasticity-robust version. Does using the heteroskedasticity-robust version change the outcome in any important way?

::: {.answer}
Answer:
:::

```{r c14_8_2}
#--- create dummy variables ---#
data_c14_8 <- data_c14_8 %>%
  mutate(
    female_belavg = female * belavg,
    female_abvavg = female * abvavg,
    female_educ = female * educ,
    female_exper = female * exper,
    female_exper2 = female * exper2
  )

#--- unrestricted model estiamtion ---#
reg_u <-
  feols(
    lwage ~ belavg + abvavg + female +
      educ + exper + exper2 + female_belavg +
      female_abvavg + female_educ + female_exper + female_exper2,
    data = data_c14_8
  )
```

::: {.answer}
F-test with default vcov estimates
:::

```{r}
# F-test with default vcov estimates
linearHypothesis(
  reg_u,
  c(
    "female_belavg=0", "female_abvavg=0", "female_educ=0",
    "female_exper=0", "female_exper2=0"
  )
)
```

::: {.answer}
F-test with robust vcov estimates
:::

```{r}
reg_u_vhet <-
  feols(
    lwage ~ belavg + abvavg + female +
      educ + exper + exper2 + female_belavg +
      female_abvavg + female_educ + female_exper + female_exper2,
    vcov = "hetero",
    data = data_c14_8
  )

# F-test with robust vcov estimates
linearHypothesis(
  reg_u_vhet,
  c(
    "female_belavg=0", "female_abvavg=0", "female_educ=0",
    "female_exper=0", "female_exper2=0"
  )
)
```

::: {.answer}
Therefore, female and male workers have statistically significantly different wage equations.
:::

------

(iii) In the full model with interactions, determine whether those involving the looks variables-$female\times belavg$ and $female\times abvavg$-are jointly significant. Are their coefficients practically small?

::: {.answer}
Answer:
:::

```{r c14_8_3}
linearHypothesis(reg_u, c("female_belavg=0", "female_abvavg=0"))
```

::: {.answer}
Therefore, I would conclude that the degree of wage discrimination by looks is not different for female and male workers. Below is the regression results of the fully flexible model:
:::

```{r c14_8_3_reg, cache=TRUE}
msummary(
  reg_u,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```

<br>

::: {.answer}
The coefficient of `r round(reg_u$coefficient['female_belavg'],digits=3)` on $female\_belavg$ suggest that female workers of below average looks make $4.4\%$ more than males workers of below average looks. This difference does not seem to e very large. Moreover, since the coefficient is not statistically significant, it is rather pointless to talk about if the coefficient is large or not.
:::

------

# Data management

In this problem, you are asked to reshape fake datasets (**corn_price_long.rds** and **corn_prod_wide.rds**) and merge them into a single dataset that is ready for statistical analysis. **corn_price_long.rds** has data on corn price by county and month for 2015. **corn_prod_wide.rds** has data on corn production by county (made-up) and month for 2015.

------

(i) **corn_prod_wide.rds** is in a wide format. Using the **gather()** function, convert **corn_prod_wide.rds** into a long format in which each row represents the production level for a specific county-year combination. 

```{r}
corn_prod_wide <- readRDS("corn_prod_wide.rds")

corn_prod_long <-
  corn_prod_wide %>%
  pivot_longer(-county_code, names_to = "month", values_to = "price")

head(corn_prod_long)
```
------

(ii) Now, it's time to merge them together. Would the single key $couty\_code$ be sufficient to merge them? Why or why not?

::: {.answer}
No. For example, for $couty\_code$=1, we have so many rows in both the corn_price and corn_prod datasets that are $couty\_code$=1. Hence, we need to use both $couty\_code$ and $month$ as keys.
:::


------

(iii) Merge them. 

```{r}
corn_price_long <- readRDS("corn_price_long.rds")

corn_price_long$month %>% class()
corn_prod_long$month %>% class()

corn_prod_long <- mutate(corn_prod_long, month = as.numeric(month))

price_prod <-
  left_join(
    corn_prod_long,
    corn_price_long,
    by = c("county_code", "month")
  ) %>%
  arrange(county_code)

head(price_prod)
```
