---
title: "01-1: Univariate Regression: Introduction"
format: 
  revealjs:
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    footer: "[back to course website with lecture slides](https://tmieno2.github.io/Applied-Econometrics-MS-Quarto/lectures/)"
    include-in-header: ../transcript-support.html
---

```{r knitr-setup}
#| include: false
#| cache: false
library(data.table)
library(fixest)
library(flextable)
library(dplyr)
library(ggplot2)

source(here::here("lectures/_lecture-theme.R"))
```

# Univariate Regression: Introduction

<html><div style='float:left'></div><hr color='#EB811B' size=2px width=1280px></html>

## Population, Sample, and Econometrics

::: {.panel-tabset}
### Population

<details class="transcript">
<summary>Transcript</summary>

Everything in econometrics starts with a question about a population, so let's be precise about that word. A population is the set of all the individuals, or farms, or firms you want to learn about. Every one of them, not the ones you happen to have data on. Look at the examples: if your question is about education and income in the United States, your population is every person in the country. Now read the Important block, because this is the part students skip. A population is not a fact about the world, it is a consequence of your question. Change the question from global to American and you have changed the population.

</details>

:::{.callout-important title=Definition}
A set of <span style = "color: red;"> ALL </span> individuals, items, or phenomena that you are interested in learning about
:::

<br>

**Example**

+ Suppose you are interested in the impact of education on income across the U.S. Then, the population is all the individuals in U.S.
+ Suppose you are interested in the impact of water pricing on irrigation water demand for farmers in NE. Then, your population is all the farmers in NE.

**Important**

Population differs depending on the scope of your interest

+ If you are interested in understanding the impact of COVID-19 on child education achievement at the global scale, then your population is every single kid in the world

+ If you are interested in understanding the impact of COVID-19 on child education achievement in U.S., then your population is every single kid in U.S.

### Sample

:::{.callout-important title=Definition}
Sample is a subset of population that you observe
:::

::: {.panel-tabset}
### Case 1

<details class="transcript">
<summary>Transcript</summary>

A sample is the subset of the population you actually get to observe, and it is always a tiny fraction of it. That gap is the entire reason statistics exists. Look at Case 1. You want to know about education and wages, and you have three hundred individuals from each state. Now ask the question in the box honestly. Every state contributes the same three hundred people, so Wyoming and California carry equal weight even though California has roughly seventy times the population. Is that the country? It depends entirely on what you intend to do with it, and that judgement is yours to make, not something the software will make for you.

</details>

+ Population: you are interested in the impact of education on wage
+ Sample (example): data on education, income, and many other things for 300 individuals from each State 

:::{.callout-note title=Question}
Is the sample representative of the population?
:::

### Case 2

<details class="transcript">
<summary>Transcript</summary>

Now the second case, and this one is sharper. Your population is farmers in Nebraska. Your sample is five hundred farmers in the Upper Republican Basin, in the southwest corner of the state. So ask the question again. Those farmers share a climate, a soil type, an aquifer, and a set of water regulations that farmers in eastern Nebraska do not. Whatever you learn about how they respond to water prices is solid for that basin. Extending it to the whole state is a claim you would have to argue for. This is not a reason to throw the data away. It is a reason to be careful about the sentence you write in your abstract.

</details>

+ Population: you are interested in the impact of water price on irrigation by farmers in Nebraska
+ Sample (example): data on water price, irrigation water use, and many other things for 500 farmers who farm in the Upper Republican Basin (southwest corner of NE)

:::{.callout-note title=Question}
Is the sample representative of the population?
:::

:::
<!--end of panel-->

:::
<!--end of panel--> 

## Simple univariate model

::: {.panel-tabset}

### What is it?

<details class="transcript">
<summary>Transcript</summary>

Here is the object the rest of this course is built on. Y equals beta-zero, plus beta-one times x, plus u. Read it as a claim about the population: this is how the world actually works, and we are saying it out loud. Y is what you want to explain, x is what you are explaining it with, and u collects everything else that moves y and that you did not put in the model. The betas are just numbers that give the sentence a size. Now read the callout, because that is the humbling part. You will never see this equation. It belongs to the population, and all you ever get is a sample.

</details>

Consider a phenomenon in the population that is correctly represented by the following model (<span style = "color: blue;"> This is the model you want to learn about using sample </span>):

```{=tex}
\begin{equation}
y=\beta_0+\beta_1 x + u 
\end{equation}
```

+ $y$: to be explained by $x$ (<span style = "color: blue;"> dependent variable</span>)
+ $x$: explain $y$ (<span style = "color: blue;"> independent variable </span>, <span style = "color: blue;"> covariate </span>, <span style = "color: blue;"> explanatory variable </span>)
+ $u$: parts of $y$ that cannot be explained by $x$ (<span style = "color: blue;"> error term </span>)
+ $\beta_0$ and $\beta_1$: real numbers that gives the model a quantitative meaning (<span style = "color: blue;"> parameters </span>)

::: {.fragment}
:::{.callout-important}
You will never know the **true** model. You can try estimating it using sample! That is what statistics is about. 
:::
:::
<!--end of the fragment-->

### What does $\beta_1$ measure?

<details class="transcript">
<summary>Transcript</summary>

So what does beta-one actually mean? Let's do the algebra rather than assert it. Write y before, with x in it, and y after, with x plus one in it. Everything else is identical, including u. Subtract the first from the second and everything cancels except beta-one. So beta-one is the change in y that comes from a one-unit change in x. But notice the assumption that did all the work: we held u fixed. That is the phrase in the callout, ceteris paribus, everything else equal. Beta-one is a causal effect only under that condition, and arranging for that condition to hold is what makes this subject hard.

</details>

```{=tex}
\begin{align}
y=\beta_0+\beta_1 x + u 
\end{align}
```

If you change $x$ by $1$ unit while holding $u$ (everything else) constant,


```{=tex}
\begin{align}
  y_{before} & = \beta_0+\beta_1 x + u \\
  y_{after} & = \beta_0+\beta_1 (x + 1) + u 
\end{align}
```

The difference in $y_{before}$ and $y_{after}$,

```{=tex}
\begin{align}
  \Delta y = \beta_1
\end{align}
```

That is, $y$ changes by $\beta_1$.

::: {.fragment}
:::{.callout-note title=So,}
+ $\beta_1$ is the change in $y$ when $x$ increases by 1
+ We call $\beta_1$ the <span style = "color: blue;"> ceteris paribus </span> (with everything else fixed) causal impact of $x$ on $y$.
:::
:::
<!--end of the fragment-->

### What does $\beta_0$ measure?

<details class="transcript">
<summary>Transcript</summary>

Beta-zero is the easy one algebraically. Set x to zero and set u to zero, and the equation collapses to y equals beta-zero, so it is the intercept. Under the zero conditional mean condition introduced later, the average error among observations with x equal to zero is also zero. Only under that condition does beta-zero equal the average value of y at x equal to zero. Now read the note, because this is where the intercept gets oversold. If zero is outside the observed range of x, beta-zero is an extrapolation and may have little useful interpretation. We normally retain an intercept so the model is not unnecessarily constrained to pass through the origin.

</details>

```{=tex}
\begin{align}
y=\beta_0+\beta_1 x + u 
\end{align}
```

When $x = 0$ and $u=0$,

```{=tex}
\begin{align}
y=\beta_0
\end{align}
```

So, $\beta_0$ represents the intercept.

Under the zero conditional mean condition $E[u|x]=0$ (see the <span style = "color: blue;"> Crucial condition </span> tab in the next section), $\beta_0 = E[y|x=0]$: the average value of $y$ among those with $x=0$.

:::{.callout-note}
$\beta_0$ is most useful when $x=0$ is substantively meaningful and within the observed range. Otherwise it is an extrapolation. The intercept should normally remain in the model so the regression is not forced through the origin, but it may not deserve a substantive story.
:::

### Visualized

<details class="transcript inline">
<summary>Transcript</summary>

The same thing as a picture, and I want you to connect each symbol to something you can see. The blue dot on the vertical axis is beta-zero. It is where the line crosses x equals zero, and here it is one. The red staircase is beta-one. Move one unit to the right along x, and the line rises by two, so beta-one is two. That is all a coefficient is: the rise for a one-unit run. Hold on to this picture, because next lecture we start estimating these two numbers from data, and the whole exercise is deciding where to put this line when the points are scattered around it.

</details>

::: {.columns}

::: {.column width="30%"}
<br>

+ $\beta_0$: intercept
+ $\beta_1$: coefficient (slope)

:::
<!--end of the 1st column-->
::: {.column width="70%"}
```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7
#| out-width: "100%"
x <- seq(0, 5, length = 1000)
y <- 1 + 2 * x
(
g_1 <-
  ggplot(data = data.table(y = y, x = x)) +
  geom_line(aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  # beta_0: where the line crosses x = 0
  annotate("point", x = 0, y = 1, size = 3, color = "blue") +
  annotate("text", x = 0.55, y = 1.7, label = "beta[0]==1", parse = TRUE, size = 5.5, color = "blue") +
  # beta_1: one step right, beta_1 steps up
  annotate("segment", x = 2, xend = 3, y = 5, yend = 5, color = "red", linewidth = 0.8) +
  annotate("segment", x = 3, xend = 3, y = 5, yend = 7, color = "red", linewidth = 0.8) +
  annotate("text", x = 2.5, y = 4.2, label = "1 unit of x", size = 4.5, color = "red") +
  annotate("text", x = 3.6, y = 6, label = "beta[1]==2", parse = TRUE, size = 5.5, color = "red") +
  # 9in wide inside a 70% column is a narrower canvas than the shared 10in one
  # shown at 75%, so it needs its own, smaller base_size to match.
  lecture_theme(14)
)
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

:::
<!--end of panel-->


```{r data-gen-1}
#| include: false
set.seed(9947398)
N <- 500
ab_A <- pmax(rnorm(N) * 1.5 + 7, 0)
ab_B <- pmax(rnorm(N) * 1.2 + 5, 0)
inc_A <- 60 + 10 * ab_A
inc_B <- 40 + 10 * ab_B
inc_A_e <- inc_A + 30 * rnorm(N)
inc_B_e <- inc_B + 30 * rnorm(N)
mean_inc_A <- mean(inc_A_e)
mean_inc_B <- mean(inc_B_e)
data_A <- data.table(x = ab_A, y = inc_A_e, yline = inc_A, University = "A")
data_B <- data.table(x = ab_B, y = inc_B_e, yline = inc_B, University = "B")
data <- rbind(data_A, data_B)
```


## Why do we want <span style = "color: blue;"> ceteris paribus </span> causal impact?

::: {.panel-tabset}

### Example

<details class="transcript">
<summary>Transcript</summary>

Let's make ceteris paribus concrete, with a decision you might actually face. You have been admitted to University A, which is better and more expensive, and to University B. You want to know what A buys you in future income. So you find the data on screen: A's graduates average about a hundred and thirty-one, B's about ninety, a difference of roughly forty-one. Now the question at the bottom is the only one that matters. Is that forty-one what <em>you</em> would gain by choosing A? Sit with it before clicking on. Almost everybody's instinct is yes, and the rest of this slide is about why the instinct is wrong.

</details>

**Quality of College**

You

+ have been admitted to University A (better, more expensive) and B (worse, less expensive)
+ are trying to decide which school to attend
+ are interested in knowing a boost in your future income to make a decision

**You have found the following data**

```{r}
#| echo: false
inc_table <-
  data[, .(
    `average income` = mean(y) %>% round(digits = 2),
    `sample size` = .N
  ),
  by = University
  ]

flextable(inc_table) %>%
  autofit()
```

**Question**

Should you assume that the observed difference of `r round(mean(inc_A_e) - mean(inc_B_e), digits = 2)` is the expected boost **you** would get if you are to attend University A instead of B?

### Why ceteris paribus impact?

<details class="transcript">
<summary>Transcript</summary>

Here are two quantities written side by side, and the gap between them is the whole lecture. Number one is the raw difference: the average income of A's students minus the average income of B's students. That is what the table just gave you. Number two holds ability fixed at six, your ability, and asks what A pays compared with B for the same person. Which do you want? Obviously the second, and the callout says why. Your innate ability does not improve because you enrolled somewhere. You want to change the school and nothing else. Number one changes the school and the student at the same time.

</details>

Let's say your ability score is $6$ out of $10$ (the higher, the better),

$$\text{(1)}\;\; E[inc|A] -E[inc|B] \;\;\; \text{(the raw difference in the table)}$$
$$\text{(2)}\;\; E[inc|A,ability=6] -E[inc|B,ability=6]$$

Which one would you like to know?

::: {.fragment}
:::{.callout-important title=""}
+ You want ability (an unobservable) to stay fixed when you change the quality of school because your innate ability is not going to miraculously increase by simply attending school A

+ You do not want the impact of school quality to be confounded with something else
:::
:::
<!--end of the fragment-->

:::{.callout-note title="Aside: Conditional Expectation"}
$E[Y|X]$ represents expected value of $Y$ conditional on $X$ (For a given value of $X$, the expected value of $Y$).
:::


### What do you observe?

<details class="transcript inline">
<summary>Transcript</summary>

Now look at what is actually in the data. Each dot is a person, ability along the horizontal axis, income up the vertical. The two sloped lines are the two universities: at any given ability the red line sits above the blue one, and that vertical gap is what A is worth. Then there are the two horizontal lines, which are just the group averages, and their gap is visibly bigger. Why? Look at where the dots sit. A's students are bunched to the right; they had higher ability before they ever enrolled. The horizontal lines compare different kinds of people. The sloped lines compare like with like.

</details>

:::{.callout-note title=""}
+ red line: $E[income|A, ability]$
+ blue line: $E[income|B, ability]$
:::

```{r cp}
#| echo: false
g_CI <-
  ggplot(data = data) +
  geom_point(aes(x = x, y = y, color = University), size = 0.4) +
  geom_smooth(aes(x = x, y = yline, color = University), method = "lm", se = FALSE, linewidth = 0.8) +
  geom_hline(yintercept = mean_inc_A, color = "red") +
  geom_hline(yintercept = mean_inc_B, color = "blue") +
  # hjust = 0 keeps the label inside the panel: x = 0.7 is left of the data, so
  # a centred label straddles the axis.
  annotate("text", x = 0.7, y = mean_inc_A + 12, label = paste(round(mean_inc_A, 0), sep = ""), family = "Times", color = "red", size = 6, hjust = 0) +
  annotate("text", x = 0.7, y = mean_inc_B - 12, label = paste(round(mean_inc_B, 0), sep = ""), family = "Times", color = "blue", size = 6, hjust = 0) +
  xlab("ability") +
  ylab("Annual Income") +
  theme_lecture +
  theme(
    legend.position = "bottom"
  )

g_CI
```

### Putting numbers on it

<details class="transcript">
<summary>Transcript</summary>

Let's put numbers on that picture, because the size of the mistake matters. A's students average an ability of seven, B's average five, and every point of ability is worth ten of income. So two points of ability is twenty of income before school does anything at all. And at the same ability, A pays sixty against B's forty, so the true ceteris paribus effect is twenty. The observed gap was about forty-one. Twenty of school, twenty-one of ability. Read the fragment: take the raw number at face value and you value University A at roughly double what it is worth. Getting from the first number to the second is what this course is for.

</details>

A's students average an ability of `r round(mean(ab_A), 1)`, B's average `r round(mean(ab_B), 1)`, and each point of ability is worth $10$ of income. At the <span style = "color: blue;"> same </span> ability, A pays $60$ and B pays $40$.

+ **(1)** the gap between the two <span style = "color: blue;"> horizontal </span> lines is `r round(mean(inc_A_e) - mean(inc_B_e), 1)` &nbsp; &larr; what you see
+ **(2)** the gap between the two <span style = "color: blue;"> sloped </span> lines is $20$ &nbsp; &larr; what attending A does for you
+ the remaining `r round(mean(inc_A_e) - mean(inc_B_e) - 20, 1)` is <span style = "color: red;"> ability, not school </span>

::: {.fragment}
:::{.callout-important title=""}
Take (1) at face value and you credit University A with roughly twice what it is worth. The whole of this course is machinery for getting from (1) to (2).
:::
:::
<!--end of the fragment-->

:::
<!--end of panel-->

## Example: corn yield and fertilizer

::: {.panel-tabset}

### Model

<details class="transcript">
<summary>Transcript</summary>

Let's switch to an agricultural example, because I want you practising how to spot the error term in the wild. Yield equals beta-zero, plus beta-one times fertilizer, plus u. The question is deliberately open: what is in u? Take a moment and actually list some. Rainfall. Soil organic matter. Drainage. Pest pressure. The farmer's skill. Whether the field got planted on time. All of it is in there, because u is defined as everything that affects yield and is not fertilizer. Notice how long that list is, and how little of it anybody measures. That is the normal situation, not a pathological one.

</details>

**Corn yield and fertilizer**

```{=tex}
\begin{align}
  yield=\beta_0+\beta_1 fertilizer+u 
\end{align}
```

**Question**

What is in the error term?

### Estimate $\beta_1$

<details class="transcript">
<summary>Transcript</summary>

So here is the position you are in. You do not know beta-zero or beta-one, because those belong to the population. What you have is a stack of observations, a yield and a fertilizer rate for each of n fields. And you want beta-one to be the ceteris paribus effect: what fertilizer does with everything else held fixed. Read the question at the bottom slowly, because it sounds impossible. You are being asked to hold constant a list of things you cannot see. Rainfall, soil, drainage, none of it is in your data. How do you hold fixed what you never measured? The answer is on the next tab, and it is more elegant than you would guess.

</details>

```{=tex}
\begin{align}
  yield=\beta_0+\beta_1 fertilizer+u 
\end{align}
```

+ you do not know $\beta_0$ and $\beta_1$, and would like to estimate them
+ you observe a series of $\{yield_i,fertilizer_i\}$ combinations $(i=1,\dots,n)$
+ you would like to estimate $\beta_1$, the impact of fertilizer on yield, **ceteris paribus** (with everything else fixed)

**Question**

How could we possibly find the **ceteris paribus** impact of fertilizer on yield when we do not observe whole bunch of other factors (error term)?

### Crucial condition

<details class="transcript">
<summary>Transcript</summary>

Here is the answer, and it is one line: the expected value of u given x is zero. Read the Why block underneath it. You need the conditional average of the unobservables not to change with x, so that when you look at fields with more fertilizer, everything else about them averages out the same as on fields with less. If that holds, differences in expected yield across fertilizer rates trace the regression function. Interpreting that relationship causally also requires x to represent a well-defined intervention and the model to be correctly specified. Remember that merely having zero correlation is weaker, and including an intercept only normalizes the unconditional average error; neither fact by itself establishes zero conditional mean.

</details>

It turns out we can identify the ceteris paribus causal impact of $x$ on $y$ as long as the following condition is satisfied:

:::{.callout-important title="Zero conditional mean"}
$E(u|x) = 0$
:::

**Why this is the condition.** $\beta_1$ is a ceteris paribus effect, so it is only recoverable if the rest of the world does not move when $x$ does. $E[u|x]=0$ says exactly that: whatever value of $x$ you look at, the unobservables average out to the same thing. Any difference in $y$ across values of $x$ is then attributable to $x$ alone.

This is satisfied when $E[u|x]=E[u]$ and the intercept normalization gives $E[u]=0$. Two distinctions matter:


:::{.callout-important}
+ $E[u|x]=0$ implies that $u$ is uncorrelated with $x$, but zero correlation alone does <span style = "color: red;">not</span> imply zero conditional mean
+ including an intercept can normalize $E[u]=0$, but it does <span style = "color: red;">not</span> make $E[u|x]=0$ true
:::

### Condition satisfied?

<details class="transcript">
<summary>Transcript</summary>

Now do it yourself, on the model at the top of the screen, with the data described in the middle: two hundred farmers, their yields and their fertilizer rates, in 2023. Two questions. First, what is in u? And read the parenthesis carefully, because only things that actually affect yield count; the farmer's favourite colour is not in the error term. Second, and this is the one that decides everything, is any of it correlated with fertilizer? Think about how a farmer picks a rate. They look at their soil, their past yields, their expectations for the season. They are choosing fertilizer using the very things sitting inside u. So no, the condition is not satisfied.

</details>

**Model**

```{=tex}
\begin{align}
  yield=\beta_0+\beta_1 fertilizer + u 
\end{align}
```

<br>

**Data**

You have collected farm-level yield-fertilizer data from 200 farmers in year 2023.

<br>

**Questions**

+ What's in $u$? (note that factors that do not affect yield are not part of $u$)
+ Is it correlated with fertilizer?


### Math asides

::: {.panel-tabset}
#### Mean independence

<details class="transcript">
<summary>Transcript</summary>

These next three tabs are the formal versions of what we just said in words. You can follow the course without them, but they are what a referee will expect you to know. Mean independence says the expected value of u given x equals the expected value of u. Verbally, as the first bullet puts it: the average of the error term is the same no matter which value of x you look at, and that common average is just the overall average. It is the precise statement of the intuition from two tabs back. In practice we use it almost interchangeably with saying the error is uncorrelated with x, and the next tab is about how much work "almost" is doing.

</details>

:::{.callout-note title="Definition: Mean Independence"}
$E[u|x]=E[u]$
:::

+ **verbally**: the average value of the error term (collection of all the unobservables) is the same at any value of $x$, and that the common average is equal to the average of $u$ over the entire population

+ **(almost) interchangeably**: the error term is not correlated with $x$

#### Correlation and mean independence

<details class="transcript">
<summary>Transcript</summary>

Here is how much work "almost" was doing. Mean independence implies zero correlation, but zero correlation does not imply mean independence. Follow the derivation. The covariance of u and x is the expectation of u times x, minus the product of the two expectations, and since the error averages to zero, that second piece drops out. Then apply the law of iterated expectations: what is left equals the expectation over x of x times the expected value of u given x. If the zero conditional mean condition holds, the inside is zero everywhere, so the whole thing is zero. The reverse fails because u can be uncorrelated with x while still depending on it non-linearly.

</details>

Mean independence of $u$ and $x$ implies no correlation. But, no correlation does not imply mean independence.

```{=tex}
\begin{aligned}
    Cov(u,x)= & E[(u-E[u])(x-E[x])] \\\\
    = & E[ux]-E[u]E[x]-E[u]E[x]+E[u]E[x]\\\\
    = & E[ux]-E[u]E[x] \\\\
    = & E[ux] \;\; \text{(since } E[u]=0 \text{)} \\\\
    = & E_x\big[x\,E_u[u|x]\big] \;\; \text{(law of iterated expectations)}
\end{aligned}
```

If zero conditional mean condition $(E(u|x)=0)$ is satisfied,

```{=tex}
\begin{aligned}
    Cov(u,x)= & E_x[x \cdot 0] = 0
\end{aligned}
```

#### $E(u)=0$

<details class="transcript">
<summary>Transcript</summary>

The last piece is the claim that the error term averages to zero, which sounds like an assumption you have to earn but is actually free. Suppose it is not zero; suppose u averages to some alpha. Watch the algebra. Add alpha and subtract alpha on the right-hand side. Fold the plus alpha into the intercept and call the result gamma-zero; fold the minus alpha into the error and call it u-two. Now u-two averages to zero by construction, and you have exactly the same model. Nothing was assumed. All it cost is that the intercept quietly absorbed the difference, which is the real reason you always include one.

</details>

Expected value of the error term is 0 $(E(u)=0)$.

This is always satisfied as long as an intercept is included in the model:

$$y = \beta_0 + \beta_1 x + u_1,\;\; \text{where}\;\; E(u_1)=\alpha$$

Rewriting the model,

```{=tex}
\begin{aligned}
y & = \beta_0 + \alpha + \beta_1 x + u_1 - \alpha \\\\
  & = \gamma_0 + \beta_1 x + u_2
\end{aligned}
```

where, $\gamma_0=\beta_0+\alpha$ and $u_2=u_1-\alpha$. 

Now, $E[u_2]=0$.
:::
<!--end of panel-->

:::
<!--end of panel-->


```{r college_income_data}
#| echo: false
# NOTE: these objects are deliberately suffixed "_r" (for "randomized") so that
# they do not overwrite the first dataset built in `data-gen-1`, which earlier
# slides still refer to in inline R.
set.seed(38943)
N_r <- 1000
college_effect <- 20
ab_A_r <- pmax(rnorm(N_r) * 1.1 + 5, 0)
ab_B_r <- pmax(rnorm(N_r) * 1.1 + 5, 0)
inc_A_r <- 40 + college_effect + 10 * ab_A_r
inc_B_r <- 40 + 10 * ab_B_r
inc_A_e_r <- inc_A_r + 20 * rnorm(N_r)
inc_B_e_r <- inc_B_r + 20 * rnorm(N_r)
mean_inc_A_r <- mean(inc_A_e_r)
mean_inc_B_r <- mean(inc_B_e_r)
data_cI <- rbind(
  data.table(x = ab_A_r, y = inc_A_e_r, yline = inc_A_r, University = "A"),
  data.table(x = ab_B_r, y = inc_B_e_r, yline = inc_B_r, University = "B")
)
```


## Going back to the college-income example

::: {.panel-tabset}

### The model

<details class="transcript">
<summary>Transcript</summary>

Now we bring the machinery back to the college question we opened with. Here is the model: income equals beta-zero, plus beta-one times College A, plus u. That College A term is a variable equal to one if you attend A and zero if you attend B. It is your first dummy variable, and it makes beta-one literally the difference between the two schools. And u, as always, is everything else, which very much includes ability. So ask the question in the second half of the slide: is the expected value of u, given which college you attend, zero? In plain English, is who ends up at A systematically related to ability?

</details>

$$
Income = \beta_0+\beta_1 College\;\; A + u
$$

where $College\;\; A$ is 1 if attending college A, 0 if attending college B, and $u$ is the error term that includes ability.

<br>

**Zero conditional mean satisfied?**

$$
E[u(ability)|college A] = 0?
$$

That is, are attending college A and ability systematically related with each other? Or, is college choice (and acceptance of course) correlated with ability?

### What you observe

<details class="transcript inline">
<summary>Transcript</summary>

Look at the figure and answer that question with your eyes. The two horizontal lines are the group averages, and the callout tells you their gap is the forty-one you started with. The two sloped lines are what A and B pay at any given ability, and their gap is twenty. Those are different numbers because the dots are not spread the same way: A's cloud sits to the right. Higher-ability students got into A and chose A. So ability, sitting inside u, is correlated with which college you attend. The zero conditional mean condition fails, and beta-one from the raw comparison is not the causal effect.

</details>

:::{.callout-note title=""}
The horizontal lines are the two group averages: their gap is the `r round(mean(inc_A_e) - mean(inc_B_e), 1)` from the table. The sloped lines are the two groups at a <span style = "color: blue;"> given </span> ability: their gap is $20$.
:::

```{r}
#| echo: false
g_CI
```

### If choice and ability were unrelated

<details class="transcript inline">
<summary>Transcript</summary>

And here is the counterfactual world, so you can see what success looks like. Same model, but now college choice has nothing to do with ability, and both clouds cover the same range. Look at the horizontal lines. Their gap is now essentially the same as the vertical gap between the sloped lines, which is the true effect of twenty. Read the callout. Nothing about our ignorance changed. We still never measured ability, and it is still sitting in the error term. What changed is that it is no longer correlated with the college variable, so the raw difference finally is the causal effect. This picture is what a randomized experiment buys you.

</details>

This is what it would look like if college choice and ability were not correlated:

```{r}
#| echo: false
ggplot(data = data_cI) +
  geom_point(aes(x = x, y = y, color = University), size = 0.5) +
  geom_smooth(aes(x = x, y = yline, color = University), method = "lm", se = FALSE) +
  geom_hline(yintercept = mean_inc_A_r, color = "red") +
  geom_hline(yintercept = mean_inc_B_r, color = "blue") +
  annotate(
    "text",
    x = 0.7,
    y = mean_inc_A_r + 6,
    label = paste(round(mean_inc_A_r, 0), sep = ""),
    family = "Times",
    color = "red",
    size = 6,
    hjust = 0
  ) +
  annotate(
    "text",
    x = 0.7,
    y = mean_inc_B_r - 6,
    label = paste(round(mean_inc_B_r, 0), sep = ""),
    family = "Times",
    color = "blue",
    size = 6,
    hjust = 0
  ) +
  xlab("ability") +
  ylab("Annual Income") +
  theme_lecture +
  theme(
    legend.position = "bottom"
  )
```

:::{.callout-important title=""}
Both groups now have the same ability distribution, so the gap between the group averages (`r round(mean_inc_A_r - mean_inc_B_r, 0)`) is, up to sampling noise, the <span style = "color: blue;"> same </span> as the $20$ gap at any fixed ability. $E[u|College\;A]=0$ holds, and the raw difference is finally the causal effect.
:::

:::
<!--end of panel-->

## Exercise

<details class="transcript">
<summary>Transcript</summary>

Before the next lecture, do this for a question of your own, and I do mean your own: something you might actually write about. Name the thing you want to explain and the thing you think explains it. Write the simple linear model down, with an intercept, a slope, and an error term. Then do the part that matters. List what is inside that error term, concretely and honestly, and name five things. Finally, for each one, ask whether it is correlated with your explanatory variable. If any of them are, you have just found the problem your thesis will have to solve, and you have found it early.

</details>

+ consider a phenomenon you are interested in understanding
  - dependent variable (variable to be explained) 
  - explanatory variable (variable to explain) 
+ construct a simple linear model
+ identify what is in the error term
+ check if they are correlated with the explanatory variable or not
