---
title: "02: Multivariate Regression"
format: 
  revealjs:
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to course website with lecture slides](https://tmieno2.github.io/Applied-Econometrics-MS-Quarto/lectures/)"
webr:
  packages: ['fixest', 'dplyr', 'data.table', 'ggplot2', 'modelsummary', 'gt']
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 10
    fig-height: 6
    out-width: 100%
filters:
  - webr
---


# Multivariate Regression: Introduction

```{r}
#| include: false
library(fixest)
library(modelsummary)

# R-side copy of the data the "Implementation (R)" webr cell simulates, so the
# regression table on the "Present results" tab reports that same regression.
# The seed matches the one in the webr cell.
set.seed(2024)
N <- 100
x1 <- rnorm(N)
x2 <- rnorm(N)
u <- rnorm(N)
y <- 1 + x1 + x2 + u
data_sim <- data.frame(y = y, x1 = x1, x2 = x2)
reg <- feols(y ~ x1 + x2, data = data_sim)
```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Univariate vs Multivariate Regression Models

```{webr-r}
#| context: setup
# NOTE: a webr setup cell draws a loading widget wherever it sits, so keep it
# inside a content slide rather than on a section divider.
# Mirror of lecture_theme() in lectures/_lecture-theme.R — WebR cannot source
# that file. Edit it first, then run `Rscript lectures/check-lecture-theme.R`,
# which fails if this copy has drifted.
lecture_theme <- function(base_size = 16) {
  ggplot2::theme_bw(base_size = base_size) +
    ggplot2::theme(
      axis.text = ggplot2::element_text(size = base_size * 0.9),
      axis.title = ggplot2::element_text(size = base_size),
      legend.text = ggplot2::element_text(size = base_size * 0.9),
      legend.title = ggplot2::element_text(size = base_size * 0.9),
      plot.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.key = ggplot2::element_rect(fill = "transparent", colour = NA),
      # theme_bw's grey92 grid was chosen against a white panel and all but
      # disappears on the warm paper once the panel stops being opaque.
      panel.grid = ggplot2::element_line(colour = "grey87")
    )
}
# The WebR canvas is 10x6in at out-width 100% of its output track
# (lectures/_metadata.yml), the same canvas the knitr figures use. It asks for
# the whole track because the 420px height cap in notebook.scss is what really
# sizes it. 18pt rather than the 16pt knitr default: on a full slide both land
# at 700x420, where 18pt reads at 17.5px and 16pt at 15.5px, and these decks
# put figures in half-width columns where the smaller size gets hard to read.
# webR draws its canvas at 72 dpi no matter what `dpi` says, so on a 216-dpi
# canvas everything ggplot measures in points or mm comes out a third of its
# size. Scale ggplot2's mm-to-point constants and the theme by the same
# factor and the figure matches the knitr ones. Keep in step with `dpi` above.
webr_scale <- 216 / 72
assignInNamespace(".pt", ggplot2::.pt * webr_scale, "ggplot2")
assignInNamespace(".stroke", ggplot2::.stroke * webr_scale, "ggplot2")
theme_lecture <- lecture_theme(18 * webr_scale) +
  # base_size also sets the theme line widths, which are in mm and already
  # pass through the scaled .pt; pin them to what 18pt gives or they come
  # out 9x thick.
  ggplot2::theme(
    line = ggplot2::element_line(linewidth = 18 / 22),
    rect = ggplot2::element_rect(linewidth = 18 / 22)
  )
ggplot2::theme_set(theme_lecture)
```

::: {.panel-tabset}


### Bi-variate vs. Uni-variate

<details class="transcript">
<summary>Transcript</summary>

Let's start with why we are doing this at all. In the univariate model on the top half of the slide, you explain an outcome, y, with only one independent variable, x. Everything else that affects y but is not represented by x is collected in the error term, u. The crucial condition for an unbiased OLS slope is that the expected value of that error, conditional on x, is zero. In words, once you specify a value of x, the omitted influences on y must average to zero. Outside a randomized experiment, that is often hard to believe. People, firms, counties, and countries differ along many dimensions, and some omitted dimension can easily move together with x.

The multivariate model in the lower half responds by including additional independent variables. Each included variable is moved out of the error term and represented explicitly by its own coefficient. That gives us a chance to compare observations that have the same values of those other included factors. It can therefore remove a particular source of omitted-variable bias. But read the word in blue carefully: the endogeneity problem becomes **less** severe, not automatically solved. Adding regressors only helps with relevant factors that you actually observe, include in the model, and specify correctly. Other omitted factors remain inside u, and any one of them may still be related to the included x variables.

</details>

**Univariate**

The most important assumption $E[u|x] = 0$ (zero conditional mean) is almost always violated (unless your data comes from randomized experiments) because all the other variables are sitting in the error term, which can be correlated with $x$.

<br>

**Multivariate**

More independent variables mean fewer factors left in the error term, which makes the endogeneity problem <span style = "color: blue;"> less </span>severe

### Example

<details class="transcript">
<summary>Transcript</summary>

Here is the move written out with wages. Look at the first line. Wage is modeled as beta-zero plus beta-one times education plus u-one. The parenthesis tells you what u-one contains in this example: u-two plus beta-two times experience. So experience really affects wages, but the univariate equation hides that effect inside its error term.

Why is that a problem for the estimated education coefficient, beta-one-hat? Experience tends to raise wages, and education and experience are generally related because an additional year in school changes the years available for work and because workers with different schooling histories can follow different career paths. When an omitted wage determinant is correlated with education, OLS cannot tell which part of their joint movement should be credited to education. The education coefficient can then absorb some of the omitted experience effect. It is unbiased only in the special case that experience is uncorrelated with education, along with the other conditions we have discussed.

Now look at the second line. Experience has been pulled out of u-one, placed explicitly in the equation, and assigned its own slope, beta-two. The remaining error is called u-two. In this bivariate model, beta-one measures the change in wage associated with one more unit of education while experience is held fixed. Likewise, beta-two measures the change associated with one more unit of experience while education is held fixed. That is the meaning of the blue phrase, "holding experience fixed." The red phrase gives our standard vocabulary: experience is **controlled for**. We are comparing workers at the same modeled experience level, rather than mixing an education comparison with systematic differences in experience. This removes experience as this particular source of omitted-variable bias, although anything still inside u-two can remain a problem.

</details>

**Uni-variate vs. bi-variate**

```{=tex}
\begin{align*}
  \text{Uni-variate}\;\; wage = & \beta_0 + \beta_1 educ + u_1 (=u_2+\beta_2 exper)\\
  \text{Bi-variate}\;\; wage = & \beta_0 + \beta_1 educ + \beta_2 exper + u_2
\end{align*}
```

<br>

**What's different?**

+ **uni-variate**: $\widehat{\beta}_1$ is biased unless experience is uncorrelated with education because experience was in error term

+ **bi-variate**: able to measure the effect of education on wage, <span style = "color: blue;"> holding experience fixed </span> because experience is modeled explicitly (<span style = "color: red;"> We say $exper$ is controlled for. </span>)


### Example

<details class="transcript">
<summary>Transcript</summary>

Same structure, now applied to a policy question you will meet in real debates: does spending more per student raise standardized test scores? In the first line, average score equals beta-zero plus beta-one times expenditure per student plus u-one. If average family income also affects scores but you leave it out, its contribution is inside u-one. The parenthesis makes that bookkeeping explicit: u-one equals u-two plus beta-two times average income.

The danger is that district income is plausibly related to both sides of the question. Higher-income districts may have a larger tax base and spend more per student. Their students may also score differently because of home resources, tutoring, nutrition, parental education, neighborhood conditions, and other channels that are not caused by school expenditure. A regression of score on expenditure alone sees spending and income move together. It can credit spending with score differences that are partly associated with income, so beta-one-hat need not isolate the effect of spending.

In the second line, average income is included explicitly. Beta-one now describes how predicted average score changes with expenditure when average income is held fixed, while beta-two describes how it changes with income when expenditure is held fixed. We say that the model controls for average income.

</details>

The impact of per student spending (`expend`) on standardized test score (`avgscore`) at the high school level

```{=tex}
\begin{align*}
avgscore= & \beta_0+\beta_1 expend + u_1 (=u_2+\beta_2 avginc) \notag \\
avgscore= & \beta_0+\beta_1 expend +\beta_2 avginc + u_2 \notag
\end{align*}
```

### Model (general)

<details class="transcript">
<summary>Transcript</summary>

Now replace the names in those examples with a general bivariate model. The equation says y equals beta-zero plus beta-one times x-one plus beta-two times x-two plus u. Here y is the dependent variable we want to explain. X-one and x-two are the two included independent variables. U contains all determinants of y that the equation does not represent explicitly.

Beta-zero is the intercept. It is the conditional mean predicted by the linear model when both x-one and x-two equal zero. Whether that is substantively meaningful depends on whether zero is a possible or useful value for those variables, but the intercept is still needed to position the regression plane.

Beta-one is the change in the model's predicted y associated with a one-unit increase in x-one, **holding x-two fixed**. You can see that directly by comparing two observations or two hypothetical cases with the same x-two and x-one values one unit apart. The beta-two x-two term is unchanged, beta-zero is unchanged, and the predicted value differs by beta-one. Beta-two has the symmetric interpretation: it is the change in predicted y for a one-unit increase in x-two while x-one is fixed. These are partial effects because each slope describes movement along one regressor while the other included regressor is held constant.

The phrase "holding other factors fixed" refers to the other factors included in the equation. It does not mean that every feature of the world has somehow been fixed. Including x-two creates a conditional comparison and prevents its modeled contribution from being mixed into the x-one slope.

</details>

More generally,

```{=tex}
\begin{align*}
  y=\beta_0+\beta_1 x_1 + \beta_2 x_2 + u
\end{align*}
```

+ $\beta_0$: intercept
+ $\beta_1$: measure the change in $y$ with respect to $x_1$, holding other factors fixed
+ $\beta_2$: measure the change in $y$ with respect to $x_2$, holding other factors fixed

:::
<!--end of panel-->

## The Crucial Condition (Assumption) for Unbiasedness of the OLS Estimator

::: {.panel-tabset}

### Uni-variate v.s. Bi-variate

<details class="transcript">
<summary>Transcript</summary>

The condition for unbiasedness generalizes from one regressor to the full set of regressors. On the univariate side, the model is y equals beta-zero plus beta-one x plus u, and the assumption is that the expected value of u conditional on x equals zero. This means that within every value of x, the positive and negative unobserved influences average out. It is not a claim that every observation has u equal to zero. Individual errors can be positive or negative. The claim is about their conditional average.

On the bivariate side, the model includes x-one and x-two, so the condition becomes: the expected value of u conditional on x-one and x-two equals zero. Read the verbal statement on the slide carefully. For **any combination** of x-one and x-two values, the average of all unobserved determinants of y must be zero. Imagine dividing the population into cells defined jointly by x-one and x-two. Within each cell, there must be no systematic positive or negative component left in u.

This is a joint conditional statement. The assumption is now stated conditional on both included variables because OLS is using variation in each variable after accounting for the other.

</details>

**Uni-variate**

$y = \beta_0 + \beta_1x + u$, 

$E[u|x]=0$

<br>

**Bi-variate**

$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u$, 

+ Mathematically: $E[u|x_1,x_2]=0$
+ Verbally: for any values of $x_1$ and $x_2$, the expected value of the unobservables is zero

### Mean independence condition: example 

<details class="transcript">
<summary>Transcript</summary>

Here is the joint mean independence condition applied to the wage model. The equation says wage equals beta-zero plus beta-one times education plus beta-two times experience plus u. Beta-one is the partial association between education and wage at a fixed experience level, and beta-two is the partial association between experience and wage at a fixed education level. For the OLS estimators of those coefficients to be unbiased, the displayed condition requires the expected value of u, given education and experience, to equal zero.

Translate that into the groups OLS is comparing. Among people with a particular education level and a particular amount of experience, the factors left inside u must average to zero. As you move to another education-experience combination, the average omitted component must still be zero. Suppose innate ability affects wages and is not observed in the data, so it remains part of u. The verbal sentence says the condition would hold if ability were, on average, unrelated to education and experience after conditioning on their values. Say that aloud and ask whether it is credible. If people with greater innate ability tend to obtain more education, then the conditional average of the omitted ability component changes with education. Beta-one-hat can then combine the education association with an ability association.

Now read the callout, because it states the honest limit of controlling for things. Adding experience moved experience **out** of the error term and gave it its own coefficient, so omitted experience can no longer bias the education coefficient through that channel. But ability is still in u. No number of other regressors can move ability itself out if ability was never observed or represented by a valid measured proxy. Multivariate regression shrinks the list of omitted determinants; it does not empty the error term.

</details>

In the following wage model,

```{=tex}
\begin{align*}
wage = & \beta_0 + \beta_1 educ + \beta_2 exper + u
\end{align*}
```

Mean independence condition is

```{=tex}
\begin{align*}
  E[u|educ,exper]=0
\end{align*}
```

**Verbally**:

This condition would be satisfied if innate ability of students is on average unrelated to education level and experience.

:::{.callout-important title="The limit of controlling for things"}
Notice what just happened. Adding $exper$ to the model moved experience <span style = "color: blue;"> out </span> of the error term, so it can no longer bias us. But ability is still in there, and no amount of extra regressors will move it out, because <span style = "color: red;"> you cannot control for something you do not observe </span>. Multivariate regression shrinks the error term. It does not empty it.
:::

:::
<!--end of panel-->

## The model with $k$ independent variables

::: {.panel-tabset}

### General model

<details class="transcript">
<summary>Transcript</summary>

The extension from two independent variables to k independent variables contains no new conceptual step, only more terms. Read the model from left to right: y equals beta-zero, the intercept, plus beta-one times x-one, beta-two times x-two, continuing through beta-k times x-k, plus the error u. The index j can label any slope from one through k. Beta-j is the change in the model's conditional mean of y associated with a one-unit increase in x-j while all of the other included x variables are held fixed. U collects determinants of y not represented by those k regressors and the chosen functional form.

The slide uses beta-OLS as shorthand for the collection of OLS estimators of all these beta coefficients. For those estimators to be unbiased under the other multiple linear regression assumptions, the expected value of u conditional on the **entire vector** x-one through x-k must equal zero. For every joint combination of the regressors, the unobserved component must average to zero. The verbal line uses "uncorrelated" as an informal description. Strictly, the displayed zero conditional mean restriction is stronger than requiring only zero pairwise correlation between u and each x. Zero pairwise correlations do not rule out a systematic nonlinear conditional pattern.

If joint zero conditional mean holds, repeated random samples center each beta-j-hat on its true beta-j. It does not say every estimate equals the truth in one sample, because sampling variation remains. If the condition fails, one or more coefficient estimators may be biased.

</details>

**Model**

```{=tex}
\begin{align*}
  y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_k x_k + u
\end{align*}
```

<br>

**Mean independence assumption?**  

$\beta_{OLS}$ (OLS estimators of $\beta$s) is unbiased if,

```{=tex}
\begin{align*}
    E[u|x_1,x_2,\dots,x_k]=0
\end{align*}
```

**Verbally**: this condition would be satisfied if the error term is uncorrelated with any of the independent variables, $x_1,x_2,\dots,x_k$. 

### Implementation (R)

<details class="transcript">
<summary>Transcript</summary>

Here is the general model implemented in R. Start at the top. The library line loads the fix-est package, which supplies fee-ols, the estimation function used throughout this course. Set-seed with twenty twenty-four fixes the pseudorandom number sequence, so every student who runs the cell gets the same simulated sample and therefore the same regression results. Capital N is set to one hundred, our sample size.

The three calls to r-norm each draw one hundred independent standard normal values. X-one and x-two are the independent variables, and u is the error term. The next line constructs y as one plus x-one plus x-two plus u. Because we wrote the data-generating process ourselves, we know the population parameters: the intercept is one, the slope on x-one is one, and the slope on x-two is one. Data-frame then combines y, x-one, and x-two into a rectangular data set with one hundred rows. Notice that u is not included in data-sim. In real data the error is unobserved, and the regression has to infer residuals after fitting the model.

The estimation line calls fee-ols. To the left of the tilde is the dependent variable y. To the right, x-one plus x-two tells R to include both regressors; an intercept is included by default. The data argument tells fee-ols to find those variable names inside data-sim. That small formula change is the syntactic move from a univariate to a multivariate regression, even though the interpretation and assumptions require much more thought.

The final line prints reg. When I run this exact code, the estimates are about one point zero seven five for the intercept, zero point nine two six for x-one, and zero point nine nine two for x-two. They are close to, but not exactly, the true values of one because this is one random sample of only one hundred observations and u creates sampling noise. Unbiasedness does not promise exact recovery in every sample. It says that if we repeatedly generated samples this way, the estimates would average to the true coefficients.

</details>

```{webr-r}
#| autorun: true

#--- load the fixest package ---#
library(fixest)

#--- generate data ---#
set.seed(2024) # so everyone sees the same numbers
N <- 100 # sample size
x1 <- rnorm(N) # independent variable
x2 <- rnorm(N) # independent variable
u <- rnorm(N) # error
y <- 1 + x1 + x2 + u # dependent variable
data_sim <- data.frame(y = y, x1 = x1, x2 = x2)

#--- OLS ---#
reg <- feols(y ~ x1 + x2, data = data_sim)

#--- print the results ---#
reg
```

Note the true values: the intercept is $1$, and both slopes are $1$. Compare them with what OLS returns.

### Present results

<details class="transcript inline">
<summary>Transcript</summary>

This is a practical reporting tab. When you hand in an assignment or write your final paper, do not paste the raw console printout from fee-ols. A regression table lets a reader see estimates, uncertainty, sample size, and model information in a standard layout. On the left is the code that produces the table, and on the right is its rendered result.

The first line loads model-summary, the package that supplies m-summary. The next fee-ols call recreates the regression from the previous tab: y is on the left of the tilde, x-one plus x-two is on the right, and data-sim supplies the variables. Then m-summary receives the fitted model object reg.

The stars equals true argument adds conventional significance markers. Read them using the key at the bottom of the table: three stars correspond to a p-value below zero point zero zero one, two below zero point zero one, one below zero point zero five, and the plus sign below zero point one. A star is not an effect size and is not evidence that an estimate is substantively important. It is only a compact statement about a hypothesis test under the reported standard error.

The g-o-f omit argument uses the pattern shown to suppress selected goodness-of-fit rows whose names contain items such as I-C, Log, Adj, F, Pseudo, or Within. The vertical bars inside the quoted pattern mean "or," so a row matching any one of those pieces is omitted. Keep that option as written for course work unless you have a reason to report a different set. It does not change the regression; it only changes which summary rows are printed.

Now read the right-hand table. Column one reports the same fitted model as the previous tab. Each coefficient estimate appears first, with its standard error directly underneath in parentheses. The intercept is about one point zero seven five with a standard error of zero point zero nine one, x-one is about zero point nine two six with a standard error of zero point zero eight nine, and x-two is about zero point nine nine two with a standard error of zero point zero eight nine. All three receive three stars in this simulation. Below them, the table retains one hundred observations, R-squared of zero point seven two zero, R-M-S-E of zero point eight nine, and the label I-I-D for the standard errors. This is the form readers expect: a clear estimate, its uncertainty, and enough model information to understand what was fitted.

</details>

When you are asked to present regression results in assignments or your final paper, use the `msummary()` function from the `modelsummary` package.

::: {.columns}

::: {.column width="50%"}

<br>

```{r}
#| eval: false
library(modelsummary)

#--- the regression from the previous tab ---#
reg <- feols(y ~ x1 + x2, data = data_sim)

#--- report regression table ---#
msummary(
  reg,
  # keep these options as they are
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r}
#| echo: false
msummary(
  reg,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|F|Pseudo|Within"
)
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


### (Math: derive OLS estimator)

::: {.panel-tabset}

#### The problem

<details class="transcript">
<summary>Transcript</summary>

The derivation is the same least-squares problem as lecture one, part two, now with more unknown coefficients. The first sentence states the goal: find the combination of beta estimates that minimizes the sum of squared residuals. A residual is the observed outcome minus the outcome predicted by the fitted equation.

Theta-hat is introduced as shorthand for the whole collection of estimates: beta-zero-hat, beta-one-hat, continuing through beta-k-hat. There are k slope coefficients because there are k independent variables, plus one intercept, so theta-hat contains k plus one unknown numbers. The word "min" under theta-hat means that we choose all k plus one of them jointly, not one at a time while treating the others as fixed estimates.

Now look inside the square brackets. For observation i, y-i is the observed dependent variable. The expression in parentheses is its fitted value: beta-zero-hat plus beta-one-hat times x-one-i, beta-two-hat times x-two-i, continuing through beta-k-hat times x-k-i. Observed y-i minus that fitted value is u-i-hat, the residual for observation i. Squaring prevents positive and negative residuals from cancelling and penalizes large misses more heavily. The summation from i equals one to n adds those squared misses across all n observations.

OLS chooses the coefficients that make this total as small as possible. Compared with the univariate problem, the criterion has not changed at all. The bracket simply has k regressor terms instead of one, and we optimize over k plus one unknowns instead of two.

</details>

**OLS**

Find the combination of $\beta$s that minimizes the sum of squared residuals

<br>

**So,**

Denoting the collection of $\widehat{\beta}$s as $\widehat{\theta} (=\{\widehat{\beta}_0,\widehat{\beta}_1,\dots,\widehat{\beta}_k\})$,

```{=tex}
\begin{align*}
    Min_{\widehat{\theta}} \sum_{i=1}^n \Big[ y_i-(\widehat{\beta}_0+\widehat{\beta}_1 x_{1,i} + \widehat{\beta}_2 x_{2,i} + \dots + \widehat{\beta}_k x_{k,i}) \Big]^2
\end{align*}
```

Exactly the univariate problem from lecture 01-2, with $k+1$ unknowns instead of two: an intercept and $k$ slope coefficients.

#### The FOCs

<details class="transcript">
<summary>Transcript</summary>

To find an interior minimum, partially differentiate the sum of squared residuals with respect to every element of theta-hat and set each derivative equal to zero. There are k plus one coefficients, so there are k plus one first-order conditions.

Start with beta-zero-hat. Each residual contains minus beta-zero-hat, so its derivative with respect to the intercept is minus one. Applying the chain rule to a squared residual gives two times the residual times minus one. Summing across observations produces minus two times the sum of the residuals. Set that equal to zero and divide by minus two. That leaves the first displayed equation: the residuals sum to zero.

For beta-one-hat, the residual contains minus beta-one-hat times x-one-i, so the derivative is minus x-one-i. The chain rule therefore gives minus two times the sum of x-one-i times the residual. After setting it to zero and dividing by minus two, we get the second equation. The same logic for beta-two-hat yields the sum of x-two-i times the residuals equal to zero. Continue this pattern through beta-k-hat, which gives the last displayed equation with x-k-i. The vertical dots mean that every intervening regressor has its own condition of exactly this form.

Look at the shared bracket in every line. It is y-i minus the fitted value, so it is u-i-hat. The intercept condition says the fitted residuals have sample mean zero because their sum is zero. Each slope condition says the sample cross-product of that regressor with the residual is zero. Since the residual mean is zero and the model includes an intercept, this is also zero sample covariance. OLS makes residuals orthogonal to every included regressor by construction.

</details>

Partially differentiate the objective function (sum of squared residuals) with respect to each of $\widehat{\theta} (=\{\widehat{\beta}_0,\widehat{\beta}_1,\dots,\widehat{\beta}_k\})$ and set each to zero. Dividing each by $-2$ leaves $k+1$ equations:

```{=tex}
\begin{align*}
  \sum_{i=1}^n \Big[ y_i-(\widehat{\beta}_0+\widehat{\beta}_1 x_{1,i} + \widehat{\beta}_2 x_{2,i} + \dots + \widehat{\beta}_k x_{k,i}) \Big] = & 0 \;\; (\widehat{\beta}_0) \\
  \sum_{i=1}^n x_{i,1}\Big[ y_i-(\widehat{\beta}_0+\widehat{\beta}_1 x_{1,i} + \widehat{\beta}_2 x_{2,i} + \dots + \widehat{\beta}_k x_{k,i}) \Big]= & 0  \;\; (\widehat{\beta}_1) \\
  \sum_{i=1}^n x_{i,2}\Big[ y_i-(\widehat{\beta}_0+\widehat{\beta}_1 x_{1,i} + \widehat{\beta}_2 x_{2,i} + \dots + \widehat{\beta}_k x_{k,i}) \Big]= & 0  \;\; (\widehat{\beta}_2) \\
  \vdots \\
  \sum_{i=1}^n x_{i,k}\Big[ y_i-(\widehat{\beta}_0+\widehat{\beta}_1 x_{1,i} + \widehat{\beta}_2 x_{2,i} + \dots + \widehat{\beta}_k x_{k,i}) \Big]= & 0  \;\; (\widehat{\beta}_k)
\end{align*}
```

#### More succinctly

<details class="transcript">
<summary>Transcript</summary>

This tab replaces the long bracket on the previous tab with its name, u-i-hat, so you can see the structure of the normal equations. The first line is the intercept condition: sum u-i-hat from i equals one to n is zero. With an intercept, the fitted residuals therefore average to zero.

Every remaining line corresponds to one slope. The sum of x-one-i times u-i-hat is zero for beta-one-hat; the sum of x-two-i times u-i-hat is zero for beta-two-hat; the pattern continues through the sum of x-k-i times u-i-hat for beta-k-hat.

Count the system. One intercept condition plus k slope conditions gives k plus one equations. The unknowns are the intercept estimate plus k slope estimates, also k plus one.

The final sentence connects this derivation to the variance estimator later in the deck. We estimate k plus one coefficients from n observations, and those fitted coefficients impose k plus one independent restrictions on the residuals. The residual sum of squares therefore has n minus k minus one residual degrees of freedom.

</details>

Every bracket above is just the residual $\widehat{u}_i$, so the whole system is:

```{=tex}
\begin{align*}
  \sum_{i=1}^n \widehat{u}_i = & 0 \;\; (\widehat{\beta}_0) \\
  \sum_{i=1}^n x_{i,1}\widehat{u}_i = & 0  \;\; (\widehat{\beta}_1) \\
  \sum_{i=1}^n x_{i,2}\widehat{u}_i = & 0  \;\; (\widehat{\beta}_2) \\
  \vdots \\
  \sum_{i=1}^n x_{i,k}\widehat{u}_i = & 0  \;\; (\widehat{\beta}_k)
\end{align*}
```

$k+1$ equations in $k+1$ unknowns. That count is what costs you $k+1$ degrees of freedom later.

:::
<!--end of panel-->

:::
<!--end of panel-->

# Small Sample Properties

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Unbiasedness of OLS Estimators

::: {.panel-tabset}

### Conditions (four)

<details class="transcript">
<summary>Transcript</summary>

The callout gives the conclusion for this section: OLS estimators in a multivariate model are unbiased if four multiple linear regression conditions are satisfied. These are the same basic conditions from lecture one, part three, now labeled M-L-R because the model has multiple regressors.

Condition one, M-L-R one, says the model is correctly specified. The outcome must be represented in a form that is linear in the unknown beta parameters, with the relevant regressors and appropriate transformations included. If the true relationship uses a log, a square, or an omitted interaction and your equation leaves that structure in u, the error can vary systematically with x. Calling the model correct is a substantive modeling claim, not something OLS can guarantee for you.

Condition two, M-L-R two, is random sampling. The n observations should be drawn randomly from the population of interest, so the sample carries the population relationship rather than a selection rule tied to the outcome or error.

Condition three, M-L-R three, is no perfect collinearity. None of the independent variables can be an exact linear combination of the others, and each regressor must have usable variation after the others are accounted for. This generalizes the univariate requirement that x varies. Without it, the data cannot separately identify all the slopes.

Condition four, M-L-R four, is zero conditional mean: expected u given x-one, x-two, through x-k equals zero. For every joint combination of the included regressors, the omitted component must average to zero. This is the condition that rules out endogeneity and gives OLS its unbiasedness. Adding regressors changes exactly what it asks: the mean of u must be zero conditional on all of them at once. These four conditions concern unbiasedness.

</details>

:::{.callout-important}
OLS estimators of multivariate models are unbiased if the following conditions are satisfied.
:::

<br>

::: {.columns}

::: {.column width="50%"}
**Condition 1**

Your model is correct (Assumption $MLR.1$)
:::
<!--end of the 1st column-->
::: {.column width="50%"}
**Condition 2**

Random sampling (Assumption $MLR.2$)
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

::: {.columns}

::: {.column width="50%"}
**Condition 3**

No perfect collinearity (Assumption $MLR.3$)
:::
<!--end of the 1st column-->
::: {.column width="50%"}
**Condition 4**

Zero Conditional Mean (Assumption $MLR.4$)

$E[u|x_1,x_2,\dots,x_k]=0$
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

### Perfect Collinearity

<details class="transcript">
<summary>Transcript</summary>

No perfect collinearity is the genuinely new-looking condition in the multivariate list. The slide states that an independent variable cannot be a linear function of the other variables. More precisely, no regressor column can be an exact linear combination of the remaining regressor columns, including the intercept. The word **perfect** matters. High correlation can make an estimate imprecise, but it does not make OLS algebraically impossible. Exact dependence does.

The displayed wage equation gives a deliberately silly example. It includes education once under the abbreviated name pronounced "ed-yook," and again as three times education. Whenever education rises by one year, the second regressor rises by exactly three. There is no observation in which those two variables move independently. The fitted contribution is beta-one times education plus beta-two times three education, which can be rewritten as beta-one plus three beta-two, all multiplied by education. The data can identify that combined coefficient, but it cannot determine a unique beta-one and beta-two separately. Infinitely many pairs give the same fitted values and the same residual sum of squares.

This becomes practically important when we study dummy variables. For now, remember the identification question: after holding all other regressors fixed, must there be some independent variation left in x-j to estimate beta-j?

</details>

**No Perfect Collinearity ($MLR.3$)**

Any variable cannot be a linear function of the other variables

<br>

**Example (silly)**

```{=tex}
\begin{align*}
  wage = \beta_0 + \beta_1 educ + \beta_2 (3\times educ) + u
\end{align*}
```

(<span style = "color: blue;"> More on this later when we talk about dummy variables</span>)

### Endogeneity (Definition)

<details class="transcript">
<summary>Transcript</summary>

Here is endogeneity written formally, although we have been circling the idea since lecture zero. The callout says that the expected value of u conditional on x-one through x-k equals some function of those regressors and is not zero. In other words, once you know the included x values, you can systematically predict part of what the model placed in the error. That violates M-L-R four. OLS then treats some movement associated with the error as if it belonged to the included regressors, so its coefficient estimators need not be unbiased or causal.

The first listed cause is functional-form misspecification. Read the two wage equations line by line. In the true equation, x-one affects wage through log x-one, while x-two enters linearly. In your equation, you mistakenly enter x-one itself. To see where the mistake goes, equate the two right-hand sides. The shared intercept and beta-two x-two terms cancel. Your error u-two must then equal the true error u-one plus beta-one times the bracket, log x-one minus x-one. That extra bracket is a deterministic function of x-one. Therefore the conditional mean of u-two changes with x-one even if the original u-one had conditional mean zero. The wrong functional form creates endogeneity by construction.

The second cause is omitting a variable that is correlated with any included regressor.

The final blue bullet warns that omitted variables and functional form are not the only sources. So endogeneity is a family of violations summarized by the same mathematical condition, not the name of one particular mistake.

</details>

:::{.callout-important title="Endogeneity: Definition"}
$$
E[u|x_1,x_2,\dots,x_k] = f(x_1,x_2,\dots,x_k) \ne 0
$$
:::

<br>

**What could cause endogeneity problem?**

+ functional form misspecification
```{=tex}
\begin{align*}
  wage = & \beta_0 + \beta_1 log(x_1) + \beta_2 x_2 + u_1 \;\;\text{(true)}\\
  wage = & \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u_2 \;\; \text{(yours)},\;\; \text{where } u_2 = u_1 + \beta_1[log(x_1)-x_1]
\end{align*}
```
+ omission of variables that are correlated with any of $x_1,x_2,\dots,x_k$ (<span style = "color: blue;"> more on this soon </span>)
+ <span style = "color: blue;"> other sources of endogeneity later </span>

:::
<!--end of panel-->

## Variance of OLS estimators

::: {.panel-tabset}

### Homoskedasticity and Variance of $\widehat{\beta}_{OLS}$

<details class="transcript">
<summary>Transcript</summary>

Condition five is homoskedasticity. The displayed assumption says that the variance of u conditional on x-one through x-k equals sigma squared. Sigma squared is one constant number, so the spread of the unobserved determinants around their conditional mean must be the same at every joint combination of regressor values. The errors themselves can differ across observations, and homoskedasticity does not say they are all small or normally distributed. It says their conditional variance does not systematically expand or contract with the x variables.

Under M-L-R one through M-L-R five, and conditional on the observed sample values of the independent variables, the variance of beta-j-hat is sigma squared divided by S-S-T-j times one minus R-squared-j. The numerator, sigma squared, measures the noise remaining in y. More noise makes the coefficient estimate vary more from sample to sample.

Now unpack the denominator. S-S-T-j is the sum, over i from one to n, of x-j-i minus x-j-bar, squared. It measures the total sample variation in regressor x-j around its sample mean. More variation in x-j gives OLS more information for estimating its slope, so a larger S-S-T-j lowers the variance.

R-squared-j comes from an auxiliary regression in which x-j is the dependent variable and all the other independent variables, plus an intercept, are the regressors. It measures the fraction of x-j's sample variation that those other regressors explain. Therefore one minus R-squared-j is the unexplained fraction, and S-S-T-j times one minus R-squared-j is the variation in x-j left after partialling out the others. That is the independent variation OLS can use to identify beta-j.

Read the note at the bottom as a limiting argument. If R-squared-j is close to zero, the other regressors explain little of x-j, and most of its variation remains useful. As R-squared-j approaches one, one minus R-squared-j approaches zero, the denominator becomes small, and the variance of beta-j-hat becomes very large. At exactly one, x-j is perfectly collinear and the coefficient is not separately estimable. Near one, it is estimable but imprecise. That is multicollinearity, the subject of lecture four.

</details>

**Condition 5** 

Error term is homoskedastic (Assumption $MLR.5$)

```{=tex}
\begin{align*}
Var(u|x_1,\dots,x_k)=\sigma^2
\end{align*}
```

<br>

Under conditions $MLR.1$ through $MLR.5$, conditional on the sample values of the independent variables,

:::{.callout-important title="Variance of $\widehat{\beta}_{OLS}$"}
```{=tex}
\begin{align*}
    Var(\widehat{\beta}_j)= \frac{\sigma^2}{SST_j(1-R^2_j)},
\end{align*}
```
:::

where 

+ $SST_j= \sum_{i=1}^n (x_{ji}-\bar{x}_j)^2$ 
+ $R_j^2$ is the R-squared from regressing $x_j$ on all other independent variables including an intercept. (<span style = "color: blue;"> We will revisit this equation</span>)

:::{.callout-note title="Why $R_j^2$ is in there"}
$R_j^2$ measures how much of $x_j$ the <span style = "color: blue;"> other </span> regressors already explain. As it approaches $1$, the term $1-R_j^2$ approaches $0$ and the variance <span style = "color: red;"> explodes </span>. Intuitively: once the others account for nearly all of $x_j$'s movement, there is almost no independent variation left to identify $\beta_j$ from. That is multicollinearity, and it is the whole of lecture 04.
:::

### Estimating $\sigma^2$

<details class="transcript">
<summary>Transcript</summary>

The variance formula on the previous tab contains sigma squared, the conditional variance of the unobserved population error. We cannot observe u, so we cannot calculate sigma squared directly. As in univariate regression, we estimate it using the fitted residuals, u-i-hat. Each residual is observed y-i minus its OLS fitted value. Squaring them and adding across observations gives the residual sum of squares.

In the univariate model, there is one slope and one intercept. Estimating those two coefficients imposes two normal-equation restrictions on the residuals, so the residual degrees of freedom are n minus two. The displayed estimator divides the sum of squared residuals by n minus two, not by n.

In the multivariate model on the lower half, k counts the independent variables, and the intercept is an additional coefficient. We estimate k plus one parameters in total. The residual degrees of freedom are therefore n minus open-parenthesis k plus one close-parenthesis, which is n minus k minus one. The estimator is the sum of u-i-hat squared divided by that quantity. Under the stated classical assumptions, this degrees-of-freedom correction makes sigma-hat squared unbiased for sigma squared. Dividing by n would tend to understate the error variance because the fitted regression has already chosen coefficients that make these in-sample residuals as small as possible.

The sentence at the bottom connects the count to the first-order conditions. The residual vector has n entries, but the intercept and k regressor orthogonality equations impose k plus one independent linear restrictions when the regressor matrix has full column rank. Thus only n minus k minus one residual directions are free. Equivalently, with a suitable full-rank subset fixed, those restrictions determine the remaining k plus one components. Every parameter estimated costs one residual degree of freedom.

</details>

Just like uni-variate regression, you need to estimate $\sigma^2$ if you want to estimate the variance (and standard deviation) of the OLS estimators.

**uni-variate regression**

```{=tex}
\begin{align*}
  \widehat{\sigma}^2=\sum_{i=1}^n \frac{\widehat{u}_i^2}{n-2}
\end{align*}
```

**multi-variate regression**

A model with $k$ independent variables with intercept.

```{=tex}
\begin{align*}
  \widehat{\sigma}^2=\sum_{i=1}^n \frac{\widehat{u}_i^2}{n-(k+1)}
\end{align*}
```

You solved $k+1$ simultaneous equations to get $\widehat{\beta}_j$ $(j=0,\dots,k)$. So, once you know the value of $n-k-1$ of the residuals, you know the rest.

### Estimator of $Var(\widehat{\beta}_j)$

<details class="transcript">
<summary>Transcript</summary>

Now put the two previous tabs together. The population variance formula for beta-j-hat contains the unknown sigma squared. Replace it with the residual-based estimator sigma-hat squared. The result is an **estimator** of the variance of the OLS estimator, which is why the word estimator is red and why there is a hat over the word variance in the callout.

Read the first equality. Estimated variance of beta-j-hat equals sigma-hat squared divided by S-S-T-j times one minus R-squared-j. S-S-T-j and R-squared-j are calculated from the observed x values. Sigma-hat squared is calculated from the OLS residuals. Everything on the right is therefore observable or computable from the sample.

The second equality substitutes the formula for sigma-hat squared explicitly. Inside the first parentheses, sum the squared residuals and divide by n minus k minus one. Then multiply by one over S-S-T-j times one minus R-squared-j. This makes the sources of uncertainty visible. Larger residual noise raises estimated variance. More total variation in x-j lowers it. More overlap between x-j and the other regressors raises it because R-squared-j grows and the usable fraction, one minus R-squared-j, shrinks.

</details>

Using the estimator of $\sigma^2$ in place of $\sigma^2$, we have the <span style = "color: red;"> estimator </span> of the variance of the OLS estimator.

:::{.callout-important title="Estimator of the variance of the OLS estimator"}
```{=tex}
\begin{align*}
\widehat{Var(\widehat{\beta}_j)} = \frac{\widehat{\sigma}^2}{SST_j(1-R^2_j)} = \left(\sum_{i=1}^n \frac{\widehat{u}_i^2}{n-k-1}\right) \cdot \frac{1}{SST_j(1-R^2_j)}
\end{align*}
```
:::

:::
<!--end of panel-->

# Frisch–Waugh–Lovell Theorem (Optional)

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Frisch–Waugh–Lovell Theorem 

<details class="transcript">
<summary>Transcript</summary>

This section is optional, but I encourage you to follow it because the Frisch-Waugh-Lovell theorem explains what "holding other factors fixed" means mechanically. The model at the top has an intercept, three regressors, and an error: y-i equals beta-zero plus beta-one x-one-i plus beta-two x-two-i plus beta-three x-three-i plus u-i. Suppose beta-one is the only coefficient you care about. It measures the partial association between x-one and y while x-two and x-three are held fixed.

Method one is regular multiple regression. Regress y on x-one, x-two, and x-three with an intercept, and estimate beta-zero through beta-three simultaneously. The coefficient attached to x-one is beta-one-hat. This is the procedure you would normally run.

Method two reaches that one coefficient in three regressions. First, regress y on x-two and x-three with an intercept, and save the residual for every observation. Call those residuals u-y-hat. They are the part of y left after its linear projection on the intercept, x-two, and x-three. Second, regress x-one on x-two and x-three with an intercept and save u-x-one-hat. These residuals are the part of x-one that the same controls do not linearly explain. Third, regress u-y-hat on u-x-one-hat. The displayed equation writes its slope as alpha-one and its new error as v-three. Because both residual series have sample mean zero, an included intercept in this final regression is estimated as zero, so omitting it from the written equation does not change the slope.

The theorem says alpha-one-hat from that third regression is **exactly equal** to beta-one-hat from method one. This is an algebraic identity for the same sample and the same set of controls, not an approximation and not an asymptotic claim.

</details>

Consider the following simple model,

```{=tex}
\begin{align*}
  y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 x_{2,i} + \beta_3 x_{3,i} + u_i
\end{align*}
```

Suppose you are interested in estimating only $\beta_1$.

Let's consider the following two methods,

<br>

**Method 1: Regular OLS**

Regress $y$ on $x_1$, $x_2$, and $x_3$ with an intercept to estimate $\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$ at the same time (just like you normally do)

<br>

**Method 2: 3-step**

+ regress $y$ on $x_2$ and $x_3$ with an intercept and get residuals, which we call $\widehat{u}_y$
+ regress $x_1$ on $x_2$ and $x_3$ with an intercept and get residuals, which we call $\widehat{u}_{x_1}$
+ regress $\widehat{u}_y$ on $\widehat{u}_{x_1}$ $(\widehat{u}_y=\alpha_1 \widehat{u}_{x_1}+v_3)$

**Frisch-Waugh–Lovell theorem**

Methods 1 and 2 produce the same coefficient estimate on $x_1$

$$\widehat{\beta}_1 = \widehat{\alpha_1}$$

## Frisch–Waugh–Lovell: see it happen

<details class="transcript">
<summary>Transcript</summary>

Rather than take the theorem on faith, run all four steps on the same simulated data. In step one, the library line loads fix-est and makes fee-ols available. Set-seed with four eight nine one fixes the random draws, and capital N equals two hundred sets the number of observations. X-two and x-three are independent standard normal draws. X-one is then built as one-half times x-two minus three-tenths times x-three plus a fresh normal draw. This construction deliberately makes x-one share variation with both controls, so partialling them out has real work to do.

The y line sets the true data-generating process: an intercept of one, a slope of two on x-one, a slope of three on x-two, a slope of negative one on x-three, and a new normal error. Data-frame collects the four observed variables into d. We know the true x-one effect is two, but sampling noise means the estimate in this particular data set need not equal two exactly.

Step two implements regular OLS. The formula y tilde x-one plus x-two plus x-three regresses y on all three variables with the default intercept. The data equals d argument tells fee-ols where to find them. The coefficient function applied to m-one extracts the coefficient vector, and brackets containing the quoted name x-one select only the x-one estimate. Running this exact cell returns about two point zero four six six seven two.

Step three implements partialling out. The first inner fee-ols call regresses y on x-two and x-three, and the resid function extracts its residuals into a new column called u underscore y. The second does the same with x-one as the dependent variable, saving u underscore x-one. Then fee-ols regresses the y residuals on the x-one residuals. Its x-one-residual slope, selected by applying the coefficient function to m-two and taking the entry named u underscore x-one, is also about two point zero four six six seven two. The final regression includes an intercept by default; it is essentially zero because both residual vectors have mean zero.

Step four uses the c function to combine the method-one and method-two coefficients side by side under descriptive labels. Look at the printed values after running it. They agree to every displayed digit. With this exact seed, method one minus method two is negative four point four four zero eight nine two times ten to the minus sixteen. That tiny discrepancy is floating-point rounding at machine precision, not an econometric difference. These are not two unrelated procedures that happen to be close. They are the same OLS coefficient computed through two algebraically equivalent routes.

</details>

Do not take it on faith. Run both methods on the same simulated data and compare.

**Step 1: generate the data.** `x1` is built from `x2` and `x3`, so they really do share variation.

```{webr-r}
#| autorun: true

library(fixest)

set.seed(4891)
N <- 200
x2 <- rnorm(N)
x3 <- rnorm(N)
x1 <- 0.5 * x2 - 0.3 * x3 + rnorm(N)
y <- 1 + 2 * x1 + 3 * x2 - x3 + rnorm(N)
d <- data.frame(y, x1, x2, x3)
```

**Step 2: Method 1.** Regress $y$ on everything at once.

```{webr-r}
#| autorun: true
#| code-track: 0.6

m1 <- feols(y ~ x1 + x2 + x3, data = d)
coef(m1)["x1"]
```

**Step 3: Method 2.** Partial $x_2$ and $x_3$ out of both $y$ and $x_1$, then regress one set of residuals on the other.

```{webr-r}
#| autorun: true
#| code-track: 0.6

d$u_y <- resid(feols(y ~ x2 + x3, data = d))
d$u_x1 <- resid(feols(x1 ~ x2 + x3, data = d))
m2 <- feols(u_y ~ u_x1, data = d)
coef(m2)["u_x1"]
```

**Step 4: compare.**

```{webr-r}
#| autorun: true
#| code-track: 0.6

c(method_1 = coef(m1)["x1"], method_2 = coef(m2)["u_x1"])
```

Not merely close: identical to machine precision. The difference is about $-4\times 10^{-16}$, which is rounding, not disagreement.

## Partialing out Interpretation from Method 2

<details class="transcript">
<summary>Transcript</summary>

Now read the three steps for meaning rather than code mechanics. In step one, regress y on x-two and x-three with an intercept. For each observation, u-y-hat is observed y minus the value predicted from those controls. It is the part of y that is orthogonal in the sample to the intercept, x-two, and x-three. The slide calls it void of their impact. Interpret that as removing their **linear fitted component**. By itself, this projection is not a causal claim about the true effects of x-two and x-three.

In step two, regress x-one on the same controls with an intercept. U-x-one-hat is observed x-one minus the part predicted by x-two and x-three. It contains the remaining variation in x-one that has zero sample covariance with each of those controls. For example, if x-two predicts that an observation should have x-one equal to five but its actual x-one is seven, this residual contributes the unexplained difference of two. This is the variation in x-one available after holding the controls fixed in the linear regression sense.

Step three regresses the y leftovers on the x-one leftovers. Any variation in y or x-one that x-two and x-three linearly account for has been removed before this comparison. The resulting slope asks how residual y changes with residual x-one. The Frisch-Waugh-Lovell theorem proves that this slope is identical to the coefficient on x-one from regressing y directly on x-one, x-two, and x-three together.

This is the partialling-out interpretation of a multiple-regression coefficient. OLS does not estimate beta-one from every bit of x-one's variation. It estimates beta-one from the component of x-one that cannot be linearly predicted by the other included regressors.

</details>

**Step 1**

Regress $y$ on $x_2$ and $x_3$ with an intercept and get residuals, which we call $\widehat{u}_y$

+ $\widehat{u}_y$ is void of the impact of $x_2$ and $x_3$ on $y$

**Step 2**

Regress $x_1$ on $x_2$ and $x_3$ with an intercept and get residuals, which we call $\widehat{u}_{x_1}$

+ $\widehat{u}_{x_1}$ is void of the impact of $x_2$ and $x_3$ on $x_1$

**Step 3**

Regress $\widehat{u}_y$ on $\widehat{u}_{x_1}$, which produces an estimate of $\beta_1$ identical to the one you get from regressing $y$ on $x_1$, $x_2$, and $x_3$

## Interpretation

<details class="transcript">
<summary>Transcript</summary>

Here is the main lesson from the whole optional section. When you regress y on x-one, x-two, and x-three, the coefficient on any one variable can be understood as a univariate regression after the linear contributions of all the other included variables have been partialled out of both that variable and y. For beta-one-hat, remove from y what x-two and x-three predict, remove from x-one what x-two and x-three predict, and relate the two residual series. The resulting slope is exactly the multiple-regression coefficient.

That is what "holding the other included variables fixed" means in OLS mechanics. The estimate uses comparisons in x-one that are not explained by the included controls. If education is the variable of interest and experience is a control, OLS uses the part of education variation not linearly predicted by experience and compares it with the part of wage variation not linearly predicted by experience.

Now read the second bullet, which gives the practical payoff. Including a relevant variable can prevent its modeled relationship with y from being confused with the relationship you want to attribute to x-one. In the spending example, controlling for average income removes the linear income-related component from both school spending and test scores before estimating the spending coefficient. That is more informative than a simple raw association.

</details>

+ Regressing $y$ on all explanatory variables $(x_1$, $x_2$, and $x_3)$ in a multivariate regression is as if you are looking at the impact of a single explanatory variable with the effects of all the other variables partialled out

+ In other words, including variables beyond your variable of interest lets you <span style = "color: red;"> control for (remove the effect of) </span> other variables, avoiding confusing the impact of the variable of interest with the impact of other variables. 
