---
title: "05: Hypothesis Testing"
format: 
  revealjs: 
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    include-in-header: ../transcript-support.html
    footer: "[back to course website with lecture slides](https://tmieno2.github.io/Applied-Econometrics-MS-Quarto/lectures/)"
execute:
  echo: true
  out-width: 80%
webr:
  packages: ['fixest', 'dplyr', 'data.table', 'ggplot2', 'wooldridge', 'broom', 'car']
  cell-options:
    editor-font-scale: 0.8
filters:
  - webr
---

# Review on Statistical Hypothesis Testing

```{r}
#| include: false
library(dplyr)
library(wooldridge)
library(data.table)
library(ggplot2)
library(fixest)
```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Hypothesis Testing

::: {.panel-tabset}

### General steps

<details class="transcript">
<summary>Lecturer transcript</summary>

Five steps, and every test in this lecture is one of them wearing different clothes, so it is worth reading them now and coming back at the end. Notice how much work step two is doing. You need the distribution of your test statistic assuming the null is true, and if you cannot get that, you have no test. Steps four and five are where judgement enters: you choose how much error you will tolerate, and then check how unlikely your number was. Everything else is mechanical. Read the red phrases twice, because "if the null is true" is the assumption the whole logic hangs on.

</details>

Here is the general step of any hypothesis testing:

+ Step 1: specify the null $(H_0)$ and alternative $(H_1)$ hypotheses

+ Step 2: find the distribution of the test statistic <span style = "color: red;"> if the null hypothesis is true </span> 

+ Step 3: calculate the test statistic based on the data and regression results

+ Step 4: define the significance level

+ Step 5: check how unlikely that you get the actual test statistic (found at Step 3) <span style = "color: red;"> if indeed the null hypothesis is true </span>

### Example

::: {.panel-tabset}

#### Setting

<details class="transcript">
<summary>Lecturer transcript</summary>

A deliberately artificial example, and the artificiality is the point. We want to test whether the expected value of x is one. Read the state of knowledge: we somehow already know the variance is four. That never happens in practice, and we assume it here only so the first pass through the logic has nothing hidden in it. Your estimator is the sample mean, which you already know is unbiased from lecture 03. So we have an estimator and a target value, and the question is how to decide whether the two are far enough apart to matter.

</details>

**Goal**

Suppose you want to test if the expected value of a normally distributed random variable $(x)$ is 1 or not.

<br>

**State of Knowledge**

We do know $x$ follows a normal distribution and its variance is 4 for some reason.

<br>

**Your Estimator**

Your estimator is the sample mean: $\theta = \sum_{i=1}^J x_i/J$

#### distribution of $\theta$

<details class="transcript">
<summary>Lecturer transcript</summary>

Two math asides and then the payoff, so follow the chain. The first says scaling a random variable scales its variance by the square of the factor. The second says shifting a Normal moves its mean but leaves its variance alone. Put them together and any Normal can be turned into a standard Normal: subtract the mean, divide by the standard deviation. Now apply that to the sample mean. Its variance is four over J, so its standard deviation is two over root J. Dividing by that gives root J over two, times theta minus alpha. That is the last line.

</details>


:::{.callout-note title="Math Aside 1"}
$Var(ax) = a^2 Var(X)$
:::

So, we know that $\theta \sim N(\alpha, 4/J)$ (of course $\alpha$ is not known).

<br>

:::{.callout-note title="Math Aside 2"}
If $x \sim N(a, b)$, then, $x-a \sim N(0, b)$ (shift) 
:::

So, $\frac{x-a}{\sqrt{b}} \sim N(0, 1)$ (combined with Math Aside 1)

<br>

**This means,**

Since $\theta = \sum_{i=1}^J x_i/J$ and $x_i \sim N(\alpha, 4)$, 

+ $Var(\theta) = J \times \frac{1}{J^2}Var(x) = 4/J$
+ $\frac{\sqrt{J}}{2} \cdot (\theta - \alpha)\sim N(0, 1)$.


#### distribution of $\theta$ if $H_0$ is true

<details class="transcript">
<summary>Lecturer transcript</summary>

Here is where the null hypothesis actually enters. Up to now alpha was unknown, so the expression on the left was not something you could compute. But if the null says alpha equals one, substitute it in, and suddenly every symbol is a number you have. Read the sentence carefully and notice the red phrase: you subtract the value under the null first, then multiply by root J, then divide by two. Students routinely drop the subtraction. The picture on the right is what that quantity follows if the null is true. It is just the standard Normal.

</details>

::: {.columns}

::: {.column width="50%"}
We established that $\frac{\sqrt{J}}{2} \cdot (\theta - \alpha)\sim N(0, 1)$.

The null hypothesis is $\alpha = 1$. 

If $\alpha = 1$ is indeed true, then $\sqrt{J} \times (\theta - 1)/2 \sim N(0, 1)$.

In other words, if you take the sample mean, <span style = "color: red;">subtract the value under the null</span>, multiply by the square root of the number of observations, and divide by 2, the result follows the standard Normal distribution like below.
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r}
#| echo: false
x <- seq(-4, 4, length = 1000)
y <- dnorm(x, sd = 1)
plot_data <- data.table(x = x, y = y)
g_n <- 
  ggplot(data = plot_data) +
  geom_line(aes(x = x, y = y), color = "black") +
  theme_bw()
g_n
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

#### Case 1

<details class="transcript">
<summary>Lecturer transcript</summary>

Now put numbers in. A hundred observations, sample mean of two, null value of one. Root of a hundred is ten, times the gap of one, divided by two, gives five. Now look at the figure and find five on the horizontal axis. It is not there; the curve has effectively no mass out that far. So either the null is true and something absurdly unlikely happened, or the null is false. That is the entire logic of hypothesis testing, and you have just done one. What we have not yet done is say how unlikely counts as unlikely enough.

</details>

::: {.columns}

::: {.column width="50%"}
Suppose you have obtained 100 samples $(J = 100)$ and calculated $\theta$ (sample mean), which turned out to be 2.

Then, your test statistic is $\sqrt{100} \times (2-1)/2 = 5$.

How unlikely is it to get the number you got (5) if the null hypothesis is indeed true?

:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r}
#| echo: false
g_n
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


#### Case 2

<details class="transcript">
<summary>Lecturer transcript</summary>

The same calculation with different numbers, and a different verdict. Four hundred observations, sample mean of one point zero two, so the gap from the null is tiny. Root of four hundred is twenty, times zero point zero two, divided by two, gives zero point two. Find that on the figure: it is right in the fat middle of the distribution. Getting zero point two is entirely ordinary if the null is true, so you have no grounds to doubt it. Note that the sample was four times larger here and still gave no evidence, because the gap was small.

</details>

::: {.columns}

::: {.column width="50%"}
Suppose you have obtained 400 samples $(J = 400)$ and calculated $\theta$ (sample mean), which turned out to be 1.02.

Then, your test statistic is $\sqrt{400} \times (1.02-1)/2 = 0.2$.

How unlikely is it to get the number you got (0.2) if the null hypothesis is indeed true?
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r}
#| echo: false
g_n
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


:::
<!--end of panel-->

### By the way

<details class="transcript inline">
<summary>Lecturer transcript</summary>

A worthwhile aside about why we bother normalising at all. You do not have to. You could just use theta minus alpha as your statistic, and the figure on the right shows what its distribution looks like with a hundred observations: a Normal centred at zero but much narrower, standard deviation zero point two. Reconsider case one with that statistic and you get one, which on this narrow curve is five standard deviations out, the same conclusion as before. The reason we normalise is at the bottom: it lets you look up one table forever instead of a new distribution for every sample size.

</details>

::: {.columns}

::: {.column width="40%"}
Note that you do not really need to use $\sqrt{J} \times (\theta - \alpha)/2$ as your test statistic. 

You could alternatively use $\theta - \alpha$. But, in that case, you need to be looking at $N(0, 4/J)$ instead of $N(0, 1)$ to see how unlikely you get the number you got. 

For example, when the number of observations is 100 $(J = 100)$, the distribution of $\theta-\alpha$ looks like the figure on the right.
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r}
#| echo: false
x <- seq(-4, 4, length = 1000)
y <- dnorm(x, sd = 2 / 10)
plot_data <- data.table(x = x, y = y)
ggplot(data = plot_data) +
  geom_line(aes(x = x, y = y), color = "black") +
  theme_bw()
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


**Reconsider the case 1**

Suppose you have obtained 100 samples $(J = 100)$ and calculated $\theta$ (sample mean), which turned out to be 2.

Then, your test statistic is $2-1 = 1$. 

Is it unlikely for you to get 1 if the null hypothesis is true? 

The conclusion would be exactly the same as using $\sqrt{J} \times (\theta - \alpha)/2$ because the distribution under the null is adjusted according to the test statistic you use.

:::{.callout-note}
We always use a normalized test statistic so that we can always look up the same distribution.
:::

### Summary

<details class="transcript">
<summary>Lecturer transcript</summary>

Two questions and their answers, and this is the slide to come back to when the later tests start blurring together. What you need: a statistic whose distribution you know under the null. What you do: transform your raw quantity, usually a coefficient or a sample mean, so that its distribution is one you can look up. Then check whether the number you got is a plausible draw from it. Note the parenthesis at the end, which is honest about what we have not done: we have no formal criterion yet for how unlikely is unlikely enough. That is the significance level.

</details>

**What do we need?**

+ test-statistic of which we know the distribution (e.g., t-distribution, Normal distribution) assuming the null hypothesis

<br>

**What do we (often) do?**

+ transform (most of the time) a raw random variable (e.g., sample mean in the example above) into a test statistic of which we know the distribution assuming that the null hypothesis is true
  - e.g., we transformed the sample mean so that it follows the standard Normal distribution.
+ check if the actual number you got from the test statistic is likely to happen or not (formal criteria has not been discussed yet)

### Exercise

::: {.panel-tabset}

#### Problem

<details class="transcript">
<summary>Lecturer transcript</summary>

Your turn, and the structure mirrors the worked example exactly, so you should be able to do all four parts. Read the setup: the difference in salaries between two universities, distributed Normal with unknown mean alpha and variance sixteen. You want to test whether alpha is ten. Work through the four questions on paper before opening the answer. Part two is the one to be careful with, because the variance of the sample mean is not sixteen; it is sixteen over four hundred. Then read the note underneath, which flags the two things this artificial setup let us skip.

</details>

You have collected data on annual salary for those who graduated from University A and B. You are interested in testing whether the difference in annual salary between the universities (call it $x$) is 10 on average. You know (for unknown reasons) that the difference is distributed as $N(\alpha, 16)$. 

1. What is the null hypothesis?
2. Under the null hypothesis, what is the distribution of the sample mean when the number of observation is 400?
3. Normalize the test statistic so that the transformed version follows $N(0, 1)$.
4. The actual difference you observed is 10.2. What is the probability that you observe a number greater than 10.2 if the null hypothesis is true? Use `pnorm()`.


:::{.callout-note}
In reality,

+ we need to find out what the distribution of the test statistic is
+ we need to formally define when we accept or not accept the null hypothesis
:::

#### Answer

<details class="transcript">
<summary>Lecturer transcript</summary>

Check yours against this. The null is alpha equals ten, and under it the sample mean has variance sixteen over four hundred, so standard deviation zero point two. Dividing by that means multiplying by root of four hundred over sixteen, which is five. Your observed difference is ten point two, so the statistic is five times zero point two, which is one. Run the cell: one minus pnorm of one is about sixteen percent. So a gap this size happens roughly one time in six when the null is true, which is not remotely unusual.

</details>

1. $\alpha = 10$
2. $\theta \sim N(\alpha, 16/400)$
3. $\sqrt{\frac{400}{16}}\cdot (\theta - \alpha)$
4. The test statistic is $5 \times (10.2 - 10) = 1$

```{webr-r}
1 - pnorm(1)
```

:::
<!--end of panel-->

:::
<!--end of panel-->

# Hypothesis Testing on the Coefficients

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Hypothesis Testing of the Coefficients

::: {.panel-tabset}

### Example

<details class="transcript">
<summary>Lecturer transcript</summary>

Now the same logic pointed at regression coefficients, which is what you actually care about. Two example hypotheses in the box: education has no effect, experience has a positive effect. Note they have different shapes, one an equality and one an inequality, and that distinction becomes one-sided versus two-sided testing shortly. Then read the three bullets, because the second is the crux. If beta-one-hat were a fixed number you would just look at it. But it is one realisation of a random variable, so even when the true beta is exactly zero you can draw an estimate far from zero.

</details>

Consider the following model,

$$wage = \beta_0 + \beta_1 educ + \beta_2 exper + u$$

:::{.callout-note title="Example Hypotheses"}
+ education has no impact on wage $(\beta_1=0)$
+ experience has a positive impact on wage $(\beta_2>0)$
:::

<br>

+ If $\widehat{\beta}_1$ is non-random, but just a scalar, all you have to do is just check if $\widehat{\beta}_1=0$ or not

+ But, the estimate you get is <span style = "color: blue;"> just one realization </span> of the range of values $\widehat{\beta}_1$ could take because it is a random variable

+ This means that even if $\beta_1=0$ in the population, it is possible to get an estimate that is very far from 0

### Distribution of $\beta_j$

::: {.panel-tabset}

#### Assumption

<details class="transcript">
<summary>Lecturer transcript</summary>

A stocktake before we can go further. We have unbiasedness from MLR one through four, and a variance formula from MLR five. But read the line in blue: we have never said anything about the *shape* of the error distribution. Unbiasedness and variance are statements about the first two moments only. To do a test you need the whole distribution of your statistic, not two summary numbers, so we need more. The parenthesis at the end is a promise worth remembering: this assumption turns out to be avoidable in large samples, which is the asymptotics lecture.

</details>

So far, we learned that:

+ OLS estimators are unbiased under MLR.1 ~ MLR.4
+ Variance of the OLS estimator of $\beta_j$ is $\frac{\sigma^2}{SST_x\cdot (1-R^2_j)}$ under MLR.1 ~ MLR.5

We have <span style = "color: blue;"> NOT </span> made any assumptions about the distribution of the error term!!

In order to perform hypothesis testing, we need to make assumptions about the distribution of error term (this is not strictly true, but more on this later)

#### Additional assumption

<details class="transcript">
<summary>Lecturer transcript</summary>

This is MLR six, and read it carefully because it is much stronger than it looks. The headline is that u is Normally distributed. But look at the other red word: independent. Independence of the error from the regressors is far more than zero correlation. Read the second callout, which spells out the consequences. Independence gives you a zero conditional mean, so MLR four comes free. It also gives you constant conditional variance, so MLR five comes free too. We are assuming a great deal here, which is why the asymptotics lecture is such a relief.

</details>

For the purpose of hypothesis testing, we will make the following assumption:

:::{.callout-note title="Normality of the error term"}
The population error $u$ is <span style = "color: red;"> independent </span> of the explanatory variables $x_1,\dots,x_k$ and is <span style = "color: red;"> normally </span> distributed with zero mean and variance $\sigma^2$:

$$u\sim N(0,\sigma^2)$$
:::


:::{.callout-note}
The normality assumption is much more than error term being distributed as Normal.

Independence of the error term implies

+ $E[u|x] = 0$
+ $Var[u|x]= \sigma^2$

So, we are necessarily assuming MLR.4 and MLR.5 hold by the independence assumption.
:::

#### Implications of the Assumptions

<details class="transcript">
<summary>Lecturer transcript</summary>

Two consequences, one per column. On the left, y conditional on x is Normal, with the regression line as its mean. That follows immediately: y is a constant plus a Normal error. On the right is the one we need. Because beta-j-hat is a weighted sum of the y values, and sums of Normals are Normal, the estimator is itself Normally distributed, centred on the true beta. Subtract the mean and divide by the standard deviation, and you have a standard Normal. Now read the question at the bottom, because the answer is no, and the next tab says why.

</details>

::: {.columns}

::: {.column width="50%"}
**distribution of the dependent variable**

The distribution of $y$ conditional on $x$ is a Normal distribution

$y|x \sim N(\beta_0+\beta_1 x_1+\dots+\beta_k x_k,\sigma^2)$

+ $E[y|x]$ is $\beta_0+\beta_1 x_1+\dots+\beta_k x_k$
+ $u|x$ is $N(0,\sigma^2)$
:::
<!--end of the 1st column-->
::: {.column width="50%"}
**distribution of the OLS estimator**

If the MLR.1 through MLR.6 are satisfied, $OLS$ estimators are also Normally distributed!

$\widehat{\beta}_j \sim N(\beta_j,Var(\widehat{\beta}_j))$

which means,

$\frac{\widehat{\beta}_j-\beta_j}{se(\widehat{\beta}_j)} \sim N(0,1)$
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


:::{.callout-note title="Question"}
Okay, so are we going to use this for testing involving $\beta_j$?
:::


:::
<!--end of panel-->

### t-test

<details class="transcript">
<summary>Lecturer transcript</summary>

And here is the catch the question was pointing at. The expression on the previous tab used the true standard error, which contains sigma squared, which you do not know. Replace it with the estimated standard error and the distribution is no longer Normal. It becomes t with n minus k minus one degrees of freedom. The extra spread in the t distribution is the cost of estimating the denominator: you have two random things in the ratio now instead of one. Note the corrected formula in the parenthesis, and the square root in particular, because the standard error is the root of the variance.

</details>

In practice, we need to estimate $se(\beta_j)$. If we use $\widehat{se(\widehat{\beta}_j)}$ instead of $se(\widehat{\beta}_j)$, then,

$\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}} \sim t_{n-k-1}$

where $n-k-1$ is the degree of freedom of residual.

(Note: $\widehat{se(\widehat{\beta}_j)} = \sqrt{\widehat{\sigma}^2/\big[SST_X\cdot (1-R_j^2)\big]}$ &mdash; the standard error is the <span style = "color: red;"> square root </span> of the estimated variance)

### Recap on notations

<details class="transcript">
<summary>Lecturer transcript</summary>

This is a reference tab rather than something to work through, and I would bookmark it. The distinction that matters most is the hat. Sigma squared is the true error variance; sigma-hat squared is what you compute from the residuals. Var of beta-j-hat is a true quantity; the version with a hat over the whole thing is your estimate of it. And se is the square root of a variance, never the variance itself. Almost every algebra mistake in this lecture comes from losing track of which of these you are holding.

</details>

+ $k$: the number of explanatory variables included except the intercept
+ $\sigma^2$: the <span style = "color: blue;"> true </span> variance of the error term
+ $\widehat{\sigma^2}$: the estimator (estimate) of the variance of the error term
  - $\widehat{\sigma^2} = \frac{\sum \widehat{u}_i^2}{n-k-1}$
+ $SST_X = \sum (x_i - \bar{x})^2$ 
+ $\widehat{\beta}_j$: OLS estimator (estimate) on explanatory variable $x_j$
  - $\widehat{\beta} = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{SST_X}$ (for simple univariate regression)
+ $var(\widehat{\beta}_j)$: the <span style = "color: blue;"> true </span> variance of $\widehat{\beta}_j$
+ $R^2_j$: $R^2$ when you regress $x_j$ on all the other covariates (mathematical expression omitted)
+ $\widehat{var(\widehat{\beta}_j)}$: Estimator (estimate) of $var(\widehat{\beta}_j)$
  - $\widehat{var(\widehat{\beta}_j)} = \frac{\widehat{\sigma^2}}{SST_X\cdot(1-R^2_j)}$
+ $se(\widehat{\beta}_j)$: square root of $var(\widehat{\beta}_j)$
+ $\widehat{se(\widehat{\beta}_j)}$: square root of $\widehat{var(\widehat{\beta}_j)}$

:::
<!--end of panel-->  

## Details of hypothesis testing on $\beta_j$

::: {.panel-tabset}

### Null and alternative hypothesis

::: {.panel-tabset}

#### Introduction

<details class="transcript">
<summary>Lecturer transcript</summary>

The courtroom analogy, and it repays taking seriously. You are the attorney. The null hypothesis is what you want to knock down: the defendant is not guilty. The alternative is what you are arguing for. Notice the asymmetry this builds in. You do not prove the null true; you either reject it or fail to reject it, exactly as a court either convicts or fails to convict. And the burden of proof sits on the alternative. That is why we will say "fail to reject" rather than "accept", and why an insignificant result is not evidence of no effect.

</details>

Statistical hypothesis testing involves two hypotheses: Null and Alternative hypotheses. 

Pretend that you are an attorney who indicted a defendant who you think committed a crime. 

<br>  

**Null Hypothesis**

Hypothesis that you would like to reject (defendant is not guilty)

<br>

**Alternative Hypothesis**

Hypothesis you are in support of (defendant is guilty) 

#### One-sided and Two-sided Alternatives Hypotheses

<details class="transcript">
<summary>Lecturer transcript</summary>

Two shapes of alternative, and which you use changes where you look. With a one-sided alternative you have a directional claim, so you only care about one tail of the distribution. With a two-sided alternative you care about departures in either direction, so you look at both tails. The practical consequence is that for a given significance level the two-sided critical value is further out, because you have to split your error budget across two tails. Then read the callout: these critical values are just quantiles, the same computation as `qnorm` in lecture 03.

</details>

**one-sided alternative**

$H_0:$ $\beta_j = 0$ 
$H_1:$ $\beta_j > 0$ 

You look at the positive end of the t-distribution to see if the t-statistic you obtained is more extreme than the level of error you accept (significance level). 

<br>

**two-sided alternative**

$H_0:$ $\beta_j = 0$ 
$H_1:$ $\beta_j \ne 0$ 

You look at both ends of the t-distribution to see if the t-statistic you obtained is more extreme than the level of error you accept (significance level).

<br>

:::{.callout-note title="This is where lecture 03 pays off"}
The numbers we are about to use as critical values are just <span style = "color: blue;">quantiles</span>, computed exactly the way we computed `qnorm(0.95) = 1.64` in the Monte Carlo lecture. The only change is which distribution you ask: `qt()` instead of `qnorm()`, because we had to estimate $\sigma$ rather than know it.
:::

:::
<!--end of panel-->

### Significance level

```{r include = F}
u <- seq(-3.5, 3.5, length = 1000)
df_fig <- 522 # n - k - 1 for the wage regression used throughout this lecture
pdf <- dt(u, df = df_fig) # default is mean=0,sd=1
data <- data.table(x = u, y = pdf)

g_t <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.05
c_rej <- qt(1 - alpha, df = df_fig)
rej_data <- data[u >= c_rej, ]

g_t_005_one <- g_t +
  geom_ribbon(
    data = rej_data,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.02,
    label = as.character(round(c_rej, digits = 2)),
    size = 4,
    family = "Times"
  ) +
  annotate("text",
    x = 2.7, y = 0.1,
    label = paste("area  =  ", round(alpha, digits = 2), sep = ""),
    size = 4,
    family = "Times"
  )


alpha <- 0.01
c_rej <- qt(1 - alpha, df = df_fig)
rej_data_001 <- data[u >= c_rej, ]

g_t_001_one <- g_t +
  geom_ribbon(
    data = rej_data_001,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.02,
    label = as.character(round(c_rej, digits = 2)),
    size = 4,
    family = "Times"
  ) +
  annotate("text",
    x = 2.7, y = 0.1,
    label = paste("area  =  ", round(alpha, digits = 2), sep = ""),
    size = 4,
    family = "Times"
  )
```


::: {.panel-tabset}

#### Definition

<details class="transcript">
<summary>Lecturer transcript</summary>

The significance level, and then the thing this slide previously left out. The definition is the probability of rejecting a true null: how often you convict an innocent defendant. Now read the table. There are two ways to be wrong, and lowering your significance level trades one for the other. Push from five percent to one percent and you convict fewer innocents, but you also acquit more of the guilty, which is Type two error. One minus that is the power of your test. The last bullet is the practical payoff: not significant does not mean no effect.

</details>

:::{.callout-note title="Definition"}
The probability of rejecting the null when the null is actually true (the probability that you wrongly claim that the null hypothesis is wrong even though it is true in reality: <span style = "color: red;">Type I error</span>)
:::

The lower the significance level, the more sure you are that the null is indeed wrong when you reject it.

<br>

**But lowering it is not free**

There are two ways to be wrong, and they trade off against each other:

|  | $H_0$ is true | $H_0$ is false |
|---|---|---|
| **You reject $H_0$** | <span style = "color: red;"> Type I error </span> (probability = significance level) | correct |
| **You do not reject $H_0$** | correct | <span style = "color: blue;"> Type II error </span> |

+ the significance level is a choice about how much <span style = "color: red;"> Type I </span> error you will tolerate
+ pushing it down from 5% to 1% moves the critical value further out, so you reject less often &mdash; which means more <span style = "color: blue;"> Type II </span> error: real effects you fail to detect
+ $1 - P(\mbox{Type II error})$ is called the <span style = "color: blue;"> power </span> of the test: the chance of detecting an effect that is genuinely there
+ this is why "not statistically significant" does <span style = "color: red;"> not </span> mean "no effect". It may just mean your test had too little power to see it

#### One-sided: 5% significance

<details class="transcript">
<summary>Lecturer transcript</summary>

Here it is as a picture. The curve is the distribution of the t-statistic assuming the null is true. The red region is the five percent of the area furthest out in the positive tail, and the number at its edge is the critical value. If the null is true, you land in the red only five percent of the time, so a statistic in the red is either bad luck at exactly that rate or evidence against the null. That is the whole decision rule. Read the two definitions underneath and keep them apart: five percent is the level, the number is the critical value.

</details>

::: {.columns}

::: {.column width="40%"}
+ Figure on the right presents the distribution of $\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}}$ if $\beta_j = 0$ (the null hypothesis is true).

+ The probability that you get a value larger than `r round(qt(0.95, df_fig), 2)` is 5% (0.05 in area).
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r echo = F}
g_t_005_one + theme_bw()
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

:::{.callout-note title="Decision Rule"}
Reject the null hypothesis if the t-statistic is greater than `r round(qt(0.95, df_fig), 2)` (95% quantile of the t-distribution)
:::

If you follow the decision rule, then you have a 5% chance that you are wrong in rejecting the null hypothesis of $\beta_j = 0$. Here, 

+ 5% is the <span style = "color: red;"> significance level</span>
+ `r round(qt(0.95, df_fig), 2)` is the <span style = "color: red;"> critical value</span> above which you will reject the null

#### One-sided: 1% significance

<details class="transcript">
<summary>Lecturer transcript</summary>

The same curve, a smaller red region. Demanding one percent rather than five pushes the critical value further out, so you need a more extreme statistic to reject. Compare the two numbers between this tab and the last. The trade-off from the definition tab is now visible: you are less likely to reject a true null, and also less likely to reject a false one. Neither level is correct in any absolute sense. Five percent is a convention, and a result that rejects at five but not at one is exactly as informative as its p-value says.

</details>

::: {.columns}

::: {.column width="40%"}
+ The figure on the right presents the distribution of $\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}}$ if $\beta_j = 0$ (the null hypothesis is true).
+ The probability that you get a value larger than `r round(qt(0.99, df_fig), 2)` is 1% (0.01 in area).
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r echo = F}
g_t_001_one + theme_bw()
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

:::{.callout-note title="Decision Rule"}
Reject the null hypothesis if the t-statistic is greater than `r round(qt(0.99, df_fig), 2)` (99% quantile of the t-distribution)
:::

If you follow the decision rule, then you have a 1% chance that you are wrong in rejecting the null hypothesis of $\beta_j = 0$. Here, 

+ 1% is the <span style = "color: red;"> significance level </span>
+ `r round(qt(0.99, df_fig), 2)` is the <span style = "color: red;"> critical value </span> above which you will reject the null

#### Two-sided: 5% significance

<details class="transcript">
<summary>Lecturer transcript</summary>

Now both tails, because the alternative allows departures in either direction. The total red area is still five percent, but it is split, so each tail gets two and a half percent, which pushes each critical value further out than the one-sided case. Read the note in the parenthesis and check it against the previous two tabs: the curve has not changed at all across these three figures. It is the same t distribution every time. All that changes is which region you shade, because that is what your alternative hypothesis decides.

</details>

```{r t-005-two, include = F}
u <- seq(-3, 3, length = 1000)
pdf <- dt(u, df = df_fig) # default is mean = 0,sd = 1
data <- data.table(x = u, y = pdf)

alpha <- 0.05

c_rej <- qt(1 - alpha / 2, df = df_fig)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_t_005_two <- g_t +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.02,
    label = as.character(round(c_rej, digits = 2)),
    size = 4,
    family = "Times"
  ) +
  annotate("text",
    x = 0, y = 0.1,
    label = paste("area  = ", round(alpha / 2, digits = 3), sep = ""),
    size = 4,
    family = "Times"
  )
```

::: {.columns}

::: {.column width="40%"}
+ The figure on the right presents the distribution of $\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}}$ if $\beta_j = 0$ (the null hypothesis is true).

+ The probability that you get a value more extreme than `r round(qt(0.975, df_fig), 2)` or `r -round(qt(0.975, df_fig), 2)` is 5% (0.05 in area combining the two areas at the edges).

(Note: irrespective of the type of test, the distribution of the t-statistic is the same &mdash; every figure on this slide is the <span style = "color: red;"> same </span> $t_{`r df_fig`}$ curve. Only the shaded region changes.)
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r echo = F}
g_t_005_two + theme_bw()
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->  

:::{.callout-note title="Decision Rule"}
Reject the null hypothesis if the absolute value of the t-statistic is greater than `r round(qt(0.975, df_fig), 2)`.
:::

If you follow the decision rule, then you have a 5% chance that you are wrong in rejecting the null hypothesis of $\beta_j = 0$. Here,

+ 5% is the <span style = "color: red;"> significance level </span>
+ `r round(qt(0.975, df_fig), 2)` is the <span style = "color: red;"> critical value </span> beyond which you will reject the null

:::
<!--end of panel-->

### p-value

<details class="transcript inline">
<summary>Lecturer transcript</summary>

The p-value inverts the question. Instead of fixing a level and asking whether you reject, it asks: at what level would this result have been on the boundary? The figure shows a statistic of two point one six, and the shaded area either side of it is three point one percent. So if you had set your level at three point one percent, this result would sit exactly at the critical value. Any level above that and you reject; any level below and you do not. Read the decision rule at the bottom, and note it always agrees with the critical value approach.

</details>

:::{.callout-note title="Definition"}
The smallest significance level at which the null hypothesis would be rejected (the probability of observing a test statistic at least as extreme as we did if the null hypothesis is true)
:::


```{r p-dist, include = F}
u <- seq(-3.5, 3.5, length = 1000)
pdf <- dt(u, df = df_fig) # default is mean = 0,sd = 1
data <- data.table(x = u, y = pdf)

g_t <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.031
c_rej <- qt(1 - alpha / 2, df = df_fig)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_t_0031 <- g_t +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.02,
    label = as.character(round(c_rej, digits = 2)),
    size = 4,
    family = "Times"
  ) +
  annotate("text",
    x = 0, y = 0.1,
    label = paste("area  =  ", round(alpha, digits = 3), sep = ""),
    size = 4,
    family = "Times"
  )
```

::: {.columns}

::: {.column width="50%"}
+ Suppose the t-statistic you got is 2.16. Then, there's a 3.1% chance you reject the null when it is actually true, if you use it as the critical value.

+ So, the lower significance level the null hypothesis is rejected is 3.1%, which is the definition of p-value.
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r echo = F}
g_t_0031 + theme_bw()
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->  
 
:::{.callout-note title="Decision rule"}
If the p-value is lower than your choice of significance level, then you reject the null.
:::

This decision rule of course results in the same test results as the one we saw that uses a t-value and critical value.

### Example

<details class="transcript">
<summary>Lecturer transcript</summary>

A real worked example, and I want you to follow the arithmetic. The estimated effect of experience is zero point zero zero four one with a standard error of zero point zero zero one seven, so the t-statistic is their ratio, two point four one. The alternative is one-sided, so we want the ninety-nine percent quantile of t with five hundred and twenty-two degrees of freedom, which is two point three three. Two point four one exceeds it, so we reject at the one percent level. Notice the coefficient is tiny in absolute terms and still highly significant, because the standard error is smaller still.

</details>

**Estimated Model**

The impact of experience on wage:

+ $log(wage) = 0.284+0.092\times educ+0.0041\times exper + 0.022 \times tenure$
+ $\widehat{se(\widehat{\beta}_{exper})} = 0.0017$ 
+ $n = 526$

<br>

**Hypothesis**

+ $H_0$: $\beta_{exper}=0$
+ $H_1$: $\beta_{exper}>0$

<br>

**Test**

t-statistic $= 0.0041/0.0017 = 2.41$

The critical value is the 99% quantile of $t_{526-3-1}$, which is $2.33$ (it can be obtained by `qt(0.99, 522)`)

Since $2.41 > 2.33$, we reject the null in favor of the alternative hypothesis at the 1% level.

:::
<!--end of panel-->


## R implementation

::: {.panel-tabset}

### Run regression

<details class="transcript">
<summary>Lecturer transcript</summary>

Now the same thing in code, on real data. Two lines: load the wage data from the Wooldridge package, and run a regression of wage on education, experience and tenure. This cell runs automatically when the page loads, because the next several slides all use the objects it creates. Nothing prints, which is correct. If you want to see the model, type the object name on its own line. From here on we are extracting pieces of this fit and doing the tests we just did by hand.

</details>

```{webr-r}
#| autorun: true
#--- get the data ---#
data("wage1", package = "wooldridge")

#--- run a regression ---#
reg_wage <- feols(wage ~ educ + exper + tenure, data = wage1)
```

### Obtain t-statistics

<details class="transcript">
<summary>Lecturer transcript</summary>

Broom's tidy function turns a regression object into an ordinary data frame, which is far easier to work with than digging through the model object. Run it and read the four columns listed underneath. Estimate is beta-hat, std dot error is the estimated standard error, statistic is the t-statistic for the null that the coefficient is zero, and p dot value is the two-sided p-value for that same null. So the most common test you will ever run is already done for you. Read the last two lines: any other null hypothesis takes work, which is the next tab.

</details>

You can apply the `broom::tidy()` function from the `broom` package to access the regression results (`reg_wage` here) as a `tibble` (`data.frame`).

```{webr-r}
#| autorun: true
#* tidy up the regression results
reg_results_table <- broom::tidy(reg_wage)

#* display
reg_results_table
```

<br>

+ `estimate`: coefficient estimates $(\widehat{\beta}_j)$
+ `std.error`: $\widehat{se(\widehat{\beta}_j)}$
+ `statistic`: t-statistic for the null of $\beta_j = 0$
+ `p.value`: p-value (for the two sided test with the null of $\beta_j = 0$)

So, for the t-test of $\beta_j = 0$ is already there. You do not need to do anything further.

For the null hypothesis other than $\beta_j = 0$, you need further work.

### R implementation (by hand)

Suppose you are testing the null hypothesis of $\beta_{educ} = 1$ against the alternative hypothesis of $\beta_{educ} \ne 1$ (so, this is a two-sided test). 

The t-value for this test is not available from the summary. 

::: {.panel-tabset}

```{r}
#| include: false
data("wage1", package = "wooldridge")

#--- run a regression ---#
reg_wage <- feols(wage ~ educ + exper + tenure, data = wage1) 
 
reg_results_table <- broom::tidy(reg_wage)

#--- coefficient estimate on educ ---#
beta_educ <-
  reg_results_table %>%
  dplyr::filter(term == "educ") %>%
  dplyr::pull(estimate)

#--- se of the coefficient on educ ---#
se_beta_educ <-
  reg_results_table %>%
  dplyr::filter(term == "educ") %>%
  dplyr::pull(std.error)

#--- get the t-value ---#
t_value <- (beta_educ - 1) / se_beta_educ
df_t <- degrees_freedom(reg_wage, "resid")
critical_value <- qt(0.975, df = df_t)
```

#### get t-value

<details class="transcript">
<summary>Lecturer transcript</summary>

Here is that work. We are testing whether the coefficient on education equals one, which is not the null R reports, so we build the statistic ourselves. Read the structure: pull out the estimate, pull out the standard error, then subtract the null value from the estimate and divide. Notice the minus one on the last line. That is the only difference from what R already gave you, and it is the whole point: the reported t-statistic assumes a null of zero. Change the null and you change the numerator.

</details>

```{webr-r}
#--- coefficient estimate on educ ---#
beta_educ <-
  reg_results_table %>%
  filter(term == "educ") %>%
  pull(estimate)

#--- se of the coefficient on educ ---#
se_beta_educ <-
  reg_results_table %>%
  filter(term == "educ") %>%
  pull(std.error)

#--- get the t-value ---#
(
  t_value <- (beta_educ - 1) / se_beta_educ
)
```



#### critical value and t-test

<details class="transcript">
<summary>Lecturer transcript</summary>

And the other half: the critical value. First get the degrees of freedom from the fit rather than counting by hand, which avoids exactly the kind of error that comes from typing a number. Then `qt` at zero point nine seven five, because this is two-sided and we want two and a half percent in each tail. Compare the absolute value of your t-statistic against it. The gap here is enormous, which should not surprise you: a return to education of one dollar per year of schooling is nothing like the estimate the data gives.

</details>

The degree of freedom $(n-k-1)$ of the t-distribution can be obtained by applying `degrees_freedom(reg_wage, "resid")`:

```{webr-r}
(
  df_t <- degrees_freedom(reg_wage, "resid")
)
```

<br>

You can get the 97.5% quantile of the $t_{522}$ using the `qt()` function:

```{webr-r}
(
  critical_value <- qt(0.975, df = df_t)
)
```

<br>

Since the absolute value of the t-value (`r t_value`) is greater than the critical value (`r critical_value`), you reject the null.

:::
<!--end of panel-->

:::
<!--end of panel-->

# Confidence Interval (CI)

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Confidence Interval (CI)

::: {.panel-tabset}

### Definition

<details class="transcript">
<summary>Lecturer transcript</summary>

Read both halves of this slide, because the second is the one people get wrong. A ninety-five percent confidence interval means that if you repeated the whole exercise on many samples, ninety-five percent of the intervals you computed would contain the true parameter. The property belongs to the procedure, not to any one interval. What it is not, on the lower half, is a probability statement about the interval you actually have. Once you have computed it, the true value is either in it or not; there is no randomness left to have a probability about.

</details>

:::{.callout-note title="Definition"}
If you calculate the 95% confidence interval on multiple different samples, 95% of the time, the calculated CI includes the true parameter
:::

<br>

**What confidence interval is not**

The probability that a realized CI calculated from specific sample data includes the true parameter

### How to get CI (in general)

::: {.panel-tabset}

#### General Procedure

<details class="transcript">
<summary>Lecturer transcript</summary>

The general recipe, and it is short. To get an A percent interval, chop off the same amount of probability from each tail: A percent in the middle means one hundred minus A percent left over, split in two. So the lower bound is the hundred minus A over two quantile and the upper bound is its mirror image. Note the parentheses carefully. It is one hundred minus A, then divided by two, not one hundred minus A over two. That distinction is the difference between two point five percent and fifty-two point five percent.

</details>

For the **assumed** distribution of statistic $x$, the $A\%$ confidence interval of $x$ is the range with

+ lower bound: $(100 - A)/2$ percent quantile of $x$
+ upper bound: $100-(100 - A)/2$ percent quantile of $x$

#### Example

<details class="transcript">
<summary>Lecturer transcript</summary>

The recipe with numbers. For a ninety-five percent interval, five percent is left over, split into two and a half at each end, so you want the two point five and ninety-seven point five percent quantiles. If your statistic is standard Normal those quantiles are minus one point nine six and one point nine six, and the figure shows exactly that: the red tails together hold five percent of the area. Recognise the number. It is the same one point nine six that was the two-sided critical value two slides ago, which is not a coincidence.

</details>

::: {.columns}

::: {.column width="40%"}
For the 95% CI (A = 95), 

+ lower bound: $2.5$ $\big((100-95)/2\big)$ percent quantile of $x$
+ upper bound: $97.5$ $\big(100-(100-95)/2\big)$ percent quantile of $x$

If $x$ follows the standard normal distribution $(x \sim N(0, 1))$, then,the 2.5% and 97.5% quantiles are -1.96 and 1.96, respectively. So, the 95% CI of $x$ is [-1.96, 1.96].

:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r norm-95, include = F}
u <- seq(-3, 3, length = 1000)
pdf <- dnorm(u) # default is mean=0,sd=1
data <- data.table(x = u, y = pdf)

g_norm <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.05
c_rej <- qnorm(1 - alpha / 2)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_norm_005 <- g_norm +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.02,
    label = as.character(round(c_rej, digits = 2)),
    size = 4,
    family = "Times"
  ) +
  annotate("text",
    x = 2.5, y = 0.15,
    label = paste("area = ", round(alpha / 2, digits = 3), sep = ""),
    size = 4,
    family = "Times"
  )
```
```{r echo = F}
g_norm_005 + theme_bw()
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->



:::
<!--end of panel-->


### How to get the CI of coefficients

<details class="transcript">
<summary>Lecturer transcript</summary>

The derivation, and it has two moves. We know the distribution of the standardised quantity, so we know its interval directly: the quantiles of the t distribution. But you want an interval for beta-j, not for that ratio, so you have to invert. Solve the ratio for beta-j and you get beta-hat minus t times the standard error. Then read the last two bullets: scale each quantile by the standard error and shift by the estimate. The lower quantile is negative and the upper positive, which is what puts the estimate in the middle.

</details>

Under the assumption of MLR.1 through MLR.6 (which includes the normality assumption of the error), we learned that

$\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}} \sim t_{n-k-1}$

So, following the general procedure we discussed in the previous slide, the A% confidence interval of $\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}}$ is 

+ lower bound: $(100 − A)/2$% quantile of the $t_{n-k-1}$ distribution (let's call this $Q_l$)
+ upper bound: $100 - (100 − A)/2$% quantile of the $t_{n-k-1}$ distribution (let's call this $Q_h$)  

But, we want the A% CI of $\beta_j$, not $\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}}$. Solving for $\beta_j$,

$\beta_j = \widehat{\beta}_j - t_{n-k-1}\times \widehat{se(\widehat{\beta}_j)}$

So, to get the A% CI of $\beta_j$, we scale the CI of $\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}}$ by $\widehat{se(\widehat{\beta}_j)}$ and then shift by $\widehat{\beta}_j$. Because the $t$ distribution is symmetric, the minus sign above just swaps which quantile lands on which end:

+ lower bound: $Q_l \times \widehat{se(\widehat{\beta}_j)} + \widehat{\beta}_j$
+ upper bound: $Q_h \times \widehat{se(\widehat{\beta}_j)} + \widehat{\beta}_j$

Note that $Q_l$ is <span style = "color: blue;"> negative </span> and $Q_h$ is <span style = "color: red;">positive</span>.

### Example (R implementation)

::: {.panel-tabset}
#### Regression

<details class="transcript">
<summary>Lecturer transcript</summary>

The same wage regression as before, and again tidied with broom, because everything we need is in that table. Both cells run on load, so the objects are there when you arrive. Run the second one and look at the row for education: an estimate and a standard error. Those two numbers, plus the degrees of freedom, are the entire input to a confidence interval. Nothing else about the regression matters for this. The next tab collects them explicitly before we do the arithmetic.

</details>

**Run OLS and extract necessary information**

```{webr-r}
#| autorun: true
wage_reg <- feols(wage ~ educ + exper + tenure, data = wage1)
```

<br>

Applying `broom::tidy()` to `wage_reg`,

```{webr-r}
#| autorun: true
(
wage_reg_coef <- broom::tidy(wage_reg)
)
```

```{r}
#| include: false
wage_reg <- feols(wage ~ educ + exper + tenure, data = wage1)
wage_reg_coef <- broom::tidy(wage_reg)
# named df_wage, not df: `df()` is base R's F density, so a bare `df` that is
# never assigned resolves to the function instead of erroring
df_wage <- degrees_freedom(wage_reg, "resid")
b_educ <- wage_reg_coef %>% filter(term == "educ") %>% pull(estimate)
se_educ <- wage_reg_coef %>% filter(term == "educ") %>% pull(std.error)
```

#### Collect necessary information

<details class="transcript">
<summary>Lecturer transcript</summary>

Three numbers, and I want you to see how few there are. The estimate is about zero point six, the standard error about zero point zero five, and the degrees of freedom are five hundred and twenty-two, which is five hundred and twenty-six observations minus three regressors minus one for the intercept. That is the whole input. Note the degrees of freedom is computed from the fit rather than typed in, which matters more than it sounds: a stale hardcoded number is exactly the kind of error that renders without complaint and quietly misleads.

</details>

We are interested in getting the 90% confidence interval of the coefficient on `educ` $(\beta_{educ})$. Under all the assumptions (MLR.1 through MLR.6), we know that in general,

$\frac{\widehat{\beta}_{educ} - \beta_{educ}}{\widehat{se(\widehat{\beta}_{educ})}} \sim t_{n-k-1}$

Specifically for this regression,

+ $\widehat{\beta}_{educ}$ = `r round(b_educ, 4)`

+ $\widehat{se(\widehat{\beta}_{educ})}$ = `r round(se_educ, 4)`

+ $n - k - 1 = `r df_wage`$ (degrees of freedom)

#### Find CI

<details class="transcript inline">
<summary>Lecturer transcript</summary>

Now the arithmetic. We want a ninety percent interval, so ten percent is left over, five at each end, and we need the five and ninety-five percent quantiles of t with five hundred and twenty-two degrees of freedom. Run the cell: they are symmetric, about minus one point six five and one point six five. Then scale each by the standard error and shift by the estimate. Compare the result against the `In practice` tab, where broom computes the same interval directly. They agree, which is the point of doing it by hand once.

</details>

```{r include = F}
u <- seq(-3, 3, length = 1000)
pdf <- dt(u, df = df_wage) # default is mean=0,sd=1
data <- data.table(x = u, y = pdf)

g_t <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.05
c_rej <- qt(1 - alpha / 2, df = df_wage)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_t_005 <- g_t +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  )
```

::: {.columns}

::: {.column width="50%"}
Now, we need to find the 5% $\big((100-90)/2\big)$ and 95% $\big(100-(100-90)/2\big)$ quantiles of $t_{`r df_wage`}$.

```{webr-r}
qt(0.05, df = 522)
qt(0.95, df = 522)
```

<br>

So, the 90% CI of $\frac{`r round(b_educ, 3)` - \beta_{educ}}{`r round(se_educ, 3)`} \sim t_{`r df_wage`}$ is [`r round(qt(0.05, df_wage), 3)`, `r round(qt(0.95, df_wage), 3)`]

By scaling and shifting, the lower and upper bounds of the 90% CI of $\beta_{educ}$ are:

+ lower bound: `r round(b_educ, 3)` + `r round(se_educ, 3)` $\times$ `r round(qt(0.05, df_wage), 3)` = `r round(b_educ + se_educ * qt(0.05, df_wage), 3)`

+ upper bound: `r round(b_educ, 3)` + `r round(se_educ, 3)` $\times$ `r round(qt(0.95, df_wage), 3)` = `r round(b_educ + se_educ * qt(0.95, df_wage), 3)`

Confirm these against the `In practice` tab, which asks `broom` for the same interval.


:::
<!--end of the 1st column-->
::: {.column width="50%"}
The distribution of $t_{`r df_wage`}$:

```{r echo = F}
g_t_005 + theme_bw()
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

#### In practice

<details class="transcript">
<summary>Lecturer transcript</summary>

And this is what you will actually do. Broom's tidy takes a `conf.int` argument, and `conf.level` lets you ask for something other than ninety-five percent. Run it and find the education row: the low and high columns should match what we computed by hand on the previous tab. One habit worth forming now. When you report a coefficient in your paper, report the interval too, not just the star. The interval tells your reader what range of effects the data is consistent with, which is almost always the more useful statement.

</details>

You can just use `broom::tidy()` with `conf.int = TRUE`, `conf.level = confidence level` like below:

```{webr-r}
broom::tidy(
    wage_reg, 
    conf.int = TRUE, 
    conf.level = 0.9
  ) %>%
  dplyr::relocate(term, conf.low, conf.high)
```

:::
<!--end of panel-->

:::
<!--end of panel-->


# Linear Combination of Multiple Coefficients

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Linear Combination of Multiple Coefficients

::: {.panel-tabset}

### Example

<details class="transcript">
<summary>Lecturer transcript</summary>

A genuinely different question, and notice that it is not about any single coefficient. Two dummy variables, one for attending a two-year college and one for a four-year college, so beta-one and beta-two are their respective effects on log wage. The question is whether four-year college has the bigger effect, which is a claim about the two coefficients *relative to each other*. Neither one on its own answers it. Read the hypotheses at the bottom, and notice the null is equality: the two kinds of college have the same effect.

</details>

**Model**

$$log(wage) = \beta_0+\beta_1 jc+ \beta_2 univ + \beta_3 exper + u$$

+ $jc$: 1 if you attended 2-year college, 0 otherwise
+ $univ$: 1 if you attended 4-year college, 0 otherwise

**Question**

Is the impact of education on wage greater if you attend a 4-year college than a 2-year college?

**Hypothesis**

The null and alternative hypotheses would be:

+ $H_1:$ $\beta_1 < \beta_2$
+ $H_0:$ $\beta_1 = \beta_2$

### Rewrite the hypothesis

<details class="transcript">
<summary>Lecturer transcript</summary>

A small piece of algebra that makes the whole problem go away. Move beta-two to the other side and the hypothesis becomes a statement about beta-one minus beta-two. Then give that difference a name, alpha, and read what you are left with: the null is that alpha equals zero, the alternative that it is negative. That is exactly the shape of every test we have already done. The last line is the key observation: alpha is a linear combination of coefficients, and the next tab says why that class of things is well behaved.

</details>

The null and alternative hypotheses are:

+ $H_1:$ $\beta_1 < \beta_2$
+ $H_0:$ $\beta_1 = \beta_2$

Rewriting them,

+ $H_1:$ $\beta_1-\beta_2 < 0$
+ $H_0:$ $\beta_1 - \beta_2 =0$

Or,  

+ $H_1:$ $\alpha < 0$
+ $H_0:$ $\alpha =0$

where $\alpha = \beta_1 - \beta_2$

Note that $\alpha$ is a linear combination of $\beta_1$ and  $\beta_2$.

### Math

<details class="transcript">
<summary>Lecturer transcript</summary>

The fact that makes this work, in the box. For any linear combination of OLS coefficients, the standardised version follows a t distribution with the same degrees of freedom as before. Not approximately, exactly, under our assumptions. Why? Because a linear combination of jointly Normal things is Normal, and the same estimated-denominator argument gives you t. So the difference between two coefficients is no harder than a single coefficient. Read the second expression and see that the only unfamiliar piece is the standard error of the difference, which is not what you might guess.

</details>

:::{.callout-note title="Important Fact"}
For any linear combination of the OLS coefficients, denoted as $\widehat{\alpha}$, the following holds:

$$\frac{\widehat{\alpha}-\alpha}{\widehat{se(\widehat{\alpha})}} \sim t_{n-k-1}$$

Where $\alpha$ is the true value (it is $\beta_1 - \beta_2$ in the example in the previous slide).
:::

<br>

So, using the example, this means that 

$$\frac{\widehat{\alpha}-\alpha}{\widehat{se(\widehat{\alpha})}} = \frac{\widehat{\beta}_1-\widehat{\beta}_2-(\beta_1 - \beta_2)}{\widehat{se(\widehat{\beta}_1-\widehat{\beta}_2)}} \sim t_{n-k-1}$$

This is great because we know how to do t-test!

### Going back to the example

<details class="transcript">
<summary>Lecturer transcript</summary>

Substituting the null into the general result. Alpha is zero under the null, so the numerator is just the difference of the two estimates. Read the red "if" carefully, because it carries the whole logic: this distribution holds only under the null. Then the last sentence lists what you need to plug in. Two of the three you can read straight off the regression output. The third, the standard error of the difference, you cannot, and that is the entire subject of the next tab.

</details>

Our null hypothesis is $\alpha = 0$ (or $\beta_1 - \beta_2 = 0$).

So, <span style = "color: red;"> If </span> indeed the null hypothesis is true, then

$$\frac{\widehat{\alpha}-0}{\widehat{se(\widehat{\alpha})}} = \frac{\widehat{\beta}_1-\widehat{\beta}_2-0}{\widehat{se(\widehat{\beta}_1-\widehat{\beta}_2)}} \sim t_{n-k-1}$$

So, all you need to do is to substitute $\widehat{\beta}_1$, $\widehat{\beta}_2$, $\widehat{se(\widehat{\beta}_1 - \widehat{\beta}_2)}$ into the formula and see if the value is beyond the critical value for your chosen level of statistical significance.

### get $se(\widehat{\beta}_1-\widehat{\beta}_2)$

<details class="transcript">
<summary>Lecturer transcript</summary>

Here is the trap. It would be very convenient if the variance of a difference were the sum of the variances, and the slide states plainly that it is not. Read the math aside: expanding the variance of a linear combination produces a covariance term in the middle. With a equals one and b equals minus one, the two variances add and the covariance term comes in with a minus two in front. So you need the covariance between the two coefficient estimators, which does not appear anywhere in a standard regression table. That is why the next slide introduces the variance-covariance matrix.

</details>

But, 

$$se(\widehat{\beta}_1-\widehat{\beta}_2)= \sqrt{Var(\widehat{\beta}_1-\widehat{\beta}_2)} \ne \sqrt{Var(\widehat{\beta}_1)+Var(\widehat{\beta}_2)}$$

If the following was true,

$$se(\widehat{\beta}_1-\widehat{\beta}_2) = \sqrt{Var(\widehat{\beta}_1)+Var(\widehat{\beta}_2)}$$

then, we could have just extracted $Var(\widehat{\beta}_1)$ and $Var(\widehat{\beta}_2)$ individually from the regression object on R, sum them up, and take a square root of it.

:::{.callout-note title="Math aside"}
$$Var(ax+by) = a^2 Var(x) + 2abCov(x,y) + b^2 Var(y)$$
:::

So, 

$$se(\widehat{\beta}_1-\widehat{\beta}_2)= \sqrt{Var(\widehat{\beta}_1-\widehat{\beta}_2)} = \sqrt{Var(\widehat{\beta}_1)-2Cov(\widehat{\beta}_1,\widehat{\beta}_2)+Var(\widehat{\beta}_2)}$$

### Demonstration using R

::: {.panel-tabset}

#### Regression

<details class="transcript">
<summary>Lecturer transcript</summary>

Real data again, this time the two-year college study. Log wage on the two college dummies and experience. The cell runs on load and prints the fit, so look at the two coefficients of interest. You will see the four-year effect is larger than the two-year effect, which is what we expected. But that is a comparison of two point estimates, and both are random. The question is whether the gap is bigger than sampling noise would produce, and answering it is what the next three tabs do.

</details>

```{webr-r}
#| autorun: true
#--- import data ---#
data("twoyear", package = "wooldridge")

(
reg_sc <- fixest::feols(lwage ~ jc + univ + exper, data = twoyear) # OLS
)
```

#### Variance covariance matrix

<details class="transcript">
<summary>Lecturer transcript</summary>

The object we were missing. Read the definition: the diagonal entries are the variances of each coefficient estimator, and the off-diagonal entries are the covariances between them. The standard errors in a regression table are just the square roots of the diagonal, which is why the covariances never appear there. Run `vcov` on the fit and look at the matrix. It is symmetric, and it is square with one row and column per coefficient including the intercept, which is why the indexing in the last line starts at two rather than one.

</details>

:::{.callout-note title="Definition"}
Variance covariance matrix is a matrix where

+ $VCOV_{i,i}$: the variance of $i$th variable's coefficient estimator
+ $VCOV_{i,j}$: the covariance between $i$th and $j$th variables' estimators
:::

<br>

You can get it by applying `vcov()` to regression results:

```{webr-r}
#| autorun: true
(
vcov_sc <- vcov(reg_sc) # variance covariance matrix
)
```

<br>

For example, $vcov_{sc}[2, 2]$ is the variance of $\widehat{\beta}_{jc}$, and $vcov_{sc}[2, 3]$ is the covariance between $\widehat{\beta}_{jc}$ and $\widehat{\beta}_{univ}$.

#### Calculate the t-statistic

<details class="transcript">
<summary>Lecturer transcript</summary>

Now assemble it. The numerator is the difference between the two coefficient estimates, straightforward. The denominator is the formula from two tabs back, read directly out of the matrix: the variance of the two-year coefficient, minus two times the covariance, plus the variance of the four-year coefficient, all under a square root. Notice we index the matrix by name rather than by number, which is worth copying as a habit; positional indexing breaks silently the moment you add a regressor. Then divide, and compare the result against the critical value for a one-sided test.

</details>

Get coefficient estimates: 

```{webr-r}
#| autorun: true
#--- reg results ---#
reg_sc_td <- broom::tidy(reg_sc)

#--- get coefficients ---#
beta_jc <-
  reg_sc_td %>%
  filter(term == "jc") %>%
  pull(estimate)

beta_univ <-
  reg_sc_td %>%
  filter(term == "univ") %>%
  pull(estimate)
```

<br>

Calculate t-statistic:

```{webr-r}
#| autorun: true
numerator <- beta_jc - beta_univ
denominator <-
  sqrt(
    vcov_sc["jc", "jc"] - 2 * vcov_sc["jc", "univ"] + vcov_sc["univ", "univ"]
  )
t_stat <- numerator / denominator
t_stat
```

#### In practice

<details class="transcript inline">
<summary>Lecturer transcript</summary>

And of course there is a function for it. `car::linearHypothesis` takes the fit and the hypothesis written as a string, and does all of that internally. Run it and read the output. Then read the note underneath, because the output is not a t-statistic: it reports a chi-squared value, which is the square of the t-statistic. Run the last cell to confirm they match. The callout is the thing to remember: squaring is a monotone transformation on the positive side, so the two tests always reach the same verdict.

</details>

In practice, you can just use `car::linearHypothesis()` to implement t-test involving a linear combination of multiple coefficients.

**Syntax**

```{r}
#| eval: false
car::linearHypothesis(regression_results, expression_of_the_test)
```

<br>

**Example**

```{webr-r}
#| autorun: true
car::linearHypothesis(reg_sc, "jc=univ")
```

<br>

Please note that it runs $\chi^2$ test. When t-statistic is squared, it follows a $\chi^2$ distribution. So, it calculates the square of the t-statistic and look up the $\chi^2$ distribution critical value to decide whether to reject the null or not. 

Let's confirm this:

```{webr-r}
#| autorun: true
t_stat^2
```

<br>

Yes, this is the same as the $\chi^2$ value reported by `car::linearHypothesis(reg_sc, "jc=univ")`

:::{.callout-important}
The test results would be the same whether you run t-test or $\chi^2$ test.
:::

:::
<!--end of panel-->

:::
<!--end of panel-->

# Multiple Linear Restrictions: F-test

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Multiple Linear Restrictions: F-test


::: {.panel-tabset}

### Example

::: {.panel-tabset}

#### Model and hypothesis

<details class="transcript">
<summary>Lecturer transcript</summary>

A new kind of question, and read the hypothesis sentence rather than just the symbols. Once years in the league and games per year are controlled for, do the three performance measures matter *collectively*? Batting average, home runs, runs batted in. The null sets all three coefficients to zero simultaneously. Notice that is a single hypothesis about three parameters, not three separate hypotheses. And read the alternative: it is simply that the null is false, which happens as soon as any one of them is non-zero.

</details>

**Model**

$$log(salary) =  \beta_0 + \beta_1 years + \beta_2 gamesyr + \beta_3 bavg + \beta_4 hrunsyr + \beta_5 rbisyr + u$$

+ $salary$: salary in 1993
+ $years$: years in the league
+ $gamesyr$: average games played per year
+ $bavg$: career batting average
+ $hrunsyr$: home runs per year
+ $rbisyr$: runs batted in per year

**Hypothesis**

Once years in the league and games per year have been controlled for, the statistics measuring performance ( $bavg$, $hrunsyr$, $rbisyr$) have no effect on salary collectively.

$H_0$: $\beta_3=0$, $\beta_4=0$, and $\beta_5=0$

$H_1$: $H_0$ is not true

#### Questions

<details class="transcript">
<summary>Lecturer transcript</summary>

Before we build new machinery, ask whether the old machinery would do. The obvious idea is to test each of the three coefficients separately with a t-test and combine the verdicts somehow. Think about why that is unsatisfying before moving on. Two reasons. First, running three tests at five percent each means your chance of at least one false rejection is well above five percent. Second, and more importantly for what follows, three individually weak pieces of evidence can add up to one strong piece, and separate tests will never see it.

</details>
  
How do we test this?

+ $H_0$ holds only if $\beta_3$, $\beta_4$, <span style = "color: red;"> and </span> $\beta_5$ are <span style = "color: red;"> all </span> zero.
+ Conduct t-test for each coefficient individually?

:::
<!--end of panel-->

### Individual t-test

::: {.panel-tabset}

#### Run regression

<details class="transcript">
<summary>Lecturer transcript</summary>

The baseball data, and the full model with all five regressors. This runs on load. Look at the tidy output and specifically at the three performance variables: batting average, home runs per year, runs batted in per year. Read their estimates and their standard errors, and form a view about each before clicking to the next tab. These are the variables the null says are all zero. You are about to discover something uncomfortable about what the individual tests say.

</details>

```{webr-r}
#| autorun: true
#--- load the mlb data ---#
data("mlb1", package = "wooldridge")

#--- run a rergession ---#
mlb_reg <- fixest::feols(log(salary) ~ years + gamesyr + bavg
  + hrunsyr + rbisyr, data = mlb1)

#--- take a look at the results ---#
broom::tidy(mlb_reg)
```

#### Question

<details class="transcript">
<summary>Lecturer transcript</summary>

Same output again, and now I want you to actually look. Go along the three performance rows and check the p-values. What do you find? Take a moment before opening the answer tab, because the surprise only works if you have looked. And ask yourself the follow-up question while you are there: does what you see match your intuition about whether batting average and home runs matter for a baseball player's salary? Because the answer you are about to get is the opposite of what any baseball fan would tell you.

</details>

```{webr-r}
#| autorun: true
broom::tidy(mlb_reg)
```

What do you find?

#### Answer

<details class="transcript">
<summary>Lecturer transcript</summary>

None of the three is significant, not even at the ten percent level. Taken at face value that says a player's batting average, home runs and runs batted in have nothing to do with his salary, which is absurd. Read the next lines carefully. If you concluded they have no joint effect you would be wrong, and the last line points at the fix: the sum of squared residuals, and equivalently R-squared, can answer the joint question even when the individual tests cannot. That is the F-test, and it is the rest of this section.

</details>

None of the coefficients on `bavg`, `hrunsyr`, and `rbisyr` is statistically significantly different from 0 even at the 10% level!!

So, does this mean that they collectively have no impact on the salary of MLB players?

If you were to conclude that they do not have statistically significant impact jointly, you would turn out to be wrong!!

$SSR$ (or $R^2$) turns out to be useful for testing their impacts jointly.

:::
<!--end of panel-->

### F-test

In doing an F-test of the null hypothesis, we compare sum of squared residuals $(SSR)$ of two models:

<br>

**Unrestricted Model**

$$log(salary) = \beta_0 + \beta_1 years + \beta_2 gamesyr + \beta_3 bavg + \beta_4 hrunsyr + \beta_5 rbisyr + u$$

<br>

**Restricted Model**

$$log(salary) = \beta_0 + \beta_1 years + \beta_2 gamesyr + u$$

The coefficients on $bavg$, $hrunsyr$, and $rbisyr$ are restricted to be 0 following the null hypothesis. 

::: {.panel-tabset}
#### Question

<details class="transcript">
<summary>Lecturer transcript</summary>

The idea behind the F-test in one question, and it is worth answering out loud. We are going to fit two models: the full one, and a restricted one that simply drops the three performance variables, which is what the null says you may do. Compare their sums of squared residuals. Which is bigger, and can you be sure? Think about it in terms of what OLS does. It picks coefficients to minimise the sum of squared residuals, so what happens when you take some of its freedom away?

</details>

If the null hypothesis is indeed true, then what do you think is going to happen if you compare the $SSR$ of the two models? Which one has a bigger $SSR$?

#### Answer

<details class="transcript">
<summary>Lecturer transcript</summary>

The restricted model has the larger sum of squared residuals, and this is not a matter of probability. Setting three coefficients to zero is a constraint, and a constrained minimum can never be lower than an unconstrained one. Worst case the restriction happens to be exactly what OLS wanted anyway and the two are equal. So the difference is always non-negative. The real question is whether it is *large*, and that is what needs a distribution: large compared to what?

</details>

$SSR$ from the restricted model should be large because the restricted model has a smaller explanatory power than the unrestricted model.

:::
<!--end of panel-->

### SSR

**SSR of the unrestricted model**: $SSR_u$

```{webr-r}
#| autorun: true
#--- run OLS ---#
res_u <- feols(log(salary) ~ years + gamesyr + bavg
  + hrunsyr + rbisyr, data = mlb1)

#--- SSR ---#
sum(res_u$residuals^2)
```

<br>

**SSR of the restricted model**: $SSR_r$

```{webr-r}
#--- run OLS ---#
res_r <- feols(log(salary) ~ years + gamesyr, data = mlb1)

#--- SSR ---#
sum(res_r$residuals^2)
```

<br>

::: {.panel-tabset}
#### Question 1

<details class="transcript">
<summary>Lecturer transcript</summary>

Both sums of squared residuals are now computed, so look at the two numbers before answering. The restricted one is larger, as we argued. Now: what does the gap between them actually measure? Think about what changed between the two fits. The only difference is that three variables were allowed to help in one and forced out of the other. So whatever extra fit the full model achieved has to be attributable to those three variables, and only to them, since everything else was identical.

</details>

What does $SSR_r - SSR_u$ measure?

#### Answer 1

<details class="transcript">
<summary>Lecturer transcript</summary>

The gap is the joint contribution of the three excluded variables to explaining the dependent variable. Not the contribution of any one of them, and not the sum of three separate contributions, but what they add together, including whatever they explain through being correlated with each other. That last part is exactly what the individual t-tests could not see. When two variables carry overlapping information, neither looks essential alone, but removing both at once costs you a great deal of fit.

</details>

The contribution from the three excluded variables in explaining the dependent variable.

#### Question 2

<details class="transcript">
<summary>Lecturer transcript</summary>

The natural next question, and the honest answer is on the following tab. You have a number, the gap between the two sums of squared residuals. Is it big? Look at it and try to decide. You cannot, and it is worth being clear about why: the number has units, it depends on how y was scaled, on the sample size, on how noisy the data is. A gap of five means nothing without something to compare it against.

</details>

Is the contribution large enough to say that the excluded variables are important? 

#### Answer 2

<details class="transcript">
<summary>Lecturer transcript</summary>

Exactly: you cannot tell, because you do not know the distribution of that difference. This is step two of the five steps from the very first slide, and it is always the hard one. We know how to compute a number; what we lack is what that number would typically look like if the null were true. The next tab supplies it, by scaling the difference into a ratio whose distribution has a name and a table.

</details>

Cannot tell at this point because we do not know the distribution of the difference!

:::
<!--end of panel-->

### F-test in general

**Setup**

Consider a following general model:

$$
y = \beta_0 +\beta_1 x_1 + \dots+\beta_k x_k +u
$$

Suppose we have $q$ restrictions to test: that is, the null hypothesis states that $q$ of the variables have zero coefficients.

$$H_0: \beta_{k-q+1} =0, \beta_{k-q+2} =0, \dots, \beta_k=0$$

When we impose the restrictions under $H_0$, the restricted model is the following:

$$y = \beta_0 +\beta_1 x_1 + \dots+\beta_{k-q} x_{k-q} + u$$


**F-statistic**

<span style = "color: red;"> If </span> the null hypothesis is true, then,

$$F = \frac{(SSR_r-SSR_u)/q}{SSR_u/(n-k-1)} \sim F_{q,n-k-1}$$

+ $q$: the number of restrictions
+ $n-k-1$: degrees of freedom of residuals

::: {.panel-tabset}

#### Question

<details class="transcript">
<summary>Lecturer transcript</summary>

A quick sanity check on the formula above before we use it. The F-statistic is a ratio, and you should be able to say something about its sign without computing anything. Look at the numerator and the denominator separately. The denominator is a sum of squares divided by a positive number, so it is positive. What about the numerator? Think back to the tab where we compared the two sums of squared residuals.

</details>

Is the above $F$-statistic always positive?

#### Answer

<details class="transcript">
<summary>Lecturer transcript</summary>

Non-negative, because the restricted sum of squared residuals can never be below the unrestricted one, for the constrained-minimum reason we gave earlier. That is worth noticing because it changes how you test. Unlike a t-statistic, an F-statistic never comes out negative, so there is no such thing as a two-sided F-test. You only ever look at the upper tail, and a large value is the only kind of evidence against the null.

</details>

Yes, because $SSR_r-SSR_u$ is always non-negative: the restricted model can never fit better than the unrestricted one.

#### Question 2

<details class="transcript">
<summary>Lecturer transcript</summary>

One more reading of the formula, and this one tells you what the test is actually sensitive to. Suppose the three variables collectively matter a great deal. What does that do to the gap between the two sums of squared residuals, and therefore to the F-statistic? Answer it from the formula rather than from intuition, then check. Getting this direction right is what makes the decision rule memorable rather than something you have to look up.

</details>

The greater the joint contribution of the $q$ variables, the (greater or smaller) the $F$-statistic?

#### Answer 2

<details class="transcript">
<summary>Lecturer transcript</summary>

Greater. The more the excluded variables contribute, the bigger the gap in the numerator, and the bigger F becomes. So large F means the restriction hurt, which means the variables mattered, which means reject the null. That is the whole decision rule and now you can reconstruct it rather than memorise it. And it explains why we only look at the upper tail: small F just means dropping the variables cost you nothing.

</details>

Greater.

:::
<!--end of panel-->

### F-test steps

<details class="transcript">
<summary>Lecturer transcript</summary>

The recipe on the right, and it is the same five steps from the first slide of the lecture, specialised. Define the null, fit both models, compute F, find the critical value, compare. On the left is the distribution itself. Notice it lives entirely on the positive side, consistent with what we just established, and that it is skewed rather than symmetric. Notice too that it takes *two* degrees of freedom parameters, the number of restrictions and the residual degrees of freedom, so you must get both right when you look up a critical value.

</details>

::: {.columns}
::: {.column width="60%"}
**F-distribution**

```{r echo = F}
x <- seq(0, 4, length = 1000)
y_1 <- df(x, df1 = 2, df2 = 100)
y_2 <- df(x, df1 = 4, df2 = 400)

f_plot <- data.table(x = x, "F(2,100)" = y_1, "F(4,400)" = y_2) %>%
  melt(id.var = "x")

ggplot(data = f_plot) +
  geom_line(aes(x = x, y = value, color = variable)) +
  scale_color_discrete(name = "") +
  theme(
    legend.position = "bottom"
  ) +
  theme_bw()
```
:::
<!--end of the 1st column-->
::: {.column width="40%"}
**F-test steps**

+ Define the null hypothesis
+ Estimate the unrestricted and restricted models to obtains their $SSR$
+ Calculate $F$-statistic
+ Define the significance level and corresponding critical value according to the F distribution with appropriate degrees of freedoms
+ Reject if your $F$-statistic is greater than the critical value, otherwise do not reject
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

### F-test by hand

::: {.panel-tabset}

#### Step 1 

<details class="transcript">
<summary>Lecturer transcript</summary>

Step one in code: fit both models. The unrestricted one has all five regressors; the restricted one drops exactly the three the null sets to zero. Then compute each sum of squared residuals by squaring the residuals and adding them up. Read the two model formulas side by side and check that the only difference is those three variables. That is what makes the comparison valid: everything else about the two fits, the data, the other regressors, the estimator, is identical.

</details>

Estimate the unrestricted and restricted models

```{webr-r}
#| autorun: true
#--- unrestricted model ---#
reg_u <- feols(log(salary) ~ years + gamesyr +
  bavg + hrunsyr + rbisyr, data = mlb1)

SSR_u <- sum(reg_u$residuals^2)

#--- restricted model ---#
reg_r <- feols(log(salary) ~ years + gamesyr, data = mlb1)

SSR_r <- sum(reg_r$residuals^2)
```

#### Step 2

<details class="transcript">
<summary>Lecturer transcript</summary>

Now assemble the statistic. Three restrictions, so `df_q` is three. The residual degrees of freedom come from the unrestricted fit. Then the numerator is the gap divided by the number of restrictions, and the denominator is the unrestricted sum of squared residuals divided by its degrees of freedom. Notice what that denominator is: it is essentially sigma-hat squared, the estimated error variance. So the F-statistic is comparing the fit you gained against the noise level, which is the natural way to make a gap unit-free.

</details>

Calculate F-stat

```{webr-r}
df_q <- 3 # the number of restrictions
df_ur <- degrees_freedom(reg_u, "resid") # degrees of freedom for the unrestricted model
F_stat_num <- (SSR_r - SSR_u) / df_q # numerator of F-stat
F_stat_denom <- SSR_u / df_ur # denominator of F-stat
F_sta <- F_stat_num / F_stat_denom # F-stat
F_sta
```

#### Steps 3 and 4

<details class="transcript">
<summary>Lecturer transcript</summary>

The critical value comes from `qf`, and note it needs both degrees of freedom arguments. Five percent significance, so we want the ninety-fifth percentile, and because the test is one-tailed by construction there is no halving. Run the comparison. It comes back true, and read the callout: the three performance variables *are* jointly significant, even though not one of them was individually significant at ten percent. That is a genuine contradiction of the naive reading, and the question at the bottom is the right one to ask.

</details>

Find the critical value

```{webr-r}
alpha <- 0.05 # 5%  significance level
c_value <- qf(1 - alpha, df1 = df_q, df2 = df_ur)
c_value
```

Is F-stat > critical value?

```{webr-r}
F_sta > c_value
```

:::{.callout-note title="Observation"}
So, the performance variables have statistically significant impacts on salary <span style = "color: red;"> jointly</span>!!

What happened?
:::

:::
<!--end of panel-->

### F-test (easier way)

::: {.panel-tabset}

#### How

<details class="transcript inline">
<summary>Lecturer transcript</summary>

The practical version. `car::linearHypothesis` takes the unrestricted fit and the restrictions written as text, and does everything we just did by hand. Read the note about the hypothesis argument: it is a character vector, one string per restriction, and each string is an equation. You are not limited to testing against zero, which is worth knowing; you can test whether a coefficient equals one, or whether two coefficients are equal, using exactly the same function.

</details>

You can use the `car::linearHypothesis()` function from the `car` package. 

<br>

**Syntax**

```{r eval = F}
car::linearHypothesis(regression, hypothesis)
```

+ `regression`: the name of the regression results of the unrestricted model

+ `hypothesis`: text of the null hypothesis. For example,

`c("x1 = 0", "x2 = 1")` means the coefficients on $x1$ and $x2$ are $0$ and $1$, respectively

#### Demonstration

<details class="transcript">
<summary>Lecturer transcript</summary>

And here it is on the baseball model. Three restrictions in one vector, all setting a coefficient to zero. Run it and compare the statistic against the one you computed by hand two tabs ago. They should match. From now on this is what you will use, but I wanted you to build it once from sums of squared residuals so that the number means something. If you ever get a surprising F-test result, knowing it is a comparison of two model fits is what lets you work out why.

</details>

The following code test if $\beta_{bavg} = 0$, $\beta_{hrunsyr} = 0$, and $\beta_{rbisyr} = 0$. 

```{webr-r}
#| autorun: true
#--- F-test ---#
car::linearHypothesis(mlb_reg, c("bavg=0", "hrunsyr=0", "rbisyr=0"))
```

:::
<!--end of panel-->

:::
<!--end of panel-->

## Simulation (multicollinearity, t-test, and F-test)


::: {.panel-tabset}

### Data generation

<details class="transcript">
<summary>Lecturer transcript</summary>

A simulation to pin down what just happened, and the construction should look familiar. `x1` and `x2` are each built mostly from a shared term `mu`, so they are strongly correlated by design. `x3` is independent. And read the line that builds y: all three variables have a coefficient of one, so all three genuinely matter, and we know that because we chose it. Run the correlation cell and see how close to one it is. This is the same data generating process as Case two in the multicollinearity lecture.

</details>

```{webr-r}
N <- 300 # num observations
mu <- runif(N) # term shared by indep vars 1 and 2
x1 <- 0.1 * runif(N) + 2 * mu # indep 1
x2 <- 0.1 * runif(N) + 2 * mu # indep 2
x3 <- runif(N) # indep 3
u <- rnorm(N) # error
y <- 1 + x1 + x2 + x3 + u # generate y
data <- data.table(y = y, x1 = x1, x2 = x2, x3 = x3) # combine into a data.table
```

<br>

`x1` and `x2` are highly correlated with each other:

```{webr-r}
cor(x1, x2) # correlation between x1 and x2
```


### Regression

<details class="transcript">
<summary>Lecturer transcript</summary>

Run the regression and look at the two rows for `x1` and `x2`. Both are insignificant, and we know for certain that both have a true coefficient of one, because we wrote it. So this is a case where the individual t-tests are unambiguously giving you the wrong impression, and unlike with the baseball data there is no arguing about it. Look at the standard errors specifically rather than the estimates. They are enormous, and that is the mechanism: not wrong estimates, but imprecise ones.

</details>

```{webr-r}
reg_u <- fixest::feols(y ~ x1 + x2 + x3, data = data) # OLS
broom::tidy(reg_u) # results
```

<br>

Both `x1` and `x2` are statistically insignificant individually. 

### F-test

<details class="transcript">
<summary>Lecturer transcript</summary>

Now the joint test, built by hand so you can see it is the same machinery. The restricted model drops both correlated variables and keeps only `x3`, so there are two restrictions. Compute both sums of squared residuals, form the ratio, and compare against the critical value from `qf`. Run it and look at the size of the F-statistic. It is not marginal; it is enormous. The two variables that individually looked like noise are, taken together, overwhelmingly important, which is exactly what we built into the data.

</details>

```{webr-r}
#--- unrestricted ---#
SSR_u <- sum(reg_u$residuals^2)

#--- restricted ---#
reg_r <- fixest::feols(y ~ x3, data = data)
SSR_r <- sum(reg_r$residuals^2)

#--- degrees of freedom of residuals ---#
df_resid <- degrees_freedom(reg_u, "resid")

#--- F ---#
F_stat <- ((SSR_r - SSR_u) / 2) / (SSR_u / df_resid)

#--- critical value ---#
alpha <- 0.05
(
  c_value <- qf(1 - alpha, df1 = 2, df2 = df_resid)
)
```

<br>

The F-statistic for the hypothesis testing is:

```{webr-r}
#--- F > critical value? ---#
F_stat
F_stat > c_value
```

<br>

The F-statistic is very high, meaning they collectively affect the dependent variable significantly.


### Important

<details class="transcript">
<summary>Lecturer transcript</summary>

The explanation, and then the connection back. The standard errors are large because the two variables are so alike that the estimation cannot tell their individual contributions apart. Read the callout, because this is lecture 04 seen from the other side. There we showed correlation inflates the variance through the one minus R-squared-j term. A bigger variance means a bigger standard error, and a bigger standard error means a smaller t-statistic. The insignificance you saw is that inflation, arriving at the testing stage. The F-test dodges it by never asking about the variables one at a time.

</details>

The standard error estimates of the coefficients on `x1` and `x2` are very high because they are so highly correlated that the estimation had a hard time distinguishing their individual impacts.

But collectively they have large impacts, and the $F$-test was able to detect the statistical significance of their impacts <span style = "color: red;"> collectively</span>.

<br>

:::{.callout-important title="This is lecture 04, seen from the other side"}
The data generating process here is Case 2 from the multicollinearity lecture: two regressors sharing a common term. There we showed that correlation inflates $Var(\widehat{\beta}_j)$ through the $1-R^2_j$ term in the denominator.

A large variance means a large standard error, and a large standard error means a <span style = "color: red;"> small </span> t-statistic &mdash; which is exactly the insignificance you see on the `Regression` tab. Multicollinearity did not make the estimates wrong; it made them too imprecise to distinguish individually. The $F$-test sidesteps the problem entirely by never asking about them one at a time.
:::

:::
<!--end of panel-->

## MLB example

::: {.panel-tabset}

### Correlation check

<details class="transcript">
<summary>Lecturer transcript</summary>

Back to the baseball data, to check whether the same mechanism explains what we saw there. Run the correlation matrix on the three performance variables. Home runs per year and runs batted in per year are strongly correlated, which makes sense: a home run drives in at least one run by definition, so a player who hits many home runs mechanically has a high runs-batted-in figure. Batting average is much less related to either. So two of the three are near-duplicates, and that is enough to produce what we saw.

</details>

Here is the correlation coefficients between the three variables:

```{webr-r}
#| autorun: true
dplyr::select(mlb1, bavg, hrunsyr, rbisyr) %>% cor()
```

As you can see, `rbisyr` and `hrunsyr` are highly correlated with each other. 

They are not so highly correlated with `bavg`.



### Multiple coefficients again

<details class="transcript">
<summary>Lecturer transcript</summary>

One last connection, and it ties the two halves of this lecture together. Testing a linear combination is just an F-test with one restriction. The fact underneath is exact: F with one numerator degree of freedom is the square of t on the same denominator degrees of freedom. Run the cell, which takes the square root of the chi-squared value, and compare against the t-statistic we computed by hand on the college slide. Same number. So there were never two procedures here, only one written two ways.

</details>

The test of a linear combination of the parameters we looked at earlier is a special case of F-test where the number of restriction is 1. 

It can be shown that square root of $F_{1, df}$ follows the $t_{df}$ distribution.

So, we can actually use the $F$-test for this type of hypothesis testing because $F_{1,n-k-1} \sim t_{n-k-1}^2$.

```{webr-r}
#--- OLS ---#
reg_sc <- feols(lwage ~ jc + univ + exper, data = twoyear) # OLS

#--- F-test ---#
F_test <- car::linearHypothesis(reg_sc, c("jc-univ=0"))

#--- t-statistic ---#
sqrt(F_test$Chisq[2])
```

<br>

Check with the previous slide of the same t-test and confirm that the t-statistics we got there and here are the same.


:::
<!--end of panel-->

## Everything in this lecture is one idea

::: {.panel-tabset}

### The recipe

<details class="transcript">
<summary>Lecturer transcript</summary>

Everything in this lecture was one recipe, and this is it, unchanged from the first slide. State the hypotheses. Find the distribution of a statistic assuming the null. Compute it. Pick a level, which fixes a critical value. Compare. Read the sentence at the bottom, because it is the useful compression: the only thing that varied from test to test was which statistic you built and which distribution you looked it up in. If a future test looks unfamiliar, ask those two questions and it usually stops being unfamiliar.

</details>

Every test we did followed the same five steps from the very first slide:

+ write down $H_0$ and $H_1$
+ find the distribution of a test statistic <span style = "color: red;"> assuming $H_0$ is true </span>
+ compute the statistic from your data
+ pick a significance level, which fixes a critical value
+ reject if your statistic is more extreme than that critical value

<br>

The only thing that changed from test to test was <span style = "color: blue;"> which statistic </span> you build and <span style = "color: blue;"> which distribution </span> you look it up in.

### The four tests

<details class="transcript">
<summary>Lecturer transcript</summary>

The four things we did, side by side. Read down the statistic column and notice how similar the first two are: an estimate, minus its value under the null, over its standard error. The only difference is what counts as the estimate. The third is genuinely different, because it compares two model fits rather than standardising a single number, and that is why it can answer a question the first two cannot. And read the two bullets: the covariance term is why the variance-covariance matrix had to appear at all.

</details>

| Question | Statistic | Distribution under $H_0$ | In R |
|---|---|---|---|
| Is one coefficient equal to a value? | $\dfrac{\widehat{\beta}_j - \beta_j^0}{\widehat{se(\widehat{\beta}_j)}}$ | $t_{n-k-1}$ | `broom::tidy()`, or by hand |
| Is a linear combination equal to a value? | $\dfrac{\widehat{\alpha} - \alpha^0}{\widehat{se(\widehat{\alpha})}}$ | $t_{n-k-1}$ | `car::linearHypothesis()` |
| Are $q$ coefficients <span style = "color: red;"> jointly </span> zero? | $\dfrac{(SSR_r-SSR_u)/q}{SSR_u/(n-k-1)}$ | $F_{q,\,n-k-1}$ | `car::linearHypothesis()` |
| What range of values is consistent with the data? | the CI, not a test | $t_{n-k-1}$ | `broom::tidy(conf.int = TRUE)` |

+ the second row needs $Cov(\widehat{\beta}_1,\widehat{\beta}_2)$, which is why the variance-covariance matrix appeared
+ the third row is the only one that cannot be built from a single coefficient and its standard error

### How they relate

<details class="transcript">
<summary>Lecturer transcript</summary>

Two connections that reduce the count. First, the F-test with one restriction is the t-test squared, which we confirmed numerically two slides ago. Second, and this one is worth internalising: a confidence interval and a two-sided test are the same statement. The ninety-five percent interval is precisely the set of null values you would fail to reject at five percent. So reporting an interval reports infinitely many tests at once, which is a good reason to prefer it. Then read the callout, which is the thing I most want you to carry away.

</details>

Two connections worth remembering, because they mean there are fewer ideas here than there appear to be:

+ **$F$ with one restriction is just $t$ squared.** $F_{1,\,n-k-1} \sim t_{n-k-1}^2$, which is why `car::linearHypothesis()` gave the same answer as the t-test we did by hand.

+ **The confidence interval and the two-sided test are the same statement.** The 95% CI is exactly the set of null values you would <span style = "color: blue;"> fail </span> to reject at the 5% level. If a CI excludes zero, the t-test rejects $\beta_j = 0$; if it contains zero, it does not.

<br>

:::{.callout-important title="The one thing to carry into your own work"}
An individually insignificant coefficient is <span style = "color: red;"> not </span> evidence of no effect. The MLB players and the simulation both showed variables that looked useless one at a time and mattered a great deal together. When variables are correlated, test them <span style = "color: blue;">jointly</span>.
:::

:::
<!--end of panel-->



