---
title: "05: Hypothesis Testing"
format: 
  revealjs: 
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to course website with lecture slides](https://tmieno2.github.io/Applied-Econometrics-MS-Quarto/lectures/)"
execute:
  echo: true
  out-width: 55%
webr:
  packages: ['fixest', 'dplyr', 'data.table', 'ggplot2', 'wooldridge', 'broom', 'car']
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 10
    fig-height: 6
    out-width: 100%
filters:
  - webr
---

# Review on Statistical Hypothesis Testing

```{r}
#| include: false
library(dplyr)
library(wooldridge)
library(data.table)
library(ggplot2)
library(fixest)

source(here::here("lectures/_lecture-theme.R"))
```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Hypothesis Testing

::: {.panel-tabset}

### General steps

<details class="transcript">
<summary>Transcript</summary>

Every test in this lecture follows these same five steps, so use this slide as a map. Step one is to state the null hypothesis, H zero, and the alternative, H one. The null is the claim you subject to a test; the alternative describes the departure that would make you reject it.

Step two is the intellectual core. Choose a test statistic and find its sampling distribution under the assumption that H zero is true. The red phrase matters twice on this slide. We are not asking for the statistic's distribution under every possible state of the world. We temporarily impose the null so that we have a reference distribution against which an observed result can be judged. If you do not know that null distribution, you do not yet have a formal test.

In step three, calculate the statistic from your sample and regression output. In step four, choose the significance level, such as five percent. That is the long-run Type I error rate you are willing to tolerate, and it determines a rejection region or critical value. Finally, in step five, ask how unusual the statistic from step three would be in the null distribution from step two. If it is sufficiently extreme, you reject H zero in favor of H one. If it is not, you fail to reject H zero. You do not prove the null true.

</details>

Here is the general step of any hypothesis testing:

+ Step 1: specify the null $(H_0)$ and alternative $(H_1)$ hypotheses

+ Step 2: find the distribution of the test statistic <span style = "color: red;"> if the null hypothesis is true </span> 

+ Step 3: calculate the test statistic based on the data and regression results

+ Step 4: define the significance level

+ Step 5: check how unlikely that you get the actual test statistic (found at Step 3) <span style = "color: red;"> if indeed the null hypothesis is true </span>

### Example

::: {.panel-tabset}

#### Setting

<details class="transcript">
<summary>Transcript</summary>

This is a deliberately artificial example, and the artificiality lets us see the testing logic without having to estimate a variance first. The goal at the top says that x is Normally distributed and that we want to test whether its population expected value is one. I will call the unknown population mean alpha. The phrase "or not" makes this a two-sided question: values of alpha either above or below one would count against the null.

The state of knowledge gives us unusually strong information. We know the population distribution is Normal and we somehow know its variance is four, so its standard deviation is two. In real applications, a population variance is almost never handed to us.

Our estimator is theta, the sample mean. Read the formula as theta equals the sum of x sub i from i equals one through J, divided by J. Here J is the number of observations, x sub i is observation i, alpha is the population mean we want to learn about, and theta is the random estimator computed from whichever sample we happen to draw. From the earlier lecture, you know theta is unbiased, so its expected value is alpha. But one observed sample mean will almost never equal alpha exactly. The testing question is therefore not simply whether theta equals one. It is whether theta is far enough from one, relative to its sampling variation, that the discrepancy would be hard to explain if alpha really were one. Flip to the distribution tab to quantify that sampling variation.

</details>

**Goal**

Suppose you want to test if the expected value of a normally distributed random variable $(x)$ is 1 or not.

<br>

**State of Knowledge**

We do know $x$ follows a normal distribution and its variance is 4 for some reason.

<br>

**Your Estimator**

Your estimator is the sample mean: $\theta = \sum_{i=1}^J x_i/J$

#### distribution of $\theta$

<details class="transcript">
<summary>Transcript</summary>

Follow the two math asides carefully, because together they produce the statistic we will test. Math Aside 1 says that if you multiply a random variable X by a constant a, its variance is multiplied by a squared. This is why standard deviations scale by the absolute value of a. Math Aside 2 says that subtracting a constant shifts a Normal distribution's mean but does not change its variance. Thus, if x follows a Normal distribution with mean a and variance b, x minus a has mean zero and variance b. Dividing that centered variable by root b, its standard deviation, gives a standard Normal variable with mean zero and variance one.

Now apply those facts to theta, the sample mean of J independent observations. Each x sub i follows Normal with mean alpha and variance four. A sum of independent Normal variables is Normal, and dividing the sum by J keeps it Normal. Its mean is alpha. For the variance, theta is one over J times each observation, so each term contributes one over J squared times four. There are J independent terms, so the variances add: J times one over J squared times four equals four over J. That justifies both the distribution shown near the top, theta follows Normal with mean alpha and variance four over J, and the first bullet at the bottom.

The standard deviation of theta is the square root of four over J, which is two over root J. Standardize theta by subtracting its mean alpha and dividing by two over root J. Dividing by a fraction is the same as multiplying by its reciprocal, so the expression becomes root J over two, times theta minus alpha. The last line says this transformed quantity follows the standard Normal distribution. Alpha is still unknown at this stage, so we know the mathematical distribution but cannot yet calculate the statistic from data. The next tab shows how imposing the null supplies the missing value.

</details>


:::{.callout-note title="Math Aside 1"}
$Var(ax) = a^2 Var(X)$
:::

So, we know that $\theta \sim N(\alpha, 4/J)$ (of course $\alpha$ is not known).

<br>

:::{.callout-note title="Math Aside 2"}
If $x \sim N(a, b)$, then, $x-a \sim N(0, b)$ (shift) 
:::

So, $\frac{x-a}{\sqrt{b}} \sim N(0, 1)$ (combined with Math Aside 1)

<br>

**This means,**

Since $\theta = \sum_{i=1}^J x_i/J$ and $x_i \sim N(\alpha, 4)$, 

+ $Var(\theta) = J \times \frac{1}{J^2}Var(x) = 4/J$
+ $\frac{\sqrt{J}}{2} \cdot (\theta - \alpha)\sim N(0, 1)$.


#### distribution of $\theta$ if $H_0$ is true

<details class="transcript">
<summary>Transcript</summary>

Here is where the null hypothesis actually enters the calculation. On the previous tab we established that root J over two, times theta minus alpha, follows a standard Normal distribution. Alpha is the unknown population mean, so that expression cannot be evaluated directly. The null hypothesis supplies a particular value: H zero says alpha equals one. If that claim is true, substitute one for alpha. The computable test statistic is therefore root J times theta minus one, divided by two.

Read the verbal recipe in the left column in exactly that order. Start with the sample mean theta. Subtract the value asserted by the null, which is one. The red phrase is there because students often forget that centering step and test the magnitude of the estimate rather than its distance from the null value. Multiply the resulting gap by the square root of the sample size J, then divide by the known population standard deviation, two. Equivalently, divide theta minus one by the standard error of the sample mean, two over root J. The statistic therefore measures how many standard errors the estimate lies from the null value.

Now look at the figure in the right column. The horizontal axis contains possible values of this standardized statistic, and the vertical height is the standard Normal density. The curve is centered at zero and symmetric. Values near zero are common under H zero, while values far into either tail are rare. Because the goal was to test whether alpha is one "or not," large positive and large negative values would both count as evidence. The curve is not the distribution of the raw x values or of theta itself. It is the distribution of the transformed statistic, conditional on the null being true. The next two tabs put actual sample means into this formula.

</details>

::: {.columns}

::: {.column width="50%"}
We established that $\frac{\sqrt{J}}{2} \cdot (\theta - \alpha)\sim N(0, 1)$.

The null hypothesis is $\alpha = 1$. 

If $\alpha = 1$ is indeed true, then $\sqrt{J} \times (\theta - 1)/2 \sim N(0, 1)$.

In other words, if you take the sample mean, <span style = "color: red;">subtract the value under the null</span>, multiply by the square root of the number of observations, and divide by 2, the result follows the standard Normal distribution like below.
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r}
#| echo: false
x <- seq(-4, 4, length = 1000)
y <- dnorm(x, sd = 1)
plot_data <- data.table(x = x, y = y)
g_n <-
  ggplot(data = plot_data) +
  geom_line(aes(x = x, y = y), color = "black") +
  theme_lecture
g_n
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

#### Case 1

<details class="transcript">
<summary>Transcript</summary>

Now substitute the numbers from Case 1. The sample contains J equals one hundred observations, and its sample mean theta is two. Under H zero, the comparison value is one. Root one hundred is ten, the difference two minus one is one, and dividing ten times one by two gives a test statistic of five. In other words, the observed sample mean is five standard errors above the mean asserted by the null, because the standard error here is two over root one hundred, or zero point two.

Look at the standard Normal curve in the right column. Its horizontal axis shows possible values of the test statistic and its vertical height shows density. The displayed range ends around four, so a value of five is not even visible. It lies extremely far into the positive tail. Under a standard Normal distribution, an observation five standard deviations from zero is extraordinarily rare. For this two-sided question, we would count an equally extreme negative result as well.

The logic is a comparison between two explanations. One possibility is that alpha truly equals one and our sample happened to produce an extraordinarily unusual statistic. The other is that alpha does not equal one, so centering the statistic at one was inappropriate and the extreme value reflects a false null. Hypothesis testing controls how unusual a result must be before we choose the second interpretation and reject. This case clearly provides strong evidence against H zero, but we have not yet established a formal cutoff.

</details>

::: {.columns}

::: {.column width="50%"}
Suppose you have obtained 100 samples $(J = 100)$ and calculated $\theta$ (sample mean), which turned out to be 2.

Then, your test statistic is $\sqrt{100} \times (2-1)/2 = 5$.

How unlikely is it to get the number you got (5) if the null hypothesis is indeed true?

:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r}
#| echo: false
g_n
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


#### Case 2

<details class="transcript">
<summary>Transcript</summary>

Case 2 uses exactly the same statistic and null distribution, but different data. Here J is four hundred and the sample mean theta is one point zero two. Root four hundred is twenty. The sample mean exceeds the null value by only zero point zero two. Multiply twenty by zero point zero two and divide by two, and the test statistic is zero point two. Equivalently, the sample mean's standard error is two over root four hundred, or zero point one, so the observed mean lies zero point two standard errors above one.

Find zero point two on the horizontal axis of the standard Normal curve. It sits very close to zero, in the high, central part of the density. Values around zero point two, and values even farther from zero, occur routinely when the null is true. This sample therefore gives us no compelling reason to doubt alpha equals one. The correct conclusion is to fail to reject the null, not to prove or accept it. A different sample could lead to a different estimate and decision.

Compare this tab with Case 1. The sample here is four times as large, which cuts the standard error in half, but the raw discrepancy from the null is only zero point zero two rather than one. A larger sample makes a given raw gap more statistically noticeable, yet it cannot manufacture evidence when the observed gap is tiny. The test statistic combines both ingredients: the size of the estimate's departure from the null and the precision with which that departure was estimated.

</details>

::: {.columns}

::: {.column width="50%"}
Suppose you have obtained 400 samples $(J = 400)$ and calculated $\theta$ (sample mean), which turned out to be 1.02.

Then, your test statistic is $\sqrt{400} \times (1.02-1)/2 = 0.2$.

How unlikely is it to get the number you got (0.2) if the null hypothesis is indeed true?
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r}
#| echo: false
g_n
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


:::
<!--end of panel-->

### By the way

<details class="transcript inline">
<summary>Transcript</summary>

This aside explains why we standardize even though standardization is not logically required. In the left column, the alternative statistic is simply theta minus alpha, the sample mean's raw estimation error. Because theta follows Normal with mean alpha and variance four over J, subtracting alpha gives a Normal distribution with mean zero and variance four over J. That is a perfectly valid null reference distribution. The complication is that it changes with J.

The figure on the right uses J equals one hundred. The horizontal axis is the raw difference theta minus alpha, and the vertical height is the density of a Normal distribution centered at zero with variance four over one hundred, which is zero point zero four, and standard deviation zero point two. That is why the curve is much narrower and taller than the standard Normal curve. Most raw differences should fall close to zero when the null is true.

Now reconsider Case 1 below the figure. Theta was two and the null value of alpha was one, so the raw statistic is two minus one, or one. On this distribution, one is five standard deviations above zero because one divided by zero point two equals five. That is exactly the standardized statistic we calculated before. The conclusion cannot depend on whether we measure the gap in original units or standard-error units, as long as we compare it with the matching distribution.

The callout gives the practical reason for normalization. Root J times theta minus alpha, divided by two, always follows the same standard Normal distribution, regardless of sample size. The unnormalized theta minus alpha requires a different variance, four over J, for every J. Standardization lets you reuse one reference distribution and its quantiles.

</details>

::: {.columns}

::: {.column width="40%"}
Note that you do not really need to use $\sqrt{J} \times (\theta - \alpha)/2$ as your test statistic. 

You could alternatively use $\theta - \alpha$. But, in that case, you need to be looking at $N(0, 4/J)$ instead of $N(0, 1)$ to see how unlikely you get the number you got. 

For example, when the number of observations is 100 $(J = 100)$, the distribution of $\theta-\alpha$ looks like the figure on the right.
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r}
#| echo: false
x <- seq(-4, 4, length = 1000)
y <- dnorm(x, sd = 2 / 10)
plot_data <- data.table(x = x, y = y)
ggplot(data = plot_data) +
  geom_line(aes(x = x, y = y), color = "black") +
  theme_lecture
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


**Reconsider the case 1**

Suppose you have obtained 100 samples $(J = 100)$ and calculated $\theta$ (sample mean), which turned out to be 2.

Then, your test statistic is $2-1 = 1$. 

Is it unlikely for you to get 1 if the null hypothesis is true? 

The conclusion would be exactly the same as using $\sqrt{J} \times (\theta - \alpha)/2$ because the distribution under the null is adjusted according to the test statistic you use.

:::{.callout-note}
We always use a normalized test statistic so that we can always look up the same distribution.
:::

### Summary

<details class="transcript">
<summary>Transcript</summary>

This summary separates what a hypothesis test requires from what we usually do to meet that requirement. First, we need a test statistic whose probability distribution is known when the null hypothesis is true. A test statistic is a random quantity before the sample is drawn and a number after we insert our data. The null distribution tells us which realized numbers are ordinary and which are unusual under H zero. The examples listed here, a t distribution and a Normal distribution, are reference distributions we can use to assign tail probabilities or find critical values.

Second, we often have to transform the raw random variable we care about. In the example, the raw object was theta, the sample mean. Its distribution depended on the unknown alpha and on the sample size. We centered it by subtracting the null value and scaled it by its standard error. Under H zero, that transformed statistic followed the standard Normal distribution.

Once the statistic has a known null distribution, compute its realized value and ask whether a draw at least that extreme is plausible under H zero. The parenthetical note in the last bullet identifies the missing piece. We have not formally defined how small a probability must be before we reject.

</details>

**What do we need?**

+ test-statistic of which we know the distribution (e.g., t-distribution, Normal distribution) assuming the null hypothesis

<br>

**What do we (often) do?**

+ transform (most of the time) a raw random variable (e.g., sample mean in the example above) into a test statistic of which we know the distribution assuming that the null hypothesis is true
  - e.g., we transformed the sample mean so that it follows the standard Normal distribution.
+ check if the actual number you got from the test statistic is likely to happen or not (formal criteria has not been discussed yet)

### Exercise

::: {.panel-tabset}

#### Problem

<details class="transcript">
<summary>Transcript</summary>

Your turn. This exercise has the same structure as the worked example, but the numbers change. Let x denote the annual-salary difference between graduates of University A and University B. The population model says x follows a Normal distribution with unknown mean alpha and known variance sixteen, so its population standard deviation is four. The substantive question is whether the average salary difference is ten, which makes the null H zero: alpha equals ten. Because the wording asks whether it is ten rather than whether it is greater or smaller, treat departures on either side as relevant.

For part two, let theta be the sample mean of four hundred independent observations. Its mean is alpha and its variance is the individual variance divided by the sample size: sixteen over four hundred, not sixteen. Its standard deviation, or standard error, is root sixteen over four hundred, which is four over root four hundred, or zero point two.

For part three, turn the sample mean into standard-error units. Subtract alpha and divide by four over root four hundred. That is the same as multiplying theta minus alpha by root four hundred over four. Under the Normal model, the result follows a standard Normal distribution. Under the null, replace alpha with ten so the statistic is computable.

Part four gives an observed sample mean of ten point two. Convert that to the standardized statistic, then use `pnorm` to find an upper-tail probability. Be precise about the question: it asks for the probability of observing a number greater than ten point two, not the two-sided probability of being at least zero point two away from ten. Try all four parts before opening the Answer tab. Finally, read the callout. In real work, the variance is not magically known, and we still need a formal rule for rejecting or failing to reject.

</details>

You have collected data on annual salary for those who graduated from University A and B. You are interested in testing whether the difference in annual salary between the universities (call it $x$) is 10 on average. You know (for unknown reasons) that the difference is distributed as $N(\alpha, 16)$. 

1. What is the null hypothesis?
2. Under the null hypothesis, what is the distribution of the sample mean when the number of observation is 400?
3. Normalize the test statistic so that the transformed version follows $N(0, 1)$.
4. The actual difference you observed is 10.2. What is the probability that you observe a number greater than 10.2 if the null hypothesis is true? Use `pnorm()`.


:::{.callout-note}
In reality,

+ we need to find out what the distribution of the test statistic is
+ we need to formally define when we reject or fail to reject the null hypothesis
:::

#### Answer

<details class="transcript">
<summary>Transcript</summary>

Check each line against your solution. Line one states the null hypothesis: alpha equals ten. An alternative is not printed in the answer, but the original goal of testing whether the average is ten would normally pair it with alpha not equal to ten.

Line two gives theta's sampling distribution as Normal with mean alpha and variance sixteen over four hundred. The variance comes from adding the variances of four hundred independent observations and then scaling their sum by one over four hundred. Numerically, sixteen over four hundred is zero point zero four, so the standard error is its square root, zero point two.

Line three standardizes theta. Read root four hundred over sixteen, times theta minus alpha. Root of the fraction is root four hundred divided by root sixteen, which is twenty divided by four, or five. That multiplier is the reciprocal of the standard error. After centering and scaling, the statistic follows the standard Normal distribution. Under H zero, alpha is replaced by ten.

Line four inserts the observed sample mean ten point two. Five times ten point two minus ten equals five times zero point two, which is one. The code cell calculates one minus `pnorm` of one. `pnorm(1)` is the standard Normal probability to the left of one, so subtracting from one gives the probability to the right. Running it returns about zero point one five nine, or fifteen point nine percent. Thus, if alpha truly equals ten, a sample mean greater than ten point two occurs about sixteen percent of the time. That is roughly one sample in six, so the upper-tail result is not unusual. The exercise explicitly asked only for the probability above ten point two, which is why the displayed code uses one tail.

</details>

1. $\alpha = 10$
2. $\theta \sim N(\alpha, 16/400)$
3. $\sqrt{\frac{400}{16}}\cdot (\theta - \alpha)$
4. The test statistic is $5 \times (10.2 - 10) = 1$

```{webr-r}
1 - pnorm(1)
```

:::
<!--end of panel-->

:::
<!--end of panel-->

# Hypothesis Testing on the Coefficients

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Hypothesis Testing of the Coefficients

::: {.panel-tabset}

### Example

<details class="transcript">
<summary>Transcript</summary>

Now apply the same testing logic to regression coefficients. The model at the top says wage equals beta zero plus beta one times education plus beta two times experience plus the error u. Beta zero is the intercept. Beta one is the population partial effect of one additional unit of education on wage, holding experience fixed, and beta two is the corresponding partial effect of experience, holding education fixed.

The callout gives two example substantive claims. "Education has no impact on wage" translates to beta one equals zero. A usual two-sided alternative would be beta one not equal to zero, because either a positive or a negative effect would contradict no effect. "Experience has a positive impact on wage" is directional. It suggests an alternative beta two greater than zero, paired with a null boundary such as beta two equals zero.

The three bullets explain why looking at the fitted coefficient is not enough. If beta one hat were a fixed scalar that exactly revealed beta one, we could simply check whether it equaled zero. But beta one hat is an estimator calculated from random sample data. Before sampling, it has a distribution; after sampling, the coefficient in your output is just one realization from that distribution, as the blue phrase emphasizes. Repeating the study with different observations would produce different estimates.

Consequently, even when the population coefficient beta one is exactly zero, sampling noise can produce an estimate away from zero, occasionally quite far away. Conversely, a nonzero population effect can yield an estimate close to zero. A valid decision must compare the observed distance from the null with the amount of sampling variation, which means we need the sampling distribution and standard error of beta one hat. The next tabs establish that distribution under the regression assumptions.

</details>

Consider the following model,

$$wage = \beta_0 + \beta_1 educ + \beta_2 exper + u$$

:::{.callout-note title="Example Hypotheses"}
+ education has no impact on wage $(\beta_1=0)$
+ experience has a positive impact on wage $(\beta_2>0)$
:::

<br>

+ If $\widehat{\beta}_1$ is non-random, but just a scalar, all you have to do is just check if $\widehat{\beta}_1=0$ or not

+ But, the estimate you get is <span style = "color: blue;"> just one realization </span> of the range of values $\widehat{\beta}_1$ could take because it is a random variable

+ This means that even if $\beta_1=0$ in the population, it is possible to get an estimate that is very far from 0

### Distribution of $\beta_j$

::: {.panel-tabset}

#### Assumption

<details class="transcript">
<summary>Transcript</summary>

This tab takes stock of what the first five multiple linear regression assumptions have given us and what they have not. Under MLR one through MLR four, the OLS coefficient estimators are unbiased. That means the expected value of beta j hat, across repeated samples generated under those assumptions, equals the true beta j. Unbiasedness tells us the center of the sampling distribution, but not its shape or how much probability lies in its tails.

Adding MLR five, homoskedasticity, gives the variance formula shown in the second bullet. Read it as sigma squared divided by SST sub x times one minus R squared sub j. Sigma squared is the error variance. SST sub x measures the sample variation in explanatory variable x sub j, and R squared sub j comes from regressing x sub j on the other covariates. More variation in x sub j reduces the estimator's variance, while greater collinearity makes R squared sub j larger, shrinks one minus R squared sub j, and increases the variance. This formula tells us the spread, but again not the full distribution.

The blue "NOT" marks the missing ingredient. MLR one through five do not specify the shape of the error term's distribution. Knowing only a random variable's mean and variance is not enough to calculate an exact critical value or p-value in a finite sample, because many different distributions share those two moments but have different tails. For exact small-sample hypothesis tests, we therefore add a distributional assumption about u. The parenthetical warning is important: Normal errors are not strictly necessary for large-sample inference, because asymptotic results can make standardized estimators approximately Normal under weaker conditions. That comes later. For the finite-sample results in this part of the course, move to the next tab and add MLR six.

</details>

So far, we learned that:

+ OLS estimators are unbiased under MLR.1 ~ MLR.4
+ Variance of the OLS estimator of $\beta_j$ is $\frac{\sigma^2}{SST_x\cdot (1-R^2_j)}$ under MLR.1 ~ MLR.5

We have <span style = "color: blue;"> NOT </span> made any assumptions about the distribution of the error term!!

In order to perform hypothesis testing, we need to make assumptions about the distribution of error term (this is not strictly true, but more on this later)

#### Additional assumption

<details class="transcript">
<summary>Transcript</summary>

This is the additional finite-sample assumption, often called MLR six, and both red words matter. The first callout says the population error u is independent of all explanatory variables x one through x k and is Normally distributed with mean zero and variance sigma squared. Read the notation as u follows Normal, zero, sigma squared. Zero is the unconditional mean and sigma squared is the population error variance.

Normality supplies a specific distributional shape. Independence supplies much more than lack of correlation. If u is independent of x, learning the values of the regressors does not change the distribution of u. Its conditional mean remains zero for every x, so expected u given x equals zero. Its conditional variance also remains sigma squared for every x, so variance of u given x equals sigma squared. These are exactly the zero conditional mean and homoskedasticity properties listed in the second callout.

That is why the slide warns that the "normality assumption" is more than merely drawing a bell-shaped marginal distribution for the errors. A marginally Normal u could still depend on x in ways that violate the conditional assumptions needed for OLS inference. The combination stated here includes independence, zero conditional mean, constant conditional variance, and a Normal conditional shape. As a result, MLR four and MLR five are automatically satisfied under this stronger assumption.

</details>

For the purpose of hypothesis testing, we will make the following assumption:

:::{.callout-note title="Normality of the error term"}
The population error $u$ is <span style = "color: red;"> independent </span> of the explanatory variables $x_1,\dots,x_k$ and is <span style = "color: red;"> normally </span> distributed with zero mean and variance $\sigma^2$:

$$u\sim N(0,\sigma^2)$$
:::


:::{.callout-note}
The normality assumption is much more than error term being distributed as Normal.

Independence of the error term implies

+ $E[u|x] = 0$
+ $Var[u|x]= \sigma^2$

So, we are necessarily assuming MLR.4 and MLR.5 hold by the independence assumption.
:::

#### Implications of the Assumptions

<details class="transcript">
<summary>Transcript</summary>

The two columns show the distributional consequences of MLR one through six. Start on the left with the regression equation. Conditional on the explanatory variables, beta zero plus beta one x one through beta k x k is fixed. All remaining randomness in y comes from u. Because u given x is Normal with mean zero and variance sigma squared, adding the fixed regression function shifts the mean without changing the variance. Therefore y given x follows a Normal distribution with mean beta zero plus beta one x one plus the remaining partial-effect terms, and variance sigma squared. The two bullets underneath identify those pieces: the conditional expectation is the regression function, while the conditional error distribution is Normal, zero, sigma squared.

The right column gives the result needed for testing. Each OLS coefficient estimator is a linear combination of the observed y values once the x values are held fixed. A linear combination of jointly Normal outcomes is Normal, so beta j hat is Normally distributed. Its mean is the true coefficient beta j because OLS is unbiased, and its variance is the OLS variance derived under homoskedasticity. Thus beta j hat follows Normal with mean beta j and variance of beta j hat.

Standardize this estimator exactly as we standardized the sample mean earlier. Subtract its mean beta j and divide by its true standard deviation, se of beta j hat, which is the square root of variance of beta j hat. The result follows the standard Normal distribution. Every symbol has a role: beta j hat is the sample estimate, beta j is the population value, and the denominator measures the estimator's sampling uncertainty.

Now read the question callout. Are we going to use this standard Normal statistic directly for testing beta j? Flip to the t-test tab for the answer.

</details>

::: {.columns}

::: {.column width="50%"}
**distribution of the dependent variable**

The distribution of $y$ conditional on $x$ is a Normal distribution

$y|x \sim N(\beta_0+\beta_1 x_1+\dots+\beta_k x_k,\sigma^2)$

+ $E[y|x]$ is $\beta_0+\beta_1 x_1+\dots+\beta_k x_k$
+ $u|x$ is $N(0,\sigma^2)$
:::
<!--end of the 1st column-->
::: {.column width="50%"}
**distribution of the OLS estimator**

If the MLR.1 through MLR.6 are satisfied, $OLS$ estimators are also Normally distributed!

$\widehat{\beta}_j \sim N(\beta_j,Var(\widehat{\beta}_j))$

which means,

$\frac{\widehat{\beta}_j-\beta_j}{se(\widehat{\beta}_j)} \sim N(0,1)$
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


:::{.callout-note title="Question"}
Okay, so are we going to use this for testing involving $\beta_j$?
:::


:::
<!--end of panel-->

### t-test

<details class="transcript">
<summary>Transcript</summary>

The catch is that the standard Normal result on the previous tab used the true standard error of beta j hat. That quantity depends on sigma squared, the population variance of the errors, and sigma squared is unknown. In practice we estimate it from the OLS residuals and use the estimated standard error, written with a hat over se of beta j hat.

Once the denominator is estimated, the standardized expression no longer has an exact standard Normal distribution. Under MLR one through six, beta j hat minus beta j, divided by the estimated standard error of beta j hat, follows a t distribution with n minus k minus one degrees of freedom. Here n is the number of observations, k is the number of explanatory variables excluding the intercept, and the additional one accounts for estimating the intercept. The same n minus k minus one is the residual degrees of freedom used to estimate sigma squared.

The t distribution is symmetric and centered at zero, like the standard Normal, but it has heavier tails when the degrees of freedom are small. Those heavier tails account for the extra uncertainty introduced by estimating the denominator. As the residual degrees of freedom increase, the estimate of sigma becomes more precise and the t distribution approaches the standard Normal.

Read the parenthetical formula carefully. The estimated standard error is the square root of estimated sigma squared divided by SST sub X times one minus R squared sub j. The red phrase "square root" prevents a common mistake: the fraction itself is an estimated variance, while a t-statistic divides by a standard error, which is measured in the same units as the coefficient. The numerator subtracts the true value beta j. When testing a specific null, replace beta j there with the value asserted by H zero. This t-statistic is the basic object for the next section.

</details>

In practice, we need to estimate $se(\beta_j)$. If we use $\widehat{se(\widehat{\beta}_j)}$ instead of $se(\widehat{\beta}_j)$, then,

$\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}} \sim t_{n-k-1}$

where $n-k-1$ is the degree of freedom of residual.

(Note: $\widehat{se(\widehat{\beta}_j)} = \sqrt{\widehat{\sigma}^2/\big[SST_X\cdot (1-R_j^2)\big]}$ &mdash; the standard error is the <span style = "color: red;"> square root </span> of the estimated variance)

### Recap on notations

<details class="transcript">
<summary>Transcript</summary>

Use this tab as a notation reference. Start with k, the number of explanatory variables excluding the intercept. Therefore a model with education, experience, and tenure has k equals three even though it estimates four coefficients. That convention is why residual degrees of freedom are n minus k minus one.

Sigma squared is the true population variance of the error u. It is a fixed but unknown parameter. Sigma hat squared is its estimator, computed as the sum of squared OLS residuals divided by n minus k minus one. Each u hat sub i is an estimated residual, the observed y sub i minus its fitted value. Dividing by residual degrees of freedom, instead of n, accounts for the coefficients estimated before the residuals were formed.

SST sub X is the total variation in the explanatory variable, the sum of x sub i minus x bar, squared. Beta j hat is the OLS estimator for the coefficient on x sub j. The simple-regression formula shown underneath, the sum of centered x times centered y divided by SST sub X, applies when there is one explanatory variable. In multiple regression the matrix formula differs, but beta j hat retains the same interpretation as the estimated partial effect.

Variance of beta j hat, without a hat over `var`, is the true sampling variance of the estimator. R squared sub j comes from the auxiliary regression of x sub j on all the other explanatory variables. It measures how well the other covariates explain x sub j and therefore enters the multicollinearity adjustment one minus R squared sub j. The estimated variance of beta j hat replaces the unknown sigma squared in the true variance formula with sigma hat squared.

Finally, keep variance and standard error distinct. Se of beta j hat is the square root of the true variance. Estimated se of beta j hat is the square root of the estimated variance, and this is the number printed in a regression table. Hats distinguish unknown population quantities from sample estimates.

</details>

+ $k$: the number of explanatory variables included except the intercept
+ $\sigma^2$: the <span style = "color: blue;"> true </span> variance of the error term
+ $\widehat{\sigma^2}$: the estimator (estimate) of the variance of the error term
  - $\widehat{\sigma^2} = \frac{\sum \widehat{u}_i^2}{n-k-1}$
+ $SST_X = \sum (x_i - \bar{x})^2$ 
+ $\widehat{\beta}_j$: OLS estimator (estimate) on explanatory variable $x_j$
  - $\widehat{\beta} = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{SST_X}$ (for simple univariate regression)
+ $var(\widehat{\beta}_j)$: the <span style = "color: blue;"> true </span> variance of $\widehat{\beta}_j$
+ $R^2_j$: $R^2$ when you regress $x_j$ on all the other covariates (mathematical expression omitted)
+ $\widehat{var(\widehat{\beta}_j)}$: Estimator (estimate) of $var(\widehat{\beta}_j)$
  - $\widehat{var(\widehat{\beta}_j)} = \frac{\widehat{\sigma^2}}{SST_X\cdot(1-R^2_j)}$
+ $se(\widehat{\beta}_j)$: square root of $var(\widehat{\beta}_j)$
+ $\widehat{se(\widehat{\beta}_j)}$: square root of $\widehat{var(\widehat{\beta}_j)}$

:::
<!--end of panel-->  

## Details of hypothesis testing on $\beta_j$

::: {.panel-tabset}

### Null and alternative hypothesis

::: {.panel-tabset}

#### Introduction

<details class="transcript">
<summary>Transcript</summary>

The courtroom analogy is useful because it makes the asymmetry between the two hypotheses explicit. Imagine that you are an attorney arguing that a defendant committed a crime. The null hypothesis corresponds to "the defendant is not guilty," the position that remains unless the evidence crosses a prespecified threshold. The alternative corresponds to "the defendant is guilty," the claim for which you are presenting evidence.

In a statistical application, the null is often a benchmark such as beta j equals zero, meaning no partial effect. The alternative states what would contradict that benchmark, such as beta j not equal to zero or beta j greater than zero. We compute how compatible the observed evidence is with the null. If the evidence would be sufficiently rare under H zero, we reject H zero in favor of H one.

Notice what the procedure does not say. A court that does not convict has not proved innocence; it has concluded that the evidence did not satisfy the burden of proof. Likewise, a statistical test that does not reject H zero has not proved the null. The proper phrase is "fail to reject," not "accept."

</details>

Statistical hypothesis testing involves two hypotheses: Null and Alternative hypotheses. 

Pretend that you are an attorney who indicted a defendant who you think committed a crime. 

<br>  

**Null Hypothesis**

Hypothesis that you would like to reject (defendant is not guilty)

<br>

**Alternative Hypothesis**

Hypothesis you are in support of (defendant is guilty) 

#### One-sided and Two-sided Alternatives Hypotheses

<details class="transcript">
<summary>Transcript</summary>

This tab contrasts two shapes of alternative hypothesis. In the one-sided example, H zero says beta j equals zero and H one says beta j is greater than zero. The alternative is directional, so only large positive t-statistics support it. A large negative statistic may contradict equality, but it points in the direction opposite to the stated alternative and therefore is not in this upper-tail rejection region. If the alternative were beta j less than zero, we would instead examine the lower tail.

In the two-sided example, H zero again says beta j equals zero, but H one says beta j does not equal zero. Now sufficiently large departures in either direction count as evidence. A positive t-statistic indicates an estimate above zero and a negative statistic indicates one below zero, so the rejection region occupies both tails of the t distribution.

Read the callout at the bottom. A critical value is simply a quantile of the null distribution. In Lecture 03, `qnorm(0.95)` returned about one point six four, the ninety-fifth percentile of the standard Normal. Here we use `qt` instead of `qnorm` because estimating sigma puts the statistic on a t distribution, and we also supply its degrees of freedom.

</details>

**one-sided alternative**

$H_0:$ $\beta_j = 0$ 
$H_1:$ $\beta_j > 0$ 

You look at the positive end of the t-distribution to see if the t-statistic you obtained is more extreme than the level of error you accept (significance level). 

<br>

**two-sided alternative**

$H_0:$ $\beta_j = 0$ 
$H_1:$ $\beta_j \ne 0$ 

You look at both ends of the t-distribution to see if the t-statistic you obtained is more extreme than the level of error you accept (significance level).

<br>

:::{.callout-note title="This is where lecture 03 pays off"}
The numbers we are about to use as critical values are just <span style = "color: blue;">quantiles</span>, computed exactly the way we computed `qnorm(0.95) = 1.64` in the Monte Carlo lecture. The only change is which distribution you ask: `qt()` instead of `qnorm()`, because we had to estimate $\sigma$ rather than know it.
:::

:::
<!--end of panel-->

### Significance level

```{r include = F}
u <- seq(-3.5, 3.5, length = 1000)
df_fig <- 522 # n - k - 1 for the wage regression used throughout this lecture
pdf <- dt(u, df = df_fig) # default is mean=0,sd=1
data <- data.table(x = u, y = pdf)

g_t <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme_lecture +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.05
c_rej <- qt(1 - alpha, df = df_fig)
rej_data <- data[u >= c_rej, ]

g_t_005_one <- g_t +
  geom_ribbon(
    data = rej_data,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.03,
    label = as.character(round(c_rej, digits = 2)),
    size = 3.7,
    family = "Times"
  ) +
  annotate("text",
    x = 2.0, y = 0.30,
    label = paste("area  =  ", round(alpha, digits = 2), sep = ""),
    size = 3.7,
    family = "Times"
  )


alpha <- 0.01
c_rej <- qt(1 - alpha, df = df_fig)
rej_data_001 <- data[u >= c_rej, ]

g_t_001_one <- g_t +
  geom_ribbon(
    data = rej_data_001,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.03,
    label = as.character(round(c_rej, digits = 2)),
    size = 3.7,
    family = "Times"
  ) +
  annotate("text",
    x = 2.0, y = 0.30,
    label = paste("area  =  ", round(alpha, digits = 2), sep = ""),
    size = 3.7,
    family = "Times"
  )
```


::: {.panel-tabset}

#### Definition

<details class="transcript">
<summary>Transcript</summary>

The significance level formalizes how unlikely a result must be before we reject. The definition callout says it is the probability of rejecting H zero when H zero is actually true. That is a Type I error, analogous to convicting an innocent defendant. If you choose a five percent level and repeatedly apply the procedure in situations where the null is true, the procedure is designed to reject about five percent of the time in the long run.

A lower level requires stronger evidence. Moving from five percent to one percent pushes the critical value farther into the tail, so fewer test statistics enter the rejection region when H zero is true. This reduces the long-run Type I error rate. It does not mean there is a one or five percent probability that the null is true. Frequentist significance treats the parameter and hypothesis as fixed and describes the behavior of the testing procedure across hypothetical repeated samples.

Now read the table by rows and columns. If H zero is true and you reject it, you make a Type I error. If it is true and you do not reject it, the decision is correct. If H zero is false and you reject it, you correctly detect a real departure. If it is false and you do not reject it, you make a Type II error.

The bullets explain the trade-off. Holding the design and sample size fixed, making rejection harder lowers Type I error but generally increases Type II error, so more real effects go undetected. One minus the probability of Type II error is power, the chance that the test rejects when a specified real effect exists. This is why "not statistically significant" does not mean "no effect." It may mean that the study had too little power to distinguish the effect from sampling noise. The next tabs show how a chosen level marks a rejection area on the null distribution.

</details>

:::{.callout-note title="Definition"}
The probability of rejecting the null when the null is actually true (the probability that you wrongly claim that the null hypothesis is wrong even though it is true in reality: <span style = "color: red;">Type I error</span>)
:::

A lower significance level requires stronger evidence before rejecting the null and reduces the long-run probability of a Type I error. It does not tell us the probability that the null hypothesis is true.

<br>

**But lowering it is not free**

There are two ways to be wrong, and they trade off against each other:

|  | $H_0$ is true | $H_0$ is false |
|---|---|---|
| **You reject $H_0$** | <span style = "color: red;"> Type I error </span> (probability = significance level) | correct |
| **You do not reject $H_0$** | correct | <span style = "color: blue;"> Type II error </span> |

+ the significance level is a choice about how much <span style = "color: red;"> Type I </span> error you will tolerate
+ pushing it down from 5% to 1% moves the critical value further out, so you reject less often &mdash; which means more <span style = "color: blue;"> Type II </span> error: real effects you fail to detect
+ $1 - P(\text{Type II error})$ is called the <span style = "color: blue;"> power </span> of the test: the chance of detecting an effect that is genuinely there
+ this is why "not statistically significant" does <span style = "color: red;"> not </span> mean "no effect". It may just mean your test had too little power to see it

#### One-sided: 5% significance

<details class="transcript">
<summary>Transcript</summary>

This picture implements a one-sided upper-tail test at the five percent significance level. The quantity in the first bullet is beta j hat minus beta j, divided by its estimated standard error. Under the null beta j equals zero, so the numerator becomes beta j hat and, under the classical assumptions, the statistic follows the t distribution shown on the right with five hundred and twenty-two residual degrees of freedom.

On the figure, the horizontal axis contains possible t-statistics and the vertical height is the t density. The curve is centered at zero and symmetric, although this test uses only its positive tail because H one is beta j greater than zero. The red ribbon is the upper five percent of the total area. Its left edge, about one point six five for this many degrees of freedom, is the ninety-fifth percentile of the t distribution. That boundary is the critical value.

The decision callout says to reject H zero if the observed t-statistic is greater than that critical value. If H zero is true, only five percent of repeated samples generate a statistic in the shaded region. Therefore the rule has a five percent Type I error probability. A statistic in the red region can still occur under a true null, so rejection is not certainty. It means the result crossed the evidence threshold we chose.

Keep the two red terms below the figure distinct. Five percent is a probability and is the significance level, chosen before the test. About one point six five is a value on the t-statistic scale and is the critical value implied by that choice and by the degrees of freedom. With different degrees of freedom, the same five percent level produces a different critical value.

</details>

::: {.columns}

::: {.column width="40%"}
+ Figure on the right presents the distribution of $\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}}$ if $\beta_j = 0$ (the null hypothesis is true).

+ The probability that you get a value larger than `r round(qt(0.95, df_fig), 2)` is 5% (0.05 in area).
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r echo = F}
g_t_005_one
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

:::{.callout-note title="Decision Rule"}
Reject the null hypothesis if the t-statistic is greater than `r round(qt(0.95, df_fig), 2)` (95% quantile of the t-distribution)
:::

If you follow the decision rule, then you have a 5% chance that you are wrong in rejecting the null hypothesis of $\beta_j = 0$. Here, 

+ 5% is the <span style = "color: red;"> significance level</span>
+ `r round(qt(0.95, df_fig), 2)` is the <span style = "color: red;"> critical value</span> above which you will reject the null

#### One-sided: 1% significance

<details class="transcript">
<summary>Transcript</summary>

This tab keeps the same one-sided hypothesis and the same t distribution but changes the significance level from five percent to one percent. The statistic is still beta j hat minus beta j, divided by its estimated standard error, and under H zero beta j equals zero. The curve on the right therefore has exactly the same shape as on the previous tab.

What changes is the shaded rejection region. The red area now contains only the upper one percent of the distribution. Its left boundary is the ninety-ninth percentile of t with five hundred and twenty-two degrees of freedom, about two point three three. The decision rule is consequently stricter: reject only if the observed t-statistic exceeds about two point three three. Compare that with the roughly one point six five boundary at the five percent level.

If H zero is true, this rule rejects only one percent of the time in repeated samples, so its Type I error probability is one percent.

Again, distinguish the level from the critical value. One percent is the chosen tail probability. About two point three three is the t-statistic cutoff created by that probability and these degrees of freedom. A result between one point six five and two point three three would reject at five percent but fail to reject at one percent. That is not a contradiction.

</details>

::: {.columns}

::: {.column width="40%"}
+ The figure on the right presents the distribution of $\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}}$ if $\beta_j = 0$ (the null hypothesis is true).
+ The probability that you get a value larger than `r round(qt(0.99, df_fig), 2)` is 1% (0.01 in area).
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r echo = F}
g_t_001_one
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

:::{.callout-note title="Decision Rule"}
Reject the null hypothesis if the t-statistic is greater than `r round(qt(0.99, df_fig), 2)` (99% quantile of the t-distribution)
:::

If you follow the decision rule, then you have a 1% chance that you are wrong in rejecting the null hypothesis of $\beta_j = 0$. Here, 

+ 1% is the <span style = "color: red;"> significance level </span>
+ `r round(qt(0.99, df_fig), 2)` is the <span style = "color: red;"> critical value </span> above which you will reject the null

#### Two-sided: 5% significance

<details class="transcript">
<summary>Transcript</summary>

Now the alternative is two-sided, so both positive and negative departures from zero count against H zero. The statistic and its null distribution have not changed: beta j hat minus beta j, divided by the estimated standard error, follows a t distribution with five hundred and twenty-two degrees of freedom under beta j equals zero. The parenthetical note emphasizes that every figure in this sequence uses the same t curve. The hypothesis changes the rejection region, not the statistic's null distribution.

The total significance level is still five percent. Because evidence in either direction matters, split that probability evenly: two point five percent in the left tail and two point five percent in the right tail. The positive boundary is the ninety-seven point five percentile of the t distribution, about one point nine six, and symmetry puts the negative boundary at about minus one point nine six. Those are farther from zero in absolute value than the one-sided five percent boundary because each tail receives less area.

Look at the figure. The horizontal axis is the t-statistic, the vertical height is density, and both extreme ends are shaded red. The unshaded center contains ninety-five percent of the null distribution. The annotation in the middle identifies the tail allocation, while the number near the right edge marks the positive critical value. The left edge is its negative counterpart even if only one label is printed.

The decision callout expresses both boundaries compactly: reject if the absolute value of the observed t-statistic is greater than about one point nine six. Taking the absolute value treats equally strong positive and negative evidence the same way. Under a true null, the probability of landing in either shaded region is five percent in total, which is the Type I error rate. Five percent is the significance level; one point nine six is the absolute critical value for these degrees of freedom. Do not halve the critical value or compare a signed statistic only with the positive boundary without taking its absolute value.

</details>

```{r t-005-two, include = F}
u <- seq(-3, 3, length = 1000)
pdf <- dt(u, df = df_fig) # default is mean = 0,sd = 1
data <- data.table(x = u, y = pdf)

alpha <- 0.05

c_rej <- qt(1 - alpha / 2, df = df_fig)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_t_005_two <- g_t +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.03,
    label = as.character(round(c_rej, digits = 2)),
    size = 3.7,
    family = "Times"
  ) +
  annotate("text",
    x = 0, y = 0.1,
    label = paste("area  = ", round(alpha / 2, digits = 3), sep = ""),
    size = 3.7,
    family = "Times"
  )
```

::: {.columns}

::: {.column width="40%"}
+ The figure on the right presents the distribution of $\frac{\widehat{\beta}_j-\beta_j}{\widehat{se(\widehat{\beta}_j)}}$ if $\beta_j = 0$ (the null hypothesis is true).

+ The probability that you get a value more extreme than `r round(qt(0.975, df_fig), 2)` or `r -round(qt(0.975, df_fig), 2)` is 5% (0.05 in area combining the two areas at the edges).

(Note: irrespective of the type of test, the distribution of the t-statistic is the same &mdash; every figure on this slide is the <span style = "color: red;"> same </span> $t_{`r df_fig`}$ curve. Only the shaded region changes.)
:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r echo = F}
g_t_005_two
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->  

:::{.callout-note title="Decision Rule"}
Reject the null hypothesis if the absolute value of the t-statistic is greater than `r round(qt(0.975, df_fig), 2)`.
:::

If you follow the decision rule, then you have a 5% chance that you are wrong in rejecting the null hypothesis of $\beta_j = 0$. Here,

+ 5% is the <span style = "color: red;"> significance level </span>
+ `r round(qt(0.975, df_fig), 2)` is the <span style = "color: red;"> critical value </span> beyond which you will reject the null

:::
<!--end of panel-->

### p-value

<details class="transcript inline">
<summary>Transcript</summary>

The p-value packages the same tail comparison into a single number. The definition callout gives two equivalent readings. It is the smallest significance level at which the observed result would lead you to reject H zero. It is also, assuming H zero is true, the probability of drawing a test statistic at least as extreme as the one actually observed.

This slide uses a two-sided example with an observed t-statistic of two point one six. The left column states that result. On the figure, the horizontal axis is the t-statistic and the vertical height is the t density under H zero. The right-tail boundary is two point one six, and the symmetric left boundary is minus two point one six. Both tails beyond those values are shaded because a statistic equally far from zero in either direction would be at least as extreme for a two-sided alternative.

The combined shaded area is about zero point zero three one, or three point one percent. That is the two-sided p-value. If you selected a significance level of exactly three point one percent, the statistic would sit at the critical boundary. At any conventional level larger than the p-value, such as five percent, the rejection region is wider and contains the observed statistic, so you reject. At a smaller level, such as one percent, the rejection boundary lies farther out and you fail to reject.

The decision callout summarizes the comparison: reject when the p-value is below the chosen significance level. This always agrees with comparing the statistic to the matching critical value, provided the same distribution, degrees of freedom, and sidedness are used.

</details>

:::{.callout-note title="Definition"}
The smallest significance level at which the null hypothesis would be rejected (the probability of observing a test statistic at least as extreme as we did if the null hypothesis is true)
:::


```{r p-dist, include = F}
u <- seq(-3.5, 3.5, length = 1000)
pdf <- dt(u, df = df_fig) # default is mean = 0,sd = 1
data <- data.table(x = u, y = pdf)

g_t <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme_lecture +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.031
c_rej <- qt(1 - alpha / 2, df = df_fig)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_t_0031 <- g_t +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.03,
    label = as.character(round(c_rej, digits = 2)),
    size = 3.7,
    family = "Times"
  ) +
  annotate("text",
    x = 0, y = 0.1,
    label = paste("area  =  ", round(alpha, digits = 3), sep = ""),
    size = 3.7,
    family = "Times"
  )
```

::: {.columns}

::: {.column width="50%"}
+ Suppose the t-statistic you got is 2.16. If the null hypothesis is true, the probability of observing a test statistic at least this extreme is 3.1%.

+ Thus, 3.1% is the smallest significance level at which the null would be rejected, which is the definition of the p-value.
:::
<!--end of the 1st column-->
::: {.column width="50%"}
```{r echo = F}
g_t_0031
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->  
 
:::{.callout-note title="Decision rule"}
If the p-value is lower than your choice of significance level, then you reject the null.
:::

This decision rule of course results in the same test results as the one we saw that uses a t-value and critical value.

### Example

<details class="transcript">
<summary>Transcript</summary>

Work through this example from the model to the decision. The estimated log-wage equation includes education, experience, and tenure. The coefficient on experience is zero point zero zero four one.

The estimated standard error of the experience coefficient is zero point zero zero one seven. We test H zero: beta exper equals zero against the one-sided alternative beta exper greater than zero. Under H zero, the t-statistic is the estimate minus zero, divided by its estimated standard error. That gives zero point zero zero four one divided by zero point zero zero one seven, or about two point four one. The statistic says the estimate is about two point four one standard errors above the null value.

The model has n equals five hundred and twenty-six observations and k equals three explanatory variables, excluding the intercept. Residual degrees of freedom are therefore five hundred and twenty-six minus three minus one, or five hundred and twenty-two. Because the alternative is positive and the significance level is one percent, use an upper-tail critical value rather than splitting the probability across two tails. `qt(0.99, 522)` returns about two point three three, the ninety-ninth percentile of the appropriate t distribution.

Finally compare on the same scale. Two point four one is greater than two point three three, so the observed statistic lies in the one-percent upper-tail rejection region. We reject H zero in favor of a positive experience effect at the one percent level.

</details>

**Estimated Model**

The impact of experience on wage:

+ $log(wage) = 0.284+0.092\times educ+0.0041\times exper + 0.022 \times tenure$
+ $\widehat{se(\widehat{\beta}_{exper})} = 0.0017$ 
+ $n = 526$

<br>

**Hypothesis**

+ $H_0$: $\beta_{exper}=0$
+ $H_1$: $\beta_{exper}>0$

<br>

**Test**

t-statistic $= 0.0041/0.0017 = 2.41$

The critical value is the 99% quantile of $t_{526-3-1}$, which is $2.33$ (it can be obtained by `qt(0.99, 522)`)

Since $2.41 > 2.33$, we reject the null in favor of the alternative hypothesis at the 1% level.

:::
<!--end of panel-->


## R implementation

::: {.panel-tabset}

### Run regression

<details class="transcript">
<summary>Transcript</summary>

This tab sets up the regression object used by the following R examples. The `autorun: true` option means the cell executes automatically when the interactive page loads, so later tabs can use its objects without requiring you to run this cell first.

The first line calls `data` with the quoted dataset name `wage1` and `package = "wooldridge"`. That loads the `wage1` data bundled with the Wooldridge package into the R session. Specifying the package makes the source explicit and avoids relying on whichever packages happen to be attached.

The second substantive line calls `feols` and assigns the fitted model to `reg_wage`. In the formula, wage appears to the left of the tilde, so it is the dependent variable. Education, experience, and tenure appear to the right, separated by plus signs, so they are the three explanatory variables. `data = wage1` tells `feols` where to find those variable names. An intercept is included automatically. The fitted object contains the coefficient estimates, residuals, variance-covariance information, sample size, degrees of freedom, and other items needed for inference.

The assignment operator means the model is stored rather than printed. That is why the cell normally produces no visible output even though it ran successfully. To inspect the standard `fixest` display, put `reg_wage` on a line by itself.

</details>

```{webr-r}
#| autorun: true
#--- get the data ---#
data("wage1", package = "wooldridge")

#--- run a regression ---#
reg_wage <- feols(wage ~ educ + exper + tenure, data = wage1)
```

### Obtain t-statistics

<details class="transcript">
<summary>Transcript</summary>

This tab extracts regression results into a rectangular table. `broom::tidy(reg_wage)` takes the `fixest` model object created on the previous tab and returns a tibble with one row per coefficient. The namespace prefix `broom::` calls `tidy` from the broom package explicitly, even if that package has not been attached. The result is assigned to `reg_results_table`, then the object name on the last line prints it. The cell runs automatically, and `code-track: 0.5` controls only the display layout, not the R calculation.

Read the table by columns. `term` identifies the intercept or explanatory variable. `estimate` is beta j hat, the OLS coefficient estimate. `std.error`, pronounced standard error, is the estimated standard error of that coefficient. `statistic` is estimate divided by standard error, which is the t-statistic for the default null beta j equals zero. More generally it is beta j hat minus zero over estimated se of beta j hat. `p.value` is the two-sided p-value for that same zero-null test, so it includes tail probability beyond both the positive and negative absolute t-statistic.

For example, the `educ` row tells you the estimated partial effect of education, its uncertainty, how many standard errors it is from zero, and the two-sided probability of a statistic at least that extreme under a zero population coefficient. The output performs this very common test automatically for every term.

The last two sentences state the limitation. If your null is beta j equals zero and your alternative is two-sided, the displayed statistic and p-value already answer the question. If the null is some other value, such as beta educ equals one, the printed t-statistic is not the statistic you need, because its numerator subtracts zero. The next tab builds a nonzero-null statistic by hand.

</details>

You can apply the `broom::tidy()` function from the `broom` package to access the regression results (`reg_wage` here) as a `tibble` (`data.frame`).

```{webr-r}
#| autorun: true
#| code-track: 0.5
#--- tidy up the regression results ---#
reg_results_table <- broom::tidy(reg_wage)

#--- display ---#
reg_results_table
```

<br>

+ `estimate`: coefficient estimates $(\widehat{\beta}_j)$
+ `std.error`: $\widehat{se(\widehat{\beta}_j)}$
+ `statistic`: t-statistic for the null of $\beta_j = 0$
+ `p.value`: p-value (for the two sided test with the null of $\beta_j = 0$)

So, for the t-test of $\beta_j = 0$ is already there. You do not need to do anything further.

For the null hypothesis other than $\beta_j = 0$, you need further work.

### R implementation (by hand)

Suppose you are testing the null hypothesis of $\beta_{educ} = 1$ against the alternative hypothesis of $\beta_{educ} \ne 1$ (so, this is a two-sided test). 

The t-value for this test is not available from the summary. 

::: {.panel-tabset}

```{r}
#| include: false
data("wage1", package = "wooldridge")

#--- run a regression ---#
reg_wage <- feols(wage ~ educ + exper + tenure, data = wage1)

reg_results_table <- broom::tidy(reg_wage)

#--- coefficient estimate on educ ---#
beta_educ <-
  reg_results_table %>%
  dplyr::filter(term == "educ") %>%
  dplyr::pull(estimate)

#--- se of the coefficient on educ ---#
se_beta_educ <-
  reg_results_table %>%
  dplyr::filter(term == "educ") %>%
  dplyr::pull(std.error)

#--- get the t-value ---#
t_value <- (beta_educ - 1) / se_beta_educ
df_t <- degrees_freedom(reg_wage, "resid")
critical_value <- qt(0.975, df = df_t)
```

#### get t-value

<details class="transcript">
<summary>Transcript</summary>

We are now testing H zero: beta educ equals one against H one: beta educ does not equal one. This is two-sided, and it is not the default zero-null test in the tidy table, so we construct the appropriate t-statistic ourselves.

The first pipeline starts with `reg_results_table`, the tibble from the previous tab. `filter(term == "educ")` keeps only the row whose term label is education. The double equals sign performs a logical equality comparison. `pull(estimate)` then extracts that row's estimate column as a single numeric value, and the assignment stores it in `beta_educ`. The pipe passes the object from each step to the next.

The second pipeline repeats the same row selection but pulls `std.error`, storing the estimated standard error in `se_beta_educ`. Filtering both values by the term name keeps the estimate and uncertainty aligned with the coefficient being tested.

The final parenthesized assignment calculates `t_value` and prints it. Read the formula as beta educ hat minus one, divided by the estimated standard error of beta educ hat. The minus one is the essential change: one is the value asserted under H zero. With this wage-level regression, the education estimate is about zero point five nine nine and its standard error about zero point zero five one, so the statistic is strongly negative, roughly minus seven point eight. The sign says the estimate lies below one; the magnitude says it is many standard errors away.

If you had used the `statistic` column from `tidy`, you would instead test beta educ equals zero and obtain a large positive value. Changing the null changes the centered numerator even though the estimate and standard error remain the same.

</details>

```{webr-r}
#| code-track: 0.5

#--- coefficient estimate on educ ---#
beta_educ <-
  reg_results_table %>%
  filter(term == "educ") %>%
  pull(estimate)

#--- se of the coefficient on educ ---#
se_beta_educ <-
  reg_results_table %>%
  filter(term == "educ") %>%
  pull(std.error)

#--- get the t-value ---#
(
  t_value <- (beta_educ - 1) / se_beta_educ
)
```



#### critical value and t-test

<details class="transcript">
<summary>Transcript</summary>

This tab completes the test by finding the correct reference distribution and critical value. The first code cell calls `degrees_freedom(reg_wage, "resid")`. The first argument is the fitted model and the second asks specifically for residual degrees of freedom. The returned value is stored in `df_t` and printed because the assignment is wrapped in parentheses. For five hundred and twenty-six observations, three explanatory variables, and an intercept, it is five hundred and twenty-two.

The second cell calls `qt(0.975, df = df_t)`. `qt` returns a quantile of a t distribution. This is a two-sided test at the five percent level, so five percent is split into two point five percent in each tail. The positive boundary therefore has ninety-seven point five percent of the distribution to its left. The `df` argument selects the t distribution with the residual degrees of freedom just obtained. The result, about one point nine six, is assigned to `critical_value` and printed.

For a two-sided decision, compare the absolute value of `t_value` with this positive critical value. The statistic from the previous tab is about minus seven point eight, so its absolute value is much greater than one point nine six. We reject H zero: beta educ equals one at the five percent level. The negative sign indicates that the estimated coefficient is below one, but either sign could have caused rejection under the not-equal alternative.

The last sentence on screen inserts the calculated values dynamically. Its logic is the same as every earlier test: place the observed standardized statistic in its null distribution and see whether it crosses the boundary. The large gap is unsurprising because a null of one dollar per additional year of education is far from the fitted wage-level coefficient of roughly zero point six.

</details>

The degree of freedom $(n-k-1)$ of the t-distribution can be obtained by applying `degrees_freedom(reg_wage, "resid")`:

```{webr-r}
#| code-track: 0.5
(
  df_t <- degrees_freedom(reg_wage, "resid")
)
```

<br>

You can get the 97.5% quantile of the $t_{522}$ using the `qt()` function:

```{webr-r}
#| code-track: 0.5
(
  critical_value <- qt(0.975, df = df_t)
)
```

<br>

Since the absolute value of the t-value (`r t_value`) is greater than the critical value (`r critical_value`), you reject the null.

:::
<!--end of panel-->

:::
<!--end of panel-->

# Confidence Interval (CI)

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Confidence Interval (CI)

::: {.panel-tabset}

### Definition

<details class="transcript">
<summary>Transcript</summary>

The definition callout gives the repeated-sampling meaning of a ninety-five percent confidence interval. Imagine repeatedly drawing new random samples using the same design and, for each sample, calculating an interval with the same formula. The endpoints change from sample to sample because the estimate and its standard error change. In the long run, ninety-five percent of those intervals will cover the fixed true parameter and five percent will miss it, provided the assumptions behind the procedure hold.

Confidence therefore describes the coverage rate of the interval-generating procedure. Before sampling, the endpoints are random and the procedure has a ninety-five percent probability of producing an interval that covers the parameter. Once a specific sample has been observed and its interval calculated, the endpoints are fixed. The population parameter is also fixed. That realized interval either contains the true parameter or it does not.

The lower half identifies the common incorrect interpretation: "there is a ninety-five percent probability that this particular realized interval includes the true parameter." In the frequentist framework used here, that is not what the statement means. We do not assign a probability distribution to the fixed beta after seeing the data. The ninety-five percent refers to what the method would do over repeated samples.

</details>

:::{.callout-note title="Definition"}
If you calculate the 95% confidence interval on multiple different samples, 95% of the time, the calculated CI includes the true parameter
:::

<br>

**What confidence interval is not**

The probability that a realized CI calculated from specific sample data includes the true parameter

### How to get CI (in general)

::: {.panel-tabset}

#### General Procedure

<details class="transcript">
<summary>Transcript</summary>

This is the general central-interval recipe for a statistic x with an assumed probability distribution. An A percent interval places A percent of that distribution's probability in the middle and leaves one hundred minus A percent outside. For a two-sided central interval, divide the omitted probability equally between the lower and upper tails.

The lower endpoint is the open parenthesis one hundred minus A close parenthesis, divided by two, percent quantile of x. The upper endpoint is one hundred minus that same tail percentage. A quantile is the value with the stated proportion of the distribution at or below it. Thus the lower endpoint leaves half of the omitted probability below, and the upper endpoint leaves the other half above.

Pay attention to the parentheses. For A equals ninety-five, calculate one hundred minus ninety-five first, obtaining five, then divide by two to obtain two point five. Writing or reading it as one hundred minus A divided by two would instead produce fifty-two point five, which is completely different. The upper percentile is one hundred minus two point five, or ninety-seven point five.

</details>

For the **assumed** distribution of statistic $x$, the $A\%$ confidence interval of $x$ is the range with

+ lower bound: $(100 - A)/2$ percent quantile of $x$
+ upper bound: $100-(100 - A)/2$ percent quantile of $x$

#### Example

<details class="transcript">
<summary>Transcript</summary>

Now insert A equals ninety-five into the general procedure. One hundred minus ninety-five leaves five percent outside the central interval. Split that evenly, so the lower endpoint is the two point five percentile and the upper endpoint is the ninety-seven point five percentile. The formulas in parentheses in the left column show both calculations explicitly.

For this example, x follows the standard Normal distribution, with mean zero and variance one. Its two point five percentile is approximately minus one point nine six, and its ninety-seven point five percentile is approximately positive one point nine six. Therefore the central ninety-five percent of possible x values lies between minus one point nine six and one point nine six.

Look at the figure in the right column. The horizontal axis shows values of x, the vertical height shows the standard Normal density, and the curve is centered at zero. The red ribbons occupy the two tails outside the interval. Each tail has area zero point zero two five, so together they contain zero point zero five, leaving zero point nine five unshaded in the middle. The annotation near the positive boundary marks about one point nine six; symmetry places the negative boundary at minus one point nine six.

</details>

::: {.columns}

::: {.column width="40%"}
For the 95% CI (A = 95), 

+ lower bound: $2.5$ $\big((100-95)/2\big)$ percent quantile of $x$
+ upper bound: $97.5$ $\big(100-(100-95)/2\big)$ percent quantile of $x$

If $x$ follows the standard normal distribution $(x \sim N(0, 1))$, then,the 2.5% and 97.5% quantiles are -1.96 and 1.96, respectively. So, the 95% CI of $x$ is [-1.96, 1.96].

:::
<!--end of the 1st column-->
::: {.column width="60%"}
```{r norm-95, include = F}
u <- seq(-3, 3, length = 1000)
pdf <- dnorm(u) # default is mean=0,sd=1
data <- data.table(x = u, y = pdf)

g_norm <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme_lecture +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.05
c_rej <- qnorm(1 - alpha / 2)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_norm_005 <- g_norm +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  annotate("text",
    x = c_rej, y = -0.03,
    label = as.character(round(c_rej, digits = 2)),
    size = 3.7,
    family = "Times"
  ) +
  annotate("text",
    x = 2.0, y = 0.30,
    label = paste("area = ", round(alpha / 2, digits = 3), sep = ""),
    size = 3.7,
    family = "Times"
  )
```
```{r echo = F}
g_norm_005
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->



:::
<!--end of panel-->


### How to get the CI of coefficients

<details class="transcript">
<summary>Transcript</summary>

This slide derives a confidence interval for beta j from the t-statistic. Under MLR one through MLR six, beta j hat minus beta j, divided by the estimated standard error of beta j hat, follows a t distribution with n minus k minus one degrees of freedom. Beta j hat is the sample coefficient estimate, beta j is the fixed population coefficient, the denominator is the estimated sampling standard deviation, n is the sample size, and k counts explanatory variables other than the intercept.

For an A percent central interval, define Q sub l as the open parenthesis one hundred minus A close parenthesis over two percentile of that t distribution. Define Q sub h as the one hundred minus open parenthesis one hundred minus A close parenthesis over two percentile. Because the t distribution is symmetric around zero, Q sub l is negative and Q sub h is the corresponding positive value.

Those quantiles initially bound the standardized statistic, but our target is beta j. Solve the displayed statistic for beta j. If t denotes a possible value of the standardized expression, multiply by the estimated standard error to get beta j hat minus beta j, then rearrange: beta j equals beta j hat minus t times the estimated standard error. The minus sign reverses which t quantile maps to which beta endpoint. Symmetry makes the familiar final result beta j hat plus a negative critical amount for the lower bound and beta j hat plus a positive critical amount for the upper bound.

That is what the final bullets state. Multiply Q sub l by the estimated standard error and add beta j hat for the lower bound. Multiply Q sub h by the same standard error and add beta j hat for the upper bound. Since Q sub l is negative and Q sub h positive, the interval is centered on beta j hat and extends the same number of standard errors in both directions. A larger standard error makes the interval wider, signaling less precision. The next tabs calculate a ninety-percent interval for the education coefficient.

</details>

Under the assumption of MLR.1 through MLR.6 (which includes the normality assumption of the error), we learned that

$\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}} \sim t_{n-k-1}$

So, following the general procedure we discussed in the previous slide, the A% confidence interval of $\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}}$ is 

+ lower bound: $(100 − A)/2$% quantile of the $t_{n-k-1}$ distribution (let's call this $Q_l$)
+ upper bound: $100 - (100 − A)/2$% quantile of the $t_{n-k-1}$ distribution (let's call this $Q_h$)  

But, we want the A% CI of $\beta_j$, not $\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}}$. Solving for $\beta_j$,

$\beta_j = \widehat{\beta}_j - t_{n-k-1}\times \widehat{se(\widehat{\beta}_j)}$

So, to get the A% CI of $\beta_j$, we scale the CI of $\frac{\widehat{\beta}_j - \beta_j}{\widehat{se(\widehat{\beta}_j)}}$ by $\widehat{se(\widehat{\beta}_j)}$ and then shift by $\widehat{\beta}_j$. Because the $t$ distribution is symmetric, the minus sign above just swaps which quantile lands on which end:

+ lower bound: $Q_l \times \widehat{se(\widehat{\beta}_j)} + \widehat{\beta}_j$
+ upper bound: $Q_h \times \widehat{se(\widehat{\beta}_j)} + \widehat{\beta}_j$

Note that $Q_l$ is <span style = "color: blue;"> negative </span> and $Q_h$ is <span style = "color: red;">positive</span>.

### Example (R implementation)

::: {.panel-tabset}
#### Regression

<details class="transcript">
<summary>Transcript</summary>

This tab reconstructs the wage-level regression for the confidence-interval example. The first cell calls `feols` with wage on the left of the tilde and education, experience, and tenure on the right. `data = wage1` supplies the dataset loaded earlier. The fitted object is assigned to `wage_reg`. Because the cell has `autorun: true`, it is available to later interactive cells as soon as the page loads, and assignment alone produces no printed model summary.

The second cell applies `broom::tidy` to `wage_reg` and assigns the returned tibble to `wage_reg_coef`. Parentheses around the assignment cause the table to print as well as be stored. This cell also runs automatically. The table has one row for the intercept and each explanatory variable, with columns for the estimate, estimated standard error, default zero-null t-statistic, and p-value.

Focus on the `educ` row. Running the code gives an estimated coefficient of about zero point five nine nine and an estimated standard error of about zero point zero five one three. The model has five hundred and twenty-six observations and three non-intercept regressors, so its residual degrees of freedom are five hundred and twenty-two. Those three pieces, estimate, standard error, and degrees of freedom, are sufficient to form the usual confidence interval for beta educ under the stated classical assumptions.

</details>

**Run OLS and extract necessary information**

```{webr-r}
#| autorun: true
wage_reg <- feols(wage ~ educ + exper + tenure, data = wage1)
```

<br>

Applying `broom::tidy()` to `wage_reg`,

```{webr-r}
#| autorun: true
(
  wage_reg_coef <- broom::tidy(wage_reg)
)
```

```{r}
#| include: false
wage_reg <- feols(wage ~ educ + exper + tenure, data = wage1)
wage_reg_coef <- broom::tidy(wage_reg)
# named df_wage, not df: `df()` is base R's F density, so a bare `df` that is
# never assigned resolves to the function instead of erroring
df_wage <- degrees_freedom(wage_reg, "resid")
b_educ <- wage_reg_coef %>%
  filter(term == "educ") %>%
  pull(estimate)
se_educ <- wage_reg_coef %>%
  filter(term == "educ") %>%
  pull(std.error)
```

#### Collect necessary information

<details class="transcript">
<summary>Transcript</summary>

We want a ninety-percent confidence interval for beta educ, the population coefficient on education in this wage-level regression. The displayed t result says beta educ hat minus beta educ, divided by the estimated standard error of beta educ hat, follows a t distribution with n minus k minus one degrees of freedom, assuming MLR one through MLR six. That result is the bridge from the regression output to an interval.

The three bullets collect the required inputs. Beta educ hat is about zero point five nine nine, shown to four decimal places on the slide. This is the center of the interval. Its estimated standard error is about zero point zero five one three, which determines the scale or width. Smaller uncertainty would produce a tighter interval around the same estimate.

The final input is five hundred and twenty-two residual degrees of freedom. The dataset supplies n equals five hundred and twenty-six observations. There are k equals three explanatory variables, education, experience, and tenure, and the intercept costs one additional degree of freedom. Thus n minus k minus one equals five hundred and twenty-six minus three minus one, or five hundred and twenty-two.

</details>

We are interested in getting the 90% confidence interval of the coefficient on `educ` $(\beta_{educ})$. Under all the assumptions (MLR.1 through MLR.6), we know that in general,

$\frac{\widehat{\beta}_{educ} - \beta_{educ}}{\widehat{se(\widehat{\beta}_{educ})}} \sim t_{n-k-1}$

Specifically for this regression,

+ $\widehat{\beta}_{educ}$ = `r round(b_educ, 4)`

+ $\widehat{se(\widehat{\beta}_{educ})}$ = `r round(se_educ, 4)`

+ $n - k - 1 = `r df_wage`$ (degrees of freedom)

#### Find CI

<details class="transcript inline">
<summary>Transcript</summary>

Now calculate the ninety-percent interval. Ninety percent in the center leaves ten percent outside, split evenly into five percent in each tail. We therefore need the five percent and ninety-five percent quantiles of a t distribution with five hundred and twenty-two degrees of freedom. The first interactive cell calls `qt(0.05, df = 522)` and `qt(0.95, df = 522)`. The first argument is the cumulative probability and the named `df` argument selects the degrees of freedom. Running it returns symmetric values of about minus one point six four eight and positive one point six four eight.

The left column then states the central ninety-percent interval for the standardized statistic. Using the rounded values displayed in the fraction, the numerator is zero point five nine nine minus beta educ and the denominator is about zero point zero five one. That ratio lies between the two t quantiles in ninety percent of repeated samples under the model assumptions.

To translate back to beta educ, scale by the estimated standard error and shift by the estimate. The lower bound is zero point five nine nine plus zero point zero five one times minus one point six four eight. The upper bound is zero point five nine nine plus zero point zero five one times positive one point six four eight. Using the underlying unrounded values, the bounds are approximately zero point five one four and zero point six eight three. The negative lower quantile moves below the estimate; the positive upper quantile moves above it.

Look at the figure in the right column. Its horizontal axis is the t-statistic, its vertical height is the t density, and the red ribbons mark symmetric tail regions. The curve is symmetric and, with five hundred and twenty-two degrees of freedom, nearly indistinguishable from standard Normal. For the ninety-percent calculation on this tab, use the five and ninety-five percent quantiles printed by the interactive cell.

The final sentence hands you to the `In practice` tab. There, broom requests the same confidence level directly. Its `conf.low` and `conf.high` values should match the hand calculation.

</details>

```{r include = F}
u <- seq(-3, 3, length = 1000)
pdf <- dt(u, df = df_wage) # default is mean=0,sd=1
data <- data.table(x = u, y = pdf)

g_t <- ggplot() +
  geom_line(data = data, aes(y = y, x = x)) +
  geom_hline(yintercept = 0) +
  xlab("") +
  ylab("") +
  theme_lecture +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha <- 0.05
c_rej <- qt(1 - alpha / 2, df = df_wage)
rej_data_high <- data[u >= c_rej, ]
rej_data_low <- data[u <= -c_rej, ]

g_t_005 <- g_t +
  geom_ribbon(
    data = rej_data_high,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  ) +
  geom_ribbon(
    data = rej_data_low,
    aes(ymax = y, ymin = 0, x = x),
    fill = "red", alpha = 0.4
  )
```

::: {.columns}

::: {.column width="50%"}
Now, we need to find the 5% $\big((100-90)/2\big)$ and 95% $\big(100-(100-90)/2\big)$ quantiles of $t_{`r df_wage`}$.

```{webr-r}
qt(0.05, df = 522)
qt(0.95, df = 522)
```

<br>

So, the 90% CI of $\frac{`r round(b_educ, 3)` - \beta_{educ}}{`r round(se_educ, 3)`} \sim t_{`r df_wage`}$ is [`r round(qt(0.05, df_wage), 3)`, `r round(qt(0.95, df_wage), 3)`]

By scaling and shifting, the lower and upper bounds of the 90% CI of $\beta_{educ}$ are:

+ lower bound: `r round(b_educ, 3)` + `r round(se_educ, 3)` $\times$ `r round(qt(0.05, df_wage), 3)` = `r round(b_educ + se_educ * qt(0.05, df_wage), 3)`

+ upper bound: `r round(b_educ, 3)` + `r round(se_educ, 3)` $\times$ `r round(qt(0.95, df_wage), 3)` = `r round(b_educ + se_educ * qt(0.95, df_wage), 3)`

Confirm these against the `In practice` tab, which asks `broom` for the same interval.


:::
<!--end of the 1st column-->
::: {.column width="50%"}
The distribution of $t_{`r df_wage`}$:

```{r echo = F}
g_t_005
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

#### In practice

<details class="transcript">
<summary>Transcript</summary>

This is the practical way to obtain the interval from a fitted model. `broom::tidy` receives `wage_reg` as its first argument. Setting `conf.int = TRUE` asks broom to add confidence-interval endpoints to the usual coefficient table. Setting `conf.level = 0.9` requests ninety-percent coverage rather than the common default of ninety-five percent. The decimal zero point nine is a proportion, not the number ninety.

The pipe sends the resulting tibble to `dplyr::relocate(term, conf.low, conf.high)`. `relocate` does not change any calculations or rows. It simply moves the term name and the two interval columns to the front so they are easier to compare. The remaining columns, including estimate, standard error, statistic, and p-value, remain in the table after them.

Run the cell and find the `educ` row. `conf.low` is approximately zero point five one four and `conf.high` approximately zero point six eight three. These agree with the by-hand calculation on the previous tab, up to rounding.

</details>

You can just use `broom::tidy()` with `conf.int = TRUE`, `conf.level = confidence level` like below:

```{webr-r}
broom::tidy(
  wage_reg,
  conf.int = TRUE,
  conf.level = 0.9
) %>%
  dplyr::relocate(term, conf.low, conf.high)
```

:::
<!--end of panel-->

:::
<!--end of panel-->


# Linear Combination of Multiple Coefficients

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Linear Combination of Multiple Coefficients

::: {.panel-tabset}

### Example

<details class="transcript">
<summary>Transcript</summary>

This section asks a question about two coefficients together rather than one coefficient in isolation. The model says log wage equals beta zero plus beta one times `jc`, beta two times `univ`, beta three times experience, and the error u. `jc` is an indicator equal to one for someone who attended a two-year college and zero otherwise. `univ` is an indicator equal to one for someone who attended a four-year college and zero otherwise. Experience is included as a control.

With log wage as the dependent variable, beta one is the partial difference in log wage associated with two-year-college attendance relative to the omitted category, holding experience and the other indicator fixed. Beta two is the analogous partial difference for four-year-college attendance.

The substantive question is whether the effect associated with four-year college is greater than the effect associated with two-year college. Looking only at whether beta one differs from zero cannot answer that. Looking only at whether beta two differs from zero cannot answer it either. We need a direct comparison between beta one and beta two.

The hypotheses at the bottom express that comparison. H one says beta one is less than beta two, which is the directional claim that four-year college has the larger effect. H zero says beta one equals beta two, meaning no difference between the two effects.

</details>

**Model**

$$log(wage) = \beta_0+\beta_1 jc+ \beta_2 univ + \beta_3 exper + u$$

+ $jc$: 1 if you attended 2-year college, 0 otherwise
+ $univ$: 1 if you attended 4-year college, 0 otherwise

**Question**

Is the impact of education on wage greater if you attend a 4-year college than a 2-year college?

**Hypothesis**

The null and alternative hypotheses would be:

+ $H_1:$ $\beta_1 < \beta_2$
+ $H_0:$ $\beta_1 = \beta_2$

### Rewrite the hypothesis

<details class="transcript">
<summary>Transcript</summary>

The algebra on this tab converts a comparison of two coefficients into a test of one constructed parameter. Start with H one: beta one is less than beta two, and H zero: beta one equals beta two. Subtract beta two from both sides of each relationship. The alternative becomes beta one minus beta two is less than zero. The null becomes beta one minus beta two equals zero.

Now define alpha as beta one minus beta two. This is only a change of notation, not a new assumption. In terms of alpha, H one says alpha is negative and H zero says alpha equals zero. We have reduced the original question to the same shape as an ordinary one-parameter t-test: is a parameter equal to a specified null value, or is it on one side of that value?

The final sentence identifies why the method generalizes. Alpha is a linear combination of regression coefficients. "Linear" means the coefficients are multiplied by fixed weights and added: here the weight on beta one is one and the weight on beta two is minus one.

</details>

The null and alternative hypotheses are:

+ $H_1:$ $\beta_1 < \beta_2$
+ $H_0:$ $\beta_1 = \beta_2$

Rewriting them,

+ $H_1:$ $\beta_1-\beta_2 < 0$
+ $H_0:$ $\beta_1 - \beta_2 =0$

Or,  

+ $H_1:$ $\alpha < 0$
+ $H_0:$ $\alpha =0$

where $\alpha = \beta_1 - \beta_2$

Note that $\alpha$ is a linear combination of $\beta_1$ and  $\beta_2$.

### Math

<details class="transcript">
<summary>Transcript</summary>

The callout states the key distributional fact. Let alpha be any fixed linear combination of population OLS coefficients and alpha hat the same combination of their estimates. Under MLR one through MLR six, alpha hat minus alpha, divided by the estimated standard error of alpha hat, follows a t distribution with n minus k minus one residual degrees of freedom. The numerator measures the estimation error in the contrast, and the denominator expresses that error in standard-error units.

Why does the result extend beyond a single coefficient? Under the Normal-error assumptions, the OLS coefficient estimators are jointly Normal, not merely Normal one at a time. Any fixed linear combination of jointly Normal random variables is also Normal. Replacing the unknown error variance in its standard deviation with the residual estimate introduces the same t-distribution adjustment as before. Thus a contrast can be tested with the same t logic, exactly under the classical finite-sample assumptions.

The second displayed equation specializes the fact to alpha equals beta one minus beta two. Alpha hat is beta one hat minus beta two hat. Subtracting the true alpha produces beta one hat minus beta two hat minus open parenthesis beta one minus beta two close parenthesis. Divide by the estimated standard error of beta one hat minus beta two hat, and the ratio follows t with n minus k minus one degrees of freedom.

Everything in that expression is familiar except the denominator. Once that standard error is available, however, the test proceeds exactly as before: impose the null value in the numerator, calculate the statistic, and compare it with the appropriate one-sided or two-sided t critical value.

</details>

:::{.callout-note title="Important Fact"}
For any linear combination of the OLS coefficients, denoted as $\widehat{\alpha}$, the following holds:

$$\frac{\widehat{\alpha}-\alpha}{\widehat{se(\widehat{\alpha})}} \sim t_{n-k-1}$$

Where $\alpha$ is the true value (it is $\beta_1 - \beta_2$ in the example in the previous slide).
:::

<br>

So, using the example, this means that 

$$\frac{\widehat{\alpha}-\alpha}{\widehat{se(\widehat{\alpha})}} = \frac{\widehat{\beta}_1-\widehat{\beta}_2-(\beta_1 - \beta_2)}{\widehat{se(\widehat{\beta}_1-\widehat{\beta}_2)}} \sim t_{n-k-1}$$

This is great because we know how to do t-test!

### Going back to the example

<details class="transcript">
<summary>Transcript</summary>

Return to the college comparison. The null is alpha equals zero, which is equivalent to beta one minus beta two equals zero. The red "If" marks the conditional logic of the test. If that equality is true, substitute zero for alpha in the general standardized contrast.

The first numerator is alpha hat minus zero. Replacing alpha hat by its definition gives beta one hat minus beta two hat minus zero. The denominator is the estimated standard error of the estimated difference, se hat of beta one hat minus beta two hat. Under H zero and the classical assumptions, this ratio follows a t distribution with n minus k minus one degrees of freedom.

The sign of the statistic matters because our alternative is directional. Alpha less than zero means beta one is smaller than beta two, so evidence for the alternative appears in the lower tail. A sufficiently negative t-statistic would lead us to reject equality in favor of a larger four-year-college effect. A positive statistic points opposite to the stated alternative.

The last sentence lists the ingredients. We need beta one hat, beta two hat, and the estimated standard error of their difference. The two coefficient estimates appear in ordinary regression output. The standard error of the difference usually does not, because a standard table reports uncertainty for each coefficient separately.

</details>

Our null hypothesis is $\alpha = 0$ (or $\beta_1 - \beta_2 = 0$).

So, <span style = "color: red;"> If </span> indeed the null hypothesis is true, then

$$\frac{\widehat{\alpha}-0}{\widehat{se(\widehat{\alpha})}} = \frac{\widehat{\beta}_1-\widehat{\beta}_2-0}{\widehat{se(\widehat{\beta}_1-\widehat{\beta}_2)}} \sim t_{n-k-1}$$

So, all you need to do is to substitute $\widehat{\beta}_1$, $\widehat{\beta}_2$, $\widehat{se(\widehat{\beta}_1 - \widehat{\beta}_2)}$ into the formula and see if the value is beyond the critical value for your chosen level of statistical significance.

### get $se(\widehat{\beta}_1-\widehat{\beta}_2)$

<details class="transcript">
<summary>Transcript</summary>

This tab derives the standard error of the coefficient difference and highlights the common mistake. A standard error is the square root of a variance, so se of beta one hat minus beta two hat equals the square root of variance of that difference. It is not generally the square root of variance beta one hat plus variance beta two hat. That simpler expression would be correct only if the covariance between the estimators were zero.

The middle paragraph explains why the tempting shortcut would be convenient. A regression table reports a standard error for beta one hat and another for beta two hat. Squaring them recovers the two estimated variances. If variances simply added for a difference, we could sum those diagonal pieces and take a square root. But both coefficients are estimated jointly from the same sample, so their estimation errors can move together.

The math-aside formula gives the general rule. Read it as variance of a times x plus b times y equals a squared variance of x, plus two a b covariance of x and y, plus b squared variance of y. For x minus y, set a equal to one and b equal to minus one. Then a squared and b squared are both one, while two a b is minus two.

Applying that result yields variance beta one hat minus beta two hat equals variance beta one hat, minus two covariance beta one hat with beta two hat, plus variance beta two hat. Taking the square root gives the required standard error. The covariance term can raise or lower the variance of the difference depending on its sign. It therefore cannot be ignored merely because it is absent from the usual coefficient table. The next R tabs obtain all three pieces from the model's variance-covariance matrix.

</details>

But, 

$$se(\widehat{\beta}_1-\widehat{\beta}_2)= \sqrt{Var(\widehat{\beta}_1-\widehat{\beta}_2)} \ne \sqrt{Var(\widehat{\beta}_1)+Var(\widehat{\beta}_2)}$$

If the following was true,

$$se(\widehat{\beta}_1-\widehat{\beta}_2) = \sqrt{Var(\widehat{\beta}_1)+Var(\widehat{\beta}_2)}$$

then, we could have just extracted $Var(\widehat{\beta}_1)$ and $Var(\widehat{\beta}_2)$ individually from the regression object on R, sum them up, and take a square root of it.

:::{.callout-note title="Math aside"}
$$Var(ax+by) = a^2 Var(x) + 2abCov(x,y) + b^2 Var(y)$$
:::

So, 

$$se(\widehat{\beta}_1-\widehat{\beta}_2)= \sqrt{Var(\widehat{\beta}_1-\widehat{\beta}_2)} = \sqrt{Var(\widehat{\beta}_1)-2Cov(\widehat{\beta}_1,\widehat{\beta}_2)+Var(\widehat{\beta}_2)}$$

### Demonstration using R

::: {.panel-tabset}

#### Regression

<details class="transcript">
<summary>Transcript</summary>

This cell fits the college model using the `twoyear` data from the Wooldridge package. The first line calls `data("twoyear", package = "wooldridge")`, which loads the dataset explicitly from that package. The parenthesized assignment on the next lines fits and prints the model.

`fixest::feols` is called with log wage, `lwage`, on the left of the formula. The regressors are `jc`, `univ`, and `exper`. `data = twoyear` identifies the data frame. The fitted object is stored as `reg_sc`. The explicit `fixest::` namespace shows which package supplies `feols`. An intercept is included automatically, and `autorun: true` ensures that later tabs can use `reg_sc` without manual setup.

Running the cell gives six thousand seven hundred sixty-three observations. The coefficient on `jc` is about zero point zero six six seven, while the coefficient on `univ` is about zero point zero seven six nine. Thus the point estimate for four-year college is larger by about zero point zero one zero two in log-wage units. Both individual coefficients are estimated precisely relative to zero, but individual significance is not our question.

We want to know whether the difference `jc` minus `univ` is sufficiently negative to distinguish the two effects, accounting for sampling uncertainty and their covariance. A raw comparison of zero point zero six six seven and zero point zero seven six nine does not answer that because both are random estimates. The next tab extracts the variance-covariance matrix needed to measure the uncertainty of their difference.

</details>

```{webr-r}
#| autorun: true
#--- import data ---#
data("twoyear", package = "wooldridge")

(
  reg_sc <- fixest::feols(lwage ~ jc + univ + exper, data = twoyear) # OLS
)
```

#### Variance covariance matrix

<details class="transcript">
<summary>Transcript</summary>

The variance-covariance matrix supplies the uncertainty information that an ordinary coefficient table hides. The definition callout uses `VCOV` sub i comma i for a diagonal entry. That entry is the estimated variance of coefficient estimator i. An off-diagonal entry, `VCOV` sub i comma j, is the estimated covariance between coefficient estimators i and j.

The code calls `vcov(reg_sc)`, assigns the result to `vcov_sc`, and prints it because the assignment is wrapped in parentheses. The argument `reg_sc` is the fitted model from the previous tab. The cell runs automatically so later calculations can index the stored matrix.

Look at the output's structure. It is square, with one row and column for every estimated coefficient, including the intercept. It is symmetric because covariance of x with y equals covariance of y with x. The diagonal values are positive estimated variances. Taking the square root of the `jc` diagonal produces the `jc` standard error shown in the regression output; the same is true for every other coefficient. The off-diagonal cells show how pairs of coefficient estimates move together across repeated samples.

The examples below the cell use positional indexing. Read `vcov_sc` bracket two comma two as row two, column two, the variance of beta `jc` hat because the intercept occupies position one. Read bracket two comma three as the covariance between beta `jc` hat and beta `univ` hat.

</details>

:::{.callout-note title="Definition"}
Variance covariance matrix is a matrix where

+ $VCOV_{i,i}$: the variance of $i$th variable's coefficient estimator
+ $VCOV_{i,j}$: the covariance between $i$th and $j$th variables' estimators
:::

<br>

You can get it by applying `vcov()` to regression results:

```{webr-r}
#| autorun: true
(
  vcov_sc <- vcov(reg_sc) # variance covariance matrix
)
```

<br>

For example, $vcov_{sc}[2, 2]$ is the variance of $\widehat{\beta}_{jc}$, and $vcov_{sc}[2, 3]$ is the covariance between $\widehat{\beta}_{jc}$ and $\widehat{\beta}_{univ}$.

#### Calculate the t-statistic

<details class="transcript">
<summary>Transcript</summary>

This tab constructs the contrast t-statistic in two stages. First, the top code cell tidies the regression and extracts the two estimates. `broom::tidy(reg_sc)` creates `reg_sc_td`, a tibble with one row per coefficient. Starting from that table, the first pipeline filters to `term == "jc"`, pulls the `estimate` column, and stores the resulting number as `beta_jc`. The second pipeline does the same for `term == "univ"`, storing `beta_univ`. The cell runs automatically so both values are ready for the calculation below.

The numerator is assigned as `beta_jc - beta_univ`, matching alpha hat equals beta one hat minus beta two hat. With the fitted results, it is about zero point zero six six seven minus zero point zero seven six nine, or minus zero point zero one zero two. The negative sign points in the direction of the alternative that the four-year effect is larger.

The denominator follows the variance formula exactly. Inside `sqrt`, `vcov_sc["jc", "jc"]` is the variance of beta `jc` hat. Subtract two times `vcov_sc["jc", "univ"]`, the covariance between the two estimates. Then add `vcov_sc["univ", "univ"]`, the variance of beta `univ` hat. Taking the square root converts the variance of the difference to its standard error.

Finally, `t_stat <- numerator / denominator` standardizes the estimated difference, and `t_stat` prints it. Running these data gives a statistic of roughly minus one point four seven. The statistic is in the predicted direction but not far enough into the tail for rejection at conventional levels. The next tab shows how a hypothesis-testing function performs the matrix calculation automatically.

</details>

Get coefficient estimates: 

```{webr-r}
#| autorun: true
#--- reg results ---#
reg_sc_td <- broom::tidy(reg_sc)

#--- get coefficients ---#
beta_jc <-
  reg_sc_td %>%
  filter(term == "jc") %>%
  pull(estimate)

beta_univ <-
  reg_sc_td %>%
  filter(term == "univ") %>%
  pull(estimate)
```

<br>

Calculate t-statistic:

```{webr-r}
#| autorun: true
numerator <- beta_jc - beta_univ
denominator <-
  sqrt(
    vcov_sc["jc", "jc"] - 2 * vcov_sc["jc", "univ"] + vcov_sc["univ", "univ"]
  )
t_stat <- numerator / denominator
t_stat
```

#### In practice

<details class="transcript inline">
<summary>Transcript</summary>

In practice, `car::linearHypothesis` can form and test the contrast without manually extracting coefficients and covariances. The syntax block shows two arguments. `regression_results` is the fitted model object, and `expression_of_the_test` describes the linear restriction. The `eval: false` option means the syntax example is displayed but not executed.

The interactive example calls `car::linearHypothesis(reg_sc, "jc=univ")`. The first argument is the unrestricted college regression. The quoted equation asks whether the coefficient on `jc` equals the coefficient on `univ`, which is algebraically the restriction `jc - univ = 0`. The function constructs the contrast, uses the model's variance-covariance matrix to obtain its standard error, and reports a Wald test. With this `fixest` object it prints a chi-squared statistic of about two point one five and a p-value around zero point one four two.

The slide then asks us to compare that statistic with `t_stat^2`. The following cell squares the signed t-statistic from the previous tab. Squaring removes the sign and gives the same numerical value, about two point one five, as the reported one-restriction Wald chi-squared statistic. That equality concerns the statistic. The chi-squared one reference distribution used in the printed Wald table is an asymptotic approximation.

Under the exact classical finite-sample assumptions, the corresponding one-restriction result is t with the residual degrees of freedom, and its square follows F with one numerator degree of freedom and the same residual denominator degrees of freedom. Chi-squared one is the large-sample limit of that F distribution. With six thousand seven hundred fifty-nine residual degrees of freedom, the resulting p-values and decisions are nearly indistinguishable, but they are not mathematically identical in every finite sample.

</details>

In practice, you can just use `car::linearHypothesis()` to implement t-test involving a linear combination of multiple coefficients.

**Syntax**

```{r}
#| eval: false
car::linearHypothesis(regression_results, expression_of_the_test)
```

<br>

**Example**

```{webr-r}
#| autorun: true
car::linearHypothesis(reg_sc, "jc=univ")
```

<br>

Please note that this method reports a Wald $\chi^2$ test. Its statistic equals the squared t-statistic for one restriction, but the $\chi^2_1$ reference distribution is an asymptotic approximation. Under the classical finite-sample assumptions, the exact result is $t_{df}^2 \sim F_{1,df}$.

Let's confirm this:

```{webr-r}
#| autorun: true
t_stat^2
```

<br>

Yes, this is the same as the $\chi^2$ value reported by `car::linearHypothesis(reg_sc, "jc=univ")`

:::{.callout-important}
The test results would be the same whether you run t-test or $\chi^2$ test.
:::

:::
<!--end of panel-->

:::
<!--end of panel-->

# Multiple Linear Restrictions: F-test

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Multiple Linear Restrictions: F-test


::: {.panel-tabset}

### Example

::: {.panel-tabset}

#### Model and hypothesis

<details class="transcript">
<summary>Transcript</summary>

This baseball example introduces a joint hypothesis involving three restrictions. The dependent variable is log salary, where salary is measured in 1993. The model includes years in the league, average games played per year, career batting average, home runs per year, and runs batted in per year. Beta zero is the intercept and beta one through beta five multiply those regressors in the order shown.

The variable list supplies the substantive meanings. `years` measures career length, and `gamesyr` measures how regularly a player appears. We treat those as controls. The remaining variables measure performance: `bavg` is career batting average, `hrunsyr` is home runs per year, and `rbisyr` is runs batted in per year.

The question asks whether the three performance measures have any effect on salary collectively after holding years and games per year fixed. The word "collectively" is essential. H zero sets beta three, beta four, and beta five all equal to zero at the same time. Under that restriction, none of the three variables adds explanatory power conditional on the controls. There are q equals three restrictions.

H one says H zero is not true. That means at least one of those three coefficients is nonzero; it does not require all three to be nonzero, and it does not specify their signs. This is one joint hypothesis about a vector of parameters, not three independent yes-or-no questions. The next tab asks why we need a new testing device for this collective claim.

</details>

**Model**

$$log(salary) =  \beta_0 + \beta_1 years + \beta_2 gamesyr + \beta_3 bavg + \beta_4 hrunsyr + \beta_5 rbisyr + u$$

+ $salary$: salary in 1993
+ $years$: years in the league
+ $gamesyr$: average games played per year
+ $bavg$: career batting average
+ $hrunsyr$: home runs per year
+ $rbisyr$: runs batted in per year

**Hypothesis**

Once years in the league and games per year have been controlled for, the statistics measuring performance ( $bavg$, $hrunsyr$, $rbisyr$) have no effect on salary collectively.

$H_0$: $\beta_3=0$, $\beta_4=0$, and $\beta_5=0$

$H_1$: $H_0$ is not true

#### Questions

<details class="transcript">
<summary>Transcript</summary>

How should we test the joint null from the previous tab? The first bullet restates its logical structure: H zero holds only when beta three, beta four, and beta five are all zero. If even one is nonzero, the joint null is false. That makes it tempting to run an ordinary t-test for each coefficient and combine the three conclusions.

</details>
  
How do we test this?

+ $H_0$ holds only if $\beta_3$, $\beta_4$, <span style = "color: red;"> and </span> $\beta_5$ are <span style = "color: red;"> all </span> zero.
+ Conduct t-test for each coefficient individually?

:::
<!--end of panel-->

### Individual t-test

::: {.panel-tabset}

#### Run regression

<details class="transcript">
<summary>Transcript</summary>

This cell loads the baseball data and fits the unrestricted model. `data("mlb1", package = "wooldridge")` loads `mlb1` from the Wooldridge package. The model formula places `log(salary)` on the left and `years`, `gamesyr`, `bavg`, `hrunsyr`, and `rbisyr` on the right. The logarithm is applied within the formula. `data = mlb1` supplies the observations, and the fitted `fixest` object is stored as `mlb_reg`.

The call to `broom::tidy(mlb_reg)` converts the coefficient results into a tibble and prints it. The cell has `autorun: true`, so the data and model objects are created for later tabs automatically. Ignore the comment's spelling and focus on the model itself: this is the full, unrestricted specification because none of the three performance coefficients is forced to zero.

Read the output rows for `bavg`, `hrunsyr`, and `rbisyr`. Their estimates are approximately zero point zero zero zero nine eight, zero point zero one four four, and zero point zero one zero eight. Their estimated standard errors are roughly zero point zero zero one one zero, zero point zero one six one, and zero point zero zero seven one seven. Dividing each estimate by its standard error gives t-statistics of about zero point eight nine, zero point nine zero, and one point five zero.

The corresponding two-sided p-values are about zero point three seven six, zero point three six nine, and zero point one three four. These values concern each coefficient separately, conditional on all other regressors remaining in the model. Before moving on, ask whether those individual results logically establish that the three coefficients are jointly zero. The next Question and Answer tabs make the distinction explicit.

</details>

```{webr-r}
#| autorun: true
#--- load the mlb data ---#
data("mlb1", package = "wooldridge")

#--- run a rergession ---#
mlb_reg <- fixest::feols(log(salary) ~ years + gamesyr + bavg
  + hrunsyr + rbisyr, data = mlb1)

#--- take a look at the results ---#
broom::tidy(mlb_reg)
```

#### Question

<details class="transcript">
<summary>Transcript</summary>

The same tidy output appears again so you can inspect it without flipping back. Go row by row through `bavg`, `hrunsyr`, and `rbisyr`. For each one, compare its two-sided p-value with ten percent, zero point one zero. Batting average has a p-value around zero point three seven six, home runs per year around zero point three six nine, and runs batted in per year around zero point one three four.

None is below ten percent. Equivalently, none of the absolute t-statistics exceeds the two-sided ten-percent critical value. Thus every individual test fails to reject that coefficient equals zero while controlling for the other performance measures and the two career controls.

</details>

```{webr-r}
#| autorun: true
broom::tidy(mlb_reg)
```

What do you find?

#### Answer

<details class="transcript">
<summary>Transcript</summary>

The result is deliberately surprising: none of the coefficients on batting average, home runs per year, or runs batted in per year is individually statistically significant, even at the ten-percent level. "Individually" means each t-test asks whether one coefficient is zero while the other two performance variables remain in the regression. The data do not estimate any one variable's unique partial effect precisely enough to reject zero at that threshold.

Do not convert those three decisions into the statement that performance collectively has no effect on salary. Failure to reject each coefficient separately is not evidence that all three population coefficients are exactly zero. The estimates may be correlated, and the group can carry substantial information even when the model cannot allocate that information precisely among its members.

The joint null is more demanding: beta `bavg`, beta `hrunsyr`, and beta `rbisyr` must all equal zero simultaneously. To evaluate it, compare the unrestricted model containing the three variables with a restricted model that imposes H zero by removing them. If the group contributes little, forcing the coefficients to zero should barely worsen the fit. If the group matters, the restricted model's residuals should become substantially larger.

The slide points specifically to SSR, the sum of squared residuals, or equivalently a comparison based on R squared.

</details>

None of the coefficients on `bavg`, `hrunsyr`, and `rbisyr` is statistically significantly different from 0 even at the 10% level!!

So, does this mean that they collectively have no impact on the salary of MLB players?

If you were to conclude that they do not have statistically significant impact jointly, you would turn out to be wrong!!

$SSR$ (or $R^2$) turns out to be useful for testing their impacts jointly.

:::
<!--end of panel-->

### F-test

In doing an F-test of the null hypothesis, we compare sum of squared residuals $(SSR)$ of two models:

<br>

**Unrestricted Model**

$$log(salary) = \beta_0 + \beta_1 years + \beta_2 gamesyr + \beta_3 bavg + \beta_4 hrunsyr + \beta_5 rbisyr + u$$

<br>

**Restricted Model**

$$log(salary) = \beta_0 + \beta_1 years + \beta_2 gamesyr + u$$

The coefficients on $bavg$, $hrunsyr$, and $rbisyr$ are restricted to be 0 following the null hypothesis. 

::: {.panel-tabset}
#### Question

<details class="transcript">
<summary>Transcript</summary>

The slide above this tab defines the two models used by the F-test. The unrestricted model contains years, games per year, batting average, home runs per year, and runs batted in per year. The restricted model keeps only years and games per year. Dropping the three performance variables imposes the joint null that their coefficients are all zero.

The question asks which model should have the larger sum of squared residuals if we fit both to the same observations. Recall the OLS objective: choose coefficient values that minimize the sum of squared residuals. The unrestricted model may choose any values for all five slope coefficients. The restricted model is forced to choose beta three, beta four, and beta five equal to zero.

Think through the constrained-minimization logic, then open the Answer tab.

</details>

If the null hypothesis is indeed true, then what do you think is going to happen if you compare the $SSR$ of the two models? Which one has a bigger $SSR$?

#### Answer

<details class="transcript">
<summary>Transcript</summary>

The restricted model's SSR must be at least as large as the unrestricted model's SSR. The unrestricted OLS fit chooses coefficients from a larger set. It could reproduce the restricted fit exactly by setting the three performance coefficients to zero, so it can never do worse. It will do better whenever nonzero choices reduce the residual sum of squares.

The slide phrases this as the restricted model having smaller explanatory power, which is the intuition. More precisely, the restricted SSR is greater than or equal to the unrestricted SSR. Equality is possible if the restrictions do not change the sample fit. Therefore SSR sub r minus SSR sub u is always nonnegative for these nested models estimated on the same data with the same dependent variable.

</details>

$SSR$ from the restricted model should be large because the restricted model has a smaller explanatory power than the unrestricted model.

:::
<!--end of panel-->

### SSR

**SSR of the unrestricted model**: $SSR_u$

```{webr-r}
#| autorun: true
#| code-track: 0.6
#--- run OLS ---#
res_u <- feols(log(salary) ~ years + gamesyr + bavg
  + hrunsyr + rbisyr, data = mlb1)

#--- SSR ---#
sum(res_u$residuals^2)
```

<br>

**SSR of the restricted model**: $SSR_r$

```{webr-r}
#| code-track: 0.6
#--- run OLS ---#
res_r <- feols(log(salary) ~ years + gamesyr, data = mlb1)

#--- SSR ---#
sum(res_r$residuals^2)
```

<br>

::: {.panel-tabset}
#### Question 1

<details class="transcript">
<summary>Transcript</summary>

The code above this tab has fitted both nested models and printed their sums of squared residuals. The unrestricted fit includes all five regressors and stores its residuals in `res_u`. `sum(res_u$residuals^2)` squares every residual and adds the squares to obtain SSR sub u. The restricted fit includes only years and games per year, and the analogous expression obtains SSR sub r.

As predicted, the restricted number is larger. The question asks what the difference SSR sub r minus SSR sub u measures. Compare the formulas for the two regressions. They use the same dependent variable, same observations, same controls, and same OLS criterion. The only change is whether batting average, home runs per year, and runs batted in per year are allowed to enter.

</details>

What does $SSR_r - SSR_u$ measure?

#### Answer 1

<details class="transcript">
<summary>Transcript</summary>

SSR sub r minus SSR sub u is the reduction in residual sum of squares produced by adding the three excluded performance variables to the restricted model. In that sense, it measures their joint contribution to explaining log salary after years and games per year are already controlled for.

It is not the unique contribution of any single performance variable, and it is not generally the sum of three separate one-variable improvements. The regressors may overlap in the information they carry. The unrestricted model chooses all three coefficients together, so the SSR difference captures their combined ability to fit the dependent variable, including contributions that cannot be cleanly allocated when the variables are correlated.

</details>

The contribution from the three excluded variables in explaining the dependent variable.

#### Question 2

<details class="transcript">
<summary>Transcript</summary>

Now that we know what SSR sub r minus SSR sub u measures, ask whether the observed contribution is large enough to conclude that the excluded variables matter in the population.

</details>

Is the contribution large enough to say that the excluded variables are important? 

#### Answer 2

<details class="transcript">
<summary>Transcript</summary>

We cannot tell at this point because we do not know the null distribution of the raw difference SSR sub r minus SSR sub u. Computing a sample number completes only the calculation step. Without a reference distribution, words such as "large" or "unlikely" have no formal meaning.

This returns us to step two of the five-step recipe. We must ask what values an appropriately standardized improvement in fit would take if the three population coefficients were all zero. Under H zero, extra regressors can still reduce sample SSR by fitting random noise, so a positive improvement is expected. The question is whether the improvement is unusually large relative to that chance behavior.

</details>

Cannot tell at this point because we do not know the distribution of the difference!

:::
<!--end of panel-->

### F-test in general

**Setup**

Consider a following general model:

$$
y = \beta_0 +\beta_1 x_1 + \dots+\beta_k x_k +u
$$

Suppose we have $q$ restrictions to test: that is, the null hypothesis states that $q$ of the variables have zero coefficients.

$$H_0: \beta_{k-q+1} =0, \beta_{k-q+2} =0, \dots, \beta_k=0$$

When we impose the restrictions under $H_0$, the restricted model is the following:

$$y = \beta_0 +\beta_1 x_1 + \dots+\beta_{k-q} x_{k-q} + u$$


**F-statistic**

<span style = "color: red;"> If </span> the null hypothesis is true, then,

$$F = \frac{(SSR_r-SSR_u)/q}{SSR_u/(n-k-1)} \sim F_{q,n-k-1}$$

+ $q$: the number of restrictions
+ $n-k-1$: degrees of freedom of residuals

::: {.panel-tabset}

#### Question 1

<details class="transcript">
<summary>Transcript</summary>

Before using the general F formula above, check its sign. The statistic is open parenthesis SSR sub r minus SSR sub u close parenthesis divided by q, all divided by SSR sub u over n minus k minus one. Here q is the positive number of restrictions, and n minus k minus one is the positive residual degrees of freedom.

The denominator contains SSR sub u, a sum of squared residuals. Except for a perfect fit it is positive, and dividing by positive residual degrees of freedom preserves that sign. It estimates the error variance in the unrestricted model.

The numerator contains the difference between restricted and unrestricted SSR. For nested OLS models fitted on the same observations, the unrestricted model's feasible coefficient choices include the restricted choices. Therefore the minimized unrestricted SSR cannot exceed the restricted SSR. The difference is nonnegative, and division by q leaves it nonnegative.

So what sign can F have? Recall that equality of the two SSR values is possible when the restrictions make no difference to the sample fit. Answer with that boundary case in mind, then open the Answer tab.

</details>

Is the above $F$-statistic always positive?

#### Answer 1

<details class="transcript">
<summary>Transcript</summary>

The F-statistic is nonnegative. The restricted sum of squared residuals can never be below the unrestricted sum for nested OLS models estimated on the same sample, so SSR sub r minus SSR sub u is greater than or equal to zero. The remaining quantities in the ratio are positive under an ordinary nondegenerate regression.

If the restriction does not change the fitted values at all, the two SSR values are equal and F is exactly zero. Otherwise, the restricted fit is worse and F is positive. This is the constrained-minimum argument from the previous tabs, not an empirical pattern that happens only most of the time.

</details>

Yes, because $SSR_r-SSR_u$ is always non-negative: the restricted model can never fit better than the unrestricted one.

#### Question 2

<details class="transcript">
<summary>Transcript</summary>

Read the F formula one more time to determine its direction. The numerator is the loss of fit from imposing q restrictions, averaged over those restrictions. The denominator is the unrestricted residual mean square, which measures the remaining noise per residual degree of freedom.

Choose between "greater" and "smaller" from the formula itself.

</details>

The greater the joint contribution of the $q$ variables, the (greater or smaller) the $F$-statistic?

#### Answer 2

<details class="transcript">
<summary>Transcript</summary>

Greater. When the q variables contribute more jointly, removing them increases the restricted model's residual sum of squares by more. The gap SSR sub r minus SSR sub u becomes larger. After dividing by q and by the unrestricted residual mean square, that larger standardized loss produces a larger F-statistic.

A small F has the opposite interpretation. It says the restricted and unrestricted models fit similarly after accounting for the number of restrictions and noise. Dropping the variables cost little, so the data are compatible with their coefficients being jointly zero. We then fail to reject H zero.

</details>

Greater.

:::
<!--end of panel-->

### F-test steps

<details class="transcript">
<summary>Transcript</summary>

This slide puts the F reference distribution beside the testing recipe. Start with the figure on the left. The horizontal axis shows nonnegative F values and the vertical axis shows probability density. The two curves are different F distributions: F with two numerator and one hundred denominator degrees of freedom, and F with four numerator and four hundred denominator degrees of freedom. Both are right-skewed and live only on zero and positive values, consistent with the SSR-ratio construction. Their shapes differ because an F distribution requires two degrees-of-freedom parameters, so there is no single universal F curve.

Now read the five steps on the right. First define H zero precisely, including every restriction. Second estimate the unrestricted model and the restricted model on the same observations, then obtain SSR sub u and SSR sub r. The unrestricted model contains all coefficients; the restricted model imposes H zero. Third calculate F as the average SSR increase per restriction divided by the unrestricted residual mean square.

Fourth choose a significance level and find the matching upper-tail critical value from F with q numerator degrees of freedom and n minus k minus one denominator degrees of freedom. Here q is the number of restrictions and k counts regressors in the unrestricted model excluding its intercept.

Fifth compare the observed statistic with the critical value. Reject H zero if F is greater, because only a large standardized loss of fit is evidence that the restrictions are false. Otherwise fail to reject. This is the same general five-step logic from the first slide, specialized to a statistic that tests several linear restrictions at once. The next tabs implement each calculation for the baseball model.

</details>

::: {.columns}
::: {.column width="60%"}
**F-distribution**

```{r echo = F}
x <- seq(0, 4, length = 1000)
y_1 <- df(x, df1 = 2, df2 = 100)
y_2 <- df(x, df1 = 4, df2 = 400)

f_plot <- data.table(x = x, "F(2,100)" = y_1, "F(4,400)" = y_2) %>%
  melt(id.var = "x")

ggplot(data = f_plot) +
  geom_line(aes(x = x, y = value, color = variable)) +
  scale_color_discrete(name = "") +
  theme(
    legend.position = "bottom"
  ) +
  theme_lecture
```
:::
<!--end of the 1st column-->
::: {.column width="40%"}
**F-test steps**

+ Define the null hypothesis
+ Estimate the unrestricted and restricted models to obtains their $SSR$
+ Calculate $F$-statistic
+ Define the significance level and corresponding critical value according to the F distribution with appropriate degrees of freedoms
+ Reject if your $F$-statistic is greater than the critical value, otherwise do not reject
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

### F-test by hand

::: {.panel-tabset}

#### Step 1 

<details class="transcript">
<summary>Transcript</summary>

Step one fits the two nested models and obtains their sums of squared residuals. In the unrestricted call, `feols` regresses log salary on years, games per year, batting average, home runs per year, and runs batted in per year, using `mlb1`. The line break after `gamesyr +` is only formatting; the plus sign continues the formula. The fitted object is stored as `reg_u`.

`reg_u$residuals` extracts the vector of unrestricted residuals. Raising it to the second power squares each element, and `sum` adds those squares. The result is stored as `SSR_u`. This is the minimized residual sum of squares when all five slope coefficients are free.

The restricted call regresses the same log salary outcome on only years and games per year. It uses the same dataset and estimator but omits `bavg`, `hrunsyr`, and `rbisyr`, thereby imposing the three zero restrictions in H zero. The fitted object is `reg_r`, and squaring and summing `reg_r$residuals` gives `SSR_r`.

Compare the formulas carefully. The dependent variable, observations, controls, intercept treatment, and OLS method are identical. Only the three jointly tested variables differ. That nested structure is required for interpreting SSR sub r minus SSR sub u as the fit lost by imposing the null. Both model cells use earlier-loaded data, and `autorun: true` ensures the objects exist for Step 2, where the raw gap is converted into an F-statistic.

</details>

Estimate the unrestricted and restricted models

```{webr-r}
#| autorun: true
#--- unrestricted model ---#
reg_u <- feols(log(salary) ~ years + gamesyr +
  bavg + hrunsyr + rbisyr, data = mlb1)

SSR_u <- sum(reg_u$residuals^2)

#--- restricted model ---#
reg_r <- feols(log(salary) ~ years + gamesyr, data = mlb1)

SSR_r <- sum(reg_r$residuals^2)
```

#### Step 2

<details class="transcript">
<summary>Transcript</summary>

Step two assembles the F-statistic from the objects created in Step 1. `df_q <- 3` records q, the number of restrictions: one each for batting average, home runs per year, and runs batted in per year. It is not a residual degree of freedom; it will be the numerator degrees of freedom of the F distribution.

`degrees_freedom(reg_u, "resid")` asks the unrestricted model for its residual degrees of freedom and stores the answer as `df_ur`. The unrestricted model is the relevant one because its SSR estimates the remaining error variance after all coefficients are freely estimated. With the available baseball observations and five regressors plus an intercept, this value is n minus five minus one.

`F_stat_num <- (SSR_r - SSR_u) / df_q` calculates the numerator mean square. Subtracting measures the fit lost under the restrictions, and dividing by three expresses the average loss per restriction. `F_stat_denom <- SSR_u / df_ur` calculates the unrestricted residual mean square. Under homoskedastic classical assumptions, this is sigma hat squared, the estimator of the error variance.

Finally, `F_sta <- F_stat_num / F_stat_denom` divides the restriction mean square by the residual mean square, and the last line prints `F_sta`. The ratio is unit-free because both pieces are measured in squared dependent-variable units. A value near the ordinary null range means the fit loss is modest relative to noise; a large value means the restrictions caused an unusually costly loss. Step 3 will use `qf` to decide how large is large at the chosen significance level.

</details>

Calculate F-stat

```{webr-r}
df_q <- 3 # the number of restrictions
df_ur <- degrees_freedom(reg_u, "resid") # degrees of freedom for the unrestricted model
F_stat_num <- (SSR_r - SSR_u) / df_q # numerator of F-stat
F_stat_denom <- SSR_u / df_ur # denominator of F-stat
F_sta <- F_stat_num / F_stat_denom # F-stat
F_sta
```

#### Steps 3 and 4

<details class="transcript">
<summary>Transcript</summary>

Steps three and four choose the threshold and make the decision. `alpha <- 0.05` sets the significance level to five percent. In this upper-tail test, the rejection region occupies the largest five percent of F values, so the critical value has ninety-five percent of the null distribution to its left.

`qf(1 - alpha, df1 = df_q, df2 = df_ur)` computes that boundary. `qf` is the F quantile function. Its first argument is zero point nine five. `df1` receives the three numerator degrees of freedom, one per restriction, and `df2` receives the unrestricted residual degrees of freedom. The result is assigned to `c_value` and printed. Do not divide alpha by two, because F is nonnegative and only the upper tail is a rejection region.

The second cell evaluates `F_sta > c_value`. This logical comparison returns `TRUE` for the baseball data, meaning the observed F-statistic lies beyond the five-percent critical value. Equivalently, its upper-tail p-value is below zero point zero five. We reject the joint null that all three performance coefficients are zero.

Read the Observation callout carefully. The variables have statistically significant impacts jointly, even though none of their individual two-sided t-tests rejected at ten percent. That is not a logical contradiction.

</details>

Find the critical value

```{webr-r}
#| code-track: 0.6
alpha <- 0.05 # 5%  significance level
c_value <- qf(1 - alpha, df1 = df_q, df2 = df_ur)
c_value
```

Is F-stat > critical value?

```{webr-r}
#| code-track: 0.6
F_sta > c_value
```

:::{.callout-note title="Observation"}
So, the performance variables have statistically significant impacts on salary <span style = "color: red;"> jointly</span>!!

What happened?
:::

:::
<!--end of panel-->

### F-test (easier way)

::: {.panel-tabset}

#### How

<details class="transcript inline">
<summary>Transcript</summary>

This is the practical interface for linear restrictions. `car::linearHypothesis` receives a fitted unrestricted regression and a description of H zero, then constructs the appropriate Wald test using the model's coefficient vector and variance-covariance matrix. For a classical OLS model and restrictions that set a group of coefficients to zero, the test corresponds to the restricted-versus-unrestricted comparison developed by hand.

The syntax block has two arguments. `regression` is the fitted model object containing the unrestricted specification. You do not pass the restricted fit. `hypothesis` is text describing one or more linear equations that must all hold under H zero. Because the code chunk has `eval = F`, it only illustrates syntax and does not try to execute placeholder names.

The example `c("x1 = 0", "x2 = 1")` uses `c` to create a character vector with two strings. The first restriction says the coefficient named `x1` equals zero. The second says the coefficient named `x2` equals one. Together they define q equals two restrictions. Quotation marks matter because the equations are parsed by the function rather than evaluated as ordinary R assignments.

The function is not limited to zero restrictions. Flip to the Demonstration tab to apply the function to the three baseball restrictions.

</details>

You can use the `car::linearHypothesis()` function from the `car` package. 

<br>

**Syntax**

```{r eval = F}
car::linearHypothesis(regression, hypothesis)
```

+ `regression`: the name of the regression results of the unrestricted model

+ `hypothesis`: text of the null hypothesis. For example,

`c("x1 = 0", "x2 = 1")` means the coefficients on $x1$ and $x2$ are $0$ and $1$, respectively

#### Demonstration

<details class="transcript">
<summary>Transcript</summary>

This code applies the convenience function to the baseball regression. `mlb_reg` is the unrestricted model containing years, games per year, batting average, home runs per year, and runs batted in per year. The second argument is a character vector made with `c`, containing `"bavg=0"`, `"hrunsyr=0"`, and `"rbisyr=0"`. The three strings jointly impose exactly the null stated at the start of the section.

`car::linearHypothesis` reads the coefficient names from the model, forms the three-restriction contrast, obtains its estimated covariance, and reports the joint test. With this `fixest` object, the displayed method reports a Wald chi-squared statistic of about twenty-eight point seven with three restrictions and a p-value around zero point zero zero zero zero zero two seven. That p-value is far below five percent, so the joint null is rejected.

Compare the conclusion with the by-hand F calculation. Both show that the performance variables matter collectively. Depending on the method used for the fitted object, the printed Wald table may use a chi-squared reference rather than the exact finite-sample F reference from the SSR derivation, so inspect the output labels rather than assuming every `linearHypothesis` call prints the same statistic. With a large residual degrees of freedom, the decisions here are the same.

</details>

The following code test if $\beta_{bavg} = 0$, $\beta_{hrunsyr} = 0$, and $\beta_{rbisyr} = 0$. 

```{webr-r}
#| autorun: true
#--- F-test ---#
car::linearHypothesis(mlb_reg, c("bavg=0", "hrunsyr=0", "rbisyr=0"))
```

:::
<!--end of panel-->

:::
<!--end of panel-->

## Simulation (multicollinearity, t-test, and F-test)


::: {.panel-tabset}

### Data generation

<details class="transcript">
<summary>Transcript</summary>

This simulation makes the multicollinearity mechanism visible because we control the data-generating process. `N <- 300` sets the sample size. `mu <- runif(N)` draws three hundred independent Uniform zero-to-one values and stores a common component that will enter two regressors.

`x1` equals zero point one times a new Uniform draw plus two times `mu`. `x2` is built the same way with its own small Uniform component but the identical large shared term. Because the coefficient two on `mu` dominates the coefficient zero point one on their separate noise, `x1` and `x2` are extremely highly correlated. `x3 <- runif(N)` draws a third regressor independently of that construction. `u <- rnorm(N)` creates standard Normal errors.

The outcome line is crucial: y equals one plus x one plus x two plus x three plus u. The intercept and all three slope coefficients are exactly one in the population process we chose. Thus x one and x two genuinely matter. Any failure to detect them individually cannot be explained by their true coefficients being zero.

`data.table(y = y, x1 = x1, x2 = x2, x3 = x3)` combines the vectors into a table for regression. The final cell calls `cor(x1, x2)`, which computes their sample correlation. It will normally be extremely close to one; one run gave about zero point nine nine seven. Because no random seed is set, your exact values and later p-values change each time the page creates new draws.

</details>

```{webr-r}
N <- 300 # num observations
mu <- runif(N) # term shared by indep vars 1 and 2
x1 <- 0.1 * runif(N) + 2 * mu # indep 1
x2 <- 0.1 * runif(N) + 2 * mu # indep 2
x3 <- runif(N) # indep 3
u <- rnorm(N) # error
y <- 1 + x1 + x2 + x3 + u # generate y
data <- data.table(y = y, x1 = x1, x2 = x2, x3 = x3) # combine into a data.table
```

<br>

`x1` and `x2` are highly correlated with each other:

```{webr-r}
cor(x1, x2) # correlation between x1 and x2
```


### Regression

<details class="transcript">
<summary>Transcript</summary>

This cell estimates the unrestricted simulation model. `fixest::feols(y ~ x1 + x2 + x3, data = data)` regresses y on all three generated regressors and stores the fit in `reg_u`. `broom::tidy(reg_u)` then prints one row per coefficient with estimates, standard errors, t-statistics, and two-sided p-values for zero-null tests.

In a typical run, both individual t-tests fail to reject zero, as the sentence below the cell states. In the run used to verify this behavior, the x one estimate was about two point two seven with a standard error about one point two nine and p-value about zero point zero seven nine, while the x two estimate was about minus zero point one four with the same size standard error and p-value about zero point nine one. Exact results vary because the simulation has no fixed seed, so occasionally one individual test can cross a conventional threshold.

The population truth is not uncertain here: the y-generating line assigned both slopes the value one.

</details>

```{webr-r}
reg_u <- fixest::feols(y ~ x1 + x2 + x3, data = data) # OLS
broom::tidy(reg_u) # results
```

<br>

Both `x1` and `x2` are statistically insignificant individually. 

### F-test

<details class="transcript">
<summary>Transcript</summary>

This tab tests H zero: beta one equals zero and beta two equals zero jointly. The unrestricted model `reg_u` already includes x one, x two, and x three. `sum(reg_u$residuals^2)` computes and stores its residual sum of squares as `SSR_u`.

The restricted model must impose both zero restrictions, so `feols(y ~ x3, data = data)` drops x one and x two but retains x three and the intercept. Its residual sum of squares is stored as `SSR_r`. `degrees_freedom(reg_u, "resid")` obtains the unrestricted residual degrees of freedom, which are three hundred minus three regressors minus one intercept, or two hundred ninety-six.

The F formula has two restrictions. The code calculates open parenthesis SSR sub r minus SSR sub u close parenthesis divided by two, then divides by SSR sub u over `df_resid`. This is the average loss of fit per excluded variable relative to the unrestricted residual mean square. `alpha <- 0.05` sets a five-percent level, and `qf(1 - alpha, df1 = 2, df2 = df_resid)` gets the ninety-fifth percentile of F with two and two hundred ninety-six degrees of freedom. Parentheses print the critical value, which is about three point zero three.

The final cell prints `F_stat` and the logical comparison `F_stat > c_value`. In the verification run, F was about two hundred fifty-three and the comparison was `TRUE`; exact values change with the random draws. The statistic is typically enormous because removing x one and x two discards their strong combined signal. Thus the joint test detects what the individual t-tests often miss: together these correlated regressors have an important effect on y. This matches the known data-generating process, where both slopes equal one.

</details>

```{webr-r}
#| code-track: 0.6
#--- unrestricted ---#
SSR_u <- sum(reg_u$residuals^2)

#--- restricted ---#
reg_r <- fixest::feols(y ~ x3, data = data)
SSR_r <- sum(reg_r$residuals^2)

#--- degrees of freedom of residuals ---#
df_resid <- degrees_freedom(reg_u, "resid")

#--- F ---#
F_stat <- ((SSR_r - SSR_u) / 2) / (SSR_u / df_resid)

#--- critical value ---#
alpha <- 0.05
(
  c_value <- qf(1 - alpha, df1 = 2, df2 = df_resid)
)
```

<br>

The F-statistic for the hypothesis testing is:

```{webr-r}
#| code-track: 0.6
#--- F > critical value? ---#
F_stat
F_stat > c_value
```

<br>

The F-statistic is very high, meaning they collectively affect the dependent variable significantly.


### Important

<details class="transcript">
<summary>Transcript</summary>

This tab explains the different individual and joint results. The opening paragraphs say the estimated standard errors for x one and x two are high because the regressors are almost copies of one another. Holding x two fixed leaves very little independent variation in x one with which to identify beta one, and the same is true in reverse. Small changes in the random sample can shift the apparent contribution between their two coefficient estimates.

Collectively, however, x one and x two contain a strong signal for y. The restricted model removes both, so it cannot use their shared variation at all. Its fit worsens sharply, and the F-test detects that loss. The test does not need to decide which member of the pair deserves which part of the effect before asking whether the pair matters.

The callout connects this to Lecture 04's variance formula. Variance beta j hat is proportional to one over open parenthesis one minus R squared sub j close parenthesis, where R squared sub j comes from regressing x sub j on the other covariates. When x one is almost perfectly predicted by x two, R squared sub j approaches one. Then one minus R squared sub j approaches zero, the variance denominator shrinks, and the coefficient variance grows.

Standard error is the square root of variance, so the inflated variance produces a large standard error. A zero-null t-statistic divides the estimate by that standard error. A larger denominator pulls the statistic toward zero in magnitude and raises its p-value. That is the insignificance seen on the Regression tab. Multicollinearity has not made OLS systematically wrong under the data-generating assumptions. It has made the separate effects too imprecise to distinguish reliably.

</details>

The standard error estimates of the coefficients on `x1` and `x2` are very high because they are so highly correlated that the estimation had a hard time distinguishing their individual impacts.

But collectively they have large impacts, and the $F$-test was able to detect the statistical significance of their impacts <span style = "color: red;"> collectively</span>.

<br>

:::{.callout-important title="This is lecture 04, seen from the other side"}
The data generating process here is Case 2 from the multicollinearity lecture: two regressors sharing a common term. There we showed that correlation inflates $Var(\widehat{\beta}_j)$ through the $1-R^2_j$ term in the denominator.

A large variance means a large standard error, and a large standard error means a <span style = "color: red;"> small </span> t-statistic &mdash; which is exactly the insignificance you see on the `Regression` tab. Multicollinearity did not make the estimates wrong; it made them too imprecise to distinguish individually. The $F$-test sidesteps the problem entirely by never asking about them one at a time.
:::

:::
<!--end of panel-->

## MLB example

::: {.panel-tabset}

### Correlation check

<details class="transcript">
<summary>Transcript</summary>

Return to the baseball data and check whether correlated regressors plausibly explain the individual-versus-joint pattern. The code starts with `dplyr::select(mlb1, bavg, hrunsyr, rbisyr)`. `select` keeps only the three performance columns from `mlb1`. The pipe passes that three-column data frame to `cor`, which computes the pairwise correlation matrix. The cell runs automatically, and `code-track: 0.6` affects only its on-slide layout.

Read the matrix by row and column. Every diagonal entry is one because a variable is perfectly correlated with itself, and the off-diagonal entries appear twice because correlation is symmetric. Running the code gives a correlation of about zero point one nine between batting average and home runs per year, about zero point three three between batting average and runs batted in per year, and about zero point eight nine between home runs per year and runs batted in per year.

The strongest relationship is therefore between `hrunsyr` and `rbisyr`. That makes substantive sense. Every home run produces at least one run batted in for the hitter, and power hitters tend to drive in other runners as well. Those two measures contain heavily overlapping information. Batting average is related much less strongly to either one.

</details>

Here is the correlation coefficients between the three variables:

```{webr-r}
#| autorun: true
#| code-track: 0.6
dplyr::select(mlb1, bavg, hrunsyr, rbisyr) %>% cor()
```

As you can see, `rbisyr` and `hrunsyr` are highly correlated with each other. 

They are not so highly correlated with `bavg`.



### Multiple coefficients again

<details class="transcript">
<summary>Transcript</summary>

This tab connects the earlier coefficient-difference test with the F-test. A restriction such as beta `jc` minus beta `univ` equals zero is one linear restriction, so q equals one. Under the classical finite-sample assumptions, the F statistic for that single restriction equals the corresponding t-statistic squared and follows F with one numerator degree of freedom and the residual denominator degrees of freedom. This is why the two-sided t-test and one-restriction F-test produce the same reject-or-fail-to-reject decision.

The code first refits the college model with log wage on `jc`, `univ`, and `exper`, storing it as `reg_sc`. It then calls `car::linearHypothesis(reg_sc, c("jc-univ=0"))`. The character vector contains one equation, so the function tests equality of the two college coefficients and stores its table in `F_test`.

The final line extracts `F_test$Chisq[2]`. The dollar sign selects the `Chisq` column and bracket two selects the test row rather than the restricted-model row. `sqrt` takes the positive square root. Because the reported one-restriction Wald chi-squared statistic is numerically the squared contrast t-statistic, this returns the magnitude of the t-statistic, about one point four seven for these data.

Be careful about two details when comparing with the earlier hand calculation. Squaring and taking the ordinary square root loses the sign, so the code recovers the absolute t-statistic, not whether `jc - univ` was positive or negative. Also, this `fixest` method labels its Wald reference as chi-squared, which is asymptotic; the exact finite-sample identity uses F with one and the residual degrees of freedom. With thousands of residual degrees of freedom, their critical values are very close. The important conceptual connection is that a one-restriction equality test contains the same evidence as the corresponding two-sided t-test.

</details>

The test of a linear combination of the parameters we looked at earlier is a special case of F-test where the number of restriction is 1. 

It can be shown that square root of $F_{1, df}$ follows the $t_{df}$ distribution.

So, we can actually use the $F$-test for this type of hypothesis testing because $F_{1,n-k-1} \sim t_{n-k-1}^2$.

```{webr-r}
#| code-track: 0.65
#--- OLS ---#
reg_sc <- feols(lwage ~ jc + univ + exper, data = twoyear)

#--- F-test ---#
F_test <- car::linearHypothesis(reg_sc, c("jc-univ=0"))

#--- t-statistic ---#
sqrt(F_test$Chisq[2])
```

<br>

Check with the previous slide of the same t-test and confirm that the t-statistics we got there and here are the same.


:::
<!--end of panel-->

## Everything in this lecture is one idea

::: {.panel-tabset}

### The recipe

<details class="transcript">
<summary>Transcript</summary>

This final section returns to the five-step recipe from the opening slide. First write H zero and H one in parameter language. Be explicit about whether the alternative is directional and whether one restriction or several must hold simultaneously. The hypothesis determines what counts as an extreme result.

Second, choose a test statistic and establish its distribution assuming H zero is true. That conditional phrase in red remains the logical foundation. For a standardized coefficient or linear contrast under the classical assumptions, the reference is t. For several joint restrictions, the standardized model-fit comparison uses F.

Third, calculate the statistic from the observed data. This might mean subtracting a null coefficient value and dividing by a standard error, or comparing restricted and unrestricted sums of squared residuals. Fourth, choose a significance level. Together with the reference distribution, degrees of freedom, and sidedness, that choice determines a critical value.

Fifth, reject if the statistic is more extreme than the critical value. For a lower-tail t-test, "more extreme" means more negative; for an upper-tail test, more positive; for a two-sided t-test, larger in absolute value; for an F-test, larger. Otherwise fail to reject.

The blue sentence gives the useful compression: what changed across this lecture was the statistic we built and the null distribution used to judge it. The inferential logic did not change.

</details>

Every test we did followed the same five steps from the very first slide:

+ write down $H_0$ and $H_1$
+ find the distribution of a test statistic <span style = "color: red;"> assuming $H_0$ is true </span>
+ compute the statistic from your data
+ pick a significance level, which fixes a critical value
+ reject if your statistic is more extreme than that critical value

<br>

The only thing that changed from test to test was <span style = "color: blue;"> which statistic </span> you build and <span style = "color: blue;"> which distribution </span> you look it up in.

### The four tests

<details class="transcript">
<summary>Transcript</summary>

Read this table across each row. The first question asks whether one coefficient equals a specified null value beta j superscript zero. Its statistic is beta j hat minus that null value, divided by the estimated standard error of beta j hat. Under H zero it follows t with n minus k minus one degrees of freedom. `broom::tidy` supplies the default zero-null statistic, and a nonzero null can be handled by changing the numerator by hand.

The second row asks whether a linear combination alpha equals a specified value alpha superscript zero. The statistic has the identical structure: alpha hat minus its null value, divided by the estimated standard error of alpha hat. It has the same t reference distribution. What changes is the estimate and its uncertainty. For a difference of coefficients, the standard error needs their covariance, which is why the variance-covariance matrix appeared. `car::linearHypothesis` can perform that calculation.

The third row tests whether q coefficients are jointly zero. Its statistic is open parenthesis SSR sub r minus SSR sub u close parenthesis divided by q, over SSR sub u divided by n minus k minus one. Under H zero it follows F with q and n minus k minus one degrees of freedom. This statistic compares two model fits rather than standardizing one coefficient. It is the only row here that cannot be built from a single coefficient estimate and its individual standard error, and it can reveal group importance hidden by multicollinearity.

The fourth row is a confidence-interval question rather than a separate hypothesis test. It asks which parameter values are consistent with the estimate at a chosen confidence level. The construction uses the same t quantiles and degrees of freedom. `broom::tidy(conf.int = TRUE)` reports the endpoints.

Across the first two rows, remember the common pattern: estimate minus value under H zero, divided by the appropriate standard error.

</details>

| Question | Statistic | Distribution under $H_0$ | In R |
|---|---|---|---|
| Is one coefficient equal to a value? | $\dfrac{\widehat{\beta}_j - \beta_j^0}{\widehat{se(\widehat{\beta}_j)}}$ | $t_{n-k-1}$ | `broom::tidy()`, or by hand |
| Is a linear combination equal to a value? | $\dfrac{\widehat{\alpha} - \alpha^0}{\widehat{se(\widehat{\alpha})}}$ | $t_{n-k-1}$ | `car::linearHypothesis()` |
| Are $q$ coefficients <span style = "color: red;"> jointly </span> zero? | $\dfrac{(SSR_r-SSR_u)/q}{SSR_u/(n-k-1)}$ | $F_{q,\,n-k-1}$ | `car::linearHypothesis()` |
| What range of values is consistent with the data? | the CI, not a test | $t_{n-k-1}$ | `broom::tidy(conf.int = TRUE)` |

+ the second row needs $Cov(\widehat{\beta}_1,\widehat{\beta}_2)$, which is why the variance-covariance matrix appeared
+ the third row is the only one that cannot be built from a single coefficient and its standard error

### How they relate

<details class="transcript">
<summary>Transcript</summary>

The first bullet links t and F testing. When there is one linear restriction, the exact classical F statistic equals the corresponding t-statistic squared. Its null distribution is F with one numerator degree of freedom and n minus k minus one denominator degrees of freedom, matching the distribution of t squared. A two-sided t-test rejects for a large absolute t, while the F-test rejects for a large t squared, so the decisions and finite-sample p-values agree when the matching exact references are used.

The second bullet links confidence intervals and two-sided tests. A ninety-five percent confidence interval is exactly the set of null parameter values that a corresponding two-sided five-percent test would fail to reject. If the interval excludes zero, zero is too far from the estimate in standard-error units and the test of beta j equals zero rejects. If the interval contains zero, the same test fails to reject. More generally, any proposed null value can be checked by seeing whether it lies inside the interval.

Finish with the callout, the main applied lesson. An individually insignificant coefficient is not evidence of no effect. It says the test did not distinguish that one partial coefficient from its null value with the available precision. The baseball example and the simulation showed correlated variables that looked weak one at a time but mattered greatly together. When your scientific question concerns a group of related regressors, state the joint null and test it jointly. Do not infer a collective absence of effect by scanning individual stars.

</details>

Two connections worth remembering, because they mean there are fewer ideas here than there appear to be:

+ **$F$ with one restriction is just $t$ squared.** $F_{1,\,n-k-1} \sim t_{n-k-1}^2$, which is why `car::linearHypothesis()` gave the same answer as the t-test we did by hand.

+ **The confidence interval and the two-sided test are the same statement.** The 95% CI is exactly the set of null values you would <span style = "color: blue;"> fail </span> to reject at the 5% level. If a CI excludes zero, the t-test rejects $\beta_j = 0$; if it contains zero, it does not.

<br>

:::{.callout-important title="The one thing to carry into your own work"}
An individually insignificant coefficient is <span style = "color: red;"> not </span> evidence of no effect. The MLB players and the simulation both showed variables that looked useless one at a time and mattered a great deal together. When variables are correlated, test them <span style = "color: blue;">jointly</span>.
:::

:::
<!--end of panel-->
