---
title: "03: Monte Carlo Simulation"
format: 
  revealjs:
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to course website with lecture slides](https://tmieno2.github.io/Applied-Econometrics-MS-Quarto/lectures/)"
execute:
  echo: true
  out-width: 55%
webr:
  packages: ['fixest', 'dplyr', 'data.table', 'ggplot2', 'modelsummary']
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 10
    fig-height: 6
    out-width: 100%
filters:
  - webr
---

```{r}
#| include: false
library(dplyr)
library(data.table)
library(ggplot2)
```

# Monte Carlo Simulation: Introduction

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Monte Carlo Simulation: Introduction

::: {.panel-tabset}

### What is it?

<details class="transcript">
<summary>Transcript</summary>

Monte Carlo simulation is how we settle arguments in this course when the algebra gets hard to trust. The idea is simple: instead of proving something, you build a fake world where you already know the answer, then check whether the method recovers it. Read the three bullets. The first is confirming theory we already proved, which is what we do today. The second is the one I actually use in research, when I suspect a problem in my data and no theorem covers it. The third is why it is in this lecture: numbers are easier to believe than Greek letters.

</details>

It is a way to test econometric theories via simulation.

<br>

**How is it used in econometrics?**

+ confirm econometric theory numerically
    - OLS estimators are unbiased if $E[u|x]=0$ along with other conditions (theory)
    - I know the above theory is right, but let's check if it is true numerically
+ You kind of sense that something in your data may cause problems, but there is no proven econometric theory about what's gonna happen (I used MC simulation for this purpose a lot)
+ assist students in understanding econometric theories by providing actual numbers instead of a series of Greek letters

### Question

<details class="transcript">
<summary>Transcript</summary>

Here is the question that motivates everything. You want to know what happens to OLS when the zero conditional mean condition fails. Could you just take real data and check? Open the answer and read it, then think about why it is so damning. With real data you never see the error term, so you cannot tell whether it is correlated with x. And you never see the true beta, so even if you got an estimate you would have nothing to compare it against. Both halves of the comparison are invisible. That is precisely what simulation gives back to you.

</details>

Suppose you are interested in checking what happens to OLS estimators if $E[u|x]=0$ (the error term and $x$ are not correlated) is violated. 

:::{.callout-note title="Question"}
Can you use the real data to do this?
:::

<br>

<br>
<details>
  <summary>Answer</summary>
No because you will never observe either error term or true value of $\beta$s.
</details>


### Key part of MC simulations

<details class="transcript">
<summary>Transcript</summary>

Read the word in blue: you. You generate the data, and that single fact is what makes the whole method work. Two consequences, both on screen. First, you know the true parameter, because you typed it in. So when an estimate comes back you can measure how far off it is, which is impossible with real data. Second, you can change exactly one thing and hold everything else fixed. That is a controlled experiment on an econometric method. In the second example today we will break one condition on purpose and watch what happens, with nothing else moving.

</details>

<span style = "color: blue;"> You </span> generate data (you have control over how data are generated)

+ You know the true parameter unlike the real data generating process
+ You can change only the part that you want to change about data generating process and econometric methods with everything else fixed

:::
<!--end of panel-->

## Generating data

::: {.panel-tabset}

### RNG

<details class="transcript">
<summary>Transcript</summary>

Before we simulate anything we need to generate random-looking numbers. The standard generators used by statistical software are pseudorandom: a deterministic algorithm produces a sequence designed to have the statistical properties of random draws. Hardware sources of physical randomness do exist, but simulation normally uses pseudorandom generators because they are fast and reproducible. Run the first cell. `runif(5)` gives five draws from a uniform distribution between zero and one. Then run the second, which draws ten thousand and plots a histogram. Look at the approximately flat shape, because equal-length intervals between zero and one have equal probability.

</details>

**Pseudo random number generators (Pseudo RNG)**

Algorithms for generating a sequence of numbers whose properties <span style = "color: blue;"> approximate </span> the properties of sequences of random numbers

<br>

**Examples** 

Draw from a uniform distribution:

```{webr-r}
#| context: setup
# NOTE: a webr setup cell renders a loading widget wherever it sits, so keep it
# inside a content slide rather than on a section divider (or, worse, above the
# first heading, where it becomes an untitled slide of its own).
# Mirror of lecture_theme() in lectures/_lecture-theme.R. WebR runs in the
# browser, against a virtual filesystem that does not hold the repo, so it
# cannot source that file. Edit the file first, then run
#     Rscript lectures/check-lecture-theme.R
# which fails if this copy has drifted from it.
lecture_theme <- function(base_size = 16) {
  ggplot2::theme_bw(base_size = base_size) +
    ggplot2::theme(
      axis.text = ggplot2::element_text(size = base_size * 0.9),
      axis.title = ggplot2::element_text(size = base_size),
      legend.text = ggplot2::element_text(size = base_size * 0.9),
      legend.title = ggplot2::element_text(size = base_size * 0.9),
      plot.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.key = ggplot2::element_rect(fill = "transparent", colour = NA),
      # theme_bw's grey92 grid was chosen against a white panel and all but
      # disappears on the warm paper once the panel stops being opaque.
      panel.grid = ggplot2::element_line(colour = "grey87")
    )
}

# The WebR canvas is 10x6in at out-width 100% of its output track
# (lectures/_metadata.yml), the same canvas the knitr figures use. It asks for
# the whole track because the 420px height cap in notebook.scss is what really
# sizes it. 18pt rather than the 16pt knitr default: on a full slide both land
# at 700x420, where 18pt reads at 17.5px and 16pt at 15.5px, and these decks
# put figures in half-width columns where the smaller size gets hard to read.
# webR draws its canvas at 72 dpi no matter what `dpi` says, so on a 216-dpi
# canvas everything ggplot measures in points or mm comes out a third of its
# size. Scale ggplot2's mm-to-point constants and the theme by the same
# factor and the figure matches the knitr ones. Keep in step with `dpi` above.
webr_scale <- 216 / 72
assignInNamespace(".pt", ggplot2::.pt * webr_scale, "ggplot2")
assignInNamespace(".stroke", ggplot2::.stroke * webr_scale, "ggplot2")
theme_lecture <- lecture_theme(18 * webr_scale) +
  # base_size also sets the theme line widths, which are in mm and already
  # pass through the scaled .pt; pin them to what 18pt gives or they come
  # out 9x thick.
  ggplot2::theme(
    line = ggplot2::element_line(linewidth = 18 / 22),
    rect = ggplot2::element_rect(linewidth = 18 / 22)
  )
ggplot2::theme_set(theme_lecture)
```

```{webr-r}
runif(5) # default is min=0 and max=1
```

<br>

```{webr-r}
x <- runif(10000)
hist(x)
```

### Pseudo?

<details class="transcript">
<summary>Transcript</summary>

Here is the practical consequence of pseudo rather than truly random. The numbers are predetermined, and you can decide which ones you get by setting a seed. Run the demonstration cell, note the five numbers, and run it again. Identical, every time. Now think about the question at the bottom. The benefit is reproducibility. Your simulation results become something I can reproduce exactly, and something you can reproduce next month when you have forgotten what you did. Any simulation that goes into a paper or an assignment should have a seed at the top. Yours and mine should match.

</details>

Numbers drawn using pseudo random number generators are not truly random

+ What numbers you will get are pre-determined
+ What numbers you will get can be determined by setting a <span style = "color: red;"> seed </span>

**Demonstration**

```{webr-r}
set.seed(2387438)
runif(5)
```

<br>

**Question**

What benefits does setting a seed have?

### Normal Distribution

<details class="transcript">
<summary>Transcript</summary>

The Normal distribution is the workhorse, and rnorm draws from it. On the left, no arguments beyond the sample size, so you get the standard Normal: mean zero, variance one. On the right, mean two and variance four. Watch the notation carefully, because this trips people up constantly. When I write N of two comma four, the second number is the variance. In the code the argument is sd, standard deviation, which is the square root, so sd equals two. The math notation and the R argument disagree, and forgetting that is a classic source of wrong simulations.

</details>

::: {.columns}

::: {.column width="50%"}
$x \sim N(0, 1)$
```{webr-r}
# default is mean = 0,sd = 1
x <- rnorm(10000)
hist(x)
```
:::
<!--end of the 1st column-->

::: {.column width="50%"}
$x \sim N(2, 4)$
```{webr-r}
# N(mean, variance), so sd = sqrt(4) = 2
x <- rnorm(10000, mean = 2, sd = 2)
hist(x)
```

:::
<!--end of the 2nd column-->

:::
<!--end of the columns-->

:::
<!--end of panel-->

## R functions for often-used distributions  

::: {.panel-tabset}

### Distribution types

<details class="transcript">
<summary>Transcript</summary>

A reference tab, and worth remembering because the pattern generalises. On the left, the distributions R knows about, which is far more than you will ever need. On the right, the four things you can ask about any one of them, each a prefix. D for density, p for the probability of being below a value, q for the quantile, which reverses that question, and r for a random draw. Attach any prefix to any distribution name and it works: dbeta, punif, qchisq. Learn the four prefixes once and you have learned hundreds of functions. The next few tabs take each in turn.

</details>

::: {.columns}

::: {.column width="50%"}
+ Normal
+ Uniform
+ Beta
+ Chi-square
+ F
+ Logistic
+ Log-normal
+ many others
:::
<!--end of the 1st column-->
::: {.column width="50%"}
For each distribution, you have four different kinds of functions:

+ <span style = "color: red;"> `d`</span>`norm`: density function
+ <span style = "color: red;"> `p`</span>`norm`: distribution function
+ <span style = "color: red;"> `q`</span>`norm`: quantile function
+ <span style = "color: red;"> `r`</span>`norm`: random draw
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->



### `dnorm()`

<details class="transcript">
<summary>Transcript</summary>

Dnorm gives you the height of the density curve at a point, and the figure shows exactly that. The red curve is the standard Normal density. The two blue dots are dnorm at minus one and dnorm at two, with dashed lines dropping to the axis so you can see where each is measured. Read the numbers: about zero point two four at minus one, and about zero point zero five at two, because two is further out in the tail. One warning that matters later. This height is not a probability. Probabilities are areas under the curve, which is the next tab.

</details>

`dnorm(x)` gives you the height of the density function at $x$.

`dnorm(-1)` and `dnorm(2)`

```{webr-r}
#| context: output

x <- seq(-3, 3, length = 1000)
pdf <- dnorm(x)
plot_data <- data.table(y = pdf, x = x)
ggplot() +
  geom_line(data = plot_data, aes(y = y, x = x), color = "red") +
  geom_point(
    data = data.table(y = dnorm(2), x = 2),
    aes(y = y, x = x),
    color = "blue", size = 2
  ) +
  annotate(
    "text",
    label = paste0("dnorm(-1) = ", round(dnorm(-1), digits = 2)),
    y = 0.25, x = -2,
    size = 4.8
  ) +
  geom_line(
    data = data.table(y = seq(0, dnorm(-1), length = 20), x = -1),
    aes(y = y, x = x),
    linetype = 2
  ) +
  geom_point(
    data = data.table(y = dnorm(-1), x = -1),
    aes(y = y, x = x),
    color = "blue", size = 2
  ) +
  annotate(
    "text",
    label = paste0("dnorm(2) = ", round(dnorm(2), digits = 2)),
    y = 0.1, x = 2.5,
    size = 4.8
  ) +
  geom_line(
    data = data.table(y = seq(0, dnorm(2), length = 20), x = 2),
    aes(y = y, x = x),
    linetype = 2
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_lecture
```

### `pnorm()`

`pnorm(x)` gives you the probability that a single random draw is <span style = "color: red;"> less </span> than $x$.

::: {.panel-tabset}
#### pnorm(-1)

<details class="transcript">
<summary>Transcript</summary>

Now the p prefix, and here is the picture to keep in your head. Pnorm of minus one is the probability that a single draw comes out below minus one, and on the figure that is the shaded blue area to the left of minus one. Read the label: about zero point one six. So roughly sixteen percent of draws from a standard Normal fall below minus one. Compare this against the previous tab. Dnorm gave you the height of the curve at a point; pnorm gives you the area under it up to that point. Height versus area is the whole distinction.

</details>

```{webr-r}
#| context: output
ggplot() +
  geom_line(data = plot_data, aes(y = y, x = x), color = "red") +
  geom_ribbon(
    data = plot_data[x <= -1, ],
    aes(ymax = y, ymin = 0, x = x),
    fill = "blue",
    alpha = 0.4
  ) +
  annotate(
    "text",
    label = paste0("pnorm(-1) = ", round(pnorm(-1), digits = 2)),
    y = 0.05, x = 0,
    size = 4.8
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_lecture
```

#### pnorm(2)

<details class="transcript">
<summary>Transcript</summary>

Same function, a value on the other side. Pnorm of two shades everything to the left of two, which is almost the entire distribution, and the number is about zero point nine eight. So a draw below two happens about ninety eight percent of the time, which also tells you that a draw above two happens about two percent of the time. That subtraction is worth noticing now, because it is how you get right-tail probabilities out of a function that only reports left tails: one minus pnorm. You will be doing exactly that when we start computing p-values in the testing lecture.

</details>

```{webr-r}
#| context: output
ggplot() +
  geom_line(data = plot_data, aes(y = y, x = x), color = "red") +
  geom_ribbon(
    data = plot_data[x <= 2, ],
    aes(ymax = y, ymin = 0, x = x),
    fill = "blue",
    alpha = 0.4
  ) +
  annotate(
    "text",
    label = paste0("pnorm(2) = ", round(pnorm(2), digits = 2)),
    y = 0.05, x = 0,
    size = 4.8
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_lecture
```

#### Exercise

<details class="transcript">
<summary>Transcript</summary>

Your turn. The question asks for the probability that a draw from a Normal with mean one and standard deviation two falls below one. Type it in the work cell before you look, because there is something to notice here. The value you are asked about, one, is the mean of the distribution. And the Normal is symmetric around its mean, so exactly half the distribution sits below it. The answer is zero point five, and the standard deviation is irrelevant. Run the answer cell to confirm, then change the sd to anything you like and watch it stay at zero point five.

</details>

What is the probability that a single random draw from a Normal distribution with `mean = 1` and `sd = 2` is less than 1?

<br>

**Work here**

```{webr-r}

```

<br>

<details>
  <summary>Answer</summary>

```{webr-r}
pnorm(1, mean = 1, sd = 2)
```

+ 1 is the mean of this distribution, and the Normal distribution is symmetric around its mean, so the answer is 0.5 whatever the `sd` is

</details>

:::
<!--end of panel-->

### `qnorm()`

::: {.panel-tabset}

#### What is it?

<details class="transcript">
<summary>Transcript</summary>

The q prefix reverses the question you have been asking. Pnorm took a value and returned a probability. Qnorm takes a probability and returns the value. Read the definition on screen: qnorm of x gives you the number q such that a draw falls below q with probability x. That is why the second bullet says it is the inverse of pnorm, and why running one inside the other gives you back what you started with. The default is again mean zero and standard deviation one, so unless you say otherwise you are getting quantiles of the standard Normal.

</details>

`qnorm(x)`, where $0 < x < 1$, gives you the number $q$ such that a single random draw is less than $q$ with probability $x$.

+ it is the <span style = "color: red;"> inverse </span> of `pnorm()`: `pnorm(qnorm(x))` $= x$

We call the output of `qnorm(x)` the $100x\%$ quantile of the standard Normal distribution (because the default is `mean = 0` and `sd = 1` for `qnorm()`).

#### qnorm(0.95)

<details class="transcript">
<summary>Transcript</summary>

Here is the quantile you will meet most often in this course. Qnorm of zero point nine five is one point six four, and the figure shows what that means: shade ninety five percent of the area starting from the left, and the point where you stop is one point six four, marked on the axis. Remember this number. When we get to hypothesis testing, one point six four and its two-tailed cousin one point nine six are the critical values that decide whether you reject. They are not magic constants; they are just quantiles of the Normal distribution, computed exactly like this.

</details>

```{webr-r}
#| context: output
 
x <- seq(-3, 3, length = 1000)
pdf <- dnorm(x)
plot_data <- data.table(y = pdf, x = x)
ggplot() +
  geom_line(data = plot_data, aes(y = y, x = x), color = "red") +
  geom_ribbon(data = plot_data[x < 1.64, ], aes(ymax = y, ymin = 0, x = x), fill = "green", alpha = 0.3) +
  annotate(
    "text",
    label = "qnorm(0.95)=1.64",
    x = 2.2, y = 0.3,
    size = 4.8
  ) +
  geom_point(aes(y = 0, x = 1.64)) +
  annotate(
    "text",
    label = "1.64",
    x = 1.64, y = 0.03,
    size = 4.8
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_lecture
```

#### Exercise

<details class="transcript">
<summary>Transcript</summary>

Your turn again, and this one has a shortcut worth seeing. You want the eighty-eight percent quantile of a Normal with mean zero and standard deviation nine. Do it directly first, in the work cell. Then read the note under the answer. The standard Normal quantile is about one point one seven, and multiplying the standard deviation by nine multiplies this zero-mean Normal quantile by nine, giving about ten point six. More generally, a Normal quantile is $\mu+\sigma z_p$. That is why standardising a Normal variable is useful: it converts it to the standard Normal whose quantiles you know.

</details>

What is the 88% quantile of Normal distribution with `mean = 0` and `sd = 9`?

<br>

**Work here**

```{webr-r}

```

<br>

<details>
  <summary>Answer</summary>

```{webr-r}
qnorm(0.88, mean = 0, sd = 9)
```

+ `qnorm(0.88)` is about 1.17 for the standard Normal, and scaling the standard deviation by 9 scales the quantile by 9 as well

</details>

:::
<!--end of panel-->

:::
<!--end of panel-->

# Monte Carlo Simulation: In Practice

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

## Monte Carlo Simulation: Steps

<details class="transcript">
<summary>Transcript</summary>

Here is the recipe, and every simulation in this course follows it. Specify the data generating process. Generate data from it. Get an estimate. Repeat many times. Compare your estimates against the true parameter you chose in step one. Now the question at the bottom, which is the important one. Why repeat? Because a single estimate from a single sample tells you nothing. Your estimator is a random variable, and one draw from a random variable cannot reveal its distribution. Repeating thousands of times is how you get to see the whole sampling distribution instead of one point from it.

</details>

+ specify the data generating process
+ generate data based on the data generating process
+ get an estimate based on the generated data (e.g. OLS, mean)
+ repeat the above steps many many times
+ compare your estimates with the true parameter

**Question**

Why do the steps $1-3$ many many times?

## Monte Carlo Simulation: Example 1

::: {.panel-tabset}

### Problem

<details class="transcript">
<summary>Transcript</summary>

Let's start with something you already believe, so you can see the machinery work before we point it at something surprising. Is the sample mean an unbiased estimator of the expected value? The formal statement is on screen: the expected value of the sample average equals the expected value of x, where each observation is an independent draw from the same distribution. You have seen this proved. Today we check it numerically. If the method cannot confirm a result we are already sure about, we have no business trusting it on the harder question that comes in Example 2.

</details>

:::{.callout-note title="Question"}
Is sample mean really an unbiased estimator of the expected value? 
:::

<br>

That is, is $E[\frac{1}{n}\sum_{i=1}^n x_i] = E[x]$, where $x_i$ is an independent random draw from the same distribution?

### Steps 1-3

<details class="transcript">
<summary>Transcript</summary>

Steps one through three in six lines. Read the comment on the third line, because it is the whole point: x is drawn from runif, so the expected value of x is zero point five. That is the truth, and we know it because we chose it. Then we compute the sample mean and print it. Run the cell and look at the number. It is close to zero point five but not equal to it, and running it again gives you a different number. That variation is not a defect. It is the sampling distribution, and it is what we are here to look at.

</details>

```{webr-r}
#--- steps 1 and 2:  ---#
# specify the data generating process and generate data
x <- runif(100) # Here, E[x]=0.5

#--- step 3 ---#
# calculate sample mean
mean_x <- mean(x)
mean_x
```

### Sample Mean: Step 4

<details class="transcript">
<summary>Transcript</summary>

Step four is where the method actually earns its keep. One estimate told us almost nothing, because we could not tell whether being off by a bit was bad luck or systematic bias. So we repeat, many times, and look at the whole collection. The word in blue is the tool: a loop. If you have never written one, the next tab walks through it slowly, because loops are the one piece of programming syntax this course genuinely requires. Everything else you will do in R is a single function call; simulation needs you to repeat things.

</details>

+ repeat the above steps many times
+ We use a <span style = "color: blue;"> loop </span> to do the same (similar) thing over and over again

### Loop: for loop 

<details class="transcript">
<summary>Transcript</summary>

A loop in isolation, doing nothing but printing. Read the R code first: B is a thousand, and then for i in one to B, print i. Now read the verbal translation underneath, which spells out what the machine does. I takes the value one, and prints one. Then i takes the value two, and prints two. And so on to a thousand. That is all a for loop is: take each value in the list in turn, and run the body with i set to it. Run the cell, watch the numbers scroll, and then we put something useful in the body.

</details>

**R code**

```{webr-r}
#--- the number of iterations ---#
B <- 1000

#--- repeat steps 1-3 B times ---#
for (i in 1:B) {
  print(i) # print i
}
```

<br>

**Verbally**

For each of $i$ in $1:B$ $(1, 2, \dots, 1000)$, do `print(i)`.

+ `i` takes the value of 1, and then `print(1)`
+ `i` takes the value of 2, and then `print(2)`
+ ...
+ `i` takes the value of 999, and then `print(999)`
+ `i` takes the value of 1000, and then `print(1000)`

### Step 4

<details class="transcript">
<summary>Transcript</summary>

Now the real loop, and notice that the body is exactly the code from the Steps one-to-three tab. Two things were added around it. Before the loop, we create a storage vector of a thousand zeros, because we need somewhere to put a thousand answers. Inside the loop, the last line writes this iteration's mean into position i of that vector. That indexing is the piece to understand: i is not just a counter, it is the address where this answer gets filed. Run it. Nothing prints, which is correct; a thousand estimates are sitting in memory waiting for the next tab.

</details>

```{webr-r}
#| autorun: true
 
#--- the number of iterations ---#
B <- 1000

#--- create a storage that stores estimates ---#
estimate_storage_mean <- rep(0, B)

#--- repeat steps 1-3 B times ---#
for (i in 1:B) {
  #--- steps 1 and 2:  ---#
  # specify the data generating process and generate data
  x <- runif(100) # Here, E[x]=0.5

  #--- step 3 ---#
  # calculate sample mean
  mean_x <- mean(x)
  estimate_storage_mean[i] <- mean_x
}
```


### Step 5

<details class="transcript">
<summary>Transcript</summary>

Step five, the comparison, and here is what a thousand estimates look like. The blue line is the truth, zero point five, which we chose. The red dashed line is the average of your estimates. They sit essentially on top of each other, which is the numerical confirmation that the sample mean is unbiased. Now read the second bullet, because it is the point students most often miss. Individual estimates are all over the place, from about zero point four to zero point six. Unbiasedness says nothing about any one of them. It is a statement about the center of this histogram.

</details>

Compare your estimates with the true parameter

+ the truth here is $E[x] = 0.5$, because $x$ is drawn from `runif()`

```{webr-r}
mean(estimate_storage_mean)

hist(estimate_storage_mean, breaks = 30, main = "1000 sample means")
abline(v = 0.5, col = "blue", lwd = 2) # the truth
abline(v = mean(estimate_storage_mean), col = "red", lty = 2, lwd = 2) # average estimate
```

+ the blue line (the truth) and the red dashed line (the average of your estimates) sit essentially on top of each other
+ note that individual estimates are <span style = "color: red;"> not </span> 0.5 &mdash; unbiasedness is a statement about the <span style = "color: blue;"> center </span> of this histogram, not about any single estimate

:::
<!--end of panel-->

## Monte Carlo Simulation: Example 2

::: {.panel-tabset}

### Question

<details class="transcript">
<summary>Transcript</summary>

Now the question we actually came for. What happens to beta-one-hat when the zero conditional mean condition fails? Note the hat carefully. The true beta-one does not change; it is a fixed feature of the world. What changes is the estimator, the thing we compute from data. This is the question we could not answer with real data, because we would need to see the error term to know the condition was violated, and see the true beta to know the estimate was wrong. In simulation we control both, so we can break the condition deliberately and watch the damage.

</details>

:::{.callout-note title="Question"}
What happens to $\hat{\beta}_1$ if $E[u|x]\ne 0$ when estimating $y=\beta_0+\beta_1 x + u$?
:::

### R code

<details class="transcript">
<summary>Transcript</summary>

Read lines two through four of the loop body, because that is where the violation is manufactured. We draw a variable called mu, and then mu goes into x, and the same mu also goes into u. So x and the error term share a common component by construction, which means they are correlated, which means the zero conditional mean condition fails. Everything else is untouched: the true intercept is one and the true slope is one, and we run ordinary OLS exactly as before. One thing broken, everything else fixed. That is the controlled experiment we promised at the start.

</details>

```{webr-r}
#--- Preparation ---#
B <- 1000 # the number of iterations
N <- 100 # sample size
estimate_storage <- rep(0, B) # estimates storage

#--- repeat steps 1-3 B times ---#
for (i in 1:B) {
  #--- steps 1 and 2:  ---#
  mu <- rnorm(N) # the common term shared by both x and u
  x <- rnorm(N) + mu # independent variable
  u <- rnorm(N) + mu # error
  y <- 1 + x + u # dependent variable
  data <- data.frame(y = y, x = x)

  #--- OLS ---#
  reg <- fixest::feols(y ~ x, data = data) # OLS
  estimate_storage[i] <- reg$coefficient["x"]
}
```


### What should we expect?

<details class="transcript">
<summary>Transcript</summary>

Before running anything, let's predict the answer, because a simulation you cannot predict is a simulation you cannot learn from. Both x and u contain mu, so the covariance between them is the variance of mu, which is one. And the variance of x is the variance of mu plus one, so two. The bias formula on screen turns those two numbers into a prediction: the estimates should center on one plus one half, which is one point five, against a true value of one. Write that number down. The next tab is the test of whether we understood our own model.

</details>

Both $x$ and $u$ contain $\mu$, so they are correlated by construction:

+ $Cov(x, u) = Var(\mu) = 1$
+ $Var(x) = Var(\mu) + 1 = 2$

<br>

And the bias of $\hat{\beta}_1$ is governed by exactly that ratio:

$$\hat{\beta}_1 \rightarrow \beta_1 + \frac{Cov(x, u)}{Var(x)} = 1 + \frac{1}{2} = 1.5$$

+ so the histogram should be centered at <span style = "color: red;">1.5</span>, not at the true value of 1

### Compare

<details class="transcript">
<summary>Transcript</summary>

And there it is. The estimates pile up around the red line at one point five, exactly as predicted, and the blue line at the true value of one sits well outside the bulk of the distribution. Compare this against the same picture in lecture 01-3, where the condition held and the histogram was centered on the truth. Now read the last bullet, which is the one to take home. More data would not save you. A larger sample tightens this histogram around one point five, so all it buys you is more and more confidence in the wrong number.

</details>

```{webr-r}
mean(estimate_storage)

hist(estimate_storage, breaks = 30, main = "1000 estimates of beta_1")
abline(v = 1, col = "blue", lwd = 2) # the truth
abline(v = 1.5, col = "red", lwd = 2) # where the estimates actually center
```

+ the estimates pile up around the red line, and the true value in blue sits well outside the bulk of the distribution
+ compare this against the same picture in lecture 01-3 (`Unbiasedness of OLS estimator` $\rightarrow$ `Seeing it`), where $E[u|x]=0$ held and the histogram <span style = "color: blue;"> was </span> centered on the truth
+ more data would <span style = "color: red;"> not </span> fix this: it would tighten the histogram around 1.5, which makes you more and more confident about the wrong number

:::
<!--end of panel-->


## Monte Carlo Simulation: Example 3 (optional)

::: {.panel-tabset}

### Question 

<details class="transcript">
<summary>Transcript</summary>

This example is optional, but it answers a question you should be asking. We have checked that beta-hat is unbiased. But every regression table also reports a standard error, which comes from an estimated variance. Is that estimate any good? On screen, the true variance of beta-one-hat, sigma squared over the sum of squared deviations of x, and underneath it the estimator we actually compute, which replaces the unknown sigma squared with the average squared residual. The question at the bottom is whether that substitution is honest. Same method as before: build a world where we know the truth, and check.

</details>

**Model**

\begin{aligned}
    y = \beta_0 + \beta_1 x + u \\
\end{aligned}

+ $x\sim N(0,1)$
+ $u\sim N(0,1)$
+ $E[u|x]=0$

<br>

**Variance of the OLS estimator**

True Variance of $\hat{\beta}_1$: $V(\hat{\beta}_1) = \frac{\sigma^2}{\sum_{i=1}^n (x_i-\bar{x})^2} = \frac{\sigma^2}{SST_X}$

Its estimator: $\widehat{V(\hat{\beta}_1)} =\frac{\hat{\sigma}^2}{SST_X} = \frac{\sum_{i=1}^n \hat{u}_i^2}{n-2} \times \frac{1}{SST_X}$

<br>

**Question**

Does the estimator really work? (Is it unbiased?)

### R code

<details class="transcript">
<summary>Transcript</summary>

One design choice here is worth pausing on. The x values are drawn once, before the loop, so they are identical in every iteration. Only the error term is redrawn. That is deliberate: the variance formula on the previous tab conditions on x, so to check it we must hold x fixed and let only u vary. Inside the loop we store two things per iteration, the coefficient and the estimated variance from vcov. Ten thousand iterations this time rather than a thousand, because we are estimating the center of a distribution that is quite skewed.

</details>

```{r}
#| include: false

# Note: this code is for generating SST_X, which is referred to in the presentation later
set.seed(903478)

#--- Preparation ---#
B <- 10000 # the number of iterations
N <- 100 # sample size
beta_storage <- rep(0, B) # estimates storage for beta
V_beta_storage <- rep(0, B) # estimates storage for V(beta)
x <- rnorm(N) # x values are the same for every iteration
SST_X <- sum((x - mean(x))^2)
```

```{webr-r}
set.seed(903478)

#--- Preparation ---#
B <- 10000 # the number of iterations
N <- 100 # sample size
beta_storage <- rep(0, B) # estimates storage for beta
V_beta_storage <- rep(0, B) # estimates storage for V(beta)
x <- rnorm(N) # x values are the same for every iteration
SST_X <- sum((x - mean(x))^2)

#--- repeat steps 1-3 B times ---#
for (i in 1:B) {
  #--- steps 1 and 2:  ---#
  u <- 2 * rnorm(N) # error
  y <- 1 + x + u # dependent variable
  data <- data.frame(y = y, x = x)

  #--- OLS ---#
  reg <- fixest::feols(y ~ x, data = data) # OLS
  beta_storage[i] <- reg$coefficient["x"]
  #* store estimated variance of beta_1_hat
  V_beta_storage[i] <- vcov(reg)["x", "x"]
}
```

### Check

<details class="transcript">
<summary>Transcript</summary>

Here is the comparison, and the true variance is computed for you at the top. Sigma squared is four, because the error was two times a standard Normal, and the sum of squared deviations of x is the number shown, so the true variance is that small number underneath. Now run the cell. It averages your ten thousand estimated variances. Compare the two numbers. They agree, which means the variance estimator is unbiased too. That matters more than it might seem: every standard error, every t-statistic, and every confidence interval you ever report rests on this being true.

</details>

**True Variance**

+ $SST_X = `r round(SST_X, digits = 2)`$
+ $\sigma^2 = 4$

$$V(\hat{\beta}) = 4/`r round(SST_X, digits = 2)` = `r round(4/SST_X, digits = 4)`$$

<br>

**Check**

Your Estimates of Variance of $\hat{\beta}_1$?

```{webr-r}
mean(V_beta_storage)
```

### Distribution

<details class="transcript">
<summary>Transcript</summary>

One last look, because averages hide things. This plots the whole distribution of your ten thousand variance estimates, with a vertical line at the true value. Notice the shape. It is not symmetric; it has a long right tail, so individual estimates overshoot badly more often than the average suggests. The average still lands on the line, which is what unbiasedness promises. But look at how wide the spread is, and remember that in real work you get exactly one draw from this distribution. Unbiased does not mean accurate in any particular sample. It never has.

</details>

```{webr-r}
ggplot(data = data.frame(x = V_beta_storage)) +
  geom_density(aes(x = x)) +
  geom_vline(xintercept = round(4 / SST_X, digits = 4)) +
  theme_lecture
```
:::
<!--end of panel-->


## Exercise (optional)

::: {.panel-tabset}

### Problem

<details class="transcript">
<summary>Transcript</summary>

An exercise to run yourself, and it isolates something we have mentioned but not shown. How does variation in x affect OLS? Two models on screen, identical except for the independent variable. X-one is standard Normal, x-two has variance nine, so three times the standard deviation. The error is the same in both, and the zero conditional mean condition holds for both, so both estimators are unbiased. Nothing is broken here. The only thing that differs is how spread out the x values are. Predict what that does to the estimates before you open the solution.

</details>

Using MC simulations, find out how the variation in $x$ affects the OLS estimators

<br>

**Model setup**

```{=tex}
\begin{align*}
  y = \beta_0 + \beta_1 x_1 + u \\
  y = \beta_0 + \beta_1 x_2 + u
\end{align*}
```

+ $x_1\sim N(0,1)$ and $x_2\sim N(0,9)$ (so $x_2$ has three times the standard deviation of $x_1$)
+ $u\sim N(0,1)$
+ $E[u|x_1]=0$ and $E[u|x_2]=0$

### Solution

<details class="transcript">
<summary>Transcript</summary>

The code runs both regressions inside one loop on the same simulated data, storing two coefficients per iteration in a matrix rather than a vector. Run it and read the four numbers at the bottom carefully. The two means are both close to one, confirming both estimators are unbiased, as they must be. Now compare the two standard deviations. The one for x-two is roughly a third of the one for x-one. That is the variance formula from lecture 01-3 in action: more variation in x means a bigger denominator, which means a smaller variance.

</details>

```{webr-r}
#--- Preparation ---#
B <- 1000 # the number of iterations
N <- 100 # sample size
estimate_storage <- matrix(0, B, 2) # estimates storage

for (i in 1:B) {
  #--- generate data ---#
  x_1 <- rnorm(N, sd = 1) # indep var 1
  x_2 <- rnorm(N, sd = 3) # indep var 2
  u <- rnorm(N) # error
  y_1 <- 1 + x_1 + u # dependent variable 1
  y_2 <- 1 + x_2 + u # dependent variable 2
  data <- data.table::data.table(y_1 = y_1, y_2 = y_2, x_1 = x_1, x_2 = x_2)

  #--- OLS ---#
  reg_1 <- fixest::feols(y_1 ~ x_1, data = data) # OLS
  reg_2 <- fixest::feols(y_2 ~ x_2, data = data) # OLS

  #--- store coef estimates ---#
  estimate_storage[i, 1] <- reg_1$coefficient["x_1"] # equation 1
  estimate_storage[i, 2] <- reg_2$coefficient["x_2"] # equation 2
}

#--- assign new names ---#
beta_1s <- estimate_storage[, 1]
beta_2s <- estimate_storage[, 2]

#--- mean ---#
mean(beta_1s)
mean(beta_2s)

#--- sd ---#
sd(beta_1s)
sd(beta_2s)
```


### Results visualization

<details class="transcript">
<summary>Transcript</summary>

And here it is as a picture, which is worth more than the four numbers. Two densities, both centered on the true value of one, so unbiasedness is not what separates them. But the distribution for equation two, with the more spread out x, is dramatically narrower. Same truth, same sample size, same conditions satisfied, and yet one estimator is far more precise. That is the practical lesson: variation in your independent variable is information. An x that barely moves gives OLS little information per observation, although increasing the sample size can still improve precision.

</details>

```{webr-r}
#| fig-width: 9
#| fig-height: 6.5

plot_data_1 <- data.table(x = beta_1s, type = "Equation 1")
plot_data_2 <- data.table(x = beta_2s, type = "Equation 2")
plot_data <- rbind(plot_data_1, plot_data_2)
ggplot(data = plot_data) +
  geom_density(aes(x = x, fill = type), alpha = 0.5) +
  scale_fill_discrete(name = "") +
  xlab("Coefficient Estimate") +
  theme_lecture +
  theme(
    legend.position = "bottom"
  )
```
:::
<!--end of panel-->
